### 2008-09-05 doc/documentation/appendix.tex
moved gringo2 branch into trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@500 3e02f596-3131-0410-8352-e04eeaec37cc

		
\begin{verbatim}
WS              = ' ' | ? ISO 6429 character Horizontal Tabulation ?;
NL              = ? ISO 6429 character Carriage Return ?;
IF              = ":-";
NOT             = 'not';
DIGIT           = '0' | '1' | '2' | '3' | '4' | '5' | '6' | '7' | '8' | '9';
DIGIT1          = DIGIT-'0';
UNDERSCORE      = '_';
LETTER          = 'a' | 'b' | 'c' | 'd' | 'e' | 'f' | 'g' | 'h'
                | 'i' | 'j' | 'k' | 'l' | 'm' | 'n' | 'o' | 'p' | 'q'
                | 'r' | 's' | 't' | 'u' | 'v' | 'w' | 'x' | 'y' | 'z';
CAP_LETTER      = 'A' | 'B' | 'C' | 'D' | 'E' | 'F' | 'G' | 'H' | 'I'
                | 'J' | 'K' | 'L' | 'M' | 'N' | 'O' | 'P' | 'Q' | 'R'
                | 'S' | 'T' | 'U' | 'V' | 'W' | 'X' | 'Y' | 'Z';
IN_STRING          = LETTER | CAP_LETTER | UNDERSCORE | DIGIT;

IDENTIFIER      = [LETTER | UNDERSCORE], {IN_STRING};
STRING          = '"', {INSTRING-('"' | 'NL')}, '"';
VARIABLE        = [CAP_LETTER],  {IN_STRING};
COMMENT         = '%', {IN_STRING-NL}, NL;
BLOCK_COMMENT   = '%*', {IN_STRING | NL}, '*%';
SHOW            = ['#'], 'show';
HIDE            = ['#'], 'hide';
CONST           = ['#'], 'const';
DOMAIN          = ['#'], 'domain';
MINIMIZE        = ['#'], 'minimize';
MAXIMIZE        = ['#'], 'maximize';
COMPUTE         = ['#'], 'compute';
BASE            = ['#'], 'base';
CUMULATIVE      = ['#'], 'cumulative';
VOLATILE        = ['#'], 'volatile';
DISJUNCTION     = '|';
LPARA           = '(';
RPARA           = ')';
LBRAC           = '{';
RBRAC           = '}';
LSBRAC          = '[';
RSBRAC          = ']';
DOT             = '.';
DOTS            = '..';
SEMI            = ';';
DDOT            = ':';
COMMA           = ',';
PLUS            = '+';
TILDE           = '~';
OR              = '?';
AND             = '&';
XOR             = '^';
MINUS           = '-';
TIMES           = '*';
DIVIDE          = '/';
ASSIGN          = '=';
EQ              = '==';
NE              = '!=';
GT              = '>';
LT              = '<';
GE              = '>=';
LE              = '<=';

START = PROGRAM;

PROGRAM = PROGRAM, RULE, DOT;
PROGRAM = PROGRAM, SHOW, SHOW_LIST, DOT;
PROGRAM = PROGRAM, HIDE, HIDE_LIST, DOT;
PROGRAM = PROGRAM, DOMAIN, DOMAIN_LIST, DOT;
PROGRAM = PROGRAM, CONST, IDENTIFIER, ASSIGN, CONST_TERM, DOT;
PROGRAM = PROGRAM, BASE, DOT;
PROGRAM = PROGRAM, CUMULATIVE, IDENTIFIER, DOT;
PROGRAM = PROGRAM, VOLATILE, IDENTIFIER, DOT;
PROGRAM = DOT;

SHOW_LIST = SHOW_LIST, COMMA, SHOW_PREDICATE;
SHOW_LIST = SHOW_PREDICATE;

HIDE_LIST = NHIDE_LIST;

NHIDE_LIST = NHIDE_LIST COMMA HIDE_PREDICATE;
NHIDE_LIST = HIDE_PREDICATE;

DOMAIN_LIST = DOMAIN_LIST COMMA DOMAIN_PREDICATE;
DOMAIN_LIST = DOMAIN_PREDICATE;

NEG_PRED = IDENTIFIER;
NEG_PRED = MINUS, IDENTIFIER;

SHOW_PREDICATE = NEG_PRED, LPARA, NEG_PRED, RPARA;
SHOW_PREDICATE = NEG_PRED;
SHOW_PREDICATE = NEG_PRED, DIVIDE, NUMBER;
HIDE_PREDICATE = NEG_PRED, LPARA VARIABLE_LIST, RPARA;
HIDE_PREDICATE = NEG_PRED;
HIDE_PREDICATE = NEG_PRED, DIVIDE, NUMBER;

DOMAIN_PREDICATE = NEG_PRED, LPARA, DOMAIN_LIST, RPARA;

VARIABLE_LIST = VARIABLE_LIST, COMMA, VARIABLE;
VARIABLE_LIST = VARIABLE;

DOMAIN_VAR = VARIABLE;
DOMAIN_VAR = DOMAIN_VAR, SEMI, VARIABLE;

DOMAIN_LIST = DOMAIN_LIST, COMMA, DOMAIN_VAR;
DOMAIN_LIST = DOMAIN_VAR(var);

RULE = HEAD_ATOM IF BODY;
RULE = HEAD_ATOM IF;
RULE = HEAD_ATOM;
RULE = IF BODY;
RULE = IF;
RULE = MAXIMIZE;
RULE = MINIMIZE;
RULE = COMPUTE;

BODY = BODY, COMMA, BODY_LITERAL;
BODY = BODY_LITERAL;

CONDITIONAL = PREDICATE;
CONDITIONAL = RELATION_LITERAL;

CONDITIONAL_LIST = CONDITIONAL_LIST, DDOT, CONDITIONAL;

BODY_LITERAL = BODY_ATOM;
BODY_LITERAL = NOT, BODY_ATOM;
BODY_LITERAL = RELATION_LITERAL;
BODY_LITERAL = VARIABLE, ASSIGN, AGGREGATE;
BODY_LITERAL = AGGREGATE_ATOM;

CONSTRAINT_LITERAL = CONSTRAINT_ATOM;
CONSTRAINT_LITERAL = NOT, CONSTRAINT_ATOM;

BODY_ATOM = PREDICATE, CONDITIONAL_LIST;

(* assignment literals are not realy relation literals but they are used like them *)
RELATION_LITERAL = VARIABLE, ASSIGN, TERM;
RELATION_LITERAL = TERM EQ TERM;
RELATION_LITERAL = TERM NE TERM;
RELATION_LITERAL = TERM GT TERM;
RELATION_LITERAL = TERM GE TERM;
RELATION_LITERAL = TERM LT TERM;
RELATION_LITERAL = TERM LE TERM;

HEAD_ATOM = AGGREGATE_ATOM;
HEAD_ATOM = DISJUNCTION;

DISJUNCTION = DISJUNCTION_LIST;
DISJUNCTION_LIST = DISJUNCTION_LIST, DISJUNCTION, CONSTRAINT_ATOM;
DISJUNCTION_LIST = CONSTRAINT_ATOM;

CONSTRAINT_ATOM = PREDICATE, CONDITIONAL_LIST;

PREDICATE = IDENTIFIER, LPARA, TERM_LIST, RPARA;
PREDICATE = IDENTIFIER;
PREDICATE = MINUS, IDENTIFIER, LPARA, TERM_LIST, RPARA;
PREDICATE = MINUS, IDENTIFIER;
AGGREGATE_ATOM = TERM, AGGREGATE, TERM;
AGGREGATE_ATOM = AGGREGATE, TERM;
AGGREGATE_ATOM = TERM, AGGREGATE;
AGGREGATE_ATOM = AGGREGATE;

TERM_LIST = TERM_LIST, COMMA, TERM;
TERM_LIST = TERM;

TERM = VARIABLE;
TERM = IDENTIFIER;
TERM = STRING;
TERM = NUMBER;
TERM = LPARA, TERM, RPARA;
TERM = TERM, MOD, TERM;
TERM = TERM, PLUS, TERM;
TERM = TERM, TIMES, TERM;
TERM = TERM, MINUS, TERM;
TERM = TERM, DIVIDE, TERM;
TERM = TERM, XOR, TERM;
TERM = TERM, AND, TERM;
TERM = TERM, OR, TERM;
TERM = TILDE, TERM;
TERM = TERM DOTS TERM;
TERM = MINUS TERM;
TERM = ABS, LPARA, TERM, RPARA;
TERM = TERM, SEMI, TERM;
TERM = IDENTIFIER, LPARA, TERM_LIST, RPARA;

CONST_TERM_LIST = CONST_TERM_LIST COMMA CONST_TERM;
CONST_TERM_LIST = CONST_TERM;

CONST_TERM = IDENTIFIER;
CONST_TERM = STRING;
CONST_TERM = NUMBER;
CONST_TERM = LPARA CONST_TERM, RPARA;
CONST_TERM = CONST_TERM, MOD, CONST_TERM;
CONST_TERM = CONST_TERM, PLUS, CONST_TERM;
CONST_TERM = CONST_TERM, TIMES, CONST_TERM;
CONST_TERM = CONST_TERM, MINUS, CONST_TERM;
CONST_TERM = CONST_TERM, DIVIDE, CONST_TERM;
CONST_TERM = MINUS, CONST_TERM;
CONST_TERM = ABS, LPARA, CONST_TERM, RPARA;
CONST_TERM = IDENTIFIER, LPARA, CONST_TERM_LIST, RPARA;

AGGREGATE = SUM, LSBRAC WEIGHT_LIST, RSBRAC;
AGGREGATE = LSBRAC, WEIGHT_LIST, RSBRAC;
AGGREGATE = COUNT, LBRAC CONSTR_LIST, RBRAC;
AGGREGATE = LBRAC, CONSTR_LIST, RBRAC;
AGGREGATE = MIN, LSBRAC, WEIGHT_LIST, RSBRAC;
AGGREGATE = MAX, LSBRAC, WEIGHT_LIST, RSBRAC;

COMPUTE  = COMPUTE, LBRAC, CONSTR_LIST, RBRAC;
COMPUTE  = COMPUTE, NUMBER, LBRAC,  CONSTR_LIST, RBRAC;
MINIMIZE = MINIMIZE, LBRAC,  CONSTR_LIST, RBRAC;
MINIMIZE = MINIMIZE, LSBRAC, WEIGHT_LIST, RSBRAC;
MAXIMIZE = MAXIMIZE, LBRAC,  CONSTR_LIST, RBRAC;
MAXIMIZE = MAXIMIZE, LSBRAC, WEIGHT_LIST, RSBRAC;

WEIGHT_LIST = NWEIGHT_LIST;
NWEIGHT_LIST = NWEIGHT_LIST, COMMA, WEIGHT_TERM;
NWEIGHT_LIST = WEIGHT_TERM;

WEIGHT_TERM = CONSTRAINT_LITERAL, ASSIGN TERM;
WEIGHT_TERM = CONSTRAINT_LITERAL;

CONSTR_LIST = NCONSTR_LIST;
NCONSTR_LIST = NCONSTR_LIST, COMMA CONSTR_TERM;
NCONSTR_LIST = CONSTR_TERM;

CONSTR_TERM = CONSTRAINT_LITERAL;
\end{verbatim}

### 2008-09-05 doc/documentation/paper.tex
moved gringo2 branch into trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@500 3e02f596-3131-0410-8352-e04eeaec37cc

\documentclass[a4paper,10pt]{article}

\usepackage[latin1]{inputenc}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage[all]{xy}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{color}

\author{Max Ostrowski}
\title{Gringo 2.0\\--Draft--}
\date{\today}

\begin{document}
\maketitle
\tableofcontents
\section{Introduction}
Answer Set Programming (ASP) \cite{baral02a} is a declarative approach for modeling and solving search problems, represented as logic programs.
An ASP system usually consists of a grounder and a solver.
First, a grounder translates a non-ground problem description into a propositional program, which can be processed by the second component of the system, the solver. This is illustrated in figure (\ref{fig:ASP}).
\begin{figure}[h]
\[
\xymatrix{*+[F]\txt{Problem\\representation}\ar[rr]^{Grounder}&&
		*+[F]\txt{Intermediate\\ representation} \ar[rr]^{Solver}&&
		*+[F]\txt{Variable\\ assignment}}
\]
\caption{Architecture of an ASP System}
\label{fig:ASP}
\end{figure}
gringo2.0 is a state of the art grounder and has many features that makes writing logic programs easy.
It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0\cite{gringopaper}.
Therefore it supports binder splitting, backjumping and all other features already implemented in gringo1.0.
A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder supports such a rich featured language.
The goal of this work is to give an overview of the features supported by gringo2.0 and its derivates.
Three different packages of gringo exist.
The first version is gringo2.0, it acts as a grounder and supports several output formats and even a human readable format for debugging purposes. This version can ground logic programs and output them to files or pipe them to a solver.
The second version is called clingo and is a grounder with a built-in solver clasp \cite{gekanesc07b}.
Output can directly be given to the solver and answer sets can be computed.
The third package is called iclingo, which stands for ``incremental clingo''. In this version the grounder gringo2.0 can be used with clasp incrementally. Detailed information about incremental solving can be found in \cite{gekaosscth08a}.
In this paper, at first, the general restrictions to the input language are stated and the supported input language is described in detail. After that, a short introduction of how to write a logic program and how to write an incremental logic program is given. In the next chapter a brief description of how to use the three programs gringo2.0, clingo and iclingo and their options for the command line is given. The paper closes with an outlook of future work that has to be done in the area of grounding logic programs.
The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.

\section{Syntax}
The basic elements of a logic program and its syntax is now described.
Disjunctive logic programs are based on an alphabet \cite{syntax} consisting of:
\begin{itemize}
\item a set $\mathcal{P}$ of predicate symbols,
\item a set $\mathcal{F}$ of function symbols,
\item a set $\mathcal{V}$ of variables,
\item a set $\mathcal{C}$ of constants,
\item a set \{ ``$\mid$'', ``,'' ,$\leftarrow$, not \} of connectives, and
\item a set \{ ``('', ``)'', ``,'' \} of punctuation symbols.
\end{itemize}

\paragraph{Predicates and Function Symbols}
Predicates and function symbols have a predicate symbol/name and an arity.
For the function symbol $f$ with arity 3 we write $f/3$.
The names of the function symbols and predicates have to start with either a lower case letter or an underscore and can contain digits and letters.
With $\mathcal{P}$ we denote to the set of predicates and $\mathcal{F}$ denotes the set of function symbols.
\paragraph{Constants}
Constants are function symbols with an arity of zero.
The name of a constant can also be encompassed by double quotation marks, which is sometimes helpful when using constants with spaces. It has to be remarked that a constant with quotation marks is never equal to a constant without, so ``\texttt{a}'' is not equal to \texttt{a}.
$\mathcal{C}$ is the set of constants.
\paragraph{Variables}
Variables have to start with a capitalized letter and can contain digits and letters.
The set of variables is denoted by $\mathcal{V}$.
The language over $\mathcal{P}$, $\mathcal{F}$ and $\mathcal{C}$ is denoted by $\mathcal{L(P,F,C)}$.
\paragraph{Terms}
The building blocks of the language are terms.
Each variable $v \in \mathcal{V}$, constant $c \in \mathcal{C}$ and each function symbol $f \in \mathcal{F}$ is a term.
Furthermore $f(t_1, \dots, t_n)$ is a term if $f/n \in \mathcal{F}$ and $t_i$ for $i \in \{1 \dots n\}$ are terms.
Since a term is defined recursively, it can be nested.
A term with the first parameter being a constant $a$, the others being terms $f(b,X)$ and $g(f(X), Y)$ would be expressed this way.
\begin{verbatim}
 p(a,f(b,X), g(f(X),Y)).
\end{verbatim}
\paragraph{Atoms}
An atom over $\mathcal{L(P, F, C)}$ is of the form $p(t_1, \dots, t_n)$ where $p/n$ is a predicate
symbol with arity $n$, and $t_i$ for $i \in \{1 \dots n\}$ are terms.
\paragraph{Rules}
A rule over the language $\mathcal{L(P,F,C)}$ has the form
$$
A_0 \mid \dots \mid A_l \leftarrow A_{l+1}, \dots, A_m, not\ A_{m+1}, \dots, not\ A_n.
$$
where $A_0$ to $A_n$ are atoms. \newline
A logic program $\Pi$ over $\mathcal{L(P,F,C)}$ is a finite set of rules over $\mathcal{L(P,F,C)}$.

\subsection{$\lambda$-Restriction}
\label{domainrestricted}
There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs. For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
gringo2.0 verifies that the input program is $\lambda$-
restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
$\lambda$-restrictedness extends the capabilities of $\omega$-restricted programs. The definition of $\lambda$-restrictedness can be found in \cite{gringopaper} and was expanded for aggregates and disjunctions.
$\lambda$-restrictedness can not be compared with the safe programs of DLV since there exist programs that are safe and not $\lambda$-restricted and vice versa.
\newline
To give a gross impression of what $\lambda$-restriction means for writing a logic program an informal description is given.
In gringo2.0 every variable used in a rule must have a complete domain. This means that this variable has to occur in at least one predicate whose domain is completed. Therefore the predicate must be in the head of rules that are already domain restricted. So the grounder has to find a valid order of the rules so that every predicate in the head is completed before it is used as a domain predicate for the next rules.
To make things more clear, an example is given
\begin{verbatim}
 zig(0) :- not zag(0).   zag(0) :- not zig(0).
 zig(1) :- not zag(1).   zag(1) :- not zig(1).
 zigzag(X,Y) :- zig(X), zag(Y).
\end{verbatim}
The rules of this example can be ordered in the following way.
At first, the rules with \texttt{zig/1} and \texttt{zag/1} are considered.
Since these are the only rules \texttt{zig/1} and \texttt{zag/1} occurs in the head, the domains of the predicates are closed. The domains \{\texttt{zig(0)}, \texttt{zig(1)}\} and \{ \texttt{zag(0)}, \texttt{zag(1)}\} are complete, since no other predicates of the form \texttt{zig/1} or \texttt{zag/1} can be derived from the program.
Now the third rule with \texttt{zigzag/2} in the head can be grounded. The variables \texttt{X} and \texttt{Y} both occur in a domain predicate, and the values $0$ respectively $1$ can substitute the variables.
\newline
To give a comparison to the $\omega$-restricted programs of lParse, an extra domain predicate has to be introduced to the example.
\begin{verbatim}
 domain(0..1).
 zig(0) :- not zag(0).   zag(0) :- not zig(0).
 zig(1) :- not zag(1).   zag(1) :- not zig(1).
 zigzag(X,Y) :- zig(X), zag(Y), domain(X), domain(Y).
\end{verbatim}
\texttt{zig/1} and \texttt{zag/1} can not be used as a domain predicate, since their truth value can not be fully evaluated.

\subsection{Basic Language}
The base building blocks of the language are rules, facts and integrity constraints.
\paragraph{Rule}
A rule normally consists of a head and a body and ends with a dot.
The body is separated from the head using ``:-'' as in
\begin{verbatim}
 head :- b1, ..., bN, not c1, ..., cN.
\end{verbatim}
The head consists of terms concatenated by a disjunction. For non disjunctive rules only one term is used.
The body is a comma separated conjunction of terms and their negation.
The variables used must be restricted to a domain in the body.
Check (\ref{domainrestricted}) to see the restrictions on domains.
\paragraph{Fact}
A rule can occur without a body as a fact, therefore the body is omitted.
\begin{verbatim}
 head.
\end{verbatim}
\paragraph{Integrity Constraint}
To express an integrity constraint, a rule with an empty body is used.
\begin{verbatim}
 :- body.
\end{verbatim}

\subsection{Advanced Language Features}
To ease the process of writing logic programs some advanced features in the input syntax are introduced and are now explained in detail.
For the most problems it is possible to write a much shorter version of the program with the extended syntax.
%Even a speedup in solving the instances of the problem can happen TODO ?
For a full syntax description in EBNF see appendix (\ref{appendix}).
\paragraph{Function Symbols}
Function symbols are fully supported with gringo2.0.
They can be finitely nested and also be used as domain predicates.
While in lParse function symbols are allowed in a very restricted way, no restrictions apply in gringo2.0.
The following program can not be grounded with lParse since $X$ is an unrestricted variable. With gringo2.0 the program is now groundable.
\begin{verbatim}
a(X) :- f(1,g(X)).
f(1,X) :- f(X).
f(g(2)).
\end{verbatim}
From the second rule the domain \texttt{f(1,g(2))} is generated.
This domain can be used for the first rule and results in the grounded rule
\begin{verbatim}
 a(2) :- f(1,g(2)).
\end{verbatim}

\paragraph{Range Operator}
To define several atoms that stretch on one or more parameters over a certain range, two dots can be used to express an increasing integer range.
Using the range operator in the head of a rule, it works as a conjunction of all the members in the range, so for every number the rule is duplicated.
\begin{verbatim}
 times(1..3, 1..2).
\end{verbatim}
will result in the facts
\begin{verbatim}
 times(1,1). times(2,1). times(3,1).
 times(1,2). times(2,2). times(3,2).
\end{verbatim}

Using the range operator in the body of a rule, a disjunction of all members in the range is created. So the rules will be duplicated for every member.
\begin{verbatim}
 a :- p(1..3).
\end{verbatim}
will result in the rules
\begin{verbatim}
 a :- p(1).
 a :- p(2).
 a :- p(3).
\end{verbatim}
If the range operator is used in an aggregate, it creates a comma separated list of atoms, so
\begin{verbatim}
 a :- {b(1..3)}.
\end{verbatim}
results in
\begin{verbatim}
 a :- {b(1), b(2), b(3)}.
\end{verbatim}

\paragraph{Expander :}
With the expander ``:'' a domain can be set.
On the left side of the operator an atom with variables is expected and on the right side the domains for the variables are given.
In the head, the atoms get connected via a disjunction, in the body they get connected via a conjunction, so no new rules are created with this operator.
\begin{verbatim}
 a(X,Y) : b(X) : c(Y) :- d(X) : e(X).
 b(1..2).
 c(3).
 e(a;b;c).
\end{verbatim}
would become
\begin{verbatim}
 a(1,3) | a(2,3) :- d(a), d(b), d(c).
\end{verbatim}
This operator can be used in aggregates to create a comma separated list of atoms.
For example
\begin{verbatim}
 n(1..9).
 1{a(X) : n(X) : X > 3}1.
\end{verbatim}
would become
\begin{verbatim}
 1 {a(4), a(5), a(6), a(7), a(8), a(9)} 1.
\end{verbatim}

\paragraph{Conjunctor ;}
The conjunctor ``;'' will create a conjunction of its elements.
Used in the head of a rule means that the rule is duplicated.
If used in the body instead, the atoms will be logically combined using a conjunction.
\begin{verbatim}
 a(X;1..2) :- b(X;a;b).
 b(foo). b(bar).
\end{verbatim}
will results in the rules\footnote{gringo2.0 would simplify the grounded rules since some literals in the body are facts}
\begin{verbatim}
 a(foo,1) :- b(foo), b(a), b(b).
 a(foo,2) :- b(foo), b(a), b(b).
 a(bar,1) :- b(bar), b(a), b(b).
 a(bar,2) :- b(bar), b(a), b(b).
\end{verbatim}
The conjunctor can not be used directly in aggregates, but for example in combination with the expander.
\begin{verbatim}
 { a(X,Y) : p(X;Y) }.
 p(1..2).
\end{verbatim}
Results in
\begin{verbatim}
 {a(1,1), a(1,2), a(2,1), a(2,2)}.
 p(1).
 p(2).
\end{verbatim}

\paragraph{Arithmetic Expressions}
The built-in arithmetics can be used for integer arithmetic.
As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
Furthermore the bitwise operators and(\&), or(?), xor(\textasciicircum) and invert($\sim$) are supported. The bitwise operators are executed on signed integers using the binary complement for negative numbers.
Arithmetic expressions can occur everywhere in the head and in the body, even in complex constructs.
As parameters, integer constants or expressions that evaluate to integers are used.
The simple example of using operators and range terms
\begin{verbatim}
 a(X-1 .. X+1) :- b(X,Y), c(abs(Y * 2 * -X)).
 b(1,2).
 c(4).
\end{verbatim}
would result in the grounded program
\begin{verbatim}
 a(0) :- b(1,2), c(4).
 a(1) :- b(1,2), c(4).
 a(2) :- b(1,2), c(4).
 b(1,2).
 c(4).
\end{verbatim}

\subparagraph{Relational Operators}
The relational operators are ``less than'' ($<$),\\ ``greater than'' ($>$), ``less equal'' ($<=$), ``greater equal'' ($>=$) , ``equal'' ($==$) and ``not equal'' ($!=$).
They can be used to constrain certain variables.
For example, the following rule is only generated for all unifications where $X$ is greater or equal $Y-1$.
\begin{verbatim}
 a(X) :- b(X,Y), X >= Y-1.
 b(0,2).
 b(4,2).
\end{verbatim}
This would let $\texttt{b(4,2)}$ be a valid substitution, since $4 >= 2-1$ but $\texttt{b(0,2)}$ not, so the grounded program could be:
\begin{verbatim}
 a(4) :- b(4,2).
 b(0,2).
 b(4,2).
\end{verbatim}

\subparagraph{Assignment Operator}
As assignment operator ``$=$'' is used.
The variable on the left side is set to the value of the right side.
\begin{verbatim}
 a(X) :- X=Y*2, b(Y).
 b(1..2).
\end{verbatim}
becomes
\begin{verbatim}
 a(2) :- b(1).
 a(4) :- b(2).
 b(1..2).
\end{verbatim}
A cyclic assignment that results in an unbound variable is not allowed and is rejected with an error . For example
\begin{verbatim}
 a(X) :- X=Y, Y=X.
\end{verbatim}
Furthermore the assignment operator can be used with aggregates. The program
\begin{verbatim}
 price(apple,5). prices(banana, 6). price(computer, 100).
 fruit(apple).
 fruit(banana).
 max_cost(1..200).
 costs(C) :- C = [price(X,Y) : fruit(X) : max_cost(Y) = Y].
\end{verbatim}
would lead to the fact \texttt{costs(11)} as the price of all fruits.
$C$ in this example becomes all values that have to be generated for the aggregate.

\paragraph{Classical Negation}
For classical negation the atom is preceded by a minus sign.
\begin{verbatim}
 -a :- not  a.
  a :- not -a. 
\end{verbatim}
\paragraph{Disjunction}
A disjunction can be expressed using the pipe symbol ``$\mid$''.
A disjunction can only occur in the head of a rule.
\begin{verbatim}
 red | green | blue :- colored.
\end{verbatim}
Using disjunctions in logic programs can increase expressiveness and complexity of the logic program.
A solver for disjunctive logic programs such as claspD \cite{claspD} or cmodels \cite{cmodels} is needed to solve such programs.

\paragraph{Aggregates}
An aggregate is a set or a multiset of atoms and has a lower and an upper bound.
The general syntax is
\begin{verbatim}
 lower aggregate_name [atom1=w1, ..., atomN=wN] upper
\end{verbatim}
If squared brackets are used, the atom list is interpreted as a multiset, otherwise, for curly brackets, as a set of atoms.
The lower and upper bound is optional, and can be omitted.

\subparagraph{Count}
A ``count'' aggregate is a set of atoms and has an optional lower and upper bound.
It evaluates to true if the number of true atoms in the set is between the lower and upper bound. Only curly brackets may be used for a count aggregate as it represents a set of atoms.
Weights are not allowed in the count aggregate.
For this aggregate a shorthand syntax exists
\begin{verbatim}
 lower {atom1, ..., atomN} upper
\end{verbatim}
The count aggregate is equal to the cardinality constraint known from smodels \cite{smodels}.
\subparagraph{Sum}
The ``sum'' aggregate is a multiset of weighted atoms and has an optional lower and upper bound. It evaluates to true if the sum of the weights of all true atoms in the multiset is between the lower an upper bound.
The sum aggregate is equal to the weight constraints known from smodels, and has therefore a shorthand syntax
\begin{verbatim}
 lower [atom1=w1, ..., atomN=wN] upper
\end{verbatim}
If the command line option ``-l'' is used the semantic may change if negative weights exist in the aggregate. The transformation used is described in \cite{lparseManual}.

\subparagraph{Min}
The ``min'' aggregate is determined by the minimum of all weights of atoms that evaluate to true.
For the minimum aggregate a multiset with weighted atoms is used. It has no shorthand syntax.
\subparagraph{Max}
The ``max'' aggregate is similar to the minimum aggregate but considers the maximum of the weights.

\paragraph{Optimize Statements}

\subparagraph{Minimize/Maximize}
An optimize statement is used to find a minimal or maximal answer set of the logic program.
The statement can be weighted or not, therefore weights can be omitted in the following syntax.
\begin{verbatim}
 minimize [atom1 = w1, ..., atomN = wN].
\end{verbatim}
The maximize statement uses the keyword ``maximize'' instead.
If there were several minimize/maximize statements in the logic program the latter will be preferenced.
Only after the n'th statement results in the same value for two different answer set, the n-1'th statement is considered.

\subsection{Meta-Syntax}
This section is about constructs that do not originally belong to the logic program but can be used to give additional information for example to the solver.
The ``\#'' sign in front of all meta expressions is optional but is it strongly recommend to use it.
It separates the meta directives from the logic program more clearly and can prevent severe mistakes,
 confusing those constructs with atoms as shown in the example.
\begin{verbatim}
 hide.
 hidden :- hide.
\end{verbatim}
This program hides all predicates from the output of the ASP solver instead of deriving the facts \texttt{hide} and \texttt{hidden}.

\paragraph{hide and show}
The hide construct is used to hide atoms from the output of the ASP solver.
Two different syntaxes are supported by gringo2.0.
The first 
\begin{verbatim}
 #hide name/arity.
\end{verbatim}
and the second
\begin{verbatim}
 #hide name(X1, ... , Xn).
\end{verbatim}
Both variants hide the predicate ``\texttt{name}''.
The arity of the predicate can be given explicit or implicit with a list of variables.
The variables are just placeholder and do not unify with other variables. For example \texttt{hide p(X,X)} would also hide the fact \texttt{p(a,b)}. The second syntax is implemented for compatibility reasons only.
If the predicate that has to be hidden has arity zero the syntax is
\begin{verbatim}
 #hide name.
\end{verbatim}
If you want to express that all but a few predicates shall be hidden the empty hide statement
in conjunction with the ``\texttt{show}'' statement can be is used.
With ``\texttt{show}'' all predicates are named that shall be visible.
So to hide all predicates but $p/3$ the following example can be used.
\begin{verbatim}
 #hide.
 #show p/3.
\end{verbatim}

\paragraph{const}
Within the ``\texttt{const}'' statement, a default value for constants can be defined in the logic program.
while they are normally defined via the command line options.
A default value for a constant $n$ can be easily be set.
\begin{verbatim}
 #const n=5.
\end{verbatim}
\paragraph{domain}
A domain statement states that a specific variable has a global domain restriction.
The statement
\begin{verbatim}
 #domain time(T).
\end{verbatim}
makes sure that the variable ``\texttt{T}'' is always bound to the domain of ``\texttt{time}''.
This has the same effect as adding the domain ``\texttt{time(T)}'' to every rule the variable ``\texttt{T}'' occurs in.
\paragraph{compute}
The compute statement is a hint for the solver which models to compute.
It has the syntax
\begin{verbatim}
 #compute number {a1, ..., aN, not aN+1, ..., not aM}.
\end{verbatim}
$number$ is the number of models that shall be computed by the answer set solver and can be omitted. If it is zero, no restriction to the number of models is applied.
All positive atoms in the compute statement must be in the computed models, and all negated atoms do not have to be in the models of the program.
The compute statement is solver dependent and not supported by every solver.
\paragraph{comments}
Two kinds of comments were supported by gringo2.0.
Line based comments start with a ``\texttt{\%}''.
Whenever a ``\texttt{\%}'' occurs in the input, the rest of the line is treated as a comment and is therefore omitted.
Block based comments can span over several lines.
Everything between ``\texttt{\%*}'' and ``\texttt{*\%}'' is treated as a comment.
\subsection{Meta-Syntax for Incremental Programs}
\label{metasyntaxinc}
Incremental programs must be splitted into 3 parts.
The base part, for the information that hold in all steps and is independent of the incremental constant.
The cumulative part, where the cumulative rules must be set that hold in all steps, but depend on the incremental parameter.
And finally the volatile part. In this part the queries to the program are given, that must hold only in the last state.
To separate the parts, three statements exists. Everything beyond such a statement is considered to belong to the specific part.
To start the base part
\begin{verbatim}
 #base.
\end{verbatim}
must be written. Now all rules of the base part can follow.
The use of the incremental constant is not allowed in this part.
For the cumulative part an additional constant has to be passed to the statement.
\begin{verbatim}
 #cumulative k.
\end{verbatim}
Now all incremental rules follow. The incremental constant used for grounding is $k$.
The same holds for the volatile part.
\begin{verbatim}
 #volatile k.
\end{verbatim}
The ``\#''-sign can not be omitted for these meta statements.
These statements can also be mixed, only the last statement is important.
If no statement is given at the beginning of the file, base is assumed.

\section{Examples}
In this section some encodings for common problems are presented and explained.

\subsection{How to Write a Logic Program}
Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
Therefore the problem must be encoded as a set of rules.
Typically these are divided into a problem description and an instance of the problem.
The solution to a problem can then be read from the answer sets of the program.
The procedure of writing a logic program shall now be explained with a little example.
We want to write a problem encoding for the 3-colorability problem of graph coloring.
Given an arbitrary graph, no two connected edges may have the same color.
The most common technique is the ``generate \& test'' approach. Therefore, in the generator, rules are created that can derive a superset of the solution to the problem. In the tester all solutions that do not fit the constraints are omitted.
\newline
Usually you start with the generator.
At first, we define the non instance domains, in our example the colors we use.
\begin{verbatim}
 color(red). color(green). color(blue).
\end{verbatim}
Now we have to color all the vertices, but we must give attention that each vertex has only one color.
We can do this by simply saying that a vertex $V$ is colored with one color $C$, and that the same vertex does not have any of the other colors $D$ or $E$.
\begin{verbatim}
 colored(V,C) :- not colored(V,D), not colored(V,E),
                 C != D, C != E, D != E,
                 color(E), color(C), color(D), vertex(V).
\end{verbatim}
As you can see, you have to give the domain of the variables in the rule.
The variable $V$ is used for vertices and the variables $C$, $D$ and $E$ are used for colors.
After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
We now need to restrict our solutions in the tester by saying that two vertices that are connected by an edge should not have the same color.
We can do this by using an integrity constraint.
An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
\begin{verbatim}
 :- edge(V,U), colored(V,C), colored(U,C).
\end{verbatim}
Here we have the edge between two vertices $V$ and $U$, and if both are colored with the same color $C$ the assignment is not a solution to our problem.
\newline
We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
We just need to define several vertices and edges that do connect the vertices.
For the following simple graph
\[
 \xymatrix{1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&2\ar@{-}[dl]\\
           3&4}
\]

we could just write
\begin{verbatim}
 vertex(1..4).
 edge(1,2). edge(1,4).
 edge(3,1). edge(3,2).
\end{verbatim}

If we give the problem description and the instance to the grounder, it will generate a variable free version of the problem. This can then be solved by a solver and in the answer sets of the solver the ``\texttt{colored/2}'' predicate states a valid coloring of all vertices.
We can do this by simply using the internal interface to clasp, to get a solution.
\begin{verbatim}
 ./clingo 3colorability.lp instance1.lp
\end{verbatim}
In the first answer set we find the ``\texttt{colored}'' predicates
\begin{verbatim}
 colored(1,green) colored(2,red) colored(3,blue) colored(4,red)
\end{verbatim}
and we can interpret them as the colors of the vertices

\[
 \xymatrix{\color{green}1\ar@{-}[dr]\ar@{-}[d]\ar@{-}[r]&\color{red}2\ar@{-}[dl]\\
           \color{blue}3&\color{red}4}
\]
To improve the readability of the answer set, all but the \texttt{colored/2} predicates can be hidden with these two lines.
\begin{verbatim}
 #hide.
 #show colored/2.
\end{verbatim}

\subsection{How to Write an Incremental Program}
As in writing a non incremental logic problem, we describe a problem rather than an algorithm.
The difference between a non incremental and an incremental program is that we can use an incremental constant.
This constant is set to zero at the beginning and the solver tries to find a solution of the problem. If it can not find a solution it starts to increment the constant by one. At each step the solver searches for a solution to the problem and increments the constant until he finds a valid answer set for the logic program.
The incremental constant can therefore be used among other things to emulate time steps.
For example, if the solution for a planning problem can not be found in 3 steps, it is tried to find one in more steps.
To make it concrete, an incremental encoding for the blocksworld problem is presented.
This is a very simple problem, where $n$ blocks are stacked on top of each other and you can move only one block at the same time. The goal is to stack the blocks in the inverse order.
Since an incremental logic program is divided into three parts, we start with the base part.
All rules that hold in all time steps that are independent of the incremental constant are put here.
At first we declare $n$ blocks and that each block is a location where a block can be put on.
Furthermore we define an extra location, the table. That is the location where the bottom most block is put.
\begin{verbatim}
 #base.

 block(1..n).
 location(X) :- block(X).
 location(table).
\end{verbatim}
Now we stack all blocks on top of each other, so block 1 is on block 2, block 2 is on block 3 etc.
The last block we put on the table.
This all holds in the initial state, so a $0$ is appended to the ``\texttt{on}'' predicate.
\begin{verbatim}
 on(X,X+1,0) :- block(X), X < n.
 on(n,table,0).
\end{verbatim}
After declaring the base part and describing the initial situation we have to describe valid moves for the blocks. This is done in the cumulative part, since moves can be done in each step. For the incremental constant we choose the name ``\texttt{k}''.
At first we want to execute at least one move every step, but not more than one. This can be done with a cardinality constraint.
\begin{verbatim}
 #cumulative k.

 1{move(X,Y,k) : block(X) : location(Y) : X != Y}1.
\end{verbatim}
A move is of the form ``\texttt{move(X,Y,k)}'' where \texttt{X} is the block that is moved, \texttt{Y} the new position of the block and \texttt{k} the step when the moved is executed.
\newline
Now that we have exactly one move per step we need to define what happens if we move a block.
Understandably if we move block \texttt{X} to position \texttt{Y} in step \texttt{k}, the new location of the block shall be \texttt{Y}.
\begin{verbatim}
 on(X,Y,k) :- move(X,Y,k).
\end{verbatim}
For all other blocks that have not moved, the position shall be the same.
In ASP a little trick is used to describe such a condition.
We first define a little helper predicate that states that a block has moved.
\begin{verbatim}
 moved(X, k) :- move(X, Y, k).
\end{verbatim}
With this helping predicate we can define that a block is on the same position in step \texttt{k}, if it was not moved.
\begin{verbatim}
 on(X,Y,k) :- on(X,Y,k-1), block(X), location(Y), not moved(X,k).
\end{verbatim}

Now we have to restrict our moves, because only one block can be moved at the same time and it can only be moved if no other block is on top of it.
First we introduce a predicate called ``\texttt{blocked}'' for each block with another block on top of it.
\begin{verbatim}
 blocked(Y, k) :- block(Y), on(X, Y, k - 1).
\end{verbatim}
Now we can restrict the moves with an integrity constraint.
\begin{verbatim}
 :- move(X, Y, k), blocked(X, k).
\end{verbatim}
So it is not allowed to move block \texttt{X} in step k, if it is blocked.
Also you can not move a block to a position where another block is on, so only one block on top of another is allowed.
\begin{verbatim}
 :- move(X, Y, k), blocked(Y, k).
\end{verbatim}
The description of the problem is nearly finished. All that has to be done is to formulate a query.
As a constraint to our solution we want to stop if all blocks are on top of each other in reversed order.
This query shall only hold in the last step, therefore we write it in the volatile part.
As always for constraints, we use an integrity constraint to express that all solutions are false where there is not block $n+1$ on the block $n$.
\begin{verbatim}
 #volatile k.

 :- not on(X+1,X,k), block(X), X < n.
\end{verbatim}
To run this example, iclingo has to be used and the constant $n$ has to be set via command line.
A potential command line call could look like
\begin{verbatim}
 iclingo --const n=10 blocksworld.lp
\end{verbatim}
and the answer to the problem can be read from the answer set.
To clean up the answer the ``\texttt{show} and \texttt{hide}'' feature of gringo can be used.
If the following lines were added to the program, the answer set only consists of \texttt{move/3} predicates.
\begin{verbatim}
 #hide.
 #show move/3.
\end{verbatim}

\section{Using Gringo}
\subsection{Command Line Interface}
All gringo tools can be downloaded as binary from \cite{download}. To build them from source, an installation guide ``INSTALL'' is included in the tar file.
The gringo tools can be started from the command line and have some parameters that can be useful.
All programs are fully backward compatible meaning, all options of gringo2.0 are valid in clingo and iclingo, and all options of clingo are valid in iclingo.
\subsection{Gringo2.0 Options}
Gringo2.0 can be called from the command line via
\begin{verbatim}
 gringo (gringo option|file)*
\end{verbatim}
If no filename is given, the standard input is used.

\paragraph{\texttt{--help,-h}}
Gives a short overview of all parameters.
\paragraph{\texttt{--version,-v}}
Prints the version information and exists.
\paragraph{\texttt{--verbose,-V}}
Prints additional information during processing.
\paragraph{\texttt{--stats}}
Prints extended statistic information.

\paragraph{\texttt{--text,-t}}
Parameter ``\texttt{--text}'' invokes the text output method, so the output is printed in human readable format, which is also the input format for gringo2.0. The output is variable free and some simplifications on the program were made. To convert a variable free program in plain format into a format readable by a solver use the \texttt{--ground} option.
\paragraph{\texttt{--lparse,-l}}
This mode prints lParse conform output. For a definition of the output format see \cite{lparseManual}.
Since some features of gringo2.0 are not natively supported for this output language, new atoms may be introduced to create an equivalent program.\newline
Sum aggregates with negative weights are handled like weight constraint in lParse for this output format. So a transformation as explained in \cite{lparseManual} is used to get rid of the negative weights.
\paragraph{\texttt{--aspils,-a 1-7}}
This parameter invokes the new output format for the intermediate language ASPils \cite{gejaosscth08b}. The optional number can range from $1$ to $7$ and refers to one of the following normal forms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7).
% If no normal form is specified, conglomeration(7)  is assumed.

\paragraph{\texttt{--ground,-g}}
If this parameter is used, gringo expects an already grounded program as input.
For converting an already grounded logic program into an output format 
supported by gringo2.0 this is the right option.
 This mode saves a lot of memory but does not ground the program.
\paragraph{\texttt{--const,-c C=v}}
You can set the value of constants defined in your logic program via the command line of gringo2.0.
After the parameter ``--const'' you have to give the name of the constant $C$ you want to set followed by a ``$=$'' and the value of that constant. The integer value $v$ can even be an arithmetic expression. For example:
\begin{verbatim}
 gringo --const n=5*2
\end{verbatim}
the constant $n$ is replaced by $10$ in the logic program.
\paragraph{\texttt{--bindersplitting=<arg>}}
With this option you can enable or disable the binder splitting used by gringo2.0
\begin{itemize}
 \item yes : Enable binder splitting
 \item no  : Disable binder splitting
\end{itemize}
\paragraph{\texttt{--ifixed=<num>}}
An incremental program is required as input. Gringo2.0 grounds the fixed number of steps of the incremental program.

\paragraph{Default Parameters}
The default command line is
\begin{verbatim}
 gringo --lparse --bindersplitting=yes
\end{verbatim}
to read from standard input and for smodels output.

\subsection{Clingo Options}

Clingo can be called from the command line via
\begin{verbatim}
 clingo (clingo option|file)*
\end{verbatim}
If no filename is given, the standard input is used.
Since clingo uses gringo2.0 as a grounder in combination with the solver clasp, all options from gringo2.0 and clasp are valid. For a detailed list of the clasp options see the clasp help (also available with \texttt{clingo --help}).
\paragraph{\texttt{--clasp,-C}}
In this mode, the output is directly parsed to clasp. For solving logic programs, no intermediate files are produced and the answer sets are printed directly.
With this option enabled, all parameters for clasp are valid and can be declared in the command line.
For example:
\begin{verbatim}
 clingo --clasp --const n=5 --supp-models 3
\end{verbatim}
Furthermore all restriction on smodels output hold, as clasps internal interface currently reads smodels format.
This is default for clingo.
\paragraph{\texttt{--solveonly=<arg>}}
This option can be used to set the working mode of clingo to use it as a solver only.
\begin{itemize}
\item yes : Pass input directly to clasp
\item no : Standard mode
\end{itemize}
The default is \texttt{--solveonly=no} to ground the logic program first.
If the input shall directly be passed to clasped, it has to be in the valid smodels input format.

\paragraph{Default Parameters}
The default is to read from standard input and search for one model.
\begin{verbatim}
 clingo 1 --clasp --solveonly=no
          --trans-ext=no --eq=5 
          --lookahead=no --lookback=yes --heuristic=Berkmin
          --rand-prop=0.0 --randomize=no --rand-watches=yes
          --restarts=100,1.5 --deletion=3,1.1,3.0
          --minimize=all --contraction=250 --loops=common
\end{verbatim}

\subsection{Iclingo Options}
Iclingo can be called from the command line via
\begin{verbatim}
 iclingo (iclingo option|file)*
\end{verbatim}
Iclingo is used for incremental solving. For a detailed description see \cite{gekaosscth08a}.
It uses the grounder gringo2.0 and the solver clasp, and therefore all options of gringo2.0, clingo and clasp are valid.
Additionally some options for controlling the incremental behavior are implemented.
\paragraph{\texttt{--incremental,-i}}
Activates the incremental solving. This is the default for iclingo.
\paragraph{\texttt{--istats}}
Prints additional statistics for each incremental step.
\paragraph{\texttt{--imin=<num>}}
At least \texttt{num} incremental steps are solved, whether an answer set has been found or not.
\paragraph{\texttt{--imax=<num>}}
At most \texttt{num} incremental steps are done. After them iclingo stops the search for answer sets.
\paragraph{\texttt{--iquery=<num>}}
With this option you can set in which steps you want to ground without solving.
Solving the incremental program starts after this step.

\paragraph{\texttt{--istop=<arg>}}
The stop condition during incremental solving can be set.
\begin{itemize}
 \item SAT    : Stop the search process if a solution was found.
 \item UNSAT  : Stop the search process if no solution was found.
\end{itemize}
The default for this option is SAT.
\paragraph{\texttt{--ilearnt=<arg>}}
This option influences the behavior of clasp when used incrementally.
\begin{itemize}
 \item keep   : Learnt nogoods are kept for the next step while solving.
 \item forget : Clasp forgets the learnt nogoods after each incremental step.
\end{itemize}
As default, clasp forgets the learnt nogoods.
\paragraph{\texttt{--iheuristic=<arg>}}
This option sets how to handle heuristic information during incremental solving.
\begin{itemize}
 \item keep   : Heuristic information is kept for the next step while solving.
 \item forget : Clasp forgets the heuristic information after each incremental step.
\end{itemize}
As default, clasp keeps the heuristic information gathered during the previous steps.
\paragraph{Default Parameters}
The default is to read from standard input and search for one model incrementally.
\begin{verbatim}
 iclingo 1 --incremental --istop=SAT --ilearnt=forget 
           --iheuristic=keep --trans-ext=no --eq=5 
           --lookahead=no --lookback=yes --heuristic=Berkmin
           --rand-prop=0.0 --randomize=no --rand-watches=yes
           --restarts=100,1.5 --deletion=3,1.1,3.0
           --minimize=all --contraction=250 --loops=common

\end{verbatim}

\subsection{Warnings}
Several warnings are produced by gringo2.0 to give hints on common mistakes.
They are all printed to standard error output.
TODO: Some new warnings and Errors were added lately by Roland

\begin{itemize}
 \item 
\begin{verbatim}
 "Warning: sum with negative bounds in the body of a rule"
\end{verbatim}
This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp. In this format negative bounds are compiled away using the transformation described in \cite{lparseManual}. This may lead to counterintuitive results.
 \item 
\begin{verbatim}
 "Warning: multiple definitions of #const n"
\end{verbatim}
This warning is printed if more than one \#const statement is used to set the same constant.
The last attached value is used for grounding.
\item
\begin{verbatim}
 "Warning: p/3 is never defined"
\end{verbatim}
This warnings states that an atom is used in the body of a rule that never occurs in the head.
The predicate literal is given in the warning. The predicate will be removed through optimization.
\item
\begin{verbatim}
 "got ground program i hope u have enough memory :)"
\end{verbatim}
Use the command line option ``--ground'' to convert a ground program into another format.
For converting already ground programs to an output language a special module has been implemented that uses far less memory. Because trying to ground an already variable free program could need a huge amount of memory.
\item
\begin{verbatim}
 "Warning: There are no #base, #lambda or #delta sections."
\end{verbatim}
If writing an incremental program, the incremental sections\ref{metasyntaxinc} have to be used.
\item
\begin{verbatim}
 "Warning: There are statements not within a #base, #lambda"
 "         or #delta section."
 "         These Statements are put into the #base section."
\end{verbatim}
If writing an incremental program, all statements that are not written in an incremental section are automatically considered to belong to the \texttt{base} part of the program.
\item
\begin{verbatim}
 "Warning: Classical negation is not handled correctly in"
 "         combination with the incremental output."
 "         You have to add rules like:"
 "         :- a, -a. on your own! (at least for now)"
\end{verbatim}
To use classical negation in an incremental program the described rules have to be added manually. This will be fixed in a later version of gringo.
\end{itemize}
Furthermore all warnings from clasp are included.

\subsection{Errors}
gringo2.0 stops the execution and prints an error message to standard error output.
\begin{itemize}
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  non domain predicates : { ... }"
\end{verbatim}
This error message gives a list of predicates that are not domain restricted.
You explicitly have to give a domain for those predicates.
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  weakly restricted variables: { ... }"
\end{verbatim}
This error message gives a list of variables that are not lambda restricted. 
\item
\begin{verbatim}
 "syntax error on line Y column Z unexpected token: X"
\end{verbatim}
The parser can not read the input format. Some unexpected token occurred in the input file.
\item
\begin{verbatim}
 "sorry clasp cannot handle disjunctive rules!"
\end{verbatim}
If using clingo or iclingo, this happens if disjunctive rules are used. Please choose another solver that support disjunctive rules.
\item
\begin{verbatim}
 "A fixed number of incremental steps is needed"
 "to ground the program."
\end{verbatim}
This error occurs if an incremental program is given to gringo without using the option \texttt{--ifixed} or iclingo.
\item
\begin{verbatim}
 "cyclic constant definition."
\end{verbatim}
A cyclic definition of a constant is not allowed as shown in the example.
\begin{verbatim}
 #const x=y.
 #const y=x.
\end{verbatim}
\item
\begin{verbatim}
 "The following statement cant be used with the"
 "incremental interface: X."
 "unsupported statement."
\end{verbatim}
It is not allowed to use an optimize (\texttt{minimize} or \texttt{maximize}) or \texttt{compute} statement in an incremental program.
\item
\begin{verbatim}
 "comparing different types"
 "comparing function symbols"
\end{verbatim}
This error occurs if trying to compare non compareable values or function symbols as in
\begin{verbatim}
 p :- 1 < s.
 p :- f(x) < f(y).
\end{verbatim}

\item
The next errors all belong to the ASPils output format.
Some constructs in the input language can not be expressed in certain normal forms.
Another normal form as to be chosen to support this feature.
\begin{verbatim}
 "Disjunction not allowed on this normal form,
  please choose normal form 2 or 4-7."
 "Count aggregate/cardinality constraint not allowed
  in this normal form,
  please choose normal form 3-7."
 "Count aggregate/cardinality constraint not allowed 
  with non trivial upper bound in this normal form,
  please choose normal form 5-7."
 "Count aggregate/cardinality constraint not allowed 
  with negative lower bound in this normal form,
  please choose normal form 5-7."
 "Sum aggregate/weight constraint not allowed
  in this normal form,
  please choose normal form 3-7."
 "Sum aggregate/weight constraint not allowed 
  with non trivial upper bound in this normal form,
  please choose normal form 5-7."
 "Sum aggregate/weight constraint not allowed 
  with negative lower bound in this normal form,
  please choose normal form 5-7"
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normal form,
  please choose normal form 5 or 7."
 "Max aggregate not allowed in this normal form,
  please choose normal form 6 or 7."
 "Max aggregate not allowed
  with negative weights in this normal form,
  please choose normal form 7."
 "Min aggregate not allowed in this normal form,
  please choose normal form 6 or 7."
 "Min aggregate not allowed 
  with negative weights in this normal form,
  please choose normal form 7."
 "Optimize statement not allowed in this normal form,
  please choose normal form 3-7."
\end{verbatim}
%\begin{verbatim}
% "Times aggregate not allowed in this normal form."
% "Times aggregate not allowed
%  with negative weights in this normal form."
%\end{verbatim}
\end{itemize}

\section{Future Work}
Further research has to be done on features in the input language. Additionally the support of incremental grounding will be improved. To speed up the grounding process research on a bottom up grounding procedure is done. To support an even less restricted input language the support for $\Pi$-restricted programs is planned in a future release of gringo. All these features will lead to more acceptance of ASP-systems as the programs become easier to write and the bottleneck of ASP, the grounding part, will be faster.
Furthermore the acceptance of a standardized intermediate format for ASP is necessary. gringo2.0 supports the ASPils output format, but many more tools for converting, shuffling and merging must be developed.
\newpage
\textbf{\appendixname}
\begin{appendix}
\section{EBNF Syntax}
\label{appendix}
\input{appendix}
\end{appendix}

\bibliographystyle{plain}
\bibliography{bibtex}
\appendix
\end{document}

### 2008-09-05 app/enumerator.cpp

@@ -1,0 +1,174 @@
+// 
+// Copyright (c) 2006-2007, Benjamin Kaufmann
+// 
+// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
+// 
+// Clasp is free software; you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation; either version 2 of the License, or
+// (at your option) any later version.
+// 
+// Clasp is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with Clasp; if not, write to the Free Software
+// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
+//
+#include "enumerator.h"
+#include <clasp/include/solver.h>
+#include <iostream>
+namespace Clasp {
+
+/////////////////////////////////////////////////////////////////////////////////////////
+// class StdOutPrinter
+/////////////////////////////////////////////////////////////////////////////////////////
+StdOutPrinter::StdOutPrinter() : index(0) {}
+void StdOutPrinter::printModel(const Solver& s) {
+	if (index) {
+		std::cout << "Answer: " << s.stats.models << "\n";
+		for (AtomIndex::size_type i = 0; i != index->size(); ++i) {
+			if (s.value((*index)[i].lit.var()) == trueValue((*index)[i].lit) && !(*index)[i].name.empty()) {
+				std::cout << (*index)[i].name << " ";
+			}
+		}
+		std::cout << std::endl;
+	}
+	else {
+		const uint32 numVarPerLine = 10;
+		std::cout << "c Model: " << s.stats.models << "\n";
+		std::cout << "v ";
+		for (Var v = 1, cnt=0; v <= s.numVars(); ++v) {
+			if (s.value(v) == value_false) std::cout << "-";
+			std::cout << v;
+			if (++cnt == numVarPerLine && v+1 <= s.numVars()) { cnt = 0; std::cout << "\nv"; }
+			std::cout << " ";
+		}
[…]

### 2008-09-05 app/enumerator.h

@@ -1,0 +1,77 @@
+// 
+// Copyright (c) 2006-2007, Benjamin Kaufmann
+// 
+// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
+// 
+// Clasp is free software; you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation; either version 2 of the License, or
+// (at your option) any later version.
+// 
+// Clasp is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with Clasp; if not, write to the Free Software
+// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
+//
+#ifndef CLASP_ENUMERATOR_H_INCLUDED
+#define CLASP_ENUMERATOR_H_INCLUDED
+
+#include <clasp/include/solve_algorithms.h>
+#include <clasp/include/smodels_constraints.h>
+#include <clasp/include/program_builder.h>
+#include <clasp/include/clause.h>
+
+#include <iosfwd>
+
+namespace Clasp {
+	
+class StdOutPrinter : public ModelPrinter {
+public:
+	StdOutPrinter();
+	void printModel(const Solver& s);
+	const AtomIndex*	index;
+};
+
+class MinimizeEnumerator : public Enumerator {
+public:
+	MinimizeEnumerator(MinimizeConstraint* c) : mini_(c) {}
+	~MinimizeEnumerator();
+	void		updateModel(Solver& s);
+	bool		backtrackFromModel(Solver& s)			{ return mini_->backtrackFromModel(s); }
+	uint64	numModels(const Solver&)		const {	return mini_->models(); }
+	bool		allowRestarts()							const { return mini_->allowRestart(); }
+private:
+	MinimizeConstraint* mini_;
+};
+
+class CBConsequences : public Enumerator {
+public:
+	enum Consequences_t { brave_consequences, cautious_consequences };
+	CBConsequences(Solver& s, AtomIndex* index, Consequences_t type, bool quiet);
+	~CBConsequences();
+	void		updateModel(Solver& s);
+	bool		backtrackFromModel(Solver& s);
[…]

### 2008-09-05 app/program_opts/detail/smartpointer.h

@@ -1,0 +1,304 @@
+// scaled down version of Jean Gressmann's smartpointer.h 
+// (see copyright notice at end of file)
+
+#ifndef PROGRAM_OPTIONS_DETAIL_SMARTPOINTER_H
+#define PROGRAM_OPTIONS_DETAIL_SMARTPOINTER_H
+
+#if defined(_MSC_VER)
+#	define for if(0);else for
+#	pragma warning(disable:4521)
+#	pragma warning(disable:4786)
+#	pragma warning(disable:4522)
+#	pragma warning(disable:4355)
+#	pragma warning(disable:4291)
+#endif
+
+#include <algorithm>
+#include <cassert>
+
+namespace ProgramOptions {
+
+// Thanks to Benjamin Kaufmann (kaufmann@cs.uni-potsdam.de)
+// who provided for compile time assertions for void pointers
+template <bool> struct StaticAssert;
+template <> struct StaticAssert<true> {};
+template <class T> struct is_void {enum {result = 0};};
+template <> struct is_void<void> {enum {result = 1};};
+
+// These freeResource() functions are required to handle their
+// specific Null indifferently, say delete 0 is
+// to do nothing.
+template<class T>
+struct PointerTraits_
+{
+	StaticAssert<!is_void<T>::result> Error_Smart_Pointer_Does_Not_Support_Void_Ptr;
+	typedef T ValueType;
+	typedef T& ReferenceType;
+	typedef const T& ConstReferenceType;
+	typedef T* PointerType;
+	typedef const T* ConstPointerType;
+	static const PointerType Null;
+
+	static void freeResource(PointerType p)
+	{
+		enum { no_incomplete_types_please = sizeof(*p) };
+		delete p;
+	}
+};
+
+template<class T>
+const typename PointerTraits_<T>::PointerType PointerTraits_<T>::Null = 0;
+
+// reference counting based on a shared counter
+template<class Traits>
+class ReferenceCountedOwner_
+{
+public:	
+	typedef typename Traits::PointerType PT;		
+	typedef typename Traits::ConstPointerType CPT;		
+	static CPT null() { return Traits::Null; }
+private:
+	class RefCount_
+	{
+		PT managed_;
+		size_t count_;		
+		RefCount_();
+		RefCount_(const RefCount_&);
[…]

### 2008-09-05 lib/clasp/include/constraint.h

@@ -1,0 +1,316 @@
+// 
+// Copyright (c) 2006-2007, Benjamin Kaufmann
+// 
+// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
+// 
+// Clasp is free software; you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation; either version 2 of the License, or
+// (at your option) any later version.
+// 
+// Clasp is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with Clasp; if not, write to the Free Software
+// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
+//
+#ifndef CLASP_CONSTRAINT_H_INCLUDED
+#define CLASP_CONSTRAINT_H_INCLUDED
+
+#ifdef _MSC_VER
+#pragma once
+#endif
+
+#include <clasp/include/literal.h>
+#include <utility>	// std::pair
+#include <cassert>
+
+/*!
+ * \file 
+ * Defines the base classes for boolean constraints.
+ * 
+ */
+namespace Clasp {
+
+class Solver;
+
+/**
+ * \defgroup constraint Boolean Constraints
+ */
+//@{
+
+//! Constraint types distinguished by a Solver.
+struct Constraint_t {
+	enum Type { native_constraint = 0, learnt_conflict = 1, learnt_loop = 2};
+};	
+typedef Constraint_t::Type ConstraintType;
+
+
+//! Base class for (boolean) constraints to be used in a Solver.
+/*!
+ * Base class for (boolean) constraints like e.g. clauses. Concrete constraint classes define
+ * representations for constraints over boolean variables.
+ * Each constraint class must define a method for inference (derive information from an assignment),
+ * it must be able to detect conflicts (i.e. detect when the current assignment makes the constraint unsatisfiable)
+ * and to return a reason for inferred literals as well as conflicts (as a set of literals). 
+ */
+class Constraint {
+public:
+	/*!
+	 * Type used as return type for Constraint::propagate.
+	 * \see Constraint::propagate for a description of the meaning of PropResult's members.
+	 */
+	typedef std::pair<bool, bool> PropResult;
+
+	Constraint();
+
+	/*!
[…]

### 2008-09-05 lib/clasp/include/program_builder.h

@@ -1,0 +1,712 @@
+// 
+// Copyright (c) 2006-2007, Benjamin Kaufmann
+// 
+// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
+// 
+// Clasp is free software; you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation; either version 2 of the License, or
+// (at your option) any later version.
+// 
+// Clasp is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with Clasp; if not, write to the Free Software
+// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
+//
+
+#ifndef CLASP_PROGRAM_BUILDER_H_INCLUDED
+#define CLASP_PROGRAM_BUILDER_H_INCLUDED
+
+#ifdef _MSC_VER
+#pragma once
+#endif
+
+#include <clasp/include/solver_types.h>
+#include <clasp/include/program_rule.h>
+#include <clasp/include/util/misc_types.h>
+#include <string>
+#include <map>
+#include <algorithm>
+
+namespace Clasp {
+
+class Solver;
+class ClauseCreator;
+class PrgNode;
+class PrgBodyNode;
+class PrgAtomNode;
+class ProgramBuilder;
+class Preprocessor;
+class DefaultUnfoundedCheck;
+class MinimizeConstraint;
+typedef PodVector<PrgAtomNode*>::type AtomList;
+typedef PodVector<PrgBodyNode*>::type BodyList;
+typedef PodVector<PrgNode*>::type NodeList;
+
+/**
+ * \defgroup problem Problem specification
+ * Classes and functions for defining/processing logic programs
+ */
+//@{
+
+//! A list of variables and their types
+/*!
+ * A VarList stores the type of each variable, which allows for easy variable-cloning.
+ * It also stores flags for each literal to be used in "mark and test"-algorithms.
+ */
+class VarList {
+public:
+	VarList() { add(Var_t::atom_var); }
+	//! adds a new var of type t to the list
+	Var		add(VarType t) {
+		vars_.push_back( static_cast<uint8>(t << 5) );
+		return (Var)vars_.size()-1;
+	}
[…]

### 2008-09-05 lib/clasp/include/solver.h

@@ -1,0 +1,977 @@
+// 
+// Copyright (c) 2006-2007, Benjamin Kaufmann
+// 
+// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
+// 
+// Clasp is free software; you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation; either version 2 of the License, or
+// (at your option) any later version.
+// 
+// Clasp is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with Clasp; if not, write to the Free Software
+// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
+//
+#ifndef CLASP_SOLVER_H_INCLUDED
+#define CLASP_SOLVER_H_INCLUDED
+#ifdef _MSC_VER
+#pragma once
+#endif
+
+#include <clasp/include/solver_types.h>
+#include <string>
+#include <utility>
+
+namespace Clasp { 
+
+///////////////////////////////////////////////////////////////////////////////
+// classes used by the Solver
+///////////////////////////////////////////////////////////////////////////////
+class DecisionHeuristic;
+class PostPropagator;
+class ClauseCreator;
+
+//! Base class for preprocessors working on clauses only
+class SatPreprocessor {
+public:
+	SatPreprocessor(Solver& s) { setSolver(s); }
+	virtual ~SatPreprocessor();
+	void setSolver(Solver& s) { solver_ = &s; }
+	virtual bool addClause(const LitVec& cl) = 0;
+	virtual bool preprocess() = 0;
+	virtual void extendModel(State& m) = 0;
+	virtual bool hasSymModel() const = 0;
+protected:
+	Solver* solver_;
+private:
+	SatPreprocessor(const SatPreprocessor&);
+	SatPreprocessor& operator=(const SatPreprocessor&);
+};
+///////////////////////////////////////////////////////////////////////////////
+
+/**
+ * \defgroup solver Solver
+ */
+//@{
+
+//! Parameter-Object for configuring a solver.
+struct SolverStrategies {
[…]

### 2008-09-05 lib/clasp/src/smodels_constraints.cpp

@@ -1,0 +1,473 @@
+// 
+// Copyright (c) 2006-2007, Benjamin Kaufmann
+// 
+// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
+// 
+// Clasp is free software; you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation; either version 2 of the License, or
+// (at your option) any later version.
+// 
+// Clasp is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with Clasp; if not, write to the Free Software
+// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
+//
+
+#include <clasp/include/smodels_constraints.h>
+#include <clasp/include/clause.h>
+#include <clasp/include/solver.h>
+#include <algorithm>
+namespace Clasp {
+
+/////////////////////////////////////////////////////////////////////////////////////////
+// WeightConstraint
+/////////////////////////////////////////////////////////////////////////////////////////
+bool BasicAggregate::newWeightConstraint(Solver& s, Literal con, WeightLitVec& lits, weight_t bound) {
+	assert(s.decisionLevel() == 0);
+	if (bound <= 0) {	// trivially SAT
+		return s.force(con, 0);
+	}
+	weight_t sumWeight = 0;
+	bool card			= true;	// cardinality constraint?
+	for (uint32 i = 0; i < lits.size();) {
+		if (s.isTrue(lits[i].first)) {
+			bound -= lits[i].second;
+			if (bound <= 0) {	// trivially SAT
+				return s.force( con, 0 );
+			}
+			lits[i] = lits.back();
+			lits.pop_back();
+		}
+		else if (s.isFalse(lits[i].first)) {
+			lits[i] = lits.back();
+			lits.pop_back();
+		}
+		else {
+			assert((sumWeight + lits[i].second) > sumWeight && "Weight-Rule: Integer overflow!");
[…]

### 2008-09-05 lib/clasp/src/solver.cpp

@@ -1,0 +1,819 @@
+// 
+// Copyright (c) 2006-2007, Benjamin Kaufmann
+// 
+// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
+// 
+// Clasp is free software; you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation; either version 2 of the License, or
+// (at your option) any later version.
+// 
+// Clasp is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with Clasp; if not, write to the Free Software
+// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
+//
+#include <clasp/include/solver.h>
+#include <clasp/include/clause.h>
+#include <clasp/include/util/misc_types.h>
+#include <functional>
+#include <stdexcept>
+
+#ifdef _MSC_VER
+#pragma warning (disable : 4355) // 'this' used in base member initializer list - intended & safe
+#pragma warning (disable : 4702) // unreachable code - intended
+#endif
+
+namespace Clasp { 
+
+uint32 randSeed_g = 1;
+
+SatPreprocessor::~SatPreprocessor() {}
+DecisionHeuristic::~DecisionHeuristic() {}
+PostPropagator::PostPropagator() {}
+PostPropagator::~PostPropagator() {}
+void PostPropagator::reset() {}
+
+/////////////////////////////////////////////////////////////////////////////////////////
+// SelectFirst selection strategy
+/////////////////////////////////////////////////////////////////////////////////////////
+// selects the first free literal
+Literal SelectFirst::doSelect(Solver& s) {
+	for (Var i = 1; i <= s.numVars(); ++i) {
+		if (s.value(i) == value_free) {
+			return s.preferredLiteralByType(i);
+		}
+	}
+	assert(!"SelectFirst::doSelect() - precondition violated!\n");
+	return Literal();
+}
+/////////////////////////////////////////////////////////////////////////////////////////
[…]

### 2008-09-05 lib/gringo/include/bindersplitter.h

@@ -1,0 +1,60 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef BINDERSPLITTER_H
+#define BINDERSPLITTER_H
+
+#include <gringo.h>
+#include <literal.h>
+#include <groundable.h>
+#include <expandable.h>
+#include <value.h>
+
+namespace NS_GRINGO
+{
+	class BinderSplitter : public Literal
+	{
+	public:
+		BinderSplitter(Domain *domain, TermVector *param, const VarVector &relevant);
+		virtual Literal* clone() const;
+		virtual SDGNode *createNode(SDG *dg, SDGNode *prev, DependencyAdd todo);
+		virtual void createNode(LDGBuilder *dg, bool head);
+		virtual void createNode(StatementChecker *dg, bool head, bool delayed);
+		virtual void print(const GlobalStorage *g, std::ostream &out) const;
+		virtual void getVars(VarSet &vars) const;
+		virtual bool checkO(LiteralVector &unsolved);
+		virtual void reset();
+		virtual bool solved();
+		virtual bool isFact(Grounder *g);
+		virtual void finish();
+		virtual void evaluate();
+		virtual IndexedDomain *createIndexedDomain(Grounder *g, VarSet &index);
+		virtual bool match(Grounder *g);
+		virtual void preprocess(Grounder *g, Expandable *e);
+		virtual NS_OUTPUT::Object *convert();
+		virtual double heuristicValue();
+		virtual void addIncParam(Grounder *g, const Value &v);
+		virtual ~BinderSplitter();
+	protected:
+		Domain *domain_;
+		TermVector *param_;
+		VarVector relevant_;
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/constant.h

@@ -1,0 +1,48 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef CONSTANT_H
+#define CONSTANT_H
+
+#include <gringo.h>
+#include <term.h>
+#include <value.h>
+
+namespace NS_GRINGO
+{
+	class Constant : public Term
+	{
+	public:
+		Constant(const Value &value);
+		Constant(const Constant &c);
+		virtual void print(const GlobalStorage *g, std::ostream &out) const;
+		virtual void getVars(VarSet &vars) const;
+		virtual bool isComplex();
+		virtual void preprocess(Literal *l, Term *&p, Grounder *g, Expandable *e);
+		virtual Value getConstValue(Grounder *g);
+		virtual Value getValue(Grounder *g);
+		virtual Term* clone() const;
+		bool unify(const GlobalStorage *g, const Value& t, const VarVector& vars, ValueVector& vals) const;
+		virtual void addIncParam(Grounder *g, Term *&p, const Value &v);
+ 		virtual ~Constant();
+	protected:
+		Value value_;
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/disjunctionaggregate.h

@@ -1,0 +1,45 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef DISJUNCTIONAGGREGATE_H
+#define DISJUNCTIONAGGREGATE_H
+
+#include <gringo.h>
+#include <aggregateliteral.h>
+
+namespace NS_GRINGO
+{
+	class DisjunctionAggregate : public AggregateLiteral
+	{
+	public:
+		DisjunctionAggregate(ConditionalLiteralVector *literals);
+		DisjunctionAggregate(const DisjunctionAggregate &a);
+		virtual void setNeg(bool neg);
+		virtual Literal *clone() const;
+		virtual void preprocess(Grounder *g, Expandable *e);
+		virtual bool match(Grounder *g);
+		virtual void match(Grounder *g, int &lower, int &upper, int &fixed);
+		virtual void print(const GlobalStorage *g, std::ostream &out) const;
+		virtual NS_OUTPUT::Object *convert();
+		virtual ~DisjunctionAggregate();
+	public:
+	       	static Literal *createHead(ConditionalLiteralVector *list);
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/expandable.h

@@ -1,0 +1,40 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef EXPANDABLE_H
+#define EXPANDABLE_H
+
+#include <gringo.h>
+
+namespace NS_GRINGO
+{
+	/// Interface which allows to expand objects by literals
+	class Expandable
+	{
+	public:
+		enum ExpansionType { MATERM, RANGETERM, COMPLEXTERM };
+	public:
+		/**
+		 * \brief Appends a literal to the object
+		 * \param l The literal
+		 */
+		virtual void appendLiteral(Literal *l, ExpansionType type) = 0;
+		virtual ~Expandable(){};
+	};
+}
+
+#endif 

### 2008-09-05 lib/gringo/include/funcsymbol.h

@@ -1,0 +1,74 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef FUNCSYMBOL_H
+#define FUNCSYMBOL_H
+
+#include <gringo.h>
+
+namespace NS_GRINGO
+{
+	/**
+	 * \brief class of function symbols
+	 *
+	 * consists of a string and a list of values 
+	 * Is used in Value
+	 */
+	class FuncSymbol
+	{
+	public:
+		/**
+		 * \brief Constructor - f(a,b,c(1,2,3))    *s="f"
+		 * doest not take ownership of s
+		 */
+		FuncSymbol(int s, const ValueVector& args);
+		/**
+		 * \brief simple hash function
+		 * Merges the hash value of the argument vector with the hash value of the function name
+		 */
+		size_t getHash() const;
+		/**
+		 * \brief Comparison operator
+		 */
+		bool operator==(const FuncSymbol& a) const;
+		/**
+		 * \brief Prints the function symbol
+		 * \param out The output stream
+		 */
+		void print(const GlobalStorage *g, std::ostream& out) const;
+		/**
+		 * \brief Get the name of the function symbol
+		 * \return the string pointer to the name
+		 */
+		int getName() const;
+		/**
+		 * \brief Get the values of the function symbol
+		 * \return a vector of values
+		 */
+		const ValueVector& getValues() const;
+		/**
+		 * \brief Virtual Destructor
+		 */
+		virtual ~FuncSymbol();
+	protected:
+		int                     name_;
+		ValueVector		args_;
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/funcsymbolterm.h

@@ -1,0 +1,48 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef FUNCSYMBOLTERM_H
+#define FUNCSYMBOLTERM_H
+
+#include <gringo.h>
+#include <term.h>
+
+namespace NS_GRINGO
+{
+	class FuncSymbolTerm : public Term
+	{
+	public:
+		FuncSymbolTerm(Grounder* g, int s, TermVector* tl);
+		FuncSymbolTerm(const FuncSymbolTerm &f);
+		virtual void getVars(VarSet &vars) const;
+		virtual bool isComplex();
+		virtual void print(const GlobalStorage *g, std::ostream &out) const;
+		virtual Value getValue(Grounder *g);
+		virtual Value getConstValue(Grounder *g);
+ 		virtual Term* clone() const;
+		virtual void preprocess(Literal *l, Term *&p, Grounder *g, Expandable *e);
+		virtual bool unify(const GlobalStorage *g, const Value& t, const VarVector& vars, ValueVector& subst) const;
+		virtual void addIncParam(Grounder *g, Term *&p, const Value &v);
+		virtual ~FuncSymbolTerm();
+	protected:
+		int          name_;
+		TermVector*  termList_;
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/literaldependencygraph.h

@@ -1,0 +1,131 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef LITERALDEPENDENCYGRAPH_H
+#define LITERALDEPENDENCYGRAPH_H
+
+#include <gringo.h>
+
+namespace NS_GRINGO
+{
+	class LDG
+	{
+		friend class LDGBuilder;
+	//private:
+	public:
+		struct LiteralNode;
+		struct VarNode;
+
+		typedef std::vector<VarNode*>     VarNodeVector;
+		typedef std::vector<LiteralNode*> LiteralNodeVector;
+
+		struct LiteralNode
+		{
+			LiteralNode(Literal *l, bool head);
+			Literal *l_;
+			int done_;
+			int in_;
+			bool head_;
+			VarNodeVector out_;
+			LDGVector sub_;
+		};
+		struct VarNode
+		{
+			VarNode(int var);
+			bool done_;
+			int var_;
+			LiteralNodeVector out_;
+		};
+		struct LiteralData
+		{
+			LiteralData(LiteralNode *n);
+			LiteralNode *n_;
+			VarVector needed_;
+			VarVector provided_;
+		};
+
+		typedef std::map<Literal*, LiteralData*> LiteralDataMap;
+	public:
+		LDG();
+
+		const VarVector &getGlobalVars() const;
+		const VarVector &getParentVars() const;
+		void check(VarVector &free);
+
+		void start(LiteralSet &list);
+		void propagate(Literal *l, LiteralSet &list);
+
+		void sortLiterals(LiteralVector *lits);
+
+		const VarVector &getNeededVars(Literal *l) const;
[…]

### 2008-09-05 lib/gringo/include/literalstatement.h

@@ -1,0 +1,61 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef LITERALSTATEMENT_H
+#define LITERALSTATEMENT_H
+
+#include <gringo.h>
+#include <statement.h>
+#include <expandable.h>
+
+namespace NS_GRINGO
+{
+	/**
+	 * \brief A statement that contains exactly one literal
+	 *
+	 * Includes compute, minimize and maximize statements.
+	 */
+	class LiteralStatement : public Statement
+	{
+	public:
+		enum Type { COMPUTE, MINIMIZE, MAXIMIZE };
+	public:
+		/// Constructor
+		LiteralStatement(Literal *lit, bool preserveOrder);
+		virtual void buildDepGraph(SDG *dg);
+		virtual void getVars(VarSet &vars) const;
+		virtual bool checkO(LiteralVector &unsolved);
+		virtual bool check(Grounder *g, VarVector &free);
+		virtual void preprocess(Grounder *g);
+		virtual void reset();
+		virtual void finish();
+		virtual void evaluate();
+		virtual bool ground(Grounder *g, GroundStep step);
+		virtual void addDomain(PredicateLiteral *pl);
+		virtual void print(const GlobalStorage *g, std::ostream &out) const;
+		virtual void grounded(Grounder *g);
+		virtual void setIncPart(Grounder *g, IncPart part, const Value &v);
+		/// Destructor
+		virtual ~LiteralStatement();
+	protected:
+		Literal *lit_;
+		bool preserveOrder_;
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/lparselexer.h

@@ -1,0 +1,36 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef LPARSELEXER_H
+#define LPARSELEXER_H
+
+#include <gringo.h>
+#include <gringolexer.h>
+
+namespace NS_GRINGO
+{
+	class LparseLexer : public GrinGoLexer
+	{
+	public:
+		LparseLexer();
+		int lex(std::string *&val);
+		virtual ~LparseLexer();
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/optimizeliteral.h

@@ -1,0 +1,62 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef OPTIMIZELITERAL_H
+#define OPTIMIZELITERAL_H
+
+#include <gringo.h>
+#include <literal.h>
+#include <expandable.h>
+
+namespace NS_GRINGO
+{
+	class OptimizeLiteral : public Literal, public Expandable
+	{
+	public:
+		enum Type {MINIMIZE, MAXIMIZE};
+	public:
+		OptimizeLiteral(Type type, ConditionalLiteralVector *literals, bool setSemantic);
+		OptimizeLiteral(const OptimizeLiteral &a);
+		virtual SDGNode *createNode(SDG *dg, SDGNode *prev, DependencyAdd todo);
+		virtual void createNode(LDGBuilder *dg, bool head);
+		virtual void createNode(StatementChecker *dg, bool head, bool delayed);
+		virtual void getVars(VarSet &vars) const;
+		virtual bool checkO(LiteralVector &unsolved);
+		void ground(Grounder *g, GroundStep step);
+		virtual bool match(Grounder *g);
+		virtual void reset();
+		virtual void finish();
+		virtual void preprocess(Grounder *g, Expandable *e);
+		virtual IndexedDomain *createIndexedDomain(Grounder *g, VarSet &index);
+		virtual void appendLiteral(Literal *l, ExpansionType type);
+		virtual double heuristicValue();
+		virtual bool isFact(Grounder *g);
+		virtual bool solved();
+		virtual Literal *clone() const;
+		virtual NS_OUTPUT::Object *convert();
+		virtual void print(const GlobalStorage *g, std::ostream &out) const;
+		virtual void addIncParam(Grounder *g, const Value &v);
[…]

### 2008-09-05 lib/gringo/include/program.h

@@ -1,0 +1,46 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef PROGRAM_H
+#define PROGRAM_H
+
+#include <gringo.h>
+#include <printable.h>
+
+namespace NS_GRINGO
+{
+	class Program : public Printable
+	{
+	public:
+		enum Type {FACT=0, BASIC=1, NORMAL=2};
+	public:
+		Program(Type type, StatementVector &rules);
+		void print(const GlobalStorage *g, std::ostream &out) const;
+		Evaluator *getEvaluator();
+		bool check(Grounder *g);
+		StatementVector *getStatements();
+		Type getType();
+		~Program();
+	private:
+		Type type_;
+		StatementVector rules_;
+		Evaluator *eval_;
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/smodelsoutput.h

@@ -1,0 +1,51 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef SMODELSOUTPUT_H
+#define SMODELSOUTPUT_H
+
+#include <gringo.h>
+#include <smodelsconverter.h>
+
+namespace NS_GRINGO
+{
+	namespace NS_OUTPUT
+	{
+		class SmodelsOutput : public SmodelsConverter
+		{
+		public:
+			SmodelsOutput(std::ostream *out);
+			void initialize(GlobalStorage *g, SignatureVector *pred);
+			void finalize(bool last);
+			~SmodelsOutput();
+		protected:
+			void printBasicRule(int head, const IntVector &pos, const IntVector &neg);
+			void printConstraintRule(int head, int bound, const IntVector &pos, const IntVector &neg);
+			void printChoiceRule(const IntVector &head, const IntVector &pos, const IntVector &neg);
+			void printWeightRule(int head, int bound, const IntVector &pos, const IntVector &neg, const IntVector &wPos, const IntVector &wNeg);
+			void printMinimizeRule(const IntVector &pos, const IntVector &neg, const IntVector &wPos, const IntVector &wNeg);
+			void printDisjunctiveRule(const IntVector &head, const IntVector &pos, const IntVector &neg);
+			void printComputeRule(int models, const IntVector &pos, const IntVector &neg);
+		private:
+			int models_;
+			IntSet compNeg_, compPos_;
+		};
+	}
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/statement.h

@@ -1,0 +1,105 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef STATEMENT_H
+#define STATEMENT_H
+
+#include <gringo.h>
+#include <printable.h>
+#include <groundable.h>
+
+namespace NS_GRINGO
+{
+	/**
+	 * \brief Base class for every construct in a logic program that has to be grounded
+	 */
+	class Statement : public Printable, public Groundable
+	{
+	public:
+		/// Constructor
+		Statement();
+		/**
+		 * \brief Adds the needed dependencies to the SDG
+		 * \param dg The SDG
+		 */
+		virtual void buildDepGraph(SDG *dg) = 0;
+		/**
+		 * \brief Returns all variables occuring in the Statement
+		 * \param vars Reference to the result
+		 */
+		virtual void getVars(VarSet &vars) const = 0;
+		/**
+		 * \brief Checks the omega restricted part of the Statement
+		 *
+		 * Some literals like conditionals may need some domain restricted
+		 * parts. This can be checked here. CheckO must be called before 
+		 * check().
+		 * \param unsolved List of literals that are not omega restricted
+		 * \return Returns false in case of failure
+		 */
+		virtual bool checkO(LiteralVector &unsolved) = 0;
+		/**
+		 * \brief Checks if the program is lambda restricted
+		 * \param free List of variables that are which are not bound by complete predicates
+		 * \return Returns false in case of failure
+		 */
+		virtual bool check(Grounder *g, VarVector &free) = 0;
+		/**
+		 * \brief Performs static preprocessing.
+		 *
+		 * Currently ; and classical negation are removed here.
+		 * \param g Reference to the grounder
+		 */
+		virtual void preprocess(Grounder *g) = 0;
+		/**
[…]

### 2008-09-05 lib/gringo/include/statementchecker.h

@@ -1,0 +1,60 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef STATEMENTCHECKER_H
+#define STATEMENTCHECKER_H
+
+#include <gringo.h>
+
+namespace NS_GRINGO
+{
+	class StatementChecker
+	{
+	private:
+		struct Delay
+		{
+			Literal *l;
+			bool head;
+			VarSet provided;
+			VarSet needed;
+		};
+		typedef std::vector<Delay> DelayVector;
+	public:
+		StatementChecker(StatementChecker *parent = 0);
+		bool check();
+		void create();
+		void getFreeVars(VarVector &vars);
+		const VarSet &getVars() const;
+		bool hasVars() const;
+
+		void createDelayedNode(Literal *l, bool head, const VarSet &needed, const VarSet &provided);
+		void createSubNode(Literal *l, bool head);
+		void createNode(const VarSet &needed, const VarSet &provided);
+	private:
+		void createNode_(const VarSet &needed, const VarSet &provided);
+	private:
+		StatementChecker *parent_;
+		VarSet vars_;
+		VarSet provided_;
+		DelayVector delayed_;
+		std::list<std::pair<VarSet, VarSet> > providedIf_;
+		std::list<std::pair<VarSet, std::pair<StatementChecker, VarSet> > > providedSub_;
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/statementdependencygraph.h

@@ -1,0 +1,96 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef DEPENDENCYGRAPH_H
+#define DEPENDENCYGRAPH_H
+
+#include <gringo.h>
+#include <printable.h>
+
+namespace NS_GRINGO
+{
+	class SDG
+	{
+		friend class SDGNode;
+	private:
+		class SCC;
+		typedef std::vector<SCC*> SCCVector;
+		typedef std::set<SCC*> SCCSet;
+
+		class SCC
+		{
+		public:
+			enum Type {FACT=0, BASIC=1, NORMAL=2};
+		public:
+			SCC();
+		public:
+			Type type_;
+			int edges_;
+			StatementVector rules_;
+			SCCSet sccs_;
+		};
+	public:
+		SDG();
+		SDGNode *createStatementNode(Statement *r, bool preserveOrder = false);
+		SDGNode *createPredicateNode(PredicateLiteral *pred);
+		void calcSCCs(Grounder *g);
+		~SDG();
+	private:
+		bool calcSCCDep(SDGNode *v);
+		void tarjan(SDGNode *v, int &index, int &back, std::vector<SDGNode*> &stack);
+	private:
+		int uids_;
+		SDGNodeVector ruleNodes_;
+		SDGNodeVector predicateNodes_;
+		SCCVector sccs_;
+		SCCSet sccRoots_;
+		SDGNode *last_;
+	};
+	
+	class SDGNode
+	{
+		friend class SDG;
+	public:
+		enum Type {STATEMENTNODE = 0, PREDICATENODE = 1};
+	public:
+		SDGNode(Statement *rule);
+		SDGNode(Domain *dom);
+		SDGNodeVector *getDependency() const;
[…]

### 2008-09-05 lib/gringo/include/sumaggregate.h

@@ -1,0 +1,40 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef SUMAGGREGATE_H
+#define SUMAGGREGATE_H
+
+#include <gringo.h>
+#include <aggregateliteral.h>
+
+namespace NS_GRINGO
+{
+	class SumAggregate : public AggregateLiteral
+	{
+	public:
+		SumAggregate(ConditionalLiteralVector *literals);
+		SumAggregate(const SumAggregate &a);
+		virtual Literal *clone() const;
+		virtual void match(Grounder *g, int &lower, int &upper, int &fixed);
+		virtual void print(const GlobalStorage *g, std::ostream &out) const;
+		virtual NS_OUTPUT::Object *convert();
+		virtual ~SumAggregate();
+	};
+}
+
+#endif
+

### 2008-09-05 lib/gringo/include/value.h

@@ -1,0 +1,161 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#ifndef VALUE_H
+#define VALUE_H
+
+#include <gringo.h>
+#include <printable.h>
+
+namespace NS_GRINGO
+{
+	/**
+	 * \brief Class to store values of terms
+	 * If strings are stored in a value this class does not take ownership of the string.
+	 * The memory of the string has to be freed elsewhere.
+	 */
+	class Value
+	{
+	public:
+		/// The type o a value
+		enum Type { INT, STRING, FUNCSYMBOL, UNDEF };
+		/// Hash function object for arrays of values
+		struct VectorHash
+		{
+			/**
+			 * \brief Hash function for arrays of values
+			 * \param key The array
+			 * \return The hash
+			 */
+			inline size_t operator() (const ValueVector &key) const;
+		};	
+		/// Comaparison function object for arrays of values
+		struct VectorEqual
+		{
+			/**
+			 * \brief Comparison function for arrays of values
+			 * \return The hash
+			 */
+			inline bool operator() (const ValueVector &a, const ValueVector &b) const;
+		};	
+	public:
+		/**
+		 * \brief Constructor initializing the value to undef
+		 */
+		Value();
+		/**
+		 * \brief Copy constructor
+		 */
+		Value(const Value &v);
+		/**
+		 * \brief Creates a value encapsulationg an object of type uid
+		 * \param intValue The int
+		 */
+		Value(Type type, int uid_);
+		/**
+		 * \brief Calculates a hash for the value
+		 * \return The Hash
+		 */
+		inline size_t hashValue() const;
+		/**
+		 * Function used to compare values
+		 * \param b The other value
[…]

### 2008-09-05 lib/gringo/src/aggregateliteral.cpp

@@ -1,0 +1,386 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "aggregateliteral.h"
+#include "predicateliteral.h"
+#include "conditionalliteral.h"
+#include "term.h"
+#include "value.h"
+#include "grounder.h"
+#include "statementdependencygraph.h"
+#include "literaldependencygraph.h"
+#include "statementchecker.h"
+#include "indexeddomain.h"
+#include "output.h"
+#include "evaluator.h"
+#include "variable.h"
+#include "dlvgrounder.h"
+
+using namespace NS_GRINGO;
+
+AggregateLiteral::AggregateLiteral(ConditionalLiteralVector *literals) : Literal(), literals_(literals), lower_(0), upper_(0), equal_(0)
+{
+}
+
+bool AggregateLiteral::solved()
+{
+	return false;
+}
+
+bool AggregateLiteral::isFact(Grounder *g)
+{
+	return fact_;
+}
+
+void AggregateLiteral::setBounds(Term *lower, Term *upper)
+{
+	equal_ = 0;
+	lower_ = lower;
+	upper_ = upper;
+}
+
+void AggregateLiteral::setEqual(Variable *equal)
+{
+	equal_ = equal;
+	lower_ = 0;
+	upper_ = 0;
+}
+
+bool AggregateLiteral::checkO(LiteralVector &unsolved) 
+{
+	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
+	{
+		if(!(*it)->checkO(unsolved))
+			return false;
+	}
+	return true;
+}
+
[…]

### 2008-09-05 lib/gringo/src/conditionalliteral.cpp

@@ -1,0 +1,517 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "conditionalliteral.h"
+#include "predicateliteral.h"
+#include "grounder.h"
+#include "term.h"
+#include "indexeddomain.h"
+#include "dlvgrounder.h"
+#include "value.h"
+#include "literaldependencygraph.h"
+#include "statementchecker.h"
+#include "aggregateliteral.h"
+#include "domain.h"
+
+using namespace NS_GRINGO;
+		
+ConditionalLiteral::ConditionalLiteral(PredicateLiteral *pred, LiteralVector *conditionals) : Literal(), pred_(pred), conditionals_(conditionals), weight_(0), grounder_(0), dg_(0), clone_(0)
+{
+	Literal::setNeg(pred_->getNeg());
+}
+
+void ConditionalLiteral::setWeight(Term *w)
+{
+	weight_ = w;
+}
+
+void ConditionalLiteral::setNeg(bool neg)
+{
+	pred_->setNeg(neg);
+	Literal::setNeg(neg);
+}
+
+void ConditionalLiteral::appendLiteral(Literal *l, ExpansionType type)
+{
+	if(!conditionals_)
+		conditionals_ = new LiteralVector();
+	conditionals_->push_back(l);
+}
+
+void ConditionalLiteral::getVars(VarSet &vars) const
+{
+	pred_->getVars(vars);
+	if(conditionals_)
+		for(LiteralVector::const_iterator it = conditionals_->begin(); it != conditionals_->end(); it++)
+			(*it)->getVars(vars);
+	if(weight_)
[…]

### 2008-09-05 lib/gringo/src/countaggregate.cpp

@@ -1,0 +1,147 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "countaggregate.h"
+#include "conditionalliteral.h"
+#include "term.h"
+#include "value.h"
+#include "output.h"
+
+using namespace NS_GRINGO;
+
+namespace
+{
+	struct Hash
+	{
+		Value::VectorHash hash;
+		size_t operator()(const std::pair<int, ValueVector> &k) const
+		{
+			return (size_t)k.first + hash(k.second);
+		}
+	};
+	struct Equal
+	{
+		Value::VectorEqual equal;
+		size_t operator()(const std::pair<int, ValueVector> &a, const std::pair<int, ValueVector> &b) const
+		{
+			return a.first == b.first && equal(a.second, b.second);
+		}
+	};
+	typedef __gnu_cxx::hash_set<std::pair<int, ValueVector>, Hash, Equal> UidValueSet;
+}
+
+CountAggregate::CountAggregate(ConditionalLiteralVector *literals) : AggregateLiteral(literals)
+{
+}
+
+void CountAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
+{
+	UidValueSet set;
+	fact_ = true;
+	lower = 0;
+	upper = 0;
+	fixed = 0;
+	maxUpperBound_ = 0;
+	minLowerBound_ = 0;
+	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
+	{
+		ConditionalLiteral *p = *it;
+		p->ground(g, GROUND);
[…]

### 2008-09-05 lib/gringo/src/disjunctionaggregate.cpp

@@ -1,0 +1,126 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "disjunctionaggregate.h"
+#include "conditionalliteral.h"
+#include "predicateliteral.h"
+#include "term.h"
+#include "value.h"
+#include "output.h"
+
+using namespace NS_GRINGO;
+
+DisjunctionAggregate::DisjunctionAggregate(ConditionalLiteralVector *literals) : AggregateLiteral(literals)
+{
+}
+
+void DisjunctionAggregate::setNeg(bool neg)
+{
+	// the literal itself will never be negative
+	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
+		(*it)->setNeg(neg);
+}
+
+bool DisjunctionAggregate::match(Grounder *g)
+{
+	fact_ = false;
+	bool match = false;
+	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
+	{
+		ConditionalLiteral *p = *it;
+		p->ground(g, GROUND);
+		for(p->start(); p->hasNext(); p->next())
+		{
+			// every literal will match :)
+			if(p->match())
+			{
+				match = true;
+				if(p->isFact())
+				{
+					fact_ = true;
+					return true;
+				}
+			}
+		}
+	}
+	return match;
+}
+
+void DisjunctionAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
+{
+	assert(false);
+}
+
+void DisjunctionAggregate::print(const GlobalStorage *g, std::ostream &out) const
+{
+	bool comma = false;
[…]

### 2008-09-05 lib/gringo/src/domain.cpp

@@ -1,0 +1,129 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "domain.h"
+#include "value.h"
+
+using namespace NS_GRINGO;
+
+Domain::Domain() : type_(UNDEFINED), defines_(0), solved_(false)
+{
+}
+
+void Domain::setType(Type type)
+{
+	type_ = type;
+}
+
+int Domain::getDefines()
+{
+	return defines_;
+}
+
+bool Domain::complete() const
+{
+	return defines_ == 0;
+}
+
+bool Domain::solved() const
+{
+	return solved_;
+}
+
+void Domain::setSolved(bool solved)
+{
+	solved_ = solved;
+/*
+// incremental grounding may fail cause of that
+// since this is only a very small optimization
+// i simply turn it off in iclingo
+#ifndef WITH_ICLASP
+	// when a domain is solved all of its entries are facts
+	if(solved_)
+	{
+		ValueVectorSet facts;
+		std::swap(facts, facts_);
+	}
+#endif
+*/
+}
+
+void Domain::evaluate()
+{
+	setSolved(complete() && (type_ == FACT  || type_ == BASIC));
+}
+
+void Domain::reset()
+{
+	defines_++;
+	solved_ = false;
+}
+
+void Domain::finish()
+{
+	defines_--;
+}
+
+bool Domain::isFact(const ValueVector &values) const
+{
+	return facts_.find(values) != facts_.end();
+}
+
+void Domain::addFact(const ValueVector &values)
+{
+	// fact programs dont need to store their facts seperatly cause 
+	// directly after grounding all values in their domain are facts
+	// and there cant be any cycles through fact programs
+	// examples:
+	// a(1).
[…]

### 2008-09-05 lib/gringo/src/funcsymbol.cpp

@@ -1,0 +1,86 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "funcsymbol.h"
+#include "value.h"
+#include "globalstorage.h"
+
+using namespace NS_GRINGO;
+
+FuncSymbol::FuncSymbol(int s, const ValueVector& args) : name_(s), args_(args)
+{
+}
+
+FuncSymbol::~FuncSymbol()
+{
+}
+
+bool FuncSymbol::operator==(const FuncSymbol& a) const
+{
+	if (name_ != a.name_)
+		return false;
+	if (args_.size() != a.args_.size())
+		return false;
+
+	for (ValueVector::const_iterator i = args_.begin(), j = a.args_.begin(); i != args_.end(); ++i,++j)
+		if (!i->equal(*j))
+			return false;
+	return true;
+}
+
+size_t FuncSymbol::getHash() const
+{
+	size_t hash = 0;
+	size_t x = 0;
+
+	Value::VectorHash t;
+	hash = t(args_);
+
+	hash = (hash << 4) + size_t(name_);
+	if((x = hash & 0xF0000000L) != 0)
+	{
+		hash ^= (x >> 24);
+	}
+	hash &= ~x;
+	return hash;
+}
+
+void FuncSymbol::print(const GlobalStorage *g, std::ostream& out) const
+{
+	out << *g->getString(name_) << "(";
+	bool comma = false;
+	for (ValueVector::const_iterator i = args_.begin(); i != args_.end(); i++)
+	{
+		if(!comma)
+			comma = true;
+		else
+			out << ",";
+		i->print(g, out);
+	}
+	out << ")";
+}
+
+int FuncSymbol::getName() const
+{
+	return name_;
+}
+
+const ValueVector& FuncSymbol::getValues() const
+{
+	return args_;
+}
+

### 2008-09-05 lib/gringo/src/globalstorage.cpp

@@ -1,0 +1,99 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include <globalstorage.h>
+#include <domain.h>
+
+using namespace NS_GRINGO;
+
+GlobalStorage::GlobalStorage()
+{
+}
+
+Domain *GlobalStorage::getDomain(int uid) const
+{
+	return domains_[uid];
+}
+
+DomainVector *GlobalStorage::getDomains() const
+{
+	return const_cast<DomainVector*>(&domains_);
+}
+
+GlobalStorage::~GlobalStorage()
+{
+	stringHash_.clear();
+	funcHash_.clear();
+	for(StringVector::iterator it = strings_.begin(); it != strings_.end(); it++)
+		delete *it;
+	for(FuncSymbolVector::iterator it = funcs_.begin(); it != funcs_.end(); it++)
+		delete *it;
+	for(DomainVector::iterator it = domains_.begin(); it != domains_.end(); it++)
+		delete *it;
+}
+
+const std::string *GlobalStorage::getString(int uid) const
+{
+	return strings_[uid];
+}
+
+const FuncSymbol  *GlobalStorage::getFuncSymbol(int uid) const
+{
+	return funcs_[uid];
+}
+
+int GlobalStorage::createString(const std::string &s2)
+{
+	return createString(new std::string(s2));
+}
+
+int GlobalStorage::createString(std::string *s)
+{
+	std::pair<StringHash::iterator, bool> res = stringHash_.insert(std::make_pair(s, strings_.size()));
+	if(!res.second)
+		delete s;
+	else
+		strings_.push_back(s);
+	return res.first->second;
+}
+
+int GlobalStorage::createFuncSymbol(FuncSymbol* fn)
+{
[…]

### 2008-09-05 lib/gringo/src/grounder.cpp

@@ -1,0 +1,421 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "grounder.h"
+#include "statement.h"
+#include "variable.h"
+#include "predicateliteral.h"
+#include "normalrule.h"
+#include "statementdependencygraph.h"
+#include "program.h"
+#include "value.h"
+#include "evaluator.h"
+#include "domain.h"
+#include "gringoexception.h"
+
+using namespace NS_GRINGO;
+
+Grounder::Grounder(const Options &opts) : opts_(opts), incremental_(false), incStep_(0), internalVars_(0), output_(0), eval_(0)
+{
+}
+
+void Grounder::setOutput(NS_OUTPUT::Output *output)
+{
+	output_ = output;
+}
+
+void Grounder::addStatement(Statement *rule)
+{
+	if(incParts_.size() > 0)
+		incParts_.back().second++;
+	rules_.push_back(rule);
+}
+
+void Grounder::addDomains(int id, std::vector<IntVector*>::iterator pos, std::vector<IntVector*>::iterator end, IntVector &list)
+{
+	if(pos == end)
+	{
+		domains_.push_back(DomainPredicate(id, new IntVector(list)));
+	}
+	else
+	{
+		for(IntVector::iterator it = (*pos)->begin(); it != (*pos)->end(); it++)
+		{
+			list.push_back(*it);
+			addDomains(id, pos + 1, end, list);
+			list.pop_back();
+		}
+		delete (*pos);
+	}
+}
+
+void Grounder::addDomains(int id, std::vector<IntVector*>* list)
+{
+	IntVector empty;
+	addDomains(id, list->begin(), list->end(), empty);
+	delete list;
+}
[…]

### 2008-09-05 lib/gringo/src/lparseoutput.cpp

@@ -1,0 +1,62 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "lparseoutput.h"
+#include "globalstorage.h"
+
+using namespace NS_GRINGO;
+using namespace NS_OUTPUT;
+		
+LparseOutput::LparseOutput(std::ostream *out) : Output(out)
+{
+}
+
+void LparseOutput::initialize(GlobalStorage *g, SignatureVector *pred)
+{
+	Output::initialize(g, pred);
+	if(hideAll_)
+		*out_ << "#hide." << std::endl;
+	for(std::map<Signature, bool>::iterator it = hide_.begin(); it  != hide_.end(); it++)
+	{
+		if(hideAll_ != it->second)
+		{
+			*out_ << (hideAll_ ? "#show " : "#hide ") << *g_->getString(it->first.first);
+			if(it->first.second > 0)
+			{
+				*out_ << "(X0";
+				for(int i=1; i < it->first.second; i++)
+					*out_ << ", X" << i;
+				*out_ << ")";
+			}
+			*out_ << "." << std::endl;
+		}
+	}
+}
+
+void LparseOutput::print(NS_OUTPUT::Object *r)
+{
+	r->print_plain(this, *out_);
+}
+
+void LparseOutput::finalize(bool last)
+{
+}
+
+LparseOutput::~LparseOutput()
+{
+}
+

### 2008-09-05 lib/gringo/src/maxaggregate.cpp

@@ -1,0 +1,214 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "maxaggregate.h"
+#include "conditionalliteral.h"
+#include "term.h"
+#include "value.h"
+#include "output.h"
+#include "dlvgrounder.h"
+#include "grounder.h"
+#include "indexeddomain.h"
+#include "variable.h"
+
+using namespace NS_GRINGO;
+
+MaxAggregate::MaxAggregate(ConditionalLiteralVector *literals) : AggregateLiteral(literals)
+{
+}
+
+void MaxAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
+{
+	fact_ = true;
+	lower = INT_MAX;
+	upper = INT_MIN;
+	fixed = INT_MIN;
+	maxUpperBound_ = INT_MAX;
+	minLowerBound_ = INT_MIN;
+	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
+	{
+		ConditionalLiteral *p = *it;
+		p->ground(g, GROUND);
+		for(p->start(); p->hasNext(); p->next())
+		{
+			if(!p->match())
+			{
+				p->remove();
+				continue;
+			}
+			int weight = p->getWeight();
+			if(p->isFact())
+				fixed = std::max(fixed, weight);
+			else
+			{
+				fact_ = false;
+				lower = std::min(lower, weight);
+				upper = std::max(upper, weight);
+			}
+			maxUpperBound_ = std::max(maxUpperBound_, weight);
[…]

### 2008-09-05 lib/gringo/src/minaggregate.cpp

@@ -1,0 +1,215 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "minaggregate.h"
+#include "conditionalliteral.h"
+#include "term.h"
+#include "value.h"
+#include "output.h"
+#include "dlvgrounder.h"
+#include "grounder.h"
+#include "indexeddomain.h"
+#include "variable.h"
+
+using namespace NS_GRINGO;
+
+MinAggregate::MinAggregate(ConditionalLiteralVector *literals) : AggregateLiteral(literals)
+{
+}
+
+void MinAggregate::match(Grounder *g, int &lower, int &upper, int &fixed)
+{
+	fact_ = true;
+	lower = INT_MAX;
+	upper = INT_MIN;
+	fixed = INT_MAX;
+	maxUpperBound_ = INT_MAX;
+	minLowerBound_ = INT_MIN;
+	for(ConditionalLiteralVector::iterator it = literals_->begin(); it != literals_->end(); it++)
+	{
+		ConditionalLiteral *p = *it;
+		p->ground(g, GROUND);
+		for(p->start(); p->hasNext(); p->next())
+		{
+			if(!p->match())
+			{
+				p->remove();
+				continue;
+			}
+			int weight = p->getWeight();
+			if(p->isFact())
+				fixed = std::min(fixed, weight);
+			else
+			{
+				fact_ = false;
+				lower = std::min(lower, weight);
+				upper = std::max(upper, weight);
+			}
+			maxUpperBound_ = std::max(maxUpperBound_, weight);
[…]

### 2008-09-05 lib/gringo/src/multipleargsterm.cpp

@@ -1,0 +1,117 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "multipleargsterm.h"
+#include "literal.h"
+#include "expandable.h"
+#include "value.h"
+#include "grounder.h"
+
+using namespace NS_GRINGO;
+
+MultipleArgsTerm::MultipleArgsTerm(Term *a, Term *b) : a_(a), b_(b)
+{
+}
+
+MultipleArgsTerm::MultipleArgsTerm(const MultipleArgsTerm &r)
+{
+	a_ = a_->clone();
+	b_ = b_->clone();
+}
+
+Term* MultipleArgsTerm::clone() const
+{
+	return new MultipleArgsTerm(*this);
+}
+
+void MultipleArgsTerm::print(const GlobalStorage *g, std::ostream &out) const
+{
+	out << "(" << pp(g, a_) << "; " << pp(g, b_) << ")";
+}
+
+void MultipleArgsTerm::getVars(VarSet &vars) const
+{
+	assert(false);
+}
+
+bool MultipleArgsTerm::isComplex()
+{
+	assert(false);
+}
+
+Value MultipleArgsTerm::getConstValue(Grounder *g)
+{
+	assert(false);
+}
+
+Value MultipleArgsTerm::getValue(Grounder *g)
+{
+	assert(false);
+}
+
+namespace
+{
+	class CloneSentinel : public Term
+	{
+	public:
+		CloneSentinel(Term *a) : a_(a) {}
+		Term* clone() const
+		{
+			Term *a = a_;
+			delete this;
+			return a;
+		}
+		~CloneSentinel() { }
+		// the rest is unused
+		void getVars(VarSet &vars) const { assert(false); }
[…]

### 2008-09-05 lib/gringo/src/pilsoutput.cpp

@@ -1,0 +1,74 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "pilsoutput.h"
+
+using namespace NS_GRINGO;
+using namespace NS_OUTPUT;
+		
+PilsOutput::PilsOutput(std::ostream *out, unsigned int normalForm) : Output(out), normalForm_(normalForm)
+{
+}
+
+const char* HEADER = "1";
+const char* VERSION = "1";
+
+void PilsOutput::initialize(GlobalStorage *g, SignatureVector *pred)
+{
+	Output::initialize(g, pred);
+
+	*out_ << HEADER << " " << "3" << VERSION << " " << normalForm_ << " " << "0 0" << NL;
+}
+
+void PilsOutput::print(NS_OUTPUT::Object *r)
+{
+	r->print(this, *out_);
+}
+
+void PilsOutput::finalize(bool last)
+{
+	if(last)
+	{
+		//write the optimize statement for all added sums
+		if (optimizedIDs_.size())
+		{
+			*out_ << "1d" << " " << optimizedIDs_.size() + 2 << " " << newUid() << " " << "1e" << " " << optimizedIDs_.size();
+			for (IntVector::const_iterator i = optimizedIDs_.begin(); i != optimizedIDs_.end(); ++i)
+			{
+				*out_ << " " << *i;
+			}
+			*out_ << " 0" << NL;
+		}
+		// end of output
+		*out_ << "0 0 0" << NL;
+	}
+}
+
+void PilsOutput::addOptimizedID(unsigned int id)
+{
+	optimizedIDs_.push_back(id);
+}
+
+unsigned int PilsOutput::getNormalForm() const
+{
+	return normalForm_;
+}
+
+PilsOutput::~PilsOutput()
+{
+}
+

### 2008-09-05 lib/gringo/src/program.cpp

@@ -1,0 +1,160 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "program.h"
+#include "statement.h"
+#include "grounder.h"
+#include "literal.h"
+#include "evaluator.h"
+#include "basicprogramevaluator.h"
+
+using namespace NS_GRINGO;
+
+Program::Program(Type type, StatementVector &rules) : type_(type), eval_(0)
+{
+	std::swap(rules, rules_);
+}
+
+Evaluator *Program::getEvaluator()
+{
+	if(!eval_)
+	{
+		switch(type_)
+		{
+			case Program::FACT:
+				eval_ = new Evaluator();
+				break;
+			case Program::BASIC:
+				eval_ = new BasicProgramEvaluator();
+				break;
+			case Program::NORMAL:
+				eval_ = new Evaluator();
+				break;
+		}
+	}
+	return eval_;
+}
+
+void Program::print(const GlobalStorage *g, std::ostream &out) const
+{
+	if(rules_.size() > 0)
+	{
+		switch(type_)
+		{
+			case Program::FACT:
+				out << "% fact program:" << std::endl;
+				break;
+			case Program::BASIC:
+				out << "% basic program:" << std::endl;
+				break;
+			case Program::NORMAL:
+				out << "% normal program:" << std::endl;
+				break;
+		}
+		for(StatementVector::const_iterator it = rules_.begin(); it != rules_.end(); it++)
+		{
+			out << pp(g, *it) << std::endl;;
+		}
+	}
+}
+
+bool Program::check(Grounder *g)
+{
+	VarVector free;
+	LiteralVector unsolved;
+	// conditionals have to be omega restricted
[…]

### 2008-09-05 lib/gringo/src/statementchecker.cpp

@@ -1,0 +1,171 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "statementchecker.h"
+#include "literal.h"
+
+using namespace NS_GRINGO;
+
+StatementChecker::StatementChecker(StatementChecker *parent) : parent_(parent)
+{
+	
+}
+
+void StatementChecker::createSubNode(Literal *l, bool head)
+{
+	providedSub_.push_back(make_pair(delayed_.back().needed, make_pair(StatementChecker(this), delayed_.back().provided)));
+	l->createNode(&providedSub_.back().second.first, head, false);
+}
+
+const VarSet &StatementChecker::getVars() const
+{
+	return vars_;
+}
+
+bool StatementChecker::hasVars() const
+{
+	return vars_.size() > 0;
+}
+
+bool StatementChecker::check()
+{
+	// also create the delayed nodes
+	while(delayed_.size() > 0)
+	{
+		delayed_.back().l->createNode(this, delayed_.back().head, true);
+		delayed_.pop_back();
+	}
+
+	bool r1, r2;
+	do
+	{
+		r1 = r2 = false;
+		bool f = providedIf_.size() > 0;
+		while(f)
+		{
+			f = false;
+			std::list<std::pair<VarSet, VarSet> >::iterator di, i = providedIf_.begin();
+			while(i != providedIf_.end())
+			{
+				di = i++;
+				VarSet::iterator dk, k = di->first.begin();
+				while(k != di->first.end())
+				{
+					dk = k++;
+					if(provided_.find(*dk) != provided_.end())
+						di->first.erase(dk);
+				}
+				if(di->first.size() == 0)
+				{
+					provided_.insert(di->second.begin(), di->second.end());
[…]

### 2008-09-05 lib/gringo/src/statementdependencygraph.cpp

@@ -1,0 +1,311 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "statementdependencygraph.h"
+#include "predicateliteral.h"
+#include "domain.h"
+#include "grounder.h"
+#include "program.h"
+
+using namespace NS_GRINGO;
+		
+// ==================== SDG::SCC ============================
+SDG::SCC::SCC() : type_(FACT), edges_(0)
+{
+}
+
+// ======================= SDG ==============================
+
+SDG::SDG() : last_(0)
+{
+}
+
+SDGNode *SDG::createStatementNode(Statement *r, bool preserveOrder)
+{
+	SDGNode *n = new SDGNode(r);
+	ruleNodes_.push_back(n);
+	if(preserveOrder)
+	{
+		if(last_)
+			n->addDependency(last_);
+		last_ = n;
+	}
+	return n;
+}
+
+SDGNode *SDG::createPredicateNode(PredicateLiteral *pred)
+{
+	if((int)predicateNodes_.size() < pred->getUid() + 1)
+		predicateNodes_.resize(pred->getUid() + 1);
+	SDGNode *&n = predicateNodes_[pred->getUid()];
+	if(!n)
+		n = new SDGNode(pred->getDomain());
+	return n;
+}
+
+namespace
+{
+	struct Call
+	{
+		Call(SDGNode *v, SDGNodeVector::iterator w) : v(v), root(true), w(w)
+		{
+		}
+
+		SDGNode *v;
+		bool root;
+		SDGNodeVector::iterator w;
+	};
+}
+
[…]

### 2008-09-05 lib/gringo/src/variable.cpp

@@ -1,0 +1,98 @@
+// Copyright (c) 2008, Roland Kaminski
+//
+// This file is part of GrinGo.
+//
+// GrinGo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// GrinGo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "variable.h"
+#include "grounder.h"
+
+using namespace NS_GRINGO;
+		
+Variable::Variable(Grounder *g, int id) : id_(id), uid_(g->registerVar(id))
+{
+}
+
+void Variable::getVars(VarSet &vars) const
+{
+	vars.insert(uid_);
+}
+
+bool Variable::isComplex()
+{
+	return false;
+}
+
+void Variable::print(const GlobalStorage *g, std::ostream &out) const
+{
+	out << *g->getString(id_);
+}
+
+Value Variable::getConstValue(Grounder *g)
+{
+	assert(false);
+}
+
+Value Variable::getValue(Grounder *g)
+{
+	return g->getValue(uid_);
+}
+
+int Variable::getUID()
+{
+	return uid_;
+}
+
+Variable::~Variable()
+{
+}
+
+void Variable::preprocess(Literal *l, Term *&p, Grounder *g, Expandable *e)
+{
+}
+
+Variable::Variable(const Variable &c) : id_(c.id_), uid_(c.uid_)
+{
+}
+
+Term* Variable::clone() const
+{
+	return new Variable(*this);
+}
+
+void  Variable::addIncParam(Grounder *g, Term *&p, const Value &v)
+{
+}
+
+bool Variable::unify(const GlobalStorage *g, const Value& t, const VarVector& vars, ValueVector& vals) const
+{
+	// TODO: constant access would be nice
+	for (size_t i = 0; i < vars.size(); ++i)
+	{
+		// if variable is bound
+		if (vars[i] == uid_)
+		{
+			// variable is not set, then set, else compare
+			if (vals[i].type_ == Value::UNDEF)
+			{
[…]

### 2008-09-08 doc/documentation/paper.tex
testing svn version information in paper.tex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@504 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \usepackage{color}
+\usepackage[nofancy]{svninfo}
 
 \author{Max Ostrowski}
-\title{Gringo 2.0\\--Draft--}
+\title{Using gringo, clingo and iclingo\\--Draft--\\Rev.\svnInfoRevision}
 \date{\today}
@@
 \begin{document}
+\svnInfo $Id$
 \maketitle
+
 \tableofcontents
@@
 \end{figure}
-gringo2.0 is a state of the art grounder and has many features that makes writing logic programs easy.
-It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0\cite{gringopaper}.
+Gringo is a state of the art grounder and has many features that makes writing logic programs easy.
+It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.0\cite{gringopaper}.
 Therefore it supports binder splitting, backjumping and all other features already implemented in gringo1.0.
@@
 The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.
+\paragraph{Versions}
+This paper references the programs with their following version numbers:
+\begin{itemize}
+ \item \texttt{gringo\ \ 2.0.0}
+ \item \texttt{clingo\ \ 1.0.0}
+ \item \texttt{iclingo\ 1.0.0}
+\end{itemize}
 

### 2008-09-08 doc/documentation/paper.tex
Doku: Added abstract, commentstyle, bugfixes, ready for review
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@505 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \usepackage[nofancy]{svninfo}
+\usepackage{comments}
 
 \author{Max Ostrowski}
-\title{Using gringo, clingo and iclingo\\--Draft--\\Rev.\svnInfoRevision}
+\title{Using gringo, clingo and iclingo
+\\\small{Rev.\svnInfoRevision}
+}
 \date{\today}
@@
 \maketitle
+\begin{center}\stamp\end{center}
+%\includecomment{comment}
+%\excludecomment{notthisone}
+
+\abstract{}
+This paper introduces the new tools gringo, clingo and iclingo of the potassco project\cite{download}.
+Since the variety of programs for grounding logic programs with first order logic variables is low, a new grounding front-end for ASP was developed.
+This front-end called gringo, has a well defined input language described in this paper and outdoes former approaches in speed, memory consumption and usability.
+Clingo is a compound of gringo and clasp, a full blown answerset solver.
+Iclingo is a program capable of incremental solving.
+All three tools are introduced and their use is recorded.
 
@@
 It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.0\cite{gringopaper}.
-Therefore it supports binder splitting, backjumping and all other features already implemented in gringo1.0.
+Therefore it supports binder splitting, backjumping and all other features already implemented in gringo1.0.0.
 A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder supports such a rich featured language.
-The goal of this work is to give an overview of the features supported by gringo2.0 and its derivates.
+The goal of this work is to give an overview of the features supported by gringo and its derivates.
 Three different packages of gringo exist.
-The first version is gringo2.0, it acts as a grounder and supports several output formats and even a human readable format for debugging purposes.
+The first version is gringo, it acts as a grounder and supports several output formats and even a human readable format for debugging purposes.
 This version can ground logic programs and output them to files or pipe them to a solver.
@@
 The third package is called iclingo, which stands for ``incremental clingo''.
-In this version the grounder gringo2.0 can be used with clasp incrementally.
+In this version the grounder gringo can be used with clasp incrementally.
 Detailed information about incremental solving can be found in \cite{gekaosscth08a}.
@@
 After that, a short introduction of how to write a logic program and how to write an incremental logic program is given.
-In the next chapter a brief description of how to use the three programs gringo2.0, clingo and iclingo and their options for the command line is given.
+In the next chapter a brief description of how to use the three programs gringo, clingo and iclingo and their options for the command line is given.
 The paper closes with an outlook of future work that has to be done in the area of grounding logic programs.
-The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.
+The syntax of the gringo input language can be found in the appendix \ref{appendix}.
 \paragraph{Versions}
-This paper references the programs with their following version numbers:
+This paper references the programs with the following version numbers:
 \begin{itemize}
@@
 \label{domainrestricted}
-There exist two major grounders for ASP besides gringo1.0, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
+There exist two major grounders for ASP besides gringo, lParse \cite{lparseManual} and  the DLV grounding module\cite{DLV}.
 To ensure the existence of an equivalent ground instantiation of a program both grounders except a subclass of logic programs.
 For lParse the logic programs must be $\omega$-restricted\cite{omega} while DLV can handle safe programs\cite{safe}.
-gringo2.0 verifies that the input program is $\lambda$-
+Gringo verifies that the input program is $\lambda$-
 restricted \cite{gringopaper}, so that the existence of a finite equivalent ground instantiation is guaranteed. 
@@
 To give a gross impression of what $\lambda$-restriction means for writing a logic program an informal description is given.
-In gringo2.0 every variable used in a rule must have a complete domain.
+In gringo every variable used in a rule must have a complete domain.
 This means that this variable has to occur in at least one predicate whose domain is completed.
@@
 To ease the process of writing logic programs some advanced features in the input syntax are introduced and are now explained in detail.
-For the most problems it is possible to write a much shorter version of the program with the extended syntax.
+For the most problems it is possible to write a much shorter and sometimes even faster version of the program with the extended syntax.
 %Even a speedup in solving the instances of the problem can happen TODO ?
@@
 \paragraph{Function Symbols}
-Function symbols are fully supported with gringo2.0.
+Function symbols are fully supported with gringo.
 They can be finitely nested and also be used as domain predicates.
-While in lParse function symbols are allowed in a very restricted way, no restrictions apply in gringo2.0.
-The following program can not be grounded with lParse since $X$ is an unrestricted variable. With gringo2.0 the program is now groundable.
+While in lParse function symbols are allowed in a very restricted way, no restrictions apply in gringo.
+The following program can not be grounded with lParse since $X$ is an unrestricted variable. With gringo the program is now groundable.
 \begin{verbatim}
@@
 \end{verbatim}
-will results in the rules\footnote{gringo2.0 would simplify the grounded rules since some literals in the body are facts}
+will results in the rules\footnote{gringo would simplify the grounded rules since some literals in the body are facts}
 \begin{verbatim}
@@
 \subparagraph{Relational Operators}
-The relational operators are ``less than'' ($<$),\\ ``greater than'' ($>$), ``less equal'' ($<=$), ``greater equal'' ($>=$) , ``equal'' ($==$) and ``not equal'' ($!=$).
+The relational operators are ``less than'' ($<$), ``greater than'' ($>$), ``less equal'' ($<=$), ``greater equal'' ($>=$) , ``equal'' ($==$) and ``not equal'' ($!=$).
 They can be used to constrain certain variables.
@@
 Using disjunctions in logic programs can increase expressiveness and complexity of the logic program.
-A solver for disjunctive logic programs such as claspD \cite{claspD} or cmodels \cite{cmodels} is needed to solve such programs.
+A solver for disjunctive logic programs such as claspD \cite{claspD} or cmodels \cite{cmodels} is needed to solve such programs. Clingo is currently not able to solve disjunctive programs.
 
@@
 If squared brackets are used, the atom list is interpreted as a multiset, otherwise, for curly brackets, as a set of atoms.
-The lower and upper bound is optional, and can be omitted.
+The lower and upper bound is optional, and can be omitted. The supported aggregates are
+\begin{verbatim}
+ 0 count {a, b, not c} 2.
+ 1 sum {a=1, b=3, not c=-2} 2.
+ 0 max {a=1, b=2, not c=3} 5.
+ 1 min {a=2, b=4, not c=1} 3.
+\end{verbatim}
+and are now explained in detail.
 
@@
 The hide construct is used to hide atoms from the output of the ASP solver.
-Two different syntaxes are supported by gringo2.0.
+Two different syntaxes are supported by gringo.
 The first 
@@
 \paragraph{comments}
-Two kinds of comments were supported by gringo2.0.
+Two kinds of comments were supported by gringo.
 Line based comments start with a ``\texttt{\%}''.
@@
 The gringo tools can be started from the command line and have some parameters that can be useful.
-All programs are fully backward compatible meaning, all options of gringo2.0 are valid in clingo and iclingo, and all options of clingo are valid in iclingo.
-\subsection{Gringo2.0 Options}
-Gringo2.0 can be called from the command line via
+All programs are fully backward compatible meaning, all options of gringo are valid in clingo and iclingo, and all options of clingo are valid in iclingo.
+\subsection{Gringo Options}
+Gringo can be called from the command line via
 \begin{verbatim}
@@
 \paragraph{\texttt{--text,-t}}
-Parameter ``\texttt{--text}'' invokes the text output method, so the output is printed in human readable format, which is also the input format for gringo2.0.
+Parameter ``\texttt{--text}'' invokes the text output method, so the output is printed in human readable format, which is also the input format for gringo.
 The output is variable free and some simplifications on the program were made.
@@
 This mode prints lParse conform output. For a definition of the output format see \cite{lparseManual}.
-Since some features of gringo2.0 are not natively supported for this output language, new atoms may be introduced to create an equivalent program.\newline
+Since some features of gringo are not natively supported for this output language, new atoms may be introduced to create an equivalent program.\newline
 Sum aggregates with negative weights are handled like weight constraint in lParse for this output format.
@@
 For converting an already grounded logic program into an output format 
-supported by gringo2.0 this is the right option.
+supported by gringo this is the right option.
  This mode saves a lot of memory but does not ground the program.
 \paragraph{\texttt{--const,-c C=v}}
-You can set the value of constants defined in your logic program via the command line of gringo2.0.
+You can set the value of constants defined in your logic program via the command line of gringo.
 After the parameter ``--const'' you have to give the name of the constant $C$ you want to set followed by a ``$=$'' and the value of that constant.
@@
 \paragraph{\texttt{--bindersplitting=<arg>}}
-With this option you can enable or disable the binder splitting used by gringo2.0
+With this option you can enable or disable the binder splitting used by gringo.
 \begin{itemize}
@@
 \paragraph{\texttt{--ifixed=<num>}}
-An incremental program is required as input. Gringo2.0 grounds the fixed number of steps of the incremental program.
+An incremental program is required as input. Gringo grounds the fixed number of steps of the incremental program.
 
@@
 If no filename is given, the standard input is used.
-Since clingo uses gringo2.0 as a grounder in combination with the solver clasp, all options from gringo2.0 and clasp are valid.
+Since clingo uses gringo as a grounder in combination with the solver clasp, all options from gringo and clasp are valid.
 For a detailed list of the clasp options see the clasp help (also available with \texttt{clingo --help}).
@@
 Iclingo is used for incremental solving. For a detailed description see \cite{gekaosscth08a}.
-It uses the grounder gringo2.0 and the solver clasp, and therefore all options of gringo2.0, clingo and clasp are valid.
+It uses the grounder gringo and the solver clasp, and therefore all options of gringo, clingo and clasp are valid.
 Additionally some options for controlling the incremental behavior are implemented.
@@
 \subsection{Warnings}
-Several warnings are produced by gringo2.0 to give hints on common mistakes.
+Several warnings are produced by gringo to give hints on common mistakes.
 They are all printed to standard error output.
@@
 \subsection{Errors}
-gringo2.0 stops the execution and prints an error message to standard error output.
+Gringo stops the execution and prints an error message to standard error output.
 \begin{itemize}
@@
 To support an even less restricted input language the support for $\Pi$-restricted programs is planned in a future release of gringo.
+For clingo, the results of research from claspD will lead to a new version of clingo, capable of handling disjunctive logic programs.
 All these features will lead to more acceptance of ASP-systems as the programs become easier to write and the bottleneck of ASP, the grounding part, will be faster.
 Furthermore the acceptance of a standardized intermediate format for ASP is necessary.
-gringo2.0 supports the ASPils output format, but many more tools for converting, shuffling and merging must be developed.
+Gringo supports the ASPils output format, but many more tools for converting, shuffling and merging must be developed.
 \newpage

### 2008-09-09 app/options.h
added --syntax option, Doku: added --syntax, fixed aggregate bug, TODO: corrent homepage url
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@508 3e02f596-3131-0410-8352-e04eeaec37cc

@@ -72,6 +72,7 @@ public:
 	bool             version;          // Defailt: false
 	bool             stats;            // Default: false
 	bool             verbose;          // Default: false
+	bool             syntax;           // Default: false
 	std::vector<std::string> files;    // Default: "" -> read from stdin
 
 	// gringo stuff
@@ -126,6 +127,7 @@ public:
 private:
 	void initOptions(ProgramOptions::OptionGroup& allOpts, ProgramOptions::OptionGroup& hidden);
 	void checkCommonOptions(const ProgramOptions::OptionValues&);
+	void printSyntax(std::ostream& os) const;
 	void printHelp(const ProgramOptions::OptionGroup& opts, std::ostream& os) const;
 	void printVersion(std::ostream& os) const;
 #ifdef WITH_CLASP

### 2008-10-06 doc/documentation/paper.tex
added --syntax option, Doku: added --syntax, fixed aggregate bug, TODO: corrent homepage url
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@508 3e02f596-3131-0410-8352-e04eeaec37cc
DOKU: updated documentation, all authors named, fixed bug in Conjunctor
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@522 3e02f596-3131-0410-8352-e04eeaec37cc
updated outdated warnings
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@526 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \usepackage[nofancy]{svninfo}
-\usepackage{comments}
-
-\author{Max Ostrowski}
+%\usepackage{comments}
+
+\author{Martin Gebser, Roland Kaminski, Benjamin Kaufmann,\\ Max Ostrowski, Torsten Schaub, Sven Thiele}
 \title{Using gringo, clingo and iclingo
@@
 \maketitle
-\begin{center}\stamp\end{center}
+%\begin{center}\stamp\end{center}
 %\includecomment{comment}
@@
 \begin{verbatim}
- a(foo,1) :- b(foo), b(a), b(b).
- a(foo,2) :- b(foo), b(a), b(b).
- a(bar,1) :- b(bar), b(a), b(b).
- a(bar,2) :- b(bar), b(a), b(b).
+ a(foo) :- b(foo), b(a), b(b).
+ a(bar) :- b(foo), b(a), b(b).
+ a(1)   :- b(bar), b(a), b(b).
+ a(2)   :- b(bar), b(a), b(b).
 \end{verbatim}
@@
  0 count {a, b, not c} 2.
- 1 sum {a=1, b=3, not c=-2} 2.
- 0 max {a=1, b=2, not c=3} 5.
- 1 min {a=2, b=4, not c=1} 3.
+ 1 sum [a=1, b=3, not c=-2] 2.
+ 0 max [a=1, b=2, not c=3] 5.
+ 1 min [a=2, b=4, not c=1] 3.
 \end{verbatim}
@@
 Prints additional information during processing.
+\paragraph{\texttt{--syntax}}
+Prints a short description of the syntax of the input language.
 \paragraph{\texttt{--stats}}
@@
 \begin{verbatim}
- "Warning: There are no #base, #lambda or #delta sections."
+ "Warning: There are no #base, #cumulative or #volatile sections."
 \end{verbatim}
@@
 \begin{verbatim}
- "Warning: There are statements not within a #base, #lambda"
- "         or #delta section."
+ "Warning: There are statements not within a #base, #cumulative"
+ "         or #volatile section."
  "         These Statements are put into the #base section."

### 2008-10-06 lib/gringo/src/grounder.cpp

@@ -164,7 +164,7 @@ void Grounder::prepare(bool incremental)
 	if(incremental_ && incParts_.size() == 0)
 	{
 		incParts_.push_back(make_pair(std::make_pair(BASE, Value()), rules_.size()));
-		std::cerr << "Warning: There are no #base, #lambda or #delta sections." << std::endl;
+		std::cerr << "Warning: There are no #base, #cumulative or #volatile sections." << std::endl;
 	}
 
 	preprocess();
@@ -218,7 +218,7 @@ void Grounder::setIncPart(IncPart part, const Value &v)
 	if(incParts_.size() == 0 && rules_.size() > 0)
 	{
 		incParts_.push_back(make_pair(std::make_pair(BASE, Value()), rules_.size()));
-		std::cerr << "Warning: There are statements not within a #base, #lambda or #delta section." << std::endl;
+		std::cerr << "Warning: There are statements not within a #base, #cumulative or #volatile section." << std::endl;
 		std::cerr << "         These Statements are put into the #base section." << std::endl;
 	}
 	incParts_.push_back(make_pair(std::make_pair(part, v), 0));

### 2008-11-03 doc/documentation/background.tex
unverified new version of guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@579 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Background}\label{sec:background}

In Section~\ref{subsec:asp},
we give a brief introduction to Answer Set Programming,
which constitutes the basic framework of the tools we describe here.
This framework deals with logic programs,
whose syntax and semantics are formally introduced in
Section~\ref{subsec:syntax} and~\ref{subsec:semantics}, respectively.
We invite the reader to merely skim or even skip these two sections
upon the first reading, and to rather look them up later on if
formal details are requested.
However, we want to stress that the syntax we use for logic programs
admits (uninterpreted) function symbols of non-zero arity,
whose full support by our tools is rather exceptional and thus innovative.

\input{figures/asp}
\subsection{Answer Set Programming}\label{subsec:asp}

Answer Set Programming (ASP)
\cite{baral03a,gelleo02a,lifschitz02a,martru99a,niemela99a}
emerged in the late 1990s as a declarative paradigm
for modeling and solving search problems.
As illustrated in Figure~\ref{fig:asp},
the basic approach of ASP is to represent a problem as a logic program~$\Pi$
such that particular Herbrand models of~$\Pi$, called answer sets,
correspond to problem solutions.
In contrast to traditional logic programming languages,
e.g., Prolog~\cite{nilmal95a},
ASP programs are sets (not lists) of rules,
and computations are oriented towards models (not proofs).
Let us illustrate this on an example.

\begin{example}\label{ex:flies}
Consider a logic program comprising the following facts:
%
\lstinputlisting{examples/bird.lp}
% \listinginput{1}{examples/bird.lp}
%
These facts express that constants \const{tux} and \const{tweety}
stand for birds.
Furthermore, \const{tux} is a penguin, 
and \const{tweety} is a chicken.
%
Note that the \emph{unique names assumption} applies, that is,
$\const{tux}\neq\const{tweety}$ holds.
Besides the two constants,
the language of the above facts contains predicates
\pred{bird}/$1$, \pred{penguin}/$1$, and \pred{chicken}/$1$
(each having arity~$1$).
%
We next augment the above facts with the following rules:%
\marginlabel{%
  Instead of \code{neg\mus flies(X)}, one could write
  \code{-flies(X)} in order to make explicit that
  instances amount to the classical (or strong) negation of
  \code{flies(X)}.}
%
\lstinputlisting[firstnumber=3]{examples/fly.lp}
%
Informally, such rules correspond to implications where the
left hand side of connective ``\code{:-}'' (called head)
is derived if all premises on the right hand side
(their conjunction called body) hold.
Connective ``\code{not}'' stands for \emph{default negation},
and uppercase letter~\var{X} is the name of a first-order variable.
Variables are local to rules, that is,
a variable name refers to a different variable in a distinct rule.
Finally, every variable is universally quantified and thus
applies to all terms in the language of a logic program.

The rule in Line~3 formalizes that any instance of~\var{X} that is
a bird \pred{flies}, as long as \pred{neg\mus flies} does not hold.
The converse relationship between \pred{neg\mus flies} and \pred{flies}
is expressed in Line~4.
At this point, observe that, though syntactically correct,
the above program ``makes no sense'' in Prolog because a query%\hspace{\listingoffset}
% \begin{alltt}
% ?- flies(tweety).
% \end{alltt}
\begin{lstlisting}[numbers=none]
?- flies(tweety).
\end{lstlisting}
can be answered neither positively nor negatively.
In ASP, however, the rules in Line~3 and~4 admit multiple belief states,
depending on whether \pred{flies} or \pred{neg\mus flies} is assumed for a bird.
In addition, the rule in Line~5 requires \pred{neg\mus flies} to hold for every penguin.
Combining this knowledge with the facts in Line~1 and~2
makes us derive \code{\pred{neg\mus flies}(\const{tux})},
and it provides us with two alternatives for \const{tweety}:
either \code{\pred{flies}(\const{tweety})} or \code{\pred{neg\mus flies}(\const{tweety})} holds.
These alternatives cannot jointly hold because, under such an assumption,
neither of them is derived, thus,
violating the justification principle of ASP (cf.\ Section~\ref{subsec:semantics}).
Without already going into formal details,
we conclude that the above program has two answer sets:
one according to which \const{tweety} flies and one where \const{tweety} does not fly.
\eexample
\end{example}

\input{figures/system}
The computation of answer sets usually consists of two phases.
First, a \emph{grounder} substitutes the variables in a logic program~$\Pi$
by variable-free terms, resulting in a propositional program~$\alt{\Pi}$.
Second, the answer sets of~$\alt{\Pi}$ are computed by a \emph{solver}.
This prototypical architecture is visualized in Figure~\ref{fig:system}.
Of course, a ground program~$\alt{\Pi}$ must be finite and equivalent to
input program~$\Pi$.
To this end, grounders like \lparse~\cite{lparseManual},
the one embedded into \dlv~\cite{dlv03a},
and \gringo\footnotemark~\cite{gescth07a}
impose certain restrictions on~$\Pi$ (cf.\ Section~\ref{subsec:lambda}).
An obtained ground program~$\alt{\Pi}$ can be piped into a solver, such as 
\assat~\cite{linzha04a}, 
\clasp\footnotemark[\value{footnote}]~\cite{gekanesc07b},
\cmodels~\cite{gilima06a},
\nomorepp~\cite{angelinesc05c},
\smodels~\cite{siniso02a}, or
\smodelscc~\cite{warsch04a},
in order to compute answer sets of~$\alt{\Pi}$, matching the ones of~$\Pi$.
All of the aforementioned solvers deal with \lparse's output format,
a numerical representation of~$\alt{\Pi}$,
while \dlv\ couples its grounding and its solving component directly
via an internal interface,
an approach also realized by \clingo\footnotemark[\value{footnote}].
Finally, \iclingo\footnotemark[\value{footnote}]~\cite{gekakaosscth08a} integrates and interleaves
grounding and solving within incremental computations.
%
%\addtocounter{footnote}{-1}%
\footnotetext{Described in this document.}

The next example illustrates the computation of answer sets for
the logic program given in Example~\ref{ex:flies}.

\begin{example}\label{ex:flies:ground}
Reconsidering the facts in Line~1 and~2 of Example~\ref{ex:flies},
we identify two constants: \const{tux} and \const{tweety}.
They can be substituted for variables~\var{X} in the rules in Line~3--5.
In fact, Line~1--5 of Example~\ref{ex:flies} are a shorthand for the ground instantiation:
%
\lstinputlisting[nolol]{examples/bird.lp}
\lstinputlisting[firstnumber=3]{examples/bfall.lp}
%
Note that the facts in Line~1 and~2 are unaltered from Example~\ref{ex:flies},
while the ground rules in Line~3--8 are obtained by instantiating variables~\var{X}.
We have that the answer sets of the above ground program match
those of the original input program in Example~\ref{ex:flies}.

In practice, grounders try to avoid producing the full ground instantiation
of an input program by applying answer-set preserving simplifications.
In particular, facts can (recursively) be eliminated.
Thus, \gringo\ computes the following ground program:%
\marginlabel{%
  The ground program in text format is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\
        \mbox{~}examples/bird.lp \textbackslash\\
        \mbox{~}examples/fly.lp}}
%
\lstinputlisting[nolol]{examples/bird.lp}
\lstinputlisting[firstnumber=3]{examples/bfeff.lp}
%
Observe that \code{\pred{neg\mus flies}(\const{tux})} has become a fact because
\code{\pred{penguin}(\const{tux})} is known.
Similarly, \code{\pred{bird}(\const{tux})} and \code{\pred{bird}(\const{tweety})}
have been removed from bodies of rules, making use of the fact that ``\textit{true} \& something'' is
equivalent to ``something.''
Applying the complementary principle that ``\textit{false} \& something'' evaluates to ``\textit{false},''
the rule with \code{not \pred{neg\mus flies}(\const{tux})} in the body has been dropped completely.
After performing these simplifications, the ground program computed by \gringo\ is still
equivalent to the input program in Example~\ref{ex:flies}, viz.,
there is one answer set such that \const{tweety} flies and one where \const{tweety} does not fly.
In fact, running \clasp\ yields:%
\marginlabel{%
  The two answer sets are computed by piping the output of \gringo\
  into \clasp:\\
  \code{\mbox{~}gringo \textbackslash\\
        \mbox{~}examples/bird.lp \textbackslash\\
        \mbox{~}examples/fly.lp | \textbackslash\\
        \mbox{~}clasp -n 0}\\
  Note that \gringo\ is called without option \code{-t},
  while option \code{-n 0} makes \clasp\ compute all answer sets
  (rather than just one, as done by default).

  Alternatively, one may use \clingo\ to compute the answer sets.
  To this end, invoke:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
        \mbox{~}examples/bird.lp \textbackslash\\
        \mbox{~}examples/fly.lp}}
%
\begin{lstlisting}[numbers=none,escapechar=@]
Answer: 1
bird(tux) penguin(tux) bird(tweety) chicken(tweety) @\textbackslash@
neg_flies(tux) flies(tweety)
Answer: 2
bird(tux) penguin(tux) bird(tweety) chicken(tweety) @\textbackslash@
neg_flies(tux) neg_flies(tweety)
\end{lstlisting}
%
Note that the order of the answer sets is incidental and 
not obliged to the order of rules and bodies in the input,
as the semantics of ASP programs is purely declarative.
\eexample
\end{example}

We have seen how an ASP system computes answer sets of a search problem
represented by a set of facts (Line~1 and~2 in Example~\ref{ex:flies})
and a set of schematic rules with first-order variables
(Line~3--5 in Example~\ref{ex:flies}).
This separation of a problem into a knowledge part, called \emph{encoding},
and a data part, called \emph{instance}, is not a coincidence
but a common methodology in ASP \cite{martru99a,niemela99a,schlipf95a},
sometimes called \emph{uniform} definition.
In fact, the possibility to describe problems in a uniform way is
a major advantage of ASP, as it promotes simplicity and flexibility in
knowledge representation.
Together with the availability of efficient reasoning engines,
this makes ASP an attractive and powerful paradigm
for declarative problem solving.
By explaining and illustrating the functionalities of
grounder \gringo, solver \clasp, and their bondings in \clingo\ and \iclingo,
this document aims at enabling the reader to make (better) use of ASP.

\subsection{Syntax of Logic Programs}\label{subsec:syntax}

This section gives a brief account of the formal syntax of logic programs,
needed for defining their semantics.
Consult, e.g., \cite{lloyd87a} for a detailed introduction.
The language of logic programs is composed from sets
%
\begin{itemize}
\item $\mathcal{P}=\{p_1/i_1,\dots,p_m/i_m\}$ of \emph{predicates}
      (with arities $i_1,\dots,i_m$),
\item $\mathcal{F}=\{f_1/j_1,\dots,f_n/j_n\}$ of \emph{functions}
      (with arities $j_1,\dots,j_n$), and
\item $\mathcal{V}=\{V_1,V_2,V_3,\dots\}$ of \emph{variables}
\end{itemize}
along with connectives (``\code{:-},'' ``\code{,},'' ``\code{not},'' etc.) and
punctuation symbols (``\code{(},'' ``\code{)},'' ``\code{.},'' etc.).
A function~$f/j$ is called a \emph{constant} if $j=0$, that is,
if~$f$ has no arguments.
The set~$\mathcal{T}$ of \emph{terms} is the smallest set containing~$\mathcal{V}$
and all expressions $f(t_1,\dots,t_k)$ such that $f/k$ is a function 
from~$\mathcal{F}$ and $t_1,\dots,t_k$ are terms in~$\mathcal{T}$.
Note that every variable and every constant is an elementary term,
while functions with non-zero arity induce compound terms.
By $\ground{\mathcal{T}}$, we denote the set of all variable-free
terms in~$\mathcal{T}$.
Note that $\ground{\mathcal{T}}$ is infinite
as soon as~$\mathcal{F}$ contains at least one constant and one function
with non-zero arity.
The set~$\mathcal{A}$ of \emph{atoms} consists of 
$\top$ (``\textit{true}''), $\bot$ (``\textit{false}''), and all expressions
$p(t_1,\dots,t_k)$ such that $p/k$ is a predicate
from~$\mathcal{P}$ and $t_1,\dots,t_k$ are terms in~$\mathcal{T}$.
As with terms, $\ground{\mathcal{A}}$ denotes the set of all variable-free
atoms in~$\mathcal{A}$.

A \emph{rule}~$r$ is an expression of the form $\mathit{Head}~\code{:-}~\mathit{Body}\code{.}$
where $\mathit{Head}$ and $\mathit{Body}$ are composed from atoms in $\mathcal{A}$
(other than $\bot$ and $\top$), punctuation symbols, and connectives other than ``\code{:-}.''
The intuitive reading of a rule is that $\mathit{Head}$ follows from $\mathit{Body}$,
and either $\mathit{Body}$ or $\mathit{Head}$ may be omitted, in which case the
rule is called a \emph{fact} or an \emph{integrity constraint}, respectively.
A fact is written in the form $\mathit{Head}\code{.}$ and understood as 
$\mathit{Head}~\code{:-}~\top\code{.}$, and an integrity constraint
$\code{:-}~\mathit{Body}\code{.}$ is a shorthand for $\bot~\code{:-}~\mathit{Body}\code{.}$
By $\vars{r}$, we denote the set of all variables in rule~$r$,
and~$r$ is called \emph{ground} if $\vars{r}=\emptyset$.
A \emph{ground substitution} is a mapping $\sigma:\vars{r}\rightarrow\ground{\mathcal{T}}$,
and~$r\sigma$ is the ground rule obtained by replacing every occurrence of a
variable $V\in\vars{r}$ with~$\sigma(V)$.
By $\ground{r}$, we denote the set of ground rules~$r\sigma$
obtained from~$r$ by applying all possible ground substitutions~$\sigma$.
A \emph{logic program}~$\Pi$ is a set of rules,
and the \emph{ground instantiation} of~$\Pi$ is
$\ground{\Pi}=\bigcup_{r\in\Pi}\ground{r}$.
By convention, we assume that~$\mathcal{P}$ and~$\mathcal{F}$ are the sets
of all predicates and functions, respectively, that occur in~$\Pi$.
Note that $\ground{\Pi}$ is infinite if
there is at least one rule $r\in\Pi$ such that $\vars{r}\neq\emptyset$ and 
if~$\mathcal{F}$ contains at least one constant and one function
with non-zero arity.

For a well-known class of logic programs, called \emph{normal},
rules are of the form:
\begin{equation}\label{eq:normal:rule}
  A_0~\code{:-}~A_1\code{,}\dots\code{,}A_m\code{,}
  \code{not}~A_{m+1}\code{,}\dots\code{,}\code{not}~A_n\code{.}
\end{equation}
%
For $0\leq k\leq n$, every~$A_k$ is an atom, that is, the head is an atom and
the body a (possibly empty) conjunction of atoms that may be preceded by ``\code{not}.''
As mentioned in Section~\ref{subsec:asp},
the order of elements in the body is not essential,
so that they can be regrouped to match the form in~(\ref{eq:normal:rule}).
For a rule~$r$ as in~(\ref{eq:normal:rule}), we define 
$\head{r}=A_0$, $\pbody{r}=\{A_1,\dots,A_m\}$, and $\nbody{r}=\{A_{m+1},\dots,A_n\}$.
If $r=A_0\code{.}$ is a fact, then
$\pbody{r}=\nbody{r}=\nolinebreak\emptyset$.
The particular structure of normal programs admits a rather intuitive 
reformulation of their semantics in the next section.

\begin{example}\label{ex:syntax}
The language of the normal program in Example~\ref{ex:flies} includes
predicates
\(
\mathcal{P}=
\{
\pred{bird}/1,\pred{penguin}/1,\pred{chicken}/1,
\pred{flies}/1,\pred{neg\mus flies}/1\}
\)
and functions
\(
\mathcal{F}=
\{
\const{tux}/0,\const{tweety}/0
\}
\).
As the arity of the latter is~$0$,
terms \const{tux} and \const{tweety} are constants.
By substituting them for variables,
we obtain the ground instantiation given in Example~\ref{ex:flies:ground}.
For instance, the ground rules
%
\begin{lstlisting}[numbers=none]
flies(tux)    :- bird(tux),    not neg_flies(tux).
flies(tweety) :- bird(tweety), not neg_flies(tweety).
\end{lstlisting}
%
are obtained from:
%
\begin{lstlisting}[numbers=none]
flies(X) :- bird(X), not neg_flies(X).
\end{lstlisting}
%
Observe that the three occurrences of variable name~\var{X}
stand for a single variable, so that substitutions
$\{\var{X}\mapsto\const{tux}\}$ and $\{\var{X}\mapsto\const{tweety}\}$
induce the above ground rules.
Notably, sets~$\mathcal{P}$ and~$\mathcal{F}$ of predicates and functions
need not be disjoint because atoms and terms are also distinguished
by the context in which they occur,
and a predicate or function symbol may be used with different arities.
For instance, we could add a rule
%
\begin{lstlisting}[numbers=none]
flies :- flies(X).
\end{lstlisting}
%
over predicates \pred{flies}/$0$ and \pred{flies}/$1$. 
In fact, such rules are not unusual to express that there is some object
with certain properties, while it does not matter which object it is.
Finally, note that an infinite ground instantiation would result from adding rule
\begin{lstlisting}[numbers=none]
bird(parent(X)) :- bird(X).
\end{lstlisting}
in view of function \const{parent}/$1$ with non-zero arity.
\eexample
\end{example}

\subsection{Semantics of Logic Programs}\label{subsec:semantics}

The semantics of logic programs is given by answer sets~\cite{gellif91a},
also called stable models~\cite{gellif88a}.
In view of the rich input language of \gringo\ (cf.\ Section~\ref{subsec:lang:gringo}),
we below recall a definition of answer sets
for propositional theories~\cite{ferraris05a} and
explain the semantics of logic programs by translation
into propositional logic.
We will also provide an alternative direct definition of answer sets
for the class of normal programs.

We consider propositional formulas composed from atoms, $\bot$,
and connectives~$\wedge$, $\vee$, and~$\rightarrow$.
Furthermore, $\top$ and~$\neg F$ are used as shorthands for
$(\bot\rightarrow\bot)$ and $(F\rightarrow\bot)$, respectively.
The \emph{reduct}~$\reduct{F}{X}$ of a propositional formula~$F$
relative to a set~$X$ of atoms is defined recursively as follows:
\begin{itemize}
\item $\reduct{F}{X}=\bot$ ~if $X\not\models F$,
\item $\reduct{F}{X}=F$ ~if $F\in X$, and
\item $\reduct{F}{X}=(\reduct{G}{X}\circ\reduct{H}{X})$ ~if 
      $X\models F$ and $F=(G\circ H)$ for $\circ\in\{\wedge,\vee,\rightarrow\}$.%
\footnote{$\models$ is the standard satisfaction relation of propositional logic
  between interpretations~$X$ and formulas~$F$.}
\end{itemize}
Essentially, the reduct relative to~$X$ is obtained by replacing
all maximal unsatisfied subformulas of~$F$ with~$\bot$.
As a matter fact, this implies that any maximal negative subformula 
of form $(G\rightarrow \bot)$ is replaced by either $\bot$ (if $X\models G$)
or $(\bot\rightarrow\bot)=\top$ (if $X\not\models G$).
Furthermore, any atom occurring in $\reduct{F}{X}$ belongs to~$X$.

A propositional theory~$\Phi$ is a set of propositional formulas,
and the reduct of~$\Phi$
relative to a set~$X$ of atoms is
$\reduct{\Phi}{X}=\{\reduct{F}{X} \mid F\in\Phi\}$.
Furthermore, $X$ is a \emph{model} of~$\Phi$ if $X\models F$
for every $F\in\Phi$.
Finally, $X$ is an \emph{answer set} of~$\Phi$ if
$X$ is a $\subseteq$-minimal model of~$\reduct{\Phi}{X}$, that is,
if $X$ is a model of~$\Phi$ such that no $Y\subset X$
is a model of~$\reduct{\Phi}{X}$.
The latter condition uses the fact that~$X$ is a model of~$\reduct{\Phi}{X}$
iff $X$ is a model of~$\Phi$.%
\footnote{If~$X$ is not a model of~$\Phi$, then 
  there is some $F\in\Phi$ such that $X\not\models F$,
  which in turn implies $\bot\in\reduct{\Phi}{X}$.}
Also note that, if~$X$ is an answer set of~$\Phi$,
it is the unique $\subseteq$-minimal model of~$\reduct{\Phi}{X}$
because all atoms occurring in~$\reduct{\Phi}{X}$ belong to~$X$
(which excludes incomparable $\subseteq$-minimal models).
This must not be confused
with uniqueness of an answer set of~$\Phi$.
In fact, $\Phi$ may have zero, one,
or multiple answer sets~$X$ (being $\subseteq$-minimal models of
distinct reducts~$\reduct{\Phi}{X}$).
In a sense, the $\subseteq$-minimality of an answer set~$X$ as a 
model of~$\reduct{\Phi}{X}$ stipulates the atoms in~$X$ to be justified by~$\Phi$
(or necessarily true) under the assumption of~$X$.

For a logic program~$\Pi$, answer sets of~$\Pi$ can now be explained
by translation to a propositional theory~$\Phi[\Pi]$.
To this end, let
\begin{itemize}
\item $\Phi[\Pi] = \bigcup_{r\in\ground{\Pi}}\phi[r]$,
\item $\phi[\mathit{Head}~\code{:-}~\mathit{Body}\code{.}] = 
       (\phi[\mathit{Body}]\rightarrow\psi[\mathit{Head}])$,
\item $\phi[A]=\psi[A]=A$ ~if $A\in\mathcal{A}$,
\item $\phi[\top]=\top$,%
\footnote{Recall that a fact $\mathit{Head}\code{.}$ is a shorthand for
          $\mathit{Head}~\code{:-}~\top\code{.}$}
\item $\phi[\code{not}~F] = \neg \phi[F]$, and
\item $\phi[G\code{,}H] = (\phi[G]\wedge\phi[H])$.
\end{itemize}
%
The translation~$\phi$ of a rule primarily consists of
straightforward syntactic conversions by replacing
``\code{:-},'' ``\code{,},'' and ``\code{not}'' with
$\rightarrow$, $\wedge$, and $\neg$, respectively.
However, it also contains a separate translation~$\psi$ for heads of rules,
which coincides with~$\phi$ on atoms by simply keeping them unchanged.
In Section~\ref{subsec:gringo:aggregate},
we will customize~$\psi$ in order to reflect the
``choice semantics'' for aggregates in heads of rules~\cite{ferlif05a,siniso02a}.
By virtue of the above translation,
we can simply define the answer sets of a logic program~$\Pi$
to be the answer sets of~$\Phi[\Pi]$.
We have thus specified answer sets for normal programs.

In order to provide more intuitions,
we now reproduce a direct definition of answer sets for normal programs.
This definition builds on the fact that the reduct for a normal program
is a set of Horn clauses, for which the $\subseteq$-minimal model (if it exists) can be constructed
systematically in a bottom-up manner~\cite{dowgal84a}.
To this end, for a normal program~$\Pi$ and a set $X\subseteq\ground{\mathcal{A}}$,
we define:
\begin{itemize}
\item $T^0_{\Pi,X} = \emptyset$ ~and
\item $T^{i+1}_{\Pi,X} = \{\head{r} \mid 
                        r\in\ground{\Pi},
                        \pbody{r}\subseteq T^i_{\Pi,X},
                        \nbody{r}\cap X=\emptyset\}$.
\end{itemize}
It is not hard to check that operator $T_{\Pi,X}$ is monotonic,
as it adds head atoms of ground rules to its previous result if the positive
body atoms have been derived,
while negative bodies are evaluated w.r.t.~$X$.
Using this operator, we can equivalently define $X\subseteq\ground{\mathcal{A}}$
as an answer set of a normal program~$\Pi$ if $X=\bigcup_{0\leq i}T^i_{\Pi,X}$.
This equilibrium requirement exhibits two fundamental aspects of answer set semantics:
\begin{enumerate}
\item Negative conditions are first evaluated w.r.t.\ an answer set candidate.
\item The candidate must be justified by the result of the preceding evaluation.
\end{enumerate}
Provided that~$X$ is a model of~$\Pi$ (that is, of~$\Phi[\Pi]$),
the second item can be understood in the sense that all atoms of~$X$
must have a proof from normal program~$\Pi$ w.r.t.~$X$.

\begin{example}\label{ex:semantics}
The normal program in Example~\ref{ex:flies}
(or its ground instantiation given in Example~\ref{ex:flies:ground})
translates into the following propositional theory:
%
\begin{lstlisting}[escapechar=@]
bird(tux)     penguin(tux)
bird(tweety)  chicken(tweety) @\\@
bird(tux)   @\rlap{\,$\wedge$}@  @\rlap{\!$\neg$}@ neg_flies(tux)   @${}\rightarrow$@ flies(tux)
bird(tweety)@\rlap{\,$\wedge$}@  @\rlap{\!$\neg$}@ neg_flies(tweety)@${}\rightarrow$@ flies(tweety)
bird(tux)   @\rlap{\,$\wedge$}@  @\rlap{\!$\neg$}@ flies(tux)       @${}\rightarrow$@ neg_flies(tux)
bird(tweety)@\rlap{\,$\wedge$}@  @\rlap{\!$\neg$}@ flies(tweety)    @${}\rightarrow$@ neg_flies(tweety)
                 penguin(tux)   @${}\rightarrow$@ neg_flies(tux)
                 penguin(tweety)@${}\rightarrow$@ neg_flies(tweety)
\end{lstlisting}
%
Let
$X=\{
\code{\pred{bird}(\const{tux})},
\code{\pred{penguin}(\const{tux})},
\code{\pred{bird}(\const{tweety})},
\code{\pred{chicken}(\const{tweety})},\linebreak[1]
\code{\pred{neg\mus flies}(\const{tux})},
\code{\pred{flies}(\const{tweety})}
\}$.
Relative to $X$, we obtain the reduct:
%
\begin{lstlisting}[escapechar=@]
bird(tux)     penguin(tux)
bird(tweety)  chicken(tweety) @\\@
              @\rlap{\!$\bot$}@ @${}\rightarrow$@ @$\bot$@
bird(tweety)@\rlap{\,$\wedge$}@  @\rlap{\!$\top$}@ @${}\rightarrow$@ flies(tweety)
bird(tux)   @\rlap{\,$\wedge$}@  @\rlap{\!$\top$}@ @${}\rightarrow$@ neg_flies(tux)
              @\rlap{\!$\bot$}@ @${}\rightarrow$@ @$\bot$@
   penguin(tux)@${}\rightarrow$@ neg_flies(tux)
              @\rlap{\!$\bot$}@ @${}\rightarrow$@ @$\bot$@
\end{lstlisting}
%
We have that~$X$ is a model of the reduct (and the original program) because
the reduct does not contain formula~$\bot$ as an element.
In addition, observe that no proper subset of~$X$ is a model of the reduct.
From this, we conclude that~$X$ is an answer set.

The latter can also be verified by determining $\bigcup_{0\leq i}T^i_{\Pi,X}$
(for~$\Pi$ as in Example~\ref{ex:flies}):
\begin{tabular}{@{}r||l}
$i$ & $T^i_{\Pi,X}$
\\\hline\hline
$0$ &
---
\\\hline
$1$ &
$
\begin{array}{@{}l}
\code{\pred{bird}(\const{tux})},
\code{\pred{penguin}(\const{tux})},
\code{\pred{bird}(\const{tweety})},
\code{\pred{chicken}(\const{tweety})}
\end{array}
$
\\\hline
$2$  & 
$
\begin{array}{@{}l@{}}
\code{\pred{bird}(\const{tux})},
\code{\pred{penguin}(\const{tux})},
\code{\pred{bird}(\const{tweety})},
\code{\pred{chicken}(\const{tweety})},\\
\code{\pred{neg\mus flies}(\const{tux})},
\code{\pred{flies}(\const{tweety})}
\end{array}
$
\\\hline
${>\,}2$ &
\textit{do.}
\end{tabular}
\\

\noindent
For $i=1$, we simply derive the head atoms of facts.
They allow us to derive the remaining atoms of~$X$, viz.,
\code{\pred{neg\mus flies}(\const{tux})} and
\code{\pred{flies}(\const{tweety})}, in step~$i=\nolinebreak2$.
In fact, two rules produce \code{\pred{neg\mus flies}(\const{tux})},
while
\begin{lstlisting}[numbers=none]
flies(tweety) :- bird(tweety), not neg_flies(tweety).
\end{lstlisting}
is the unique rule with head \code{\pred{flies}(\const{tweety})} that ``fires.''
No further rule applies in step~$i=3$,
so that the set of atoms derived for~$i=2$ equals $\bigcup_{0\leq i}T^i_{\Pi,X}$.
We have thus reproduced~$X$ by bottom-up derivation,
which again shows that~$X$ is an answer set.
\eexample
\end{example}

We have above seen two definitions of answer sets that start
from an answer set candidate $X\subseteq\ground{\mathcal{A}}$
that is then verified in some way.
Such definitions must not be confused with the computation of answer sets,
as (explicitly) enumerating all subsets of $\ground{\mathcal{A}}$
is infeasible in practice.
Rather, the computational schemes of solvers are similar
to the ``Davis-Putnam-Logemann-Loveland'' procedure (DPLL) \cite{dalolo62a,davput60}
or to ``Conflict-Driven Clause Learning'' (CDCL) \cite{marsak99a,mitchell05a,momazhzhma01a}.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End:

### 2008-11-03 doc/documentation/errors.tex
unverified new version of guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@579 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Errors and Warnings}\label{sec:error:warn}

This section explains the most frequent errors and warnings
related to inappropriate inputs or command line options that,
if they occur, lead to messages sent to the standard error stream.
The difference between errors and warnings is that the former
involve immediate termination,
while the latter are hints pointing at possibly corrupt input
that can still be processed further.
In the below description of errors (Section~\ref{subsec:error})
and warnings (Section~\ref{subsec:warn}),
we refrain from attributing them to a particular one among the tools
\gringo, \clasp, \clingo, and \iclingo,
in view of the fact that they share a number of functionalities.

\subsection{Errors}\label{subsec:error}

We start our description with errors that may be encountered during grounding,
where the following one indicates a syntax error in the input:
%
\begin{lstlisting}[numbers=none,escapechar=@]
syntax error on line @\textit{L}@ column @\textit{C}@ unexpected token: @\textit{T}@
\end{lstlisting}
%
To correct this error, please investigate the line~\code{\textit{L}}
and check whether something looks strange there
(like a missing period, an unmatched parenthesis, etc.).

The next error occurs if an input program is not level-restricted
(cf.\ Section~\ref{subsec:lambda}):
%
\begin{lstlisting}[numbers=none,escapechar=@]
the following rule cannot be grounded, \
weakly restricted variables: { @\textit{V}@, ... }
        @\textit{rule}@
\end{lstlisting}
%
Along with the error message, the \code{\textit{rule}} and the
name~\code{\textit{V}} of at least one variable causing the problem
are reported.
The first action to take usually consists of checking whether
variable~\code{\textit{V}} is actually in the scope of any atom
(in the positive body of \code{\textit{rule}}) that can bind it.%
\footnote{%
  Recall from Section~\ref{subsec:gringo:arith} and~\ref{subsec:gringo:comp}
  that a variable in the scope of a built-in arithmetic function may not be bound
  by a corresponding atom and that built-in comparison predicates do not bind
  any variable.}
If~\code{\textit{V}} is a local variable belonging to an atom~$A$ 
on the left-hand side of a condition (cf.\ Section~\ref{subsec:gringo:condition})
or to an aggregate (cf.\ Section~\ref{subsec:gringo:aggregate}),
an atom over some domain predicate might be included in a condition
to bind~\code{\textit{V}}.
In particular, if~$A$ itself is over a domain predicate,
the problem is often easily fixed by writing ``\code{$A$:$A$}.''
If the problem is not as simple,
then the predicates of all atoms that in principle may be used
to bind~\code{\textit{V}} are involved in positive recursion
through the predicate of some atom in the head of \code{\textit{rule}}.
In this case, the positive bodies of some of the rules subject to recursion
must be augmented with additional atoms over non-recursive predicates that
can bind variables, in order to eventually establish level-restrictedness.
As breaking recursion is a rather global matter, the reported
\code{\textit{rule}} is not necessarily the best place to tackle it.

The following error is related to conditions
(cf.\ Section~\ref{subsec:gringo:condition}):
%
\begin{lstlisting}[numbers=none,escapechar=@]
the following rule cannot be grounded, \
non domain predicates : { @\textit{p}@(...), ... }
        @\textit{rule}@
\end{lstlisting}
%
The problem is that an atom \code{\textit{p}(...)} such
that its predicate \code{\textit{p}/\!\textit{i}} is not a domain predicate
(cf.\ Section~\ref{subsec:strat}) is used on the right-hand side of a condition
within \code{\textit{rule}}.
The error is corrected by either removing the atom or
by replacing it with another atom over a domain predicate.

The next errors may occur within an arithmetic evaluation
(cf.\ Section~\ref{subsec:gringo:arith}):
%
\begin{lstlisting}[numbers=none,escapechar=@]
trying to convert string to int
trying to convert functionsymbol to int
\end{lstlisting}
%
It means that either a (symbolic) constant or a compound term
(over an uninterpreted function with non-zero arity)
has occurred in the scope of some built-in arithmetic function.
Unfortunately, no context information is provided yet,
which is left as an issue to future work
(cf.\ Section~\ref{sec:future}).

The below errors are related to built-in comparison predicates
(cf.\ Section~\ref{subsec:gringo:comp}):
%
\begin{lstlisting}[numbers=none,escapechar=@]
comparing different types
comparing function symbols
\end{lstlisting}
%
The first error indicates that one of the built-in comparison predicates
\code{<}, \code{<=}, \code{>}, and \code{>=}
has been applied to an integer or a (symbolic) constant
and a term of a different sort,
while the second error means that compound terms have occurred in such a comparison.
The mere reason for these errors is that the required functionalities
have not been implemented yet,
so that the errors are likely to disappear in the future
(cf.\ Section~\ref{sec:future}).

An error occurs if a constant~\code{\textit{c}}
ought to be replaced with a term~\code{\textit{t}}
containing~\code{\textit{c}}:
%
\begin{lstlisting}[numbers=none,escapechar=@]
cyclic constant definition
\end{lstlisting}
%
To resolve such an error, check the occurrences of \const{\#const}
in the input as well as \code{--const} or \code{-c} options
in the command line, and change them in a way such that a constant
that is to be replaced does not (transitively) occur within a term
to be inserted.

The next error may occur if a statement from the input language
of \iclingo\ (cf.\ Section~\ref{subsec:lang:iclingo})
is used in the input of \gringo, \clingo, or
\iclingo\ run non-incrementally
(via one of options \code{--text}, \code{--lparse}, \code{--aspils},
 or \code{--clingo}):
%
\begin{lstlisting}[numbers=none,escapechar=@]
A fixed number of incremental steps is needed to ground the program.
\end{lstlisting}
%
This problem is solved by providing a step number
via option \code{--ifixed}.

The following error may be reported by \iclingo:
%
\begin{lstlisting}[numbers=none,escapechar=@]
The following statement cant be used with the incremental interface:
        @\textit{rule}@
\end{lstlisting}
%
The \code{\textit{rule}} contains a \const{\#maximize},
\const{\#minimize}, or \const{\#compute} statement,
which are currently not supported within incremental computations.
Their support by \iclingo\ is an issue to future work
(cf.\ Section~\ref{sec:future}).

The next error may occur if \emph{ASPils}~\cite{gejaosscth08a}
output is requested via option \code{--aspils}:
%
\begin{lstlisting}[numbers=none,escapechar=@]
@\textit{e}@ not allowed in this normal form, \ 
please choose normal form @\textit{n}@.
\end{lstlisting}
%
The chosen normal form (in-between~\code{1} and~\code{7})
does not cover the input language expression~\code{\textit{e}},
and one among the normal forms~\code{\textit{n}} suggested
in the error message ought to be provided as an argument instead.

The following error message is issued by (embedded) \clasp:
%
\begin{lstlisting}[numbers=none,escapechar=@]
Input Error at line @\textit{L}@: Symbol table expected!
\end{lstlisting}
%
This error means that the input does not comply with \lparse's
numerical format~\cite{lparseManual}.
It is not unlikely that the input can be processed
by \gringo, \clingo, or \iclingo.

The next error indicates that input in
\lparse's numerical format~\cite{lparseManual} is corrupt:
%
\begin{lstlisting}[numbers=none,escapechar=@]
Input Error at line @\textit{L}@: Atom out of bounds
\end{lstlisting}
%
There is no way to resolve this problem.
If the input has been generated by \gringo, \clingo, or \iclingo,
please report the problem to the authors of this guide.

The following error message is issued by (embedded) \clasp:
%
\begin{lstlisting}[numbers=none,escapechar=@]
Input Error at line @\textit{L}@: Unsupported rule type!
\end{lstlisting}
%
It means that some rule type in \lparse's
numerical format~\cite{lparseManual} is not supported.
Most likely, the program under consideration contains
rules with disjunction in the head.

A similar error may occur with \clingo\ or \iclingo:
%
\begin{lstlisting}[numbers=none,escapechar=@]
sorry clasp cannot handle disjunctive rules!
\end{lstlisting}
%
The program under consideration contains rules with disjunction in the head,
which are currently not supported by \clasp,
but by \claspD~\cite{drgegrkakoossc08a}.
The integration of \clasp\ and \claspD\ is a subject to future work
(cf.\ Section~\ref{sec:future}).

All of the tools \gringo, \clasp, \clingo, and \iclingo\
try to expand incomplete (long) options to recognized ones.
Parsing command line options may nonetheless fail due to the following three reasons:
%
\begin{lstlisting}[numbers=none,escapechar=@]
unknown option: @\textit{o}@
ambiguous option: '@\!\textit{o}@' could be:
  @\textit{o1}@
  @\textit{o2}@
  ...
'@\!\textit{a}@': invalid value for Option '@\!\textit{o}@'
\end{lstlisting}
%
The first error means that a provided option~\code{\textit{o}}
could not be expanded to one that is recognized,
while the second error expresses that the result of expanding~\code{\textit{o}}
is ambiguous.
Finally, the third error occurs if a provided argument~\code{\textit{a}}
is invalid for option~\code{\textit{o}}.
In either case, option \code{--help} can be used to see 
the recognized options and their arguments.

\subsection{Warnings}\label{subsec:warn}

The following warnings may be raised by \gringo, \clingo, or \iclingo:
%
\begin{lstlisting}[numbers=none,escapechar=@]
Warning: @\textit{p}@/@\!\textit{i}@ is never defined.
Warning: sum with negative bounds in the body of a rule
Warning: multiple definitions of #const @\textit{c}@
\end{lstlisting}
%
The first one, stating that a predicate \code{\textit{p}/\!\textit{i}}
has occurred in some rule body, but not in the head of any rule,
might point at a mistyped predicate.
The second warning means that negative weights within a \const{sum} aggregate
(cf.\ Section~\ref{subsec:gringo:aggregate})
have been compiled away~\cite{siniso02a},
which is semantically delicate~\cite{ferraris05a}.
Thus, we strongly recommend to avoid negative weights
within a \const{sum} aggregate in the positive body of a rule.
Finally, the third warning recognizes that multiple \const{\#const}
statements over the same constant~\code{\textit{c}} are contained in the input,
while only the first among them will be considered.

The next four warnings may be issued by \iclingo:
\begin{lstlisting}[numbers=none,escapechar=@]
Warning: There are no #base, #cumulative or #volatile sections.
Warning: Option ifixed will be ignored!
Warning: There are statements not within a #base, \
         #cumulative or #volatile section.
         These Statements are put into the #base section.
Warning: Classical negation is not handled correctly \
         in combination with the incremental output.
         You have to add rules like: :- a, -a. on your own! \
         (at least for now)
\end{lstlisting}
%
For omitting the first warning, one may use option \code{--clingo},
as the input program contains no statements particular to \iclingo\
(cf.\ Section~\ref{subsec:lang:iclingo}).
Similarly, option \code{--clingo} may be used to suppress the second warning,
stating that a provided option \code{--ifixed} is ignored in incremental mode
(if \iclingo\ is not run as a grounder 
 via one of options \code{--text}, \code{--lparse}, or \code{--aspils}).
The third warning states that rules have been inserted into the static
program part without a 	preceding \const{\#base} statement.
This warning can be avoided by ending an encoding with \const{\#base.}
and by supplying an instance file afterwards, as illustrated in Section~\ref{subsec:ex:block}.
The fourth warning refers to the fact that there currently is no
built-in support for classical negation (cf.\ Section~\ref{subsec:gringo:negation})
by \iclingo.
Hence, integrity constraints of the form ``$\code{:-}~A\code{,}\code{-}A\code{.}$''
have to be added explicitly by a user.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End:

### 2008-11-03 doc/documentation/examples.tex
unverified new version of guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@579 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Examples}\label{sec:examples}

We exemplarily solve the following problems in ASP:
$N$-Coloring (Section~\ref{subsec:ex:color}),
Traveling Salesperson (Section~\ref{subsec:ex:tsp}), and
Blocks-World Planning (Section~\ref{subsec:ex:block}).
While the first problem could likewise be solved within neighboring paradigms,
the second one requires checking reachability,
something that is rather hard to encode in either
Boolean Satisfiability~\cite{bihemawa08a} or
Constraint Programming~\cite{robewa06a}.
The third problem coming from the area of planning
illustrates incremental solving with \iclingo.

\subsection{$N$-Coloring}\label{subsec:ex:color}

As already mentioned in Section~\ref{subsec:asp},
it is custom in ASP to provide a \emph{uniform}
problem definition~\cite{martru99a,niemela99a,schlipf95a}.
We follow this methodology and separate the encoding
from an instance of the following problem:
given a (directed) graph, decide whether each node can be assigned
one of~$N$ colors such that any pair of adjacent nodes is colored differently.
Note that this problem is NP-complete for~$N\geq 3$
(see, e.g.,~\cite{papadimitriou94a}),
and thus it seems unlikely that a worst-case polynomial algorithm
can be found.
In view of this,
it is convenient to reduce the particular problem to
a declarative problem solving paradigm like ASP,
where efficient off-the-shelf tools like \gringo\ and \clasp\
are ready to solve the problem reasonably well.
Such a reduction is now exemplified.

\subsubsection{Problem Instance}\label{subsec:color:instance}
\input{figures/graph}

We consider directed graphs specified via facts over predicates
\pred{node}/$1$ and \pred{edge}/$2$.%
\footnote{%
  Directedness is not an issue in $N$-Coloring,
  but we will reuse our directed example graph in Section~\ref{subsec:ex:tsp}.}
The graph shown in Figure~\ref{fig:graph} is represented by the following set of facts:
%
\lstinputlisting{examples/graph.lp}
%
Recall from Section~\ref{subsec:lang:gringo} that~``\code{..}'' and~``\code{;}''
in the head expand to multiple rules, which are facts here.
Thus, the instance contains six nodes and 17 directed edges among them.

\subsubsection{Problem Encoding}\label{subsec:color:encoding}

We now proceed by encoding $N$-coloring via non-ground rules that are
independent of particular instances.
Typically, an encoding consists of a \emph{Generate}, a \emph{Define},
and a \emph{Test} part~\cite{lifschitz02a}.
As $N$-Coloring has a rather simple pattern, the following encoding does
not contain any Define part:
%
\lstinputlisting{examples/color.lp}
%
In Line~2, we use the \code{\#const} declarative,
described in Section~\ref{subsec:gringo:meta},
to install~\const{3} as default value for constant~\const{n} that is to be replaced
with the number~$N$ of colors.
(The default value can be overridden by invoking \gringo\ with option
 \code{--const n=$N$}.)
The Generate rule in Line~4 makes use of the \const{count} aggregate
(cf.\ Section~\ref{subsec:gringo:aggregate}).
For our example graph and~\const{1} substituted for~\var{X},
we obtain the following ground rule:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\
    \mbox{~}examples/color.lp \textbackslash\\
    \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
1 { color(1,1), color(1,2), color(1,3) } 1.
\end{lstlisting}
%
Note that~\code{\pred{node}(\const{1})} has been removed from the body,
as it is derived via a corresponding fact,
and similar ground instances are obtained for the other nodes~\const{2} to~\const{6}.
Furthermore, for each instance of \pred{edge}/2,
we obtain~\code{n} ground instances of the integrity constraint in Line~6,
prohibiting that the same color~\var{C} is assigned to the adjacent nodes.
Given~\code{\const{n}=\const{3}},
we get the following ground instances due to \code{\pred{edge}(\const{1},\const{2})}:
%
\begin{lstlisting}[numbers=none]
:- color(1,1), color(2,1).
:- color(1,2), color(2,2).
:- color(1,3), color(2,3).
\end{lstlisting}
%
Again note that \code{\pred{edge}(\const{1},\const{2})},
derived via a fact, has been removed from the body.

\subsubsection{Problem Solution}\label{subsec:color:solution}
\input{figures/color}

Provided that a given graph is colorable with~\const{n} colors,
a solution can be read off an answer set of the program consisting
of the instance and the encoding.
For the graph in Figure~\ref{fig:graph},
the following answer set can be computed:%
\marginlabel{%
  To find an answer set, invoke:\\
  \code{\mbox{~}gringo \textbackslash\\
        \mbox{~}examples/color.lp \textbackslash\\
        \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
        \mbox{~}clasp}\\
  or alternatively:\\
  \code{\mbox{~}clingo \textbackslash\\
        \mbox{~}examples/color.lp \textbackslash\\
        \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
Answer: 1
... color(1,2) color(2,1) color(3,1) color(4,3) color(5,2) color(6,3)
\end{lstlisting}
%
Note that we have omitted the six instances of \pred{node}/$1$ and the
17 instances of \pred{edge}/$2$ in order to emphasize the actual solution,
which is depicted in Figure~\ref{fig:color}.
Such output projection can also be specified within a logic program file by
using the declaratives \code{\#hide} and \code{\#show},
described in Section~\ref{subsec:gringo:meta}.

\subsection{Traveling Salesperson}\label{subsec:ex:tsp}

We now consider the well-known Traveling Salesperson Problem (TSP),
where the task is to decide whether there is a round trip that visits
each node in a graph exactly once (viz., a Hamiltonian cycle) and whose
accumulated edge costs must not exceed some budget~$B$.
We tackle a slightly more general variant of the problem by not
a priori fixing~$B$ to any integer.
Rather,
we want to compute a minimum budget~$B$ along with a round trip of cost~$B$.
This problem is FP$^\textrm{NP}$-complete (cf.~\cite{papadimitriou94a}),
that is, it can be solved with a polynomial number of queries to an NP-oracle.
As with $N$-Coloring,
we provide a uniform problem definition by separating the encoding from instances.

\subsubsection{Problem Instance}\label{subsec:tsp:instance}
\input{figures/costs}

We reuse graph specifications in terms of predicates \pred{node}/$1$ and \pred{edge}/$2$
as in Section~\ref{subsec:color:instance}.
In addition, facts over predicate \pred{cost}/$3$ are used to define edge costs:
%
\lstinputlisting{examples/costs.lp}
%
Figure~\ref{fig:costs} shows the (directed) graph from Figure~\ref{fig:graph}
along with the associated edge costs.
Symmetric edges have the same costs here,
but differing costs would also be possible.

\subsubsection{Problem Encoding}\label{subsec:tsp:encoding}

The first subproblem consists of describing a Hamiltonian cycle,
constituting a candidate for a minimum-cost round trip.
Using the Generate-Define-Test pattern~\cite{lifschitz02a},
we encode this subproblem via the following non-ground rules:
%
\lstinputlisting{examples/ham.lp}
%
The Generate rules in Line~2 and~3 assert that every node must have
exactly one outgoing and exactly one incoming edge, respectively,
that belong to the cycle.
By inserting the available edges, for node~\const{1},
Line~2 and~3 are grounded as follows:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
1 { cycle(1,2), cycle(1,3), cycle(1,4) } 1.
1 { cycle(3,1), cycle(4,1) } 1.
\end{lstlisting}
%
Observe that the first rule groups all outgoing edges of node~\const{1},
while the second one does the same for incoming edges.
We proceed by considering the Define rules in Line~5 and~6,
which recursively check whether nodes are reached by a cycle candidate
produced via the Generate part.
Note that the rule in Line~5 builds on the assumption that the cycle
``starts'' at node~\code{1}, that is,
any successor~\var{Y} of~\code{1} is reached by the cycle.
The second rule in Line~6 states that, from a reached node~\var{X},
an adjacent node~\var{Y} can be reached via a further edge in the cycle.
Note that this definition admits positive recursion
among the ground instances of \pred{reached}/$1$,
in which case a ground program is called \emph{non-tight}~\cite{erdlif03a,fages94a}.
The ``correct'' treatment of (positive) recursion due to the justification
requirement w.r.t.\ the reduct (cf.\ Section~\ref{subsec:semantics})
is a particular feature of answer set semantics,%
\marginlabel{%
  To compute the six Hamiltonian cycles
  for the graph in Figure~\ref{fig:graph}, invoke:\\
  \code{\mbox{~}gringo \textbackslash\\
        \mbox{~}examples/ham.lp \textbackslash\\
        \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
        \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
        \mbox{~}examples/ham.lp \textbackslash\\
        \mbox{~}examples/graph.lp}}
which is hard to mimic in either
Boolean Satisfiability~\cite{bihemawa08a} or
Constraint Programming~\cite{robewa06a}.
In our present problem, this feature makes sure that all nodes are reached
by a global cycle from node~\code{1}, thus, excluding isolated subcycles.
In fact, the Test in Line~8 stipulates that every node in the given graph
is reached, that is, the instances of \pred{cycle}/$2$ in an answer set
must be the edges of a Hamiltonian cycle.
Finally, the additional Display part in Line~10 and~11 states that answer
sets should be projected to instances of \pred{cycle}/$2$, as only they
describe a solution.
We have so far not considered edge costs,
and answer sets for the above 
part of the encoding correspond to Hamiltonian cycles.

In order to minimize costs,
we add the following optimization statement:
%
\lstinputlisting[firstnumber=13]{examples/min.lp}
%
Here, edges belonging to the cycle are weighted according to their costs.
After grounding, the minimization in Line~14 ranges over~17 instances
of \pred{cycle}/$2$, one for each weighted edge in Figure~\ref{fig:costs}.

\subsubsection{Problem Solution}\label{subsec:tsp:solution}
\input{figures/tsp}

Finally, we explain how the unique minimum-cost round trip
(depicted in Figure~\ref{fig:tsp}) can be computed.
The catch is that we are now interested in optimal answer sets,
rather than in arbitrary ones.
In order to determine the optimum, we can start by gradually
decreasing the costs associated to answer sets
until we cannot find a strictly better one anymore.
To this end, it is important to invoke \clasp\ (or \clingo)
with option ``\code{-n~0}'' (cf.\ Section~\ref{subsec:opt:clasp}),
making it enumerate successively better answer sets
w.r.t.\ the provided optimization statements (cf.\ Section~\ref{subsec:gringo:optimize}).
Any answer set is printed as soon as it has been computed,
and the last one is optimal.
If there are multiple optimal answer sets, an arbitrary one among them is computed.
For the graph in Figure~\ref{fig:costs},
the optimal answer set (cf.\ Figure~\ref{fig:tsp}) is unique,
and its computation can proceed as follows:%
\marginlabel{%
  To compute the minimum-cost round trip
  for the graph in Figure~\ref{fig:costs}, invoke:\\
  \code{\mbox{~}gringo \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
    \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
Answer: 1
cycle(1,3) cycle(2,4) cycle(3,5) \
cycle(4,1) cycle(5,6) cycle(6,2)
Optimization: 13
Answer: 2
cycle(1,2) cycle(2,5) cycle(3,4) \
cycle(4,1) cycle(5,6) cycle(6,3)
Optimization: 11
\end{lstlisting}
%
Given that no answer is obtained after the second one,
we know that~\code{11} is the optimum value,
but there might be more optimal answer sets that have not been computed yet.
In order to find them too, we can use the following command line options
of \clasp\ (cf.\ Section~\ref{subsec:opt:clasp}):
``\code{--opt-value=11}'' in order to initialize the optimum and
``\code{--opt-all}'' to compute also equitable
(rather than only strictly better) answer sets.%
\marginlabel{%
  The full invocation is:\\
  \code{\mbox{~}gringo \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
    \mbox{~}clasp -n 0 \textbackslash\\
    \mbox{~}--opt-value=11 \textbackslash\\
    \mbox{~}--opt-all}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
    \mbox{~}--opt-value=11 \textbackslash\\
    \mbox{~}--opt-all \textbackslash\\
    \mbox{~}examples/ham.lp \textbackslash\\
    \mbox{~}examples/min.lp \textbackslash\\
    \mbox{~}examples/costs.lp \textbackslash\\
    \mbox{~}examples/graph.lp}}
After obtaining only the second answer given above,
we are sure that this is the unique optimal answer set,
whose associated edge costs (cf.\ Figure~\ref{fig:tsp}) correspond to
the reported optimization value~\code{11}.
Note that, with \const{maximize} statements in the input, this correlation
might be less straightforward because they are compiled into \const{minimize}
statements in the process of generating \lparse's output format~\cite{lparseManual}.
Furthermore, if there are multiple optimization statements,
\clasp\ (or \clingo) will report separate optimization values ordered by significance.
Finally, the two-stage invocation scheme exercised above,
first determining the optimum and afterwards all (and only) optimal answer sets,
is recommended for general use.
Otherwise,
if using option ``\code{--opt-all}'' right away without knowing the optimum,
one risks the enumeration of plenty suboptimal answer sets.
We also invite everyone to explore command line option ``\code{--opt-restart}''
(cf.\ Section~\ref{subsec:opt:clasp})
in order to see whether it improves search efficiency on optimization problems.

\subsection{Blocks-World Planning}\label{subsec:ex:block}

The Blocks-World is a well-known planning domain,
where finding \emph{shortest} plans has received particular attention~\cite{gupnau92a}.
In an eventually propositional formalism like ASP,
a bound on the plan length must be fixed before search can proceed.
This is usually accomplished by including some constant~\const{t}
in an encoding, which is then replaced with the actual bound during grounding.
Of course, if the length of a shortest plan is unknown,
an ASP system must repeatedly be queried while varying the bound.
With a traditional ASP system, processing
the same planning problem with a different bound
involves grounding and solving from scratch.
In order to reduce such redundancies,
the incremental ASP system \iclingo~\cite{gekakaosscth08a}
can gradually increase a bound, % thereby,
doing only the necessary additions in each step.
Note that planning is a natural application domain for \iclingo,
but other problems including some mutable bound can be addressed too, thus,
\iclingo\ is not a specialized planning system.
However, we use Blocks-World Planning to illustrate the exploitation of
\iclingo's incremental computation mode.

\subsubsection{Problem Instance}\label{subsec:block:instance}

As with the other two problems above,
an instance is given by a set of facts, here,
over predicates \pred{block}/$0$ (declaring blocks),
\pred{init}/$1$ (defining the initial state),
\pred{goal}/$1$ (specifying the goal state).
A well-known Blocks-World instance is described by:%
\footnote{%
  Blocks-World instances ``\code{examples/world$i$.lp}'' for
  $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$
  are adaptations of the instances provided at~\cite{erdemBW}.}
%
\lstinputlisting{examples/world0.lp}
%
Note that the facts in Line~13--15 and~24--26 specify the initial
and the goal state depicted in Line~9-11 and~19--22, respectively.
We here use (uninterpreted) function \const{on}/$2$ to illustrate another
important feature available in \gringo, \clingo, and~\iclingo, namely,
the possibility of instantiating variables to compound terms.

\subsubsection{Problem Encoding}\label{subsec:block:encoding}

Our Blocks-World Planning encoding for \iclingo\ makes use of statements
\const{\#base}, \const{\#cumulative}, and \const{\#volatile},
separating the encoding into a static, a cumulative, and a volatile (query) part.
Each of them can be further refined into Generate, Define, Test, and Display constituents, 
as indicated in the comments below:
%
\lstinputlisting{examples/blocks.lp}
%
In the initial \const{\#base} part (Line~1--5),
we define blocks and constant \const{table}
as instances of predicate \pred{location}/$1$.
Moreover, we use instances of \pred{init}/$1$
to initialize predicate \pred{holds}/$2$ for step~\const{0}, thus,
defining the conditions before the first incremental step.
Note that variable~\var{F} is instantiated to compound terms
over function \const{on}/$2$.

The \const{\#cumulative} statement in Line~7 declares constant~\code{t}
as a placeholder for step numbers in the cumulative encoding part below.
Here, the Generate rule in Line~9 states that exactly one block~\var{X}
must be moved to a location~\var{Y} (different from~\var{X}) at each step~\code{t}.
The integrity constraint in Line~11-13 is used to test whether
moving block~\var{X} to location~\var{Y} is possible at step~\code{t}
by denying~\code{\pred{move}(X,Y,t)} to hold if there
is either some block~\var{A} on~\var{X} or some block~\var{B} distinct from~\var{X} on~\var{Y}
(this condition is only checked if~\var{Y} is a block, viz., different from constant \const{table})
at step~\code{t-1}.
Finally, the Define part in Line~15 propagates
a move to the state at step~\code{t},
while the rule in Line~16--17 states that a block~\var{X} stays on a location~\var{Z}
if it is not moved to any other location~\var{Y}.
Note that we require atoms \code{\pred{block}(\var{X})} and \code{\pred{location}(\var{Z})}
to bind variables~\var{X} and~\var{Z} in the body of the latter rule,
as recursive predicate \pred{holds}/$2$ cannot be used for this purpose
(cf.\ the definition of level-restrictedness in Section~\ref{subsec:lambda}).

The \const{\#volatile} statement in Line~19 declares the next part as a
query depending on step number~\const{t}, but not accumulated over
successive steps.
In fact, the integrity constraint in Line~21 tests whether goal conditions are
satisfied at step~\const{t}.

Our incremental encoding concludes with a second \const{\#base} part,
as specified in Line~23.
Note that, for the meta-statements with Display functionality (Line~25-26),
it is actually unimportant whether they belong to a static, cumulative, or
volatile program part, as answer sets are projected
(to instances of \pred{move}/$3$) in either case.
However, by ending the encoding file with a \const{\#base} statement,
we make sure that the contents of a concatenated instance file
is included in the static program part.
This is also the default of \iclingo\
(as well as of \gringo\ and \clingo\ that can be used for non-incremental computations),
but applying this default behavior triggers a warning (cf.\ Section~\ref{subsec:warn}).%
\marginlabel{%
  To observe the ground program dealt with internally \iclingo\
  at a step $t$, invoke:\\
  \code{\mbox{~}gringo -t \textbackslash\\
        \mbox{~}--ifixed=$t$ \textbackslash\\
        \mbox{~}examples/blocks.lp \textbackslash\\
        \mbox{~}examples/world$i$.lp}
  where $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$.}

Finally, let us stress an important prerequisite for obtaining
a well-defined incremental computation result from \iclingo.
First, the ground instances of head atoms of rules in the
static, cumulative, and volatile program part must be pairwisely disjoint.
Furthermore, ground instances of head atoms in the volatile part
must not occur in the cumulative part,
and those of the cumulative part must in turn not be used in the static part.
Finally, ground instances of head atoms in either the cumulative or the volatile part
must be different for each pair of distinct steps.
This is granted for our encoding because both atoms over \pred{move}/$3$
and those over \pred{holds}/$2$ include~\const{t} as an argument in the
heads of the rules in Line~9, 15, and~16--17.
Notably, the smallest step number to replace~\const{t} with is~\const{1},
so that there is no clash with the ground atoms over \pred{holds}/$2$
obtained from the head of the static rule in Line~5.
Further details on these requirements and their formal background can
be found in~\cite{gekakaosscth08a}.
Arguably, many problems including some mutable bound are naturally encoded
such that the above conditions apply.
Some attention should of course be spent on
putting rules into the right program parts.

\subsubsection{Problem Solution}\label{subsec:block:solution}

We can now use \iclingo\ to \emph{incrementally} compute the shortest
sequence of moves that brings us from the initial to the goal state
depicted in the instance in Section~\ref{subsec:block:instance}:%
\marginlabel{%
  To this end, invoke:\\
  \code{\mbox{~}iclingo -n 0 \textbackslash\\
        \mbox{~}examples/blocks.lp \textbackslash\\
        \mbox{~}examples/world0.lp}}
%
\begin{lstlisting}[numbers=none]
Answer: 1
move(b2,table,1) move(b1,b0,2) move(b2,b1,3)
\end{lstlisting}
%
This unique answer set tells us that the given problem instance can
be solved by moving block~\const{b2} to the~\const{table} in order to
then put~\const{b1} on top of~\const{b0} and finally~\const{b2} on top of~\const{b1}.
This solution is computed by \iclingo\ in three grounding and solving steps,
where, starting from the \const{\#base} program, constant~\const{t}
is successively replaced with step numbers~\const{1}, \const{2}, and~\const{3}
in the~\const{\#cumulative} and in the~\const{\#volatile} part.
While the query postulated in the~\const{\#volatile} part cannot be
fulfilled in steps~\const{1} and~\const{2}, \iclingo\ 
stops its incremental computation after finding some answer set in step~\const{3}.
The scheme of iterating steps until finding at least one answer set
is the default behavior of~\iclingo,
which can be customized via command line options
(cf.\ Section~\ref{subsec:opt:iclingo}).

Finally, let us describe how solutions obtained via an incremental computation
can be computed in the standard way, that is, in a single pass.
To this end, the step number can be fixed to some~$n$ via
option ``\code{--ifixed=$n$}'' (cf.\ Section~\ref{subsec:opt:gringo}),
\marginlabel{%
  For non-incremental solving, invoke:\\
  \code{\mbox{~}gringo --ifixed=$n$ \textbackslash\\
        \mbox{~}examples/blocks.lp \textbackslash\\
        \mbox{~}examples/world$i$.lp |\rlap{\textbackslash}\\
        \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
        \mbox{~}--ifixed=$n$ \textbackslash\\
        \mbox{~}examples/blocks.lp \textbackslash\\
        \mbox{~}examples/world$i$.lp}
  where $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$.}
enabling \gringo\ or \clingo\ to generate the ground program present
inside \iclingo\ at step~$n$.
Note that \const{\#volatile} parts are here only instantiated for the final step~$n$,
while \const{\#cumulative} rules are added for all steps $\const{1},\dots,n$.
Option ``\code{--ifixed=$n$}'' can be useful for investigating the contents
of a ground program dealt with at step~$n$ or for using an external solver
(other than \clasp).
In the latter case, repeated invocations with varying~$n$
are required if the bound of an optimal solution is a priori unknown.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End:

### 2008-11-03 doc/documentation/introduction.tex
unverified new version of guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@579 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Introduction}\label{sec:introduction}

The ``Potsdam Answer Set Solving Collection'' (Potassco)~\cite{potassco}
by now gathers a variety of tools for Answer Set Programming.
Among them, we find grounder \gringo, solver \clasp, and
combinations thereof within integrated systems \clingo\ and \iclingo.
All these tools are published under GNU General Public License(s)~\cite{GNUgpl}.
Source packages as well as precompiled binaries for Linux and Windows are available at~\cite{potassco}.
Note that there currently are two source packages:
one containing \clasp, and another one grouping \gringo, \clingo, and \iclingo.
For building one of the tools from sources,
please download the most recent source package and consult the
included \code{README} or \code{INSTALL} text file, respectively.
Please make sure that the platform to build on has the
required software installed.
If you nonetheless encounter problems in the building process,
please do not hesitate to contact the authors of this guide.

After downloading (and possibly building) a tool,
one can check whether everything works fine by invoking the tool
with flag \code{--version} (to get version information) or
with flag \code{--help} (to see the available command line options).
For instance, assuming that a binary called \gringo\ is in the path
(similarly, with the other tools),
the following command line calls should be responded by \gringo:
%
\begin{lstlisting}[numbers=none]
gringo --version
gringo --help
\end{lstlisting}
If grounder \gringo, solver \clasp, as well as integrated systems
\clingo\ and \iclingo\ are all available,
one usually provides the filenames of input text files to either
\gringo, \clingo, or \iclingo, while the output of \gringo\ is
typically piped into \clasp.
Thus, the standard invocation schemes are as follows:
\begin{lstlisting}[numbers=none]
gringo  [ options | filenames ] | clasp [ options | number ]
clingo  [ options | filenames | number ]
iclingo [ options | filenames | number ]
\end{lstlisting}
Note that a numerical argument provided to either \clasp, \clingo, or \iclingo\
determines the maximum number of answer sets to be computed,
where \code{0} stands for ``compute all answer sets.''
By default, only one answer set is computed (if it exists).

This guide introduces the fundamentals of using
\gringo, \clasp, \clingo, and \iclingo.
In particular, it tries to enable the reader to benefit from them
by significantly reducing the ``time to solution'' on difficult problems.
The outline is as follows.
In Section~\ref{sec:background},
we describe the basics of Answer Set Programming,
and we formally introduce the syntax and semantics of logic programs.
Section~\ref{sec:restricted} details restrictedness notions,
important when dealing with logic programs containing first-order variables.
The probably most important part for a user, Section~\ref{sec:language},
is dedicated to the input languages of our tools,
where the joint input language of \gringo\ and \clingo\
claims the main share (later on, it is extended by \iclingo).
For illustrating the application of our tools,
three well-known example problems are solved in Section~\ref{sec:examples}.
Practical aspects are also in the focus of Section~\ref{sec:options} and~\ref{sec:error:warn},
where we elaborate and give some hints on the available command line options
as well as input-related errors and warnings that may be reported.
Finally, we conclude with some remarks on future work in Section~\ref{sec:future}.

For readers familiar with \lparse~\cite{lparseManual}
(a grounder that constitutes the traditional front end of solver \smodels~\cite{siniso02a}),
Appendix~\ref{sec:lparse} lists the most prominent differences to our tools.
Otherwise, \gringo, \clingo, and \iclingo\ should accept most inputs recognized by \lparse,
while the input of solver \clasp\ can also be generated by \lparse\ instead of \gringo.
Throughout this guide, we will provide quite a number of examples.
Many of them can actually be run, and instructions how to accomplish this
(or sometimes meta-remarks)
are provided in margin boxes, where an occurrence of ``\code{\char`\\}''
usually means that a text line broken for space reasons is actually continuous.
After all these preliminaries, it is time to start our guided tour
through Potassco~\cite{potassco}.
We sincerely hope that you will find it enjoyable and helpful!

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End:

### 2008-11-03 doc/documentation/language.tex
unverified new version of guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@579 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Input Languages}\label{sec:language}

This section provides an overview of the input languages of
grounder \gringo, combined grounder and solver \clingo,
incremental grounder and solver \iclingo, and of solver \clasp.
The joint input language of \gringo\ and \clingo\ is detailed in
Section~\ref{subsec:lang:gringo}.
It is extended by \iclingo\ with a few directives described in Section~\ref{subsec:lang:iclingo}.
Finally, Section~\ref{subsec:lang:clasp} is dedicated to the inputs
handled by \clasp.

\subsection{Input Language of \gringo\ and \clingo}\label{subsec:lang:gringo}

The tool \gringo~\cite{gescth07a} is a grounder capable of translating
logic programs provided by users into equivalent ground programs.
The output of \gringo\ can be piped into solver \clasp~\cite{gekanesc07b},
which then computes answer sets.
System \clingo\ internally couples \gringo\ and \clasp, thus,
it takes care of both grounding and solving.
% The input of \gringo\ and \clingo\ is the same,
% but 
In contrast to \gringo\ outputting ground programs, 
\clingo\ returns answer sets.
Both \gringo\ and \clingo\ can handle level-restricted input programs
(cf.\ Section~\ref{subsec:lambda}),
usually specified in one or more text files whose names are passed via the command line
in an invocation of either \gringo\ or \clingo.
We below provide a description of constructs belonging to the input language of \gringo\ and \clingo.

\subsubsection{Normal Programs and Integrity Constraints}\label{subsec:gringo:normal}

In the previous sections, we have already seen a number of normal programs.
Now also considering integrity constraints, we get the following basic rule types:

\begin{tabular}{rl@{}l}
\textbf{Fact:} & $A_0$&\code{.}
\\
\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\\
\textbf{Integrity Constraint:} & & $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
\end{tabular}
\\

\noindent
The head~$A_0$ of a fact or a rule is an \emph{atom} of the form
$p\code{(}t_1\code{,}\dots\code{,}t_n\code{)}$,
where $p$ is the name of some predicate, that is, a sequence of letters and digits
starting with a lowercase letter,
and any $t_i$ is a term.%
\footnote{An atom~$p$ without arguments is simply a sequence of letters and digits
  (starting with a lowercase letter).
  For such an atom~$p$, parentheses after the name are skipped, e.g.,
  \pred{p42X} could be an atom without arguments.
  Also note that lowercase letters include ``\code{\mus},''
  which can thus be used at any position within a predicate name.}
A \emph{term} $t$ starting with an uppercase letter followed by
a sequence of letters and digits (e.g., \var{X08x15}) is a variable name,
and integers are written as sequences of digits possibly preceded by ``\code{-}.''
In addition, a term can have the same syntactic structure as an atom
(e.g., \code{\pred{p}(\const{a},\const{1},\const{f}(\var{X}))}
 can be either an atom or a term, depending on where it occurs in a rule),
and functions can be nested within a term.
There are also built-in constructs
(cf.\ Section~\ref{subsec:gringo:arith} and~\ref{subsec:gringo:comp})
having particular representations.
Finally, any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
for an atom~$A$.

In Section~\ref{subsec:semantics},
we have already provided a translation~$\psi$
from the head of a rule to a propositional formula.
By viewing an integrity constraint
$\code{:-}~\mathit{Body}\code{.}$ as a shorthand for $\bot~\code{:-}~\mathit{Body}\code{.}$,
we can now explain the semantics of integrity constraints by adding the following case:
\begin{itemize}
\item $\psi[\bot]=\bot$.
\end{itemize}
Note the role of integrity constraints is to eliminate answer set candidates.
In fact, given an integrity constraint~$\code{:-}~\mathit{Body}\code{.}$,
any answer set~$X$ is such that $X\not\models\phi[\mathit{Body}]$, or
in other words, some literal~$L$ in $\mathit{Body}$ must be unsatisfied w.r.t.~$X$.
Elaborate examples on the usage of facts, rules, and integrity constraints
will be provided in Section~\ref{sec:examples}.

\subsubsection{Classical Negation}\label{subsec:gringo:negation}

In logic programs, connective \code{not} expresses default negation,
that is, a literal $\code{not}~A$ is assumed to hold unless~$A$ is derived.
In contrast, the classical (or strong) negation of some proposition
holds if the complement of the proposition is objectively derived~\cite{gellif91a}.
Classical negation, indicated by symbol \code{-}, is permitted in front of atoms.
That is, if~$A$ is an atom, then~$\code{-}A$ is the complement of~$A$.
Semantically, $\code{-}A$ is simply a new atom,
with the additional condition that~$A$ and~$\code{-}A$ must not jointly hold.
A logic program~$\Pi$ containing complementary atoms is thus translated
to propositional theory
$\Phi[\Pi \cup \{\code{:-}~A\code{,}\code{-}A \mid A\in\ground{\mathcal{A}}\}]$.%
\footnote{Recall that $\ground{\mathcal{A}}$ consists of all variable-free atoms
built from predicates and functions in~$\Pi$.}
Observe that classical negation is merely a syntactic feature that can be
implemented via integrity constraints whose effect is to eliminate
any answer set candidate containing complementary atoms.

\begin{example}\label{ex:flies:neg}
The following logic program reformulates the one in Example~\ref{ex:flies}:
%
\lstinputlisting[nolol]{examples/bird.lp}
\lstinputlisting[firstnumber=3]{examples/flycn.lp}
%
Logically, classical negation is reflected by
(implicit) integrity constraints as follows:%
\marginlabel{By invoking\\
  \code{\mbox{~}gringo -t \textbackslash\\
        \mbox{~}examples/bird.lp \textbackslash\\
        \mbox{~}examples/flycn.lp}
the reader can observe that \gringo\ indeed produces the integrity constraint
in Line~7.}
%
\begin{lstlisting}[firstnumber=6]
:- flies(tux),    -flies(tux).
:- flies(tweety), -flies(tweety).
\end{lstlisting}
The additional integrity constraints do not yet change the semantics of the
program, which still has the answer sets already provided in Example~\ref{ex:flies:ground}
(of course, identifying
\code{\pred{neg\mus flies}(\const{tux})} with
\code{\pred{-flies}(\const{tux})} and
\code{\pred{neg\mus flies}(\const{tweety})} with
\code{\pred{-flies}(\const{tweety})}).
This situation changes if we add the following fact:
\begin{lstlisting}[firstnumber=8]
flies(tux).
\end{lstlisting}
While the program from Example~\ref{ex:flies} still admits two answer sets,
both containing 
\code{\pred{flies}(\const{tux})} and
\code{\pred{neg\mus flies}(\const{tux})},
there no longer is any answer set for our new program using classical negation.
In fact, answer set candidates that contain both
\code{\pred{flies}(\const{tux})} and
\code{\pred{-flies}(\const{tux})} violate the integrity constraint in Line~6.
\eexample
\end{example}

\subsubsection{Disjunction}\label{subsec:gringo:disjunction}

Disjunctive logic programs permit connective~\code{|} between atoms in rule heads.
An additional case of translation~$\psi$ from Section~\ref{subsec:semantics}
reflects the semantics of disjunction:
\begin{itemize}
\item $\psi[G\code{|}H]=\psi[G]\vee\psi[H]$.
\end{itemize}
%
The concept of level-restrictedness (cf.\ Section~\ref{subsec:lambda}) is extended
to disjunctive programs by,
for a disjunctive rule~$r$, every predicate~$p/i$
of some atom in the head of~$r$, and each $V\in\vars{r}$,
requiring that there is some $A\in\binders{V}{r}$ such that
$\lambda(p/i)>\lambda(q/j)$ for predicate~$q/j$ of~$A$.
Furthermore, a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
must not contain any rule with a (non-trivial) disjunctive head.

The following rule combines the ones in Line~3 and~4 from Example~\ref{ex:flies:neg}:
\begin{lstlisting}[numbers=none]
flies(X) | -flies(X) :- bird(X).
\end{lstlisting}
In general, the use of disjunction however increases
computational complexity~\cite{eitgot95a}.
This is why \clingo%
\footnote{Run as a monolithic system performing both grounding and solving.}
and solvers like 
\assat~\cite{linzha04a},
\clasp~\cite{gekanesc07b},
\nomorepp~\cite{angelinesc05c},
\smodels~\cite{siniso02a}, and
\smodelscc~\cite{warsch04a}
do not work on disjunctive programs.
Rather,
\claspD~\cite{drgegrkakoossc08a},
\cmodels~\cite{gilima06a,lierler05a}, or
\gnt~\cite{janisesiyo06a}
needs to be used for solving a disjunctive program.%
\footnote{System \dlv~\cite{dlv03a} also deals with disjunctive programs,
  but it uses a different syntax than presented here.}
We thus suggest to use ``choice constructs'' (cf.\ Section~\ref{subsec:gringo:aggregate})
instead of disjunction, unless the latter is required for complexity reasons
(see~\cite{eitpol06a} for an implementation methodology).

\subsubsection{Built-In Arithmetic Functions}\label{subsec:gringo:arith}

\gringo\ and \clingo\ support a number of arithmetic functions that
are evaluated during grounding.
The following symbols are used for these functions:
\code{+} (addition),
\code{-} (subtraction),
\code{*} (multiplication),
\code{/} or \code{div} (integer division),
\code{mod} (modulo function),
\code{abs} (absolute value),
\code{\textasciitilde} (bitwise complement),
\code{\&}  (bitwise AND),
\code{?}   (bitwise OR), and
\code{\^} (bitwise exclusive OR).

\begin{example}\label{ex:arith:fun}
The usage of arithmetic functions is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program,
  obtained after evaluating all arithmetic functions,
  can be inspected by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithf.lp}}
%
\lstinputlisting{examples/arithf.lp}
%
Note that variables~\var{L} and~\var{R} are instantiated to~\const{7} and~\const{2},
respectively, before arithmetic evaluations.
Consecutive and non-separative (e.g., before ``\code{(}'')
spaces can also be dropped,
while spaces around tokens \code{div} and \code{mod} are mandatory.
Furthermore, the argument of function \code{abs} must be enclosed in parentheses.
In Line~9, observe that there is a unary version of \code{-},
``\code{- \var{R}}'' standing for ``\code{0 - \var{R}}.''
The four bitwise functions apply to signed integers,
using the complement on two of a negative integer.
\eexample
\end{example}

It is important to note that variables
in the scope of an arithmetic function are not necessarily bound 
(in the sense of Section~\ref{subsec:lambda}) by a corresponding atom.
For instance, atom \code{\pred{p}(\var{X}+1,\var{X})} belongs to
$\binders{\var{X}}{r}$, while atom \code{\pred{p}(\var{X}+1,\var{Y})}
does not belong to $\binders{\var{X}}{r}$ because it contains~\var{X}
only in the scope of an arithmetic function.

\subsubsection{Built-In Comparison Predicates}\label{subsec:gringo:comp}

The following built-in predicates permit term comparisons
within the bodies of rules and on the right-hand side of conditions
(cf.\ Section~\ref{subsec:gringo:condition}):
\code{==} (equal),
\code{!=} (not equal),
\code{<} (less than),
\code{<=} (less than or equal),
\code{>} (greater than),
\code{>=} (greater than or equal).

\begin{example}\label{ex:arith:pred}
The usage of comparison predicates is illustrated by the logic program:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithc.lp}}
%
\lstinputlisting{examples/arithc.lp}
%
The last two lines hint at the fact that arithmetic functions are evaluated
before comparison predicates, so that the latter actually compare integers.

All comparison predicates can also be used with constants,
as in the next program:%
\marginlabel{%
  As above, invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/symbc.lp}
  yields the unique answer set of the program in terms of facts.}
%
\lstinputlisting{examples/symbc.lp}
%
Finally, note that~\code{==} and~\code{!=} also apply to
general terms over functions of non-zero arity as well as to
mixed integer and non-integer arguments.%
\footnote{%
  Such functionalities are currently not supported for the other four
  comparison predicates,
  where (after instantiation and arithmetic evaluation)
  both arguments must be either integers or (symbolic) constants.
}
\eexample
\end{example}

Importantly, a comparison predicate cannot bind variables,
that is, it does not belong to $\binders{\var{X}}{r}$ for any variable~\var{X}.
Also note that built-in predicates are not considered by level mappings~$\xi$
in the context of stratification (cf.\ Section~\ref{subsec:strat}), or equivalently,
one may assume that built-in predicates are mapped to level~$0$ and all other
predicates to levels greater than~$0$.

\subsubsection{Assignments}\label{subsec:gringo:assign}

Built-in predicate~\code{=} can be used in the body of a rule
(or on the right-hand side of a condition)
to assign a term on its right-hand side to a variable on its left-hand side.
Note that ``\code{\var{X} = }$t$'' belongs to~$\binders{\var{X}}{r}$,
provided that~$t$ is variable-free or that its variables are bound
by atoms other than assignments mentioning~\var{X} on their right-hand side.
Thus, cyclic assignments over otherwise unbound variables are prohibited.

\begin{example}\label{ex:assign}
The next program demonstrates how terms can be assigned to variables:%
\marginlabel{%
  The unique answer set of the program is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/assign.lp}}
%
\lstinputlisting{examples/assign.lp}
%
Line~3 contains four assignments, where the left-hand sides directly or indirectly
depend on~\var{X} and~\var{Y}.
These two variables are bound in Line~4 via atoms of predicate \pred{num}/$1$.
Also observe the different usage and role of built-in comparison predicate~\code{==}.
\eexample
\end{example}

\subsubsection{Intervals}\label{subsec:gringo:interval}

In Line~1 of Example~\ref{ex:assign},
there are five facts~\code{\pred{num}($k$)}
over consecutive integers~$k$.
For a more compact representation,
\gringo\ and \clingo\ support integer intervals of the form $i$\code{..}$j$,
where~$i$ and~$j$ are integers.
Such an interval represents each integer~$k$ such that $i\leq k\leq j$,
and intervals are expanded during grounding.

\begin{example}\label{ex:int}
The next program makes use of integer intervals:
%
\lstinputlisting{examples/int.lp}
%
The facts in Line~1 and~2 are expanded as follows:
%
\begin{lstlisting}[numbers=none]
num(1).    num(2).   num(3).   num(4).   num(5).
top5(5).  top5(6).  top5(7).  top5(8).  top5(9).
\end{lstlisting}
%
By instantiating~\var{X} to~\const{9},
the rule in Line~4 becomes:
%
\begin{lstlisting}[numbers=none]
top5num(1..5,5..9) :- num(5..9), top5(1..5), top(9).
\end{lstlisting}
%
It is expanded to the cross product
$(\code{\const{1}..\const{5}})\times
 (\code{\const{5}..\const{9}})\times
 (\code{\const{5}..\const{9}})\times
 (\code{\const{1}..\const{5}})$
of intervals:
%
\begin{lstlisting}[numbers=none,escapechar=@]
top5num(1,5) :- num(5), top5(1), top(9).
top5num(2,5) :- num(5), top5(1), top(9).
        @\rlap{\vdots}@
top5num(5,5) :- num(5), top5(1), top(9).
top5num(1,6) :- num(5), top5(1), top(9).
top5num(2,6) :- num(5), top5(1), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@
top5num(5,9) :- num(5), top5(1), top(9).
top5num(1,5) :- num(6), top5(1), top(9).
top5num(2,5) :- num(6), top5(1), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(1), top(9).
top5num(1,5) :- num(5), top5(2), top(9).
top5num(2,5) :- num(5), top5(2), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@         @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(4), top(9).
top5num(1,5) :- num(5), top5(5), top(9).
top5num(2,5) :- num(5), top5(5), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@
top5num(5,9) :- num(5), top5(5), top(9).
top5num(1,5) :- num(6), top5(5), top(9).
top5num(2,5) :- num(6), top5(5), top(9).
        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
top5num(5,9) :- num(9), top5(5), top(9).
\end{lstlisting}
%
Note that only the rules with
\code{\pred{num}(\const{5})} and
\code{\pred{top5}(\const{5})}
\marginlabel{%
  Again the unique answer set is obtained via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/int.lp}}
%
in the body actually contribute to the unique answer set
of the above program by deriving all atoms \code{\pred{top5num}($m$,$n$)}
for $\const{1}\leq m\leq\const{5}$ and $\const{5}\leq n\leq\const{9}$.
\eexample
\end{example}

As with built-in comparison functions,
an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
cannot be used to bind the variable.

\subsubsection{Conditions}\label{subsec:gringo:condition}

Conditions allow for instantiating variables to
collections of terms within a single rule.
This is particularly useful for encoding conjunctions or disjunctions over
arbitrarily many ground atoms as well as for the compact representation of aggregates
(cf.\ Section~\ref{subsec:gringo:aggregate}).
The symbol~``\code{:}'' is used to formulate conditions.

\begin{example}\label{ex:cond}
The following program uses conditions in a rule body and in a rule head:
%
\lstinputlisting{examples/cond.lp}
%
We are particularly interested in the rules in Line~5 and~6,
instantiated as follows:%
%
\begin{lstlisting}[firstnumber=5]
meet :- available(jane), available(john).
on(mon) | on(tues) | on(wed) | on(thurs) | on(fri) :- meet.
\end{lstlisting}
%
The conjunction in Line~5 is obtained by replacing~\var{X} in
\marginlabel{%
  The reader can reproduce these ground rules by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/cond.lp}}
\code{\pred{available}(\var{X})} with all ground terms~$t$ such that
\code{\pred{person}($t$)} holds, namely, $t=\const{jane}$ and $t=\const{john}$.
Furthermore, the condition in the head of the rule in Line~6 turns into
a disjunction over all ground instances of
\code{\pred{on}(\var{X})} where~\var{X} is substituted by some term~$t$
such that
\code{\pred{day}($t$)} holds.
That is, conditions in the body and in the head of a rule
are expanded to different basic language constructs.

Several conditions may be connected via~``\code{:},''
as in the following additional rules:
%
\lstinputlisting[firstnumber=7]{examples/twocond.lp}
%
Observe that we may use the same atom, viz., \code{\pred{day}(\var{X})},
both on the left-hand and on the right-hand side of~``\code{:}.''
Furthermore, negative literals like \code{not \pred{weekend}(\var{X})}
can occur on both sides of a condition.
Note that literals on the right-hand side of a condition are connected conjunctively, that is,
all of them must hold for ground instances of an atom in front
of the condition.
Thus, the instantiated rule in Line~8 looks as follows:
%
\begin{lstlisting}[firstnumber=8]
on_week :- day(mon), day(tues), day(wed), day(thurs), day(fri).
\end{lstlisting}
%
The atoms in the body of this rule follow from facts, so that
the rule can be simplified to a fact \code{\pred{on\_week}.}
(as done by \gringo).
\eexample
\end{example}

There are three important issues about the correct usage of conditions.
First, all predicates of atoms on the right-hand side of a condition
must be either domain predicates (cf.\ Section~\ref{subsec:strat}) or built-in,
which is due to the fact that conditions are evaluated during grounding.%
\footnote{%
  The bodies of rules in a stratified subprogram (cf.\ Section~\ref{subsec:strat})
  may contain conditions.
  For a rule~$r$ and
  $L_0\code{:}L_1\code{:}\dots\code{:}L_n$
  in the body of~$r$,
  let $A_0\in\nbody{r}$ if $L_0=\code{not}~A_0$
  and $L_0\in\pbody{r}$ otherwise,
  and for $1\leq i\leq n$, 
  assume $A_i\in\nbody{r}$ for atom~$A_i$
  such that $L_i=A_i$ or $L_i=\code{not}~A_i$, respectively.
  }
Second, any variable occurring within a condition is considered as \emph{local},
that is, a condition cannot be used to bind variables outside itself.
In turn, variables outside conditions are \emph{global}, and each variable
within an atom in front of a condition must occur on the right-hand side or
be global.
Third, global variables take priority over local ones, that is,
they are instantiated first.
As a consequence, a local variable that also occurs globally is substituted by a term
before the ground instances of a condition are determined.
Hence, the names of local variables must be chosen with care,
making sure that they do not accidentally match the names of global variables.

\subsubsection{Pooling}\label{subsec:gringo:pool}

Symbol~``\code{;}'' allows for pooling alternative terms to be used as
argument within an atom, thus, specifying rules more compactly.
An atom written in the form \code{\pred{p}($\dots$,X;Y,$\dots$)}
abbreviates two options:
\code{\pred{p}($\dots$,X,$\dots$)} and \code{\pred{p}($\dots$,Y,$\dots$)}.
Pooled arguments in an atom of a rule body 
(or on the right-hand side of a condition) are expanded to a conjunction
of the options within the same body (or within the same condition),
while they are expanded to multiple rules when occurring in the head.

\begin{example}\label{ex:pool}
The following logic program makes use of pooling:
%
\lstinputlisting{examples/pool.lp}
%
Let us consider instantiations of the rule in Line~3 obtained with substitution 
$\{\var{A}\mapsto\const{a},\linebreak[1]\var{B}\mapsto\const{b},
   \var{M}\mapsto\const{1},\var{N}\mapsto\const{2}\}$.
First note that \pred{mix}/$2$ and \pred{-mix}/$2$ each admit four options,
corresponding to the cross product of $\{\const{a},\const{b}\}$ substituted
for~\var{A} and~\var{B}, respectively, together with $\{\const{1},\const{2}\}$
substituted for~\var{M} and~\var{N}.
While the instances obtained for \pred{mix}/$2$ give rise to four rules,
the instances for \pred{-mix}/$2$ jointly belong to the body.
The (repeated) body also contains two instance each of \pred{sym}/$1$ and of \pred{num}/$1$.
We thus get the rules:%
\marginlabel{%
  Simplified versions of these rules are produced via call:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/pool.lp}}
%
\begin{lstlisting}[numbers=none]
mix(a,1) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
mix(a,2) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
mix(b,1) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
mix(b,2) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
  not -mix(1,b), not -mix(2,a), not -mix(2,b).
\end{lstlisting}
%
Finally, we note that pooling is also possible with arguments of built-in predicates.
\eexample
\end{example}

\subsubsection{Aggregates}\label{subsec:gringo:aggregate}

An aggregate is an operation on a multiset of weighted literals that evaluates to an integer.
In combination with arithmetic comparisons, we can extract a truth value from
an aggregate's evaluation, thus, obtaining an aggregate atom.
Customizing the notation in~\cite{ferraris05a},
we consider ground \emph{aggregate atoms} of the form:
%
\begin{equation}\label{eq:aggregate:atom}
  l~ 
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
\end{equation}
%
where~$l$ and~$u$ are integers each of which can possibly be omitted,
$\mathrm{op}$ is a function from multisets of integers to an integer,
each $L_i$ is a literal of the form~$A$ or $\code{not}~A$
for some atom~$A\in\ground{\mathcal{A}}$,
and each~$w_i$ is an integer.
The intuitive reading of an aggregate atom is that the value returned by~$\mathrm{op}$
applied to all weights~$w_i$ such that~$L_i$ holds should be in-between
lower bound~$l$ and upper bound~$u$ (inclusively).
If~$l$ or~$u$ is omitted,
it means that the aggregate's value is not constrained from below or above,
respectively.

For an aggregate atom~$\mathit{Agg}$ as in~(\ref{eq:aggregate:atom}),
we let $\atom{\mathit{Agg}}=\{L_i\in\nolinebreak\ground{\mathcal{A}}\mid 1\leq i\leq n\}$
denote the set of atoms occurring positively in~$\mathit{Agg}$.
Currently, \gringo\ and \clingo\ support aggregate operations
\const{sum}, \const{max}, and \const{min}.%
\footnote{%
  The additional operation \const{count} is a shorthand for a \const{sum}
  aggregate in which all weights are~\const{1}.}
We can now extend translations~$\phi$ and~$\psi$ from Section~\ref{subsec:semantics}
to aggregate atoms, 
using the approach in~\cite{ferraris05a} but
applying the ``choice semantics'' for aggregates
in heads of rules~\cite{ferlif05a,siniso02a}:%
\footnote{%
  The translation~$\phi$ we provide for \const{sum} aggregates
  assumes that there are no negative weights for literals,
  while the more sophisticated translation in~\cite{ferraris05a} also
  captures negative weights.
  All current solvers work with positive weights only, and thus
  \gringo\ and \lparse\ use a compilation technique~\cite{siniso02a}
  to eliminate negative weights, which are not permitted in
  \lparse's output format~\cite{lparseManual}.
  As this technique may lead to counterintuitive results~\cite{ferraris05a},
  we strongly recommend not to use negative weights within \const{sum} aggregates.}
\begin{enumerate}
\item % op %
$\phi[
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ] = 
 \top$,
\item % l op u %
$\begin{array}[t]{@{}r@{}l@{}}
  \phi[
  l~
 &
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ] = 
 {} \\ 
 &
  \phi[
  l~
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ]
  \wedge
  \phi[
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ]\text{,}
 \end{array}$
\item % l sum %
$\begin{array}[t]{@{}r@{}l@{}}
  \phi[
  l~
 &
  \const{sum} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ] = 
 {} \\ 
 &
  \big(
  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},l\leq w_{i_1}+\dots+w_{i_j}}
  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
  \big)\text{,}
 \end{array}$
\item % sum u %
$\begin{array}[t]{@{}r@{}l@{}}
  \phi
  [
 &
  \const{sum} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ] = 
 {} \\ 
 &
  \neg
  \big(
  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},u< w_{i_1}+\dots+w_{i_j}}
  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
  \big)\text{,}
 \end{array}$
\item % l max %
 $\phi[
  l~
  \const{max} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ] = 
 \hspace{6pt}
  \big(
  \bigvee_{i\in\{1,\dots,n\},l\leq w_i}
  \phi[L_i]
  \big)$,
\item % max u %
 $\phi[
  \const{max} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ] = 
  \neg
  \big(
  \bigvee_{i\in\{1,\dots,n\},u< w_i}
  \phi[L_i]
  \big)$,
\item % l min %
 $\phi[
  l~
  \const{min} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  ] = 
  \neg
  \big(
  \bigvee_{i\in\{1,\dots,n\},w_i< l}
  \phi[L_i]
  \big)
 $,
\item % min u %
 $\phi[
  \const{min} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  u
  ] = 
  \hspace{6pt}
  \big(
  \bigvee_{i\in\{1,\dots,n\},w_i\leq u}
  \phi[L_i]
  \big)
 $, and
\item % psi %
$\begin{array}[t]{@{}r@{}l@{}}
  \psi[
  \lhd
 &~
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  \rhd
  ] = 
% {} \\ 
% &
   \phi[
  \lhd~
  \mathrm{op} 
  \code{[}
  L_1\code{=}w_1\code{,}
  \dots\code{,}
  L_n\code{=}w_n
  \code{]}
  \rhd
  ]
  \wedge
 {} \\ 
 &
  \big(
  \bigwedge_{A\in\atom{
    \lhd~
    \mathrm{op} 
    \code{[}
    L_1\code{=}w_1\code{,}
    \dots\code{,}
    L_n\code{=}w_n
    \code{]}
    \rhd
  }}
  (A\vee\neg A)
  \big)\text{.}
 \end{array}$
\end{enumerate}
%
The first item reflects that an unconstrained aggregate atom always holds,
while aggregate atoms with a lower and an upper bound are split up into two
conjunctively connected parts (Item~2).
Item~3--8 specify these parts individually for~\const{sum},~\const{max}, and~\const{min}
aggregates, respectively.
Provided that only non-negative weights are used within \const{sum}s,
all three aggregates are subject to one comparison that is monotone,
meaning that by default it evaluates to \textit{false} so that truth must be established.
For \const{sum} and \const{max} aggregates, these monotone comparisons are related
to lower bounds (Item~3 and~5),
while an upper bound behaves monotonically for \const{min} (Item~8).
Vice versa, the opposite comparison is antimonotone, that is,
by default it evaluates to \textit{true} unless too many or improper literals hold.
This is the case for comparisons to upper bounds with aggregates
\const{sum} and \const{max} (Item~4 and~6), while it concerns 
the lower bound for \const{min} (Item~7).
Finally, translation~$\psi$ of an aggregate atom in the head of a rule consists of two parts:
first, $\phi$ is used like in the body of a rule
(in Item~9, $\lhd$ and~$\rhd$ are used as ``wildcards'' for provided or omitted 
 lower and upper bounds),
second, a disjunction $(A\vee\neg A)$ is added for each atom~$A$ occurring positively.
Even though such disjunctions are tautological,
they are important under answer set semantics (cf.\ Section~\ref{subsec:semantics}),
as they permit deriving an arbitrary subset of atoms from an aggregate in the head
(which explains the name  ``choice semantics'').
Beyond this special treatment in rule heads,
the truth values of aggregate atoms result from their literals as
may be intuitively expected,
and Item~1--8 basically do nothing but formalizing the standard meanings
of aggregate operations.
Note that the above translations merely define the semantics of aggregates,
which does not imply that solvers unfold them in this way.
In fact, \clasp\ natively supports aggregates without translating them.

As regards syntactic representation,
weight~\const{1} is considered a default,
so that $L_i\code{=}\const{1}$ can simply be written as $L_i$.
For instance, the following sets of (weighted) literals are the same
when combined with any kinds of aggregate operation and bounds:
\begin{lstlisting}[numbers=none,escapechar=@]
[a=1, not b=1, c=2]  @\textnormal{and}@
[a,   not b,   c=2]@\textnormal{.}@
\end{lstlisting}
%
Furthermore,
keyword \const{sum} may be omitted, which in a sense makes \const{sum} the default
aggregate operation.
In fact, the following aggregate atoms are synonyms:
\begin{lstlisting}[numbers=none,escapechar=@]
2 sum [a, not b, c=2] 3  @\textnormal{and}@
2     [a, not b, c=2] 3@\textnormal{.}@
\end{lstlisting}
By omitting keyword \const{sum}, we obtain the same notation as the one of 
so-called ``weight constraints''~\cite{siniso02a,lparseManual}, which are actually aggregate atoms
whose operation is addition.

It is important to note that the (weighted) literals within an aggregate
belong to a multiset.
In particular, if there are multiple occurrences
$L\code{=}w_1,\dots,L\code{=}w_k$
of a literal~$L$, in combination with \const{min} and \const{max},
it is not the same like having $L\code{=}w_1+\dots+w_k$.
To see this, note that the program consisting of the facts:
\begin{lstlisting}[numbers=none]
2 max [a=2].  2 min [a=2].
\end{lstlisting}
has $\{\code{a}\}$ as its unique answer set, while there is no answer set for:
\begin{lstlisting}[numbers=none]
2 max [a,a].  2 min [a,a].
\end{lstlisting}

If literals ought not to be repeated,
we can use \const{count} instead of \const{sum}.
Syntactically, \const{count} requires curly instead of square
brackets, and there must not be any weights within a \const{count} aggregate.
Regarding semantics,
% \begin{equation*}%\label{eq:aggregate:atom}
$(
  l~ 
  \const{count} 
  \code{\char`\{}
  L_1\code{,}
  \dots\code{,}
  L_n
  \code{\char`\}}
  u
)$ 
%\end{equation*}
reduces to
$(
  l~ 
  \const{sum} 
  \code{[}
  L_1\code{=}\const{1}\code{,}
  \dots\code{,}
  L_m\code{=}\const{1}
  \code{]}
  u
)$,
where $\{L_1,\dots,L_m\}=\{L_1,\dots,L_n\}$ is obtained by dropping
repeated literals.
Of course, the use of~$l$ and~$u$ is optional also with \const{count}.
As an example, note that the following aggregate atoms express the same:
\begin{lstlisting}[numbers=none,escapechar=@]
1 sum   [a=1, not b=1]     1  @\textnormal{and}@ 
1 count {a,a, not b,not b} 1@\textnormal{.}@  
\end{lstlisting}
Keyword \const{count} can be omitted (like \const{sum}),
so that the following are synonyms:
\begin{lstlisting}[numbers=none,escapechar=@]
1 count {a, not b} 1  @\textnormal{and}@ 
1       {a, not b} 1@\textnormal{.}@  
\end{lstlisting}
The last notation is similar to the one of so-called
``cardinality constraints''~\cite{siniso02a,lparseManual},
which are aggregate atoms using counting as their operation.

After considering the syntax and semantics of ground aggregate atoms,
we now turn our attention to non-ground aggregates.
Regarding contained variables, an atom occurring in an aggregate behaves
similar to an atom on the left-hand side of a condition
(cf.\ Section~\ref{subsec:gringo:condition}).
That is, any variable occurring within an aggregate is a priori local,
and it must be bound via a variable of the same name that is global
or that occurs on the right-hand side of a condition
(with the atom containing the variable in front).
As with local variables of conditions, global variables take priority
during grounding, so that the names of local variables must be chosen
with care to avoid accidental clashes.
Beyond conditions (which are more or less the natural construct to use
for instantiating variables within an aggregate),
classical negation (cf.\ Section~\ref{subsec:gringo:negation}),
built-in arithmetic functions (cf.\ Section~\ref{subsec:gringo:arith}),
intervals (cf.\ Section~\ref{subsec:gringo:interval}), and 
pooling (cf.\ Section~\ref{subsec:gringo:pool})
can be incorporated as usual within aggregates,
where intervals and pooling are expanded like in the body of a rule.%
\footnote{%
  Assignments (cf.\ Section~\ref{subsec:gringo:assign}) and (currently)
  also built-in comparison predicates (cf.\ Section~\ref{subsec:gringo:comp})
  are permitted on the right-hand side of a condition only.}
The notions of level-restrictedness and stratification
(cf.\ Section~\ref{subsec:lambda} and~\ref{subsec:strat}) are extended
to programs with aggregates by
considering the predicates of all atoms (except for those on the right-hand side of a condition)
in an aggregate that is the head of a rule~$r$,
and by assuming $A\in\nbody{r}$ for all atoms~$A$ occurring in an aggregate in the body of~$r$.
Finally, note that aggregates without bounds are also permitted on the
right-hand sides of assignments, but using this feature is only recommended
for aggregates whose atoms belong to domain predicates because space blow-up
can become a bottleneck otherwise.
The following example, making exhaustive use of aggregates,
nonetheless demonstrates this and other features.

\begin{example}\label{ex:aggr}
Consider a situation where an informatics student
wants to enroll for a number of courses at the beginning of the new term.
In the university calendar, eight courses are found eligible,
and they are represented by the following facts:
%
\lstinputlisting[xrightmargin=-20pt,lastline=8]{examples/aggr.lp}
%
In an instance of \const{course}/$3$,
the first argument is a number identifying one of the eight courses,
and the third argument provides the course's contact hours per week.
The second argument stands for a subject area:
\const{1} corresponding to ``theoretical informatics,''
\const{2}               to ``practical informatics,''
\const{3}               to ``technical informatics,''
and~\const{4}               to ``applied informatics.''
For instance, atom \code{\const{course}(\const{1},\const{2},\const{5})}
expresses that course~\const{1} accounts for~\const{5} contact hours per week
that may be credited to subject area~\const{2} (``practical informatics'').
Observe that a single course is usually eligible for multiple
subject areas.

After specifying the above facts,
the student starts to provide personal constraints on the courses to enroll.
The first condition is that~\const{3} to~\const{6} courses should be enrolled:
%
\lstinputlisting[nolol,firstline=11,lastline=11,firstnumber=9]{examples/aggr.lp}
%
Instantiating the above \const{count} aggregate
yields the following ground rule:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/aggr.lp}}
%
\begin{lstlisting}[firstnumber=9,stepnumber=9]
3 { enroll(1), enroll(2), enroll(3), enroll(4), 
    enroll(5), enroll(6), enroll(7), enroll(8) } 6.
\end{lstlisting}
%
Observe that an instance of atom~\code{\const{enroll}(\var{C})} is included for each
instantiation of~\var{C} such that \code{\const{course}(\var{C},\var{S},\var{H})}
holds for some values of~\var{S} and~\var{H}.
Duplicates resulting from distinct values for~\var{S} are removed, thus,
obtaining the above set of ground atoms.

The next constraints of the student regard the subject areas of enrolled courses:
%
\lstinputlisting[nolol,firstline=13,lastline=15,firstnumber=10]{examples/aggr.lp}
%
Each of the three integrity constraints above contains a \const{sum} aggregate,
using default weight~\const{1} for literals.
Recalling that \const{sum} aggregates operate on multisets,
duplicates are not removed.
Thus, the integrity constraint in Line~10 is instantiated as follows:
%
\begin{lstlisting}[firstnumber=10,stepnumber=10]
:- [ enroll(1) = 1, enroll(1) = 1, 
     enroll(2) = 1, enroll(2) = 1,
     enroll(3) = 1, enroll(3) = 1,
     enroll(4) = 1, enroll(4) = 1, enroll(4) = 1,
     enroll(5) = 1, enroll(5) = 1,
     enroll(6) = 1, enroll(6) = 1,
     enroll(7) = 1, enroll(7) = 1, enroll(7) = 1,
     enroll(8) = 1, enroll(8) = 1 ] 10.
\end{lstlisting}
%
Note that courses~\const{4} and~\const{5} count three times because
they are eligible for three subject areas, viz., there are three
distinct instantiations for~\var{S} in
\code{\const{course}(\var{4},\var{S},\const{3})} and
\code{\const{course}(\var{5},\var{S},\const{4})}, respectively.
Comparing the above ground instance,
the meaning of the integrity constraint in Line~10 is that the 
number of eligible subject areas over all enrolled courses
must be more than~\const{10}.
Similarly, the integrity constraint in Line~11 expresses the requirement
that at most one course of subject area~\const{2} (``practical informatics'')
is not enrolled,
while Line~12 stipulates that the enrolled courses
amount to less than six nominations of
subject area~\const{3} (``technical informatics'') 
or~\const{4} (``applied informatics'').
Also note that, given the facts in Line~1--8,
we could equivalently have used \const{count} rather than \const{sum} in Line~11,
but not in Line~10 and~12.

The remaining constraints of the student deal with contact hours.
To express them, we first introduce an auxiliary rule and a fact:
%
\lstinputlisting[nolol,firstline=17,lastline=18,firstnumber=13]{examples/aggr.lp}
%
The rule in Line~13 projects instances of \pred{course}/$3$ to
\pred{hours}/$2$, thereby, dropping courses' subject areas.
This is used to not consider the same course multiple times within the following
integrity constraints:
%
\lstinputlisting[nolol,firstline=20,lastline=22,firstnumber=15]{examples/aggr.lp}
%
As Line~15 shows,
we may use default negation via ``\code{not}'' in front of aggregate atoms,
and bounds may be specified in terms of variables.
In fact, by instantiating~\var{M} to~\const{20},
we obtain the following ground instance of the integrity constraint in Line~15:
%
\begin{lstlisting}[firstnumber=15,stepnumber=15]
:- not 18 [ enroll(1) = 5, enroll(2) = 4, 
            enroll(3) = 6, enroll(4) = 3,
            enroll(5) = 4, enroll(6) = 2,
            enroll(7) = 4, enroll(8) = 5 ] 20.
\end{lstlisting}
%
The above integrity constraint states that the \const{sum} of contact hours per week
must lie in-between~\const{18} and~\const{20}.
Note that the \const{min} and \const{max} aggregates in Line~16 and~17, respectively,
work on the same (multi)set of weighted literals as in Line~15.
While the integrity constraint in Line~16 stipulates that any course to enroll
must include more than~\const{2} contact hours,
the one in Line~17 prohibits enrolling for courses of~\const{6} or more contact hours.
Of course, the last two requirements could also be formulated as follows:
%
\begin{lstlisting}[firstnumber=16]
:- enroll(C), hours(C,H), H <= 2.
:- enroll(C), hours(C,H), H >= 6.
\end{lstlisting}

Finally, the following rules illustrate the use of aggregates within assignments:
%
\lstinputlisting[nolol,firstline=24,lastline=25,firstnumber=18]{examples/aggr.lp}
%
Note that the above aggregates have already been used in Line~9 and~15, respectively,
where keywords \const{count} and \const{sum} have been omitted for convenience.
These keywords can be dropped here too, and we merely include them to show the
more verbose notations of \const{count} and \const{sum} aggregates.
However, the usage of aggregates in the last two lines is different from before,
as they now serve to assign an integer to a variable~\var{N}.
In this context, bounds are not permitted, and so none are provided in Line~18 and~19.
The effect of these two lines is that the student can read off the number of
courses to enroll and the amount of contact hours per week from instances of
\pred{courses}/$1$ and \pred{hours}/$1$ belonging to an answer set.%
\marginlabel{%
  To compute the unique answer set of the program, invoke:\\
  \code{\mbox{~}gringo \textbackslash\\
        \mbox{~}examples/aggr.lp | \textbackslash\\
        \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
        \mbox{~}examples/aggr.lp}}
%
In fact, running \clasp\ shows the student that a unique
collection of~\const{5} courses to enroll satisfies all requirements:
the courses~\const{1}, \const{2}, \const{4}, \const{5}, and~\const{7},
amounting to~\const{20} contact hours per week.

Although the above program does not reflect this possibility,
it should be noted that (as has been mentioned in Section~\ref{subsec:gringo:condition})
multiple literals may be connected via ``\code{:}'' in order to construct
composite conditions within an aggregate.
As before, the predicates of atoms on the right-hand side of such conditions
must be either domain predicates or built-in.
Furthermore, the usage of non-domain predicates within an aggregate
on the right-hand side of an assignment (like \pred{enroll}/$1$ in Line~18 and~19 above)
is not recommended in general because the space blow-up may be significant.
\eexample
\end{example}

\subsubsection{Optimization}\label{subsec:gringo:optimize}

Optimization statements extend the basic question of
whether a set of atoms is an answer set to
whether it is an optimal answer set.
To support this reasoning mode, \gringo\ and \clingo\ adopt the
optimization statements of \lparse~\cite{lparseManual},
indicated via keywords \const{maximize} and \const{minimize}.
Syntactically, a \const{maximize} or \const{minimize} statement
is similar to a fact whose head is a \const{count} or a \const{sum}
aggregate (without bounds),
viz.,
``$\textrm{opt}\code{\char`\{}\dots\code{\char`\}}\code{.}$'' or
``$\textrm{opt}\code{[}\dots\code{]}\code{.}$'' where
$\textrm{opt}\in\{\const{maximize},\linebreak[1]\const{minimize}\}$.
As an optimization statement does not admit a body, any (local) variable
in it must also occur in an atom (over a domain or built-in predicate)
on the right-hand side of a condition (cf.\ Section~\ref{subsec:gringo:condition})
within the optimization statement.
In multiset notation (square brackets),
weights may be provided as with \const{sum} aggregates. %,
% where~\const{1} is used as default for omitted weights.
In set notation (curly brackets),
duplicates of literals are removed
as with \const{count} aggregates.

The semantics of an optimization statement is intuitive:
an answer set is \emph{optimal} if the sum of weights
(using~\const{1} for unsupplied weights)
of literals that hold is maximal or minimal, as required by the statement,
among all answer sets of the given program.
This definition is sufficient if a single optimization statement is specified
along with a logic program.
Unfortunately, the syntax used by \lparse\ and adopted by \gringo\ and \clingo\
does not provide any means to explicitly declare priorities among multiple
optimization statements.
A technique how to compile multiple optimization statements into a single one
is discussed in~\cite{siniso02a},
but it can lead to large integers
(probably rather unnatural for a user to assign).
Rather than restricting to a single optimization statements,
a sequence of them is thus permitted,
where a statement provided later on takes priority over previous ones.
That is, the last optimization statement in program is the most significant one,
and previous ones are only considered if answer sets agree on the sum of 
weights of its literals.
In this way, multiple optimization statements are arranged into a total order,
which is the inverse order of occurrence.
In contrast to the declarative answer set semantics, the semantics of
optimization statements is thus order-dependent,
requiring special care when providing more than one of them.

\begin{example}\label{ex:opt}
To illustrate optimization, we consider a hotel booking situation
where we want to chose one among five available hotels.
The hotels are identified via numbers assigned in descending order of stars.
Of course, the more stars a hotel has the more it costs per night.
As an ancillary information, we know that hotel~\const{4} is located
on a main street, which is why we expect its rooms to be noisy.
This knowledge is specified in Line~1--5 of the following program:
%
\lstinputlisting[xrightmargin=-26pt]{examples/opt.lp}
%
Line~6--8 contribute optimization statements in inverse order of significance,
according to which we want to chose the best hotel to book.
The most significant optimization statement in Line~8 states that
avoiding noise is our main priority.
The secondary optimization criterion in Line~7 consists of
minimizing the cost per star.
Finally, the third optimization statement in Line~6 specifies that we want
to maximize the number of stars among hotels that are otherwise indistinguishable.
The optimization statements in Line~6--8 are instantiated as follows:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/opt.lp}}
%
\begin{lstlisting}[firstnumber=6,breaklines,breakindent=48pt]
maximize [ hotel(1) = 5, hotel(2) = 4, hotel(3) = 3,    hotel(4) = 3, hotel(5) = 2 ].
minimize [ hotel(1) = 34, hotel(2) = 35, hotel(3) = 30,   hotel(4) = 25, hotel(5) = 30 ].
minimize [ noisy = 1 ].
\end{lstlisting}
If we now use \clasp\ to compute an optimal answer set,%
\marginlabel{%
  To compute the unique optimal answer set, invoke:\\
  \code{\mbox{~}gringo \textbackslash\\
        \mbox{~}examples/opt.lp | \textbackslash\\
        \mbox{~}clasp -n 0}\\
  or alternatively:\\
  \code{\mbox{~}clingo -n 0 \textbackslash\\
        \mbox{~}examples/opt.lp}}
we find that hotel~\const{4} is not eligible because it implies \pred{noisy}.
Thus, hotel~\const{3} and~\const{5} remain as optimal w.r.t.\ the second most
significant optimization statement in Line~7.
This tie is broken via the least significant optimization statement in Line~6
because hotel~\const{3} has one star more than hotel~\const{5}.
We thus decide to book hotel~\const{3} offering~\const{3} stars
to cost~\const{90} per night.
\eexample
\end{example}

\subsubsection{Meta-Statements}\label{subsec:gringo:meta}

After considering the language of logic programs,
we now introduce external features going beyond the contents of a program.

\paragraph{Comments.}
To keep records of the contents of a logic program,
a logic program file may include comments.
A comment until the end of a line is initiated by symbol ``\code{\%},''
and a comment within one or over multiple lines is enclosed
by ``\code{\%*}'' and ``\code{*\%}.''
As an abstract example, consider:
%
\begin{lstlisting}[numbers=none,escapechar=@]
@\textit{logic program}@  %* enclosed comment *%  @\textit{logic program}@
@\textit{logic program}@  % comment till end of line
@\textit{logic program}@
%*
comment over multiple lines
*%
@\textit{logic program}@
\end{lstlisting}

\paragraph{Hiding Predicates.}
Sometimes, one may be interested only in a subset of the atoms belonging
to an answer set.
In order to suppress the atoms of ``irrelevant'' predicates from the output,
the \code{\#hide} declarative (in which~\code{\#} is optional) can be used.
The meanings of the following statements are indicated via accompanying comments:
%
\begin{lstlisting}[numbers=none]
#hide.          % Suppress all atoms in output
 hide.          % Same as "#hide."
#hide p/3.      % Suppress all atoms of predicate p/3 in output
 hide p/3.      % Same as "#hide p/3."
#hide p(X,Y,Z). % Same as "#hide p/3."
 hide p(X,Y,Z). % Same as "#hide p/3."
\end{lstlisting}
%
In order to selectively include the atoms of a certain predicate in the output,
one may use the \code{\#show} declarative (in which~\code{\#} is again optional),
Here are some examples:
%
\begin{lstlisting}[numbers=none]
#show p/3.      % Include all atoms of predicate p/3 in output
 show p/3.      % Same as "#show p/3."
#show p(X,Y,Z). % Same as "#show p/3."
 show p(X,Y,Z). % Same as "#show p/3."
\end{lstlisting}
%
A typical usage of \code{\#hide} and \code{\#show} is to hide all predicates
via ``\code{\#hide.}'' and to selectively re-add atoms of certain predicates
\code{p/$n$} to the output via ``\code{\#show p/$n$.}''

\paragraph{Constant Replacement.}
Constants appearing in a logic program may actually be placeholders for
concrete values to be provided by a user.
An example of this will be given in Section~\ref{subsec:ex:color}.
Via the \code{\#const} declarative (\code{\#}~is optional),
one may define a default value to be inserted for a constant.
Such a default value can still be overridden via command line option
\code{--const} (cf.\ Section~\ref{subsec:opt:gringo}).
Syntactically, \code{\#const} must be followed by an assignment having
a (symbolic) constant on the left-hand side and a term on the right-hand side.
Some exemplary \code{\#const} declarations are:
%
\begin{lstlisting}[numbers=none]
#const x = 42.
 const y = f(g,h).
\end{lstlisting}

\paragraph{Domain Declarations.}
Usually, variable names are local to a rule,
where they must be bound via appropriate atoms (cf.\ Section~\ref{subsec:lambda}).
This locality can be undermined by using
\code{\#domain} declaratives (\code{\#}~is optional)
that globally associate variable names to atoms.
An associated atom is then simply added to the body of a rule in which
such a predefined variable name occurs in.
The following is a made-up example:
%
\begin{lstlisting}[numbers=none]
p(1,1). p(1,2).
#domain p(X,Y).
 domain p(Y,Z).
q(Z,X) :- not p(Z,X).
\end{lstlisting}
%
The above program is a priori not level-restricted because
variables~\var{X} and~\var{Z} are unbound in the last rule.
However, as they belong to \code{\#domain} declarations,
\gringo\ and \clingo\ expand the last rule to:
%
\begin{lstlisting}[numbers=none]
q(Z,X) :- p(X,Y), p(Y,Z), not p(Z,X).
\end{lstlisting}
%
Observe that the resulting program is level-restricted (and stratified).

\paragraph{Compute Statements.}
These statements are artefacts supported for backward compatibility.
Although we strongly recommend to avoid compute statements,
we now describe their syntax.
A compute statement is of the form
``$\const{\#compute}~n \code{\char`\{}\dots\code{\char`\}}\code{.}$''
(\code{\#}~and non-negative integer $n$ are optional),
where the ``$\code{\char`\{}\dots\code{\char`\}}$'' part
is similar to a \const{count} aggregate.
The meaning is that all literals contained in
``$\code{\char`\{}\dots\code{\char`\}}$'' must hold w.r.t.\ answer sets
that are to be computed,
while~$n$ specifies a number of answer sets to compute.
As \clasp\ and \clingo\ provide command line option
\code{--number} (cf.\ Section~\ref{subsec:opt:clasp})
to specify how many answer sets are to be computed,
they simply ignore~$n$.
Furthermore,
the ``$\code{\char`\{}\dots\code{\char`\}}$'' part can equivalently
be expressed in terms of integrity constraints,
as indicated in the comments provided along with the following example:
%
\begin{lstlisting}[xrightmargin=-20pt,numbers=none]
q(1;2).
{ p(1..5) }.
#compute 0 { p(X) : q(X) }.       % :- 1 { not p(X) : q(X) }.
 compute   { not p(X) : X=4..5 }. % :- 1 { p(X) : X=4..5 }.
\end{lstlisting}

\subsection{Input Language of \iclingo}\label{subsec:lang:iclingo}

System \iclingo~\cite{gekakaosscth08a} extends \clingo\ by an
\emph{incremental} computation mode that incorporates both grounding and solving.
Hence, its input language includes all constructs described
in Section~\ref{subsec:lang:gringo}.%
\footnote{%
  In its current version, \iclingo\ has no built-in support for
  classical negation (cf.\ Section~\ref{subsec:gringo:negation}).
  That is, complementary atoms may jointly belong to answer sets,
  unless explicitly prohibited by integrity constraints.}
In addition, \iclingo\ deals with statements of the following form:
%
\begin{lstlisting}[numbers=none,escapechar=@]
#base.
#cumulative @\textit{constant}@.
#volatile   @\textit{constant}@.
\end{lstlisting}
%
Via ``\code{\const{\#base}.}'',
the subsequent part of a logic program is declared as static, that is,
it is processed only once at the beginning of an incremental computation.
In contrast, ``\code{\const{\#cumulative} \textit{constant}.}''
and ``\code{\const{\#volatile} \textit{constant}.}'' are used to declare
a (symbolic) \code{\textit{constant}} as a placeholder for incremental step numbers.
In the parts of a logic program below a \const{\#cumulative} statement,
\code{\textit{constant}} is in each step replaced with the current step number,
and the resulting rules, facts, and integrity constraints are accumulated
over a whole incremental computation.
While the replacement of \code{\textit{constant}} is similar,
a logic program part below a \const{\#volatile} statement is local to steps,
that is, all rules, facts, and integrity constraints computed in one step
are dismissed before the next incremental step.
Note that the type of a logic program part (static, cumulative, or volatile)
is determined by the last
\const{\#base}, \const{\#cumulative}, or \const{\#volatile} statement
preceding it.

During an incremental computation, all static program parts are grounded first,
while cumulative and volatile parts are grounded step-wise,
replacing \code{\textit{constant}}s with successive step numbers starting from~\const{1}.
After a grounding step, \clasp\ is usually invoked via an internal interface
(like with \clingo), and the incremental computation stops after a step
in which at least one answer set has been found by \clasp.
This default behavior can be readapted via command line 
options (cf.\ Section~\ref{subsec:opt:iclingo}).
For obtaining a well-defined incremental computation result,
it is important that (ground) head atoms within static, cumulative, and volatile program parts
are distinct from each other, and they must also be different from step to step
(see~\cite{gekakaosscth08a} for details).
In Section~\ref{subsec:ex:block},
we will provide a typical example in which these conditions naturally hold.

\subsection{Input Language of \clasp}\label{subsec:lang:clasp}

Solver \clasp~\cite{gekanesc07b} works on logic programs in \lparse's
output format~\cite{lparseManual}.
This numerical format, which is not supposed to be human-readable,
is output by \gringo\ and can be piped into \clasp.
Such an invocation of \clasp\ looks as follows:
%
\begin{lstlisting}[numbers=none]
gringo [options] [files] | clasp [number] [options]
\end{lstlisting}
%
Note that \code{number} may be provided to specify a maximum number of answer sets
to be computed, where~\code{0} makes \clasp\ compute all answer sets.
This maximum number can also be set via
option \code{--number} or its abbreviation \code{-n}
(cf.\ Section~\ref{subsec:opt:clasp}).
By default, \clasp\ computes one answer set (if it exists).
If a logic program in \lparse's output format has been stored in a \code{file},
then it can be redirected into \clasp\ as follows:
%
\begin{lstlisting}[numbers=none]
clasp [number] [options] < file
\end{lstlisting}
%
Via option \code{--dimacs},
\clasp\ can also be instructed to compute models of a propositional formula
in DIMACS/CNF format~\cite{dimacs}.
If such a formula is contained in \code{file},
then \clasp\ can be invoked in the following way:
%
\begin{lstlisting}[numbers=none]
clasp [number] [options] --dimacs < file
\end{lstlisting}
%
Finally, \clasp\ may be used as a library, as done within \clingo\ and \iclingo.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End:

### 2008-11-03 doc/documentation/lparse.tex
unverified new version of guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@579 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Differences to the Language of \lparse}\label{sec:lparse}

We below provide a (most likely incomplete) list of differences between the
input languages of \gringo~\cite{gescth07a} and \lparse~\cite{lparseManual}.
First of all, it is important to note that the language of \gringo\
significantly extends the one of \lparse.
For instance, \const{min} and \const{max} aggregates (cf.\ Section~\ref{subsec:gringo:aggregate})
are not supported by \lparse.
Furthermore, \gringo\ provides full support for variables within compound terms
(over uninterpreted functions with non-zero arity).
That is, an atom like \code{p(\const{f}(\var{X}))} in the positive body of a rule
can potentially be used to bind~\var{X} (cf.\ Section~\ref{subsec:lambda}),
while \lparse\ would treat \code{\const{f}(\var{X})} like an arithmetic
function (cf.\ Section~\ref{subsec:gringo:arith}) whose variables cannot be bound.
Finally, \gringo\ deals with level-restricted programs (cf.\ Section~\ref{subsec:lambda}),
while \lparse\ requires programs to be $\omega$-restricted~\cite{syrjanen01a}.
As the latter are more restrictive,
programs for \lparse\ tend to be more verbose than the ones for \gringo.
Thus, we do not suggest writing programs in the input language of \gringo\
for compatibility to \lparse.

However, a bulk of existing encodings are written for \lparse\
(see, e.g.,~\cite{asparagus}),
and \gringo\ (likewise, \clingo\ and \iclingo) should actually be able to deal with most of them.
If this is not case, one of the following might be the reason:
%
\begin{itemize}
\item
The input contains primed atoms like \code{\pred{p}'}, which are (currently) not supported by \gringo.
\item
Symbolic names are used for built-in constructs, e.g., \const{plus} or \pred{eq}
for built-in arithmetic function~\code{+} or predicate~\code{==}, respectively.
With a few exceptions (e.g., \const{abs} introduced in Section~\ref{subsec:gringo:arith}),
such names are not associated to built-in constructs by \gringo,
as they may accidentally clash with users' names otherwise.
\item
The input contains (or is instantiated to) 
a \const{count} aggregate (or a cardinality constraints, respectively)
containing duplicates of literals.
Such duplicates are not removed by \lparse, e.g.,
it treats \code{2\char`\{\pred{p}(\const{c}),\pred{p}(\const{c})\char`\}}
as a synonym for \code{2[\pred{p}(\const{c})=1,\pred{p}(\const{c})=1]}.
In contrast, \gringo\ associates curly brackets to a set
(and square brackets to a multiset), as described in Section~\ref{subsec:gringo:aggregate}, so that
\code{2\char`\{\pred{p}(\const{c}),\pred{p}(\const{c})\char`\}} is the same as
\code{2\char`\{\pred{p}(\const{c})\char`\}},
where the latter can clearly not hold.
\item
Pooling is expanded differently,
e.g.,
\lparse\ interprets
\code{\pred{p}(\var{X},\var{Y};\var{X},\var{Z})} 
as a shorthand for
\code{\pred{p}(\var{X},\var{Y}),\pred{p}(\var{X},\var{Z})},
while \gringo\ expands it to
\code{\pred{p}(\var{X},\var{Y},\var{Z}),\linebreak[1]\pred{p}(\var{X},\var{X},\var{Z})},
as explained in Section~\ref{subsec:gringo:pool}.
\item
The input contains some (rather unusual) meta-statement, e.g.,
\code{\#external}, not supported by \gringo.
\end{itemize}
%
As indicated above, the provided list is probably incomplete.
If you would like some difference to be added,
please contact the authors of this guide.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End:

### 2008-11-03 doc/documentation/options.tex
unverified new version of guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@579 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Command Line Options}\label{sec:options}

In this section, we briefly describe the meanings of
command line options supported by
\gringo\ (Section~\ref{subsec:opt:gringo}),
\clingo\ (Section~\ref{subsec:opt:clingo}),
\iclingo\ (Section~\ref{subsec:opt:iclingo}), and
\clasp\ (Section~\ref{subsec:opt:clasp}).
Each of these tools displays its available options
when invoked with flag \code{--help} or \code{-h},
respectively.%
\footnote{%
  Note that our description of command line options
  is based on Version~2.0.2 of \gringo, \clingo, and \iclingo\
  as well as Version~1.1.2 of \clasp.
  While it is rather unlikely that command line options will
  disappear in future versions,
  additional ones might be introduced.
  We will try to keep this document up-to-date,
  but checking the help information shipped 
  with a new version is always a good idea.}
The approach of distinguishing long options, starting with ``\code{--},''
and short ones of the form ``\code{-\textit{l}},''
where \code{\textit{l}} is a letter,
follows the GNU Coding Standards~\cite{GNUcoding}.
For obvious reasons,
short forms are made available only for the most common (long) options.
Some options, also called flags, do not take any argument,
while others require arguments.
An argument \code{\textit{arg}} is provided to a (long) option \code{\textit{opt}}
by writing 
``\code{--\textit{opt}=\textit{arg}}'' or
``\code{--\textit{opt} \textit{arg}},''
while only
``\code{-\textit{l} \textit{arg}}''
is accepted for a short option \code{\textit{l}}.
For each command line option,
we below indicate whether it requires an argument,
and if so, we also describe its meaning.

\subsection{\gringo\ Options}\label{subsec:opt:gringo}

An abstract invocation of \gringo\ looks as follows:
%
\begin{lstlisting}[numbers=none]
gringo [ options | filenames ]
\end{lstlisting}
%
Note that options and filenames do not need to be passed to \gringo\
in any particular order.
If neither a filename nor an option that makes \gringo\ exit (see below)
is provided, \gringo\ reads from the standard input.
In the following, we list and describe the options accepted by \gringo\
along with their particular arguments (if required):
%
\begin{description}
\item[\code{--help,-h}]~\\
Print help information and exit.
\item[\code{--version,-v}]~\\
Print version information and exit.
\item[\code{--syntax}]~\\
Print syntax information (about the input language) and exit.
\item[\code{--stats}]~\\
Print (extended) statistic information before termination.
\item[\code{--verbose,-V}]~\\
Print additional (progress) information during computation.
\item[\code{--debug}]~\\
Print internal representations of rules during grounding.
(This may be used to identify either semantic errors in an input program or
 performance bottlenecks.)
\item[\code{--const,-c \textit{c}=\textit{t}}]~\\
Replace occurrences (in the input program) of a constant~\code{\textit{c}}
with a term~\code{\textit{t}}.
\item[\code{--text,-t}]~\\
Output ground program in (human-readable) text format.
\item[\code{--lparse,-l}]~\\
Output ground program in \lparse's numerical format~\cite{lparseManual}.
\item[\code{--aspils,-a 1|2|3|4|5|6|7}]~\\
Output ground program in \emph{ASPils} format~\cite{gejaosscth08a},
where the argument specifies one of the seven normal forms
defined in~\cite{gejaosscth08a}.
\item[\code{--ground,-g}]~\\
Enable lightweight mode for processing a ground input program.
(This option is recommended to 
 omit unnecessary overhead if the input program is already ground,
 but it leads to a syntax error (cf.\ Section~\ref{subsec:error}) otherwise.)
\item[\code{--bindersplit yes|no}]~\\
Enable or disable binder-splitting~\cite{gescth07a} in the instantiation of rules.
(This option is included mainly for comparison purposes, and generally,
 it is recommended to keep the default argument value \code{yes}.)
\item[\code{--ifixed \textit{n}}]~\\
Use \code{\textit{n}} as fix step number if the input program
contains \const{\#cumulative} or \const{\#volatile} statements.
(This option permits the handling of programs written for \iclingo\
 in a traditional single pass computation.)
\item[\code{--ibase}]~\\
Process only the static part (that can be initiated by a \const{\#base} statement)
of an input program.
(This option may be used to investigate the basic setting of a problem
 including some mutable bound.)
\end{description}
%
The default command line when invoking \gringo\ is as follows:
%
\begin{lstlisting}[numbers=none]
gringo --lparse --bindersplit=yes
\end{lstlisting}
That is, \gringo\ usually outputs a ground program in \lparse's numerical format,
dealt with by various solvers (cf.\ Section~\ref{subsec:asp}),
and it performs binder-splitting in rule instantiation.

\subsection{\clingo\ Options}\label{subsec:opt:clingo}

ASP system \clingo\ combines grounder \gringo\ and solver \clasp\
via an internal interface.
An abstract invocation of \clingo\ looks as follows:
%
\begin{lstlisting}[numbers=none]
clingo [ number | options | filenames ]
\end{lstlisting}
%
A numerical argument is permitted for backward compatibility to
the usage of solver \smodels~\cite{siniso02a},
where it specifies the maximum number of answer sets to be computed
(\code{0}~standing for all answer sets).
As with \gringo, a number, options, and filenames do not need to be passed to \clingo\
in any particular order.
Given that \clingo\ combines \gringo\ and \clasp,
it accepts all options described in the previous section
and in Section~\ref{subsec:opt:clasp}.
In particular, (long) options \code{--help}, \code{--version}, and
\code{--syntax} make \clingo\ print the desired information and exit,
while \code{--text}, \code{--lparse}, and \code{--aspils}
instruct \clingo\ to output a ground program (rather than solving it) like \gringo.
If neither a filename nor an option that makes \clingo\ exit
(see Section~\ref{subsec:opt:gringo}) is provided, \clingo\ reads from the standard input.
Beyond the options described in Section~\ref{subsec:opt:gringo} and~\ref{subsec:opt:clasp},
\clingo\ has a single additional option:
%
\begin{description}
\item[\code{--clasp}]~\\
Run \clingo\ as a plain solver (using embedded \clasp).
\end{description}
%
Finally, the default command line when invoking \clingo\
consists of all \clasp\ defaults (cf.\ Section~\ref{subsec:opt:clasp})
and option \code{--bindersplit=yes} of \gringo.

\subsection{\iclingo\ Options}\label{subsec:opt:iclingo}

Incremental ASP system \iclingo\ extends \clingo\ by interleaving
grounding and solving for problems including a mutable bound,
and an abstract invocation of \iclingo\ is as with \clingo:
%
\begin{lstlisting}[numbers=none]
iclingo [ number | options | filenames ]
\end{lstlisting}
%
The external behavior of \iclingo\ is similar to \clingo,
described in the previous section, except for the fact that
option \code{--ifixed} is ignored by \iclingo\ if not run as a grounder
(via one of (long) options \code{--text}, \code{--lparse}, or \code{--aspils}).
However, option \code{--clingo} (see below) may be used to let \iclingo\
work like \clingo.
The additional options of \iclingo\ focus on customizing
incremental computations:
%
\begin{description}
\item[\code{--istats}]~\\
Print statistic information for each incremental solving step.
\item[\code{--imin \textit{n}}]~\\
Perform at least \code{\textit{n}} incremental solving steps before termination.
(This may be used to force steps regardless of the termination condition
 set via \code{--istop}.)
\item[\code{--imax \textit{n}}]~\\
Perform at most \code{\textit{n}} incremental solving steps before termination.
(This may be used to limit steps regardless of the termination condition
 set via \code{--istop}.)
\item[\code{--istop SAT|UNSAT}]~\\
Terminate after an incremental solving step 
in which some (\code{SAT}) or no (\code{UNSAT})
answer set has been found.
\item[\code{--iquery \textit{n}}]~\\
Start with incremental solving at step number~\code{\textit{n}}.
(This may be used to skip some solving steps,
 still accumulating static and cumulative rules for these steps.)
\item[\code{--ilearnt keep|forget}]~\\
Maintain (\code{keep}) or delete (\code{forget}) learnt constraints
in-between incremental solving steps.
(This option configures the behavior of embedded \clasp.)
\item[\code{--iheuristic keep|forget}]~\\
Maintain (\code{keep}) or delete (\code{forget}) heuristic information
in-between incremental solving steps.
(This option configures the behavior of embedded \clasp.)
\item[\code{--clingo}]~\\
Run \iclingo\ as a non-incremental ASP system (like \clingo).
\end{description}
%
As with \clingo, the default command line when invoking \iclingo\
consists of all \clasp\ defaults, explained in the next section,
option \code{--bindersplit=yes} of \gringo\ (cf.\ Section~\ref{subsec:opt:gringo}),
along with \code{--istop=SAT}, \code{--iquery=1}, \code{--ilearnt=keep}, and
\code{--iheuristic=forget}.
That is, incremental solving starts at step number~\code{1} and stops
after a step in which some answer set has been found.
In-between incremental solving steps, embedded \clasp\ maintains learnt constraints
but deletes heuristic information.

\subsection{\clasp\ Options}\label{subsec:opt:clasp}

Stand-alone \clasp~\cite{gekanesc07b} is a solver for ground programs in
\lparse's numerical format~\cite{lparseManual}.
An abstract invocation of \clasp\ looks as follows:
%
\begin{lstlisting}[numbers=none]
clasp [ number | options ]
\end{lstlisting}
%
As with \clingo\ and \iclingo,
a numerical argument specifies the maximum number of answer sets to be computed,
where \code{0}~stands for all answer sets.
(The number of requested answer sets can likewise be set via long option
 \code{--number} or its short form~\code{-n}.)
If neither a filename (via long option \code{--file} or its short form \code{-f})
nor an option that makes \clasp\ exit (see below) is provided,
\clasp\ reads from the standard input.
In fact, it is typical to use \clasp\ in a pipe with \gringo\ in the following way:
\begin{lstlisting}[numbers=none]
gringo [ ... ] | clasp [ ... ]
\end{lstlisting}
In such a pipe, \gringo\ instantiates an input program and outputs the ground
rules in \lparse's numerical format,
which is then consumed by \clasp\ that computes and outputs answer sets.
Note that \clasp\ offers plenty of options to configure its behavior.
We thus categorize them according to their functionalities in the following description.

\subsubsection{General Options}\label{subsec:clasp:general}

We below group general options of \clasp,
used to configure its global behavior.
%
\begin{description}
\item[\code{--help,-h}]~\\
Print help information and exit.
\item[\code{--version,-v}]~\\
Print version information and exit.
\item[\code{--stats}]~\\
Print (extended) statistic information before termination.
\item[\code{--file,-f \textit{filename}}]~\\
Read from file \code{\textit{filename}} (rather than from standard input).
\item[\code{--number,-n \textit{n}}]~\\
Compute at most~\code{\textit{n}} answer sets,
$\code{\textit{n}}=\code{0}$ standing for compute all answer sets.
\item[\code{--quiet,-q}]~\\
Do not print computed answer sets.
(This is useful for benchmarking.)
\item[\code{--seed \textit{n}}]~\\
Use seed~\code{\textit{n}} for random number generator.
(This may be used to obtain a reproducible behavior
 when running \clasp\ with randomized features.)
\item[\code{--brave}]~\\
Compute the brave consequences (union of all answer sets) of a logic program.
\item[\code{--cautious}]~\\
Compute the cautious consequences (intersection of all answer sets) of a logic program.
\item[\code{--opt-all}]~\\
Compute all optimal answer sets (cf.\ Section~\ref{subsec:gringo:optimize}).
(This is implemented by enumerating~\cite{gekanesc07c} answer sets that are not worse
 than the best one found so far.)
\item[\code{--opt-restart}]~\\
Restart the search from scratch (rather than doing enumeration~\cite{gekanesc07c})
after finding a provisionally optimal answer set.
(This may speed up finding the optimum.)
\item[\code{--opt-value \textit{n1}[,\textit{n2},\textit{n3}...]}]~\\
Initialize objective function(s) to minimize with \code{\textit{n1}[,\textit{n2},\textit{n3}...]}.
Depending on the reasoning mode,
only equitable or strictly better answer sets are computed.
\item[\code{--supp-models}]~\\
Compute supported models~\cite{apblwa88a} (rather than answer sets).
\item[\code{--dimacs}]~\\
Compute models of a propositional formula in DIMACS/CNF format~\cite{dimacs}.
\item[\code{--trans-ext all|choice|weight|no}]~\\
Compile extended rules~\cite{siniso02a} into normal rules of form (\ref{eq:normal:rule}).
Arguments \code{choice} and \code{weight}
state that all ``choice rules'' or all ``weight rules,'' respectively,
are to be compiled into normal rules,
while \code{all} means that both and \code{no} that none of them
are subject to compilation.
\item[\code{--eq \textit{n}}]~\\
Run equivalence reasoning~\cite{gekanesc08a} for \code{\textit{n}} iterations,
$\code{\textit{n}}=\code{-1}$ and $\code{\textit{n}}=\code{0}$ standing for
run to fixpoint or do not run equivalence reasoning, respectively.
\item[\code{--sat-prepro yes|no|\textit{n1}[,\textit{n2},\textit{n3}]}]~\\
Run \emph{SatElite}-like preprocessing~\cite{eenbie05a}
for at most~\code{\textit{n1}} iterations 
($\code{\textit{n1}}=\code{-1}$ standing for run to fixpoint),
using cutoff~\code{\textit{n2}} for variable elimination
($\code{\textit{n2}}=\code{-1}$ standing for no cutoff), and
for no longer than~\code{\textit{n3}} seconds
($\code{\textit{n3}}=\code{-1}$ standing for no time limit).
Arguments \code{yes} and \code{no} mean
$\code{\textit{n1}}=\code{\textit{n2}}=\code{\textit{n3}}=\code{-1}$
(that is, run to fixpoint)
or that \emph{SatElite}-like preprocessing is not to be run at all, respectively.
\item[\code{--rand-watches yes|no}]~\\
Initially choose watched literals randomly (with argument \code{yes})
or systematically (with argument \code{no}).
\end{description}
%
Having introduced the general options of \clasp,
let us note that the options below \code{--dimacs} in the above list
are quite low-level and more or less an issue of fine-tuning.
More important is the fact that virtually all optimization functionalities
are only provided by \clasp\ if the maximum number of answer sets to be
computed is set to~\code{0} (standing for all answer sets),
as it is likely to stop search too early otherwise.
The same applies to computing either brave or cautious consequences
(via one of the flags \code{--brave} and \code{--cautious}).

\subsubsection{Search Options}\label{subsec:clasp:search}

The options listed below can be used to configure the main search
strategies of \clasp.
%
\begin{description}
\item[\code{--lookback yes|no}]~\\
Enable or disable lookback techniques (cf.\ Section~\ref{subsec:clasp:back}).
(This option is included mainly for comparison purposes, and generally,
 it is recommended to keep the default argument value \code{yes}.)
\item[\code{--lookahead atom|body|hybrid|auto|no}]~\\
Apply failed-literal detection~\cite{freeman95}
to atoms (with argument \code{atom}),
to rule bodies (with argument \code{body}),
to atoms and rule bodies like in \nomorepp~\cite{angelinesc05c} (with argument \code{hybrid}),
or let \clasp\ pick one of the previous three possibilities
based on program properties (with argument \code{auto}).
Failed-literal detection is switched off via argument~\code{no}.
\item[\code{--initial-lookahead}]~\\
Apply failed-literal detection (to atoms) in preprocessing,
but not (necessarily) during search.
\item[\code{--heuristic Berkmin|Vmtf|Vsids|Unit|None}]~\\
Use \emph{BerkMin}-like decision heuristic~\cite{golnov02a}
(with argument~\code{Berkmin}),
\emph{Siege}-like decision heuristic~\cite{ryan04a}
(with argument~\code{Vmtf}),
\emph{Chaff}-like decision heuristic~\cite{momazhzhma01a}
(with argument~\code{Vsids}),
\emph{Smodels}-like decision heuristic~\cite{siniso02a}
(with argument~\code{Unit}),
or (arbitrary) static variable ordering
(with argument~\code{None}).
\item[\code{--rand-freq \textit{p}}]~\\
Perform random (rather than heuristic) decisions
with probability $0\leq\code{\textit{p}}\leq 1$.
\item[\code{--rand-prob yes|no|\textit{n1},\textit{n2}}]~\\
Run \emph{Satzoo}-like random probing~\cite{een03a},
initially performing~\code{\textit{n1}} passes of up to~\code{\textit{n2}} conflicts
making random decisions.
Arguments \code{yes} and \code{no} mean
$\code{\textit{n1}}=\code{50},\linebreak[1]\code{\textit{n2}}=\code{20}$
or that random probing is not to be run at all, respectively.
\end{description}
%
Let us note that switching the above options can have dramatic effects
(both positively and negatively) on the search performance of \clasp.
Sticking to the default argument values \code{yes} for \code{--lookback}
and \code{no} for \code{--lookahead} is generally recommended.
If nonetheless performance bottlenecks are observed,
it is worthwhile to give \code{Vmtf} and \code{Vsids} for \code{--heuristic}
a try, in particular,
when the program under consideration is huge 
but scarcely yields conflicts during search.

\subsubsection{Lookback Options}\label{subsec:clasp:back}

The following options have an effect only if \code{--lookback}
is turned on, that is, its argument value must be \code{yes}.
%
\begin{description}
\item[\code{--restarts,-r \textit{n1}[,\textit{n2},\textit{n3}]|no}]~\\
Choose and parameterize a restart policy.
If a single argument \code{\textit{n1}} is provided,
\clasp\ restarts search from scratch after a number of conflicts
determined by a universal sequence~\cite{lusizu93a}, where \code{\textit{n1}}
constitutes the base unit.
If two arguments \code{\textit{n1},\textit{n2}} are specified,
\clasp\ runs a geometric sequence~\cite{eensor03a},
restarting every $\code{\textit{n1}}*\code{\textit{n2}}^i$ conflicts,
where~$i$ is the number of restarts performed so far.
Given three arguments \code{\textit{n1},\textit{n2},\textit{n3}},
\clasp\ repeats geometric restart sequence $\code{\textit{n1}}*\code{\textit{n2}}^i$
when it reaches an outer limit~$\code{\textit{n1}}*\code{\textit{n3}}^j$~\cite{biere08a},
where~$j$ counts how often the outer limit has been hit so far.
Finally, restarts are disabled via argument~\code{no}.
\item[\code{--local-restarts}]~\\
Count conflicts locally~\cite{ryvstr08a} (rather than globally)
for deciding when to restart.
\item[\code{--bounded-restarts}]~\\
Perform bounded restarts in answer set enumeration~\cite{gekanesc07c}.
\item[\code{--save-progress}]~\\
Use cached (rather than heuristic) decisions~\cite{pipdar07a} if available.
\item[\code{--shuffle,-s \textit{n1},\textit{n2}}]~\\
Shuffle internal data structures after~\code{\textit{n1}} restarts
($\code{\textit{n1}}=\code{0}$ standing for no shuffling)
and then reshuffle every~\code{\textit{n2}} restarts 
($\code{\textit{n2}}=\code{0}$ standing for no reshuffling).
\item[\code{--deletion,-d \textit{n1}[,\textit{n2},\textit{n3}]}]~\\
Limit the number of learnt constraints to 
$\min\{(c/\code{\textit{n1}})*\code{\textit{n2}}^i,c*\code{\textit{n3}}\}$,
where~$c$ is the initial number of constraints and~$i$ is the number of restarts
performed so far.
\item[\code{--reduce-on-restart}]~\\
Delete a portion of learnt constraints after every restart.
\item[\code{--strengthen bin|tern|all|no}]~\\
Check binary (with argument \code{bin}),
binary and ternary (with argument \code{tern}), or 
all (with argument \code{all}) antecedents for
self-subsumption~\cite{eenbie05a}
in order to strengthen a constraint to learn.
Strengthening is disabled via argument~\code{no}.
\item[\code{--recursive-str}]~\\
Recursively apply strengthening, as proposed in~\cite{soreen05a}.
\item[\code{--loops common|distinct|shared|no}]~\\
Learn loop nogood~\cite{gekanesc07a} per atom in an unfounded set~\cite{gerosc91a}
(with argument \code{common}),
shrink unfounded set before learning another loop nogood
(with argument \code{distinct}),
learn loop formula~\cite{linzha04a} for atoms in an unfounded set
(with argument \code{shared}),
or do not record unfounded sets at all
(with option \code{no}).
\item[\code{--contraction \textit{n}}]~\\
Temporarily truncate learnt constraints over more than \code{\textit{n}} variables~\cite{ryan04a}.
\end{description}
%
As with search options described in the previous section,
switching lookback options may have a significant impact on the
performance of \clasp.
In particular, we suggest trying the universal restart sequence~\cite{lusizu93a}
with different base units~\code{\textit{n1}} or even disabling restarts
(both via (long) option \code{--restarts} or its short form~\code{-r})
in case that performance needs to be improved.

Finally, let us consider the default command line of \clasp:
%
\begin{lstlisting}[numbers=none]
clasp 1 --trans-ext=no --eq=5 --sat-prepro=no --rand-watches=yes
        --lookback=yes --lookahead=no --heuristic=Berkmin
        --rand-freq=0.0 --rand-prob=no
        --restarts=100,1.5 --shuffle=0,0 --deletion=3.0,1.1,3.0
        --strengthen=all --loops=common --contraction=250
\end{lstlisting}
%
Considering only the most significant defaults,
numeric argument~\code{1} instructs \clasp\ to terminate immediately
after finding an answer set.
Together with \code{--lookback=yes}, \code{--restarts=100,1.5} lets \clasp\ apply a geometric
restart policy.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End:

### 2008-11-03 doc/documentation/restrictions.tex
unverified new version of guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@579 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Restrictedness Notions}\label{sec:restricted}

In view of function symbols with non-zero arity,
we may be confronted with logic programs over an infinite Herbrand base.
In order to maintain decidability of reasoning tasks,
it is thus important to identify language fragments for which
finite equivalent ground programs are guaranteed to exist.
Level-restricted logic programs constitute such a fragment,
where finiteness is manifested in the requirement that any variable in a rule
must be bound to a finite set of ground terms via a predicate not
subject to positive recursion through the rule.
The notion of level-restrictedness is complemented by stratification,
which by disallowing negative recursion among predicates describes a
class of logic programs having unique answer sets.
The formal definition of level-restricted programs
that can be grounded by \gringo, \clingo, and \iclingo\
is provided in Section~\ref{subsec:lambda}.
Section~\ref{subsec:strat} introduces stratified programs and the related
concept of domain predicates, which can serve particular purposes during grounding.
Both sections may be skimmed or even skipped upon the first reading,
and rather be looked up later on to find out details.

\subsection{Level-Restricted Logic Programs}\label{subsec:lambda}

The main task of a grounder is to substitute the variables in a logic program~$\Pi$
by terms such that the result is a finite equivalent ground program~$\alt{\Pi}$.%
\footnote{$\Pi$ and~$\alt{\Pi}$ are equivalent if they have the same answer sets.}
Of course, a necessary condition for this is that~$\Pi$ possesses
only finitely many finite answer sets.
Unfortunately, such properties are undecidable in general~\cite{daeigovo01a}.
Instead of semantic properties, grounders do thus impose rather simple syntactic
conditions to guarantee the existence of finite equivalent ground programs
(which are then also computed).
For instance, the \dlv\ system~\cite{dlv03a} requires programs to be \emph{safe} for
establishing finiteness in the absence of functions with non-zero arity and under
limited arithmetic.
Grounder \lparse~\cite{lparseManual} deals with \emph{$\omega$-restricted}
programs~\cite{syrjanen01a}, which are more restrictive than safe programs
but guarantee finiteness also in the presence of functions with non-zero arity.
Finally, \gringo~\cite{gescth07a} requires programs to be \emph{$\lambda$-restricted},
a property more general than $\omega$-restrictedness but likewise applicable
to programs over the same languages.
We below reproduce the concept of $\lambda$-restrictedness,
called level-restrictedness here.

The idea underlying level-restrictedness is that the structure of a logic program~$\Pi$
must be such that, for each rule $r\in\Pi$,
the set of potentially applicable ground instances of~$r$ is known in advance.
This is the case if, for each variable~$V\in\vars{r}$,
we find an atom~$A$ in the body of~$r$ with
$V\in\vars{A}$, where $\vars{A}$ is that set of variables occurring in~$A$,
such that the potentially derivable ground instances of~$A$ are limited (see below).
In fact, with such an atom~$A$ at hand,
the set of ground terms to which~$V$ needs to be instantiated is known a priori,
and no further ground terms need to be considered.
The sketched approach is justified by the $\subseteq$-minimality of answer sets
in the sense of Section~\ref{subsec:semantics},
as it allows us to restrict the attention to rules whose bodies are derivable w.r.t.\
some answer set.
Beyond outputting only ``relevant'' rules,
grounders can apply answer-set preserving simplifications,
some of which are discussed in Section~\ref{subsec:strat}.

We next provide a more precise account of level-restrictedness,
where we start by considering normal programs~$\Pi$.
For a rule $r\in\Pi$ and $V\in\vars{r}$,
we let $\binders{V}{r}$ contain all elements of~$\pbody{r}$ that may be used to
limit~$V$ to a certain subset of~$\ground{\mathcal{T}}$.
Although some exceptions will be explained later on,
% for normal rules~$r$ of form~(\ref{eq:normal:rule}),
it is convenient to assume that $\binders{V}{r}$ consists of all atoms $A\in\pbody{r}$
such that $V\in\vars{A}$.
We then % consider level mappings $\lambda$ and 
define~$\Pi$ as \emph{level-restricted}
if there is some level mapping~$\lambda:\mathcal{P}\rightarrow\mathbb{N}$ such that,%
\footnote{Recall that $\mathcal{P}$ is the set of all predicates that occur in~$\Pi$.}
for every rule~$r\in\Pi$ and each $V\in\vars{r}$,
there is some $A\in\binders{V}{r}$ such that $\lambda(p/i)>\lambda(q/j)$
for predicate~$p/i$ of $\head{r}$ and predicate~$q/j$ of~$A$.
Note that an appropriate~$\lambda$ (if it exists) is easy to determine,
and \gringo\ exploits it to schedule the order in which rules will be instantiated.
In fact, if rules are processed in the order of levels of their head predicates,
ground  terms to be substituted for variables can be restricted
on the basis of atoms identified as (un)derivable before.

\begin{example}\label{ex:lambda}
Consider the following logic program:
%
\lstinputlisting{examples/zigzag.lp}
%
The set~$\mathcal{P}$ of predicates contains
\pred{zig}/$1$, \pred{zag}/$1$,
\pred{zigzag}/$2$, and \pred{zagzig}/$2$.
The rules~$r$ with head predicates \pred{zig}/$1$ or \pred{zag}/$1$ in Line~1 and~2
are ground ($\vars{r}=\nolinebreak\emptyset$).
Hence, the level-restrictedness condition is trivially satisfied for these rules,
and we may map their head predicates by~$\lambda$ as follows:
$\pred{zig}/1 \mapsto 0$ and $\pred{zag}/1 \mapsto 0$.
For the rule in Line~3,
we have to respect that
$\lambda(\pred{zigzag}/2) > \lambda(\pred{zig}/1) = \lambda(\pred{zag}/1) =\nolinebreak 0$.
We can thus choose $\pred{zigzag}/2\mapsto 1$.
Also note that \code{\pred{zig}(\var{X})} and \code{\pred{zag}(\var{Y})} are
the only atoms in $\binders{\var{X}}{}$ and $\binders{\var{Y}}{}$, respectively.
Finally, we let $\pred{zagzig}/2\mapsto 2$ in view of the rule in Line~4,
where \code{\pred{zigzag}(\var{X},\var{Y})} belongs to
$\binders{\var{X}}{}$ and $\binders{\var{Y}}{}$.
The total mapping
\(
\lambda=
\{
\pred{zig}/1 \mapsto 0,
\pred{zag}/1 \mapsto 0,
\pred{zigzag}/2\mapsto 1,
\pred{zagzig}/2\mapsto 2
\}
\)
witnesses that the above program is level-restricted.

Note that the given mapping~$\lambda$ would no longer be appropriate
if we add rule
%
\begin{lstlisting}[firstnumber=5]
zigzag(X,Y) :- zagzig(Y,X).
\end{lstlisting}
%
In fact, the program consisting of all rules in Line~1--5 is not
level-restricted because Line~4 and~5 require
$\lambda(\pred{zagzig}/2) > \lambda(\pred{zigzag}/2)$ and
$\lambda(\pred{zigzag}/2) > \lambda(\pred{zagzig}/2)$, respectively.
Clearly, there cannot be any level mapping~$\lambda$ jointly
satisfying these two conditions.
However, replacing Line~4 with
%
\begin{lstlisting}[firstnumber=4]
zagzig(Y,X) :- zigzag(X,Y), zig(X), zag(Y).
\end{lstlisting}
%
would restore level-restrictedness because
\code{\pred{zig}(\var{X})} and \code{\pred{zag}(\var{Y})} belong to
$\binders{\var{X}}{}$ and $\binders{\var{Y}}{}$, respectively.
Instead of the previous~$\lambda$, the following level mapping could then be used:
\(
%\lambda=
\{
\pred{zig}/1 \mapsto 0,
\pred{zag}/1 \mapsto 0,
\pred{zagzig}/2\mapsto 1,
\pred{zigzag}/2\mapsto 2
\}
\).\nolinebreak
\eexample
\end{example}

The fundamental property of a level-restricted
program is that it admits a finite equivalent ground program, not to be confused with
the (full) ground instantiation because $\ground{\mathcal{T}}$ may still be infinite.
Furthermore, level-restrictedness provides a syntactic criterion that is not
difficult to check.
In fact, \gringo\ performs such a check and accepts an input program if it is level-restricted.
Generalizations of level-restrictedness to the rich input language of \gringo\
will be discussed in Section~\ref{sec:language}.

\subsection{Stratified Logic Programs}\label{subsec:strat}

Another relevant restriction is stratification (cf.~\cite{minker88}),
as every stratified (normal) program has a unique answer.
Grounder \gringo\ exploits stratification in two respects:
first, it fully evaluates stratified (sub)programs during grounding, and second,
predicates that are subject to stratification can serve as domain predicates (see below).
As with level-restrictedness, stratification can be characterized in terms
of level mappings, here, witnessing the absence of recursion through ``\code{not}.''
We start by introducing stratification on normal programs, and in a second step,
we consider stratified subprograms.

A normal program~$\Pi$ is \emph{stratified}
if there is some level mapping~$\xi:\mathcal{P}\rightarrow\mathbb{N}$ such that,
for every rule~$r\in\Pi$ and predicate~$p/i$ of~$\head{r}$, we have
\begin{itemize}
\item
$\xi(p/i)\geq\max\{\xi(q/j) \mid q(t_1,\dots,t_j)\in\pbody{r}\}$ ~and
\item
$\xi(p/i)>   \max\{\xi(q/j) \mid q(t_1,\dots,t_j)\in\nbody{r}\}$.
\end{itemize}
%
Observe that the levels of predicates in the positive body of a rule
can be equal to the level of the head predicate,
while the levels of predicates in the negative body must be strictly less.
That is, positive recursion is allowed among predicates of the same level,
but negative dependencies must obey a strict order.

\begin{example}\label{ex:strat:prog}
We construct a level mapping~$\xi$ for the predicates in the logic program:
%
\lstinputlisting{examples/strat.lp}
%
Observe that predicate \pred{bitter}/$1$ does not occur in the head on any rule, thus,
we may map
$\pred{bitter}/1 \mapsto 0$.
Furthermore, the facts in Line~1 and the rule in Line~2 do not have
a negative body, so that their head predicates can also be mapped to the lowest level:
$\pred{fruit}/1 \mapsto 0$,
$\pred{foul}/1 \mapsto 0$, and
$\pred{natural}/1 \mapsto\nolinebreak 0$.
We still have to map predicates \pred{healthy}/$1$ and \pred{tasty}/$1$.
As \pred{foul}/$1$ occurs in the negative body of the rule in Line~3,
we require 
$\xi(\pred{healthy}/1)>\xi(\pred{foul}/1)=0$.
We can thus choose 
$\pred{healthy}/1 \mapsto 1$.
Finally, the occurrence of \pred{healthy}/$1$ in the positive body of the rule in Line~4
necessitates $\xi(\pred{tasty}/1)\geq\xi(\pred{healthy}/1)=1$, realized by
$\pred{tasty}/1 \mapsto 1$.
In this way, we also satisfy $\xi(\pred{tasty}/1)>\xi(\pred{bitter}/1)=0$.
The obtained total mapping 
$
\xi
=
\{
\pred{bitter}/1 \mapsto 0,\linebreak[1]
\pred{fruit}/1 \mapsto 0,\linebreak[1]
\pred{foul}/1 \mapsto 0,\linebreak[1]
\pred{natural}/1 \mapsto 0,\linebreak[1]
\pred{healthy}/1 \mapsto 1,\linebreak[1]
\pred{tasty}/1 \mapsto 1
\}
$
witnesses that the above program is stratified.
As mentioned before, every stratified normal program has a unique answer set~$X$.
\marginlabel{%
By invoking\\
  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/strat.lp}
the reader can observe that \gringo\ computes this unique answer set~$X$
and represents it in terms of facts.}
Here, we get
$
X
=
\{
\code{\pred{fruit}(\const{apple})},\linebreak[1]
\code{\pred{fruit}(\const{peach})},\linebreak[1]
\code{\pred{foul}(\const{peach})},\linebreak[1]
\code{\pred{natural}(\const{apple})},\linebreak[1]
\code{\pred{natural}(\const{peach})},\linebreak[1]
\code{\pred{healthy}(\const{apple})},\linebreak[1]
\code{\pred{tasty}(\const{apple})}
\}
$.

The program would no longer be stratified if we add rule
%
\begin{lstlisting}[firstnumber=5]
bitter(X) :- healthy(X), not tasty(X).
\end{lstlisting}
%
In fact, the conditions $\xi(\pred{tasty}/1)>\xi(\pred{bitter}/1)$
(because of Line~4) and $\xi(\pred{bitter}/1)>\xi(\pred{tasty}/1)$
(because of Line~5) cannot jointly be satisfied by any level mapping~$\xi$.
Note that the non-stratified program containing all rules in Line~1--5
has two answer sets: $X$ and
$(X\setminus\{\code{\pred{tasty}(\const{apple})}\})\cup\{\code{\pred{bitter}(\const{apple})}\}$.
\eexample
\end{example}

After dealing with stratified programs, we consider subprograms.
For a given logic program~$\Pi$, some~$\pi\subseteq\Pi$ is a 
\emph{stratified subprogram} of~$\Pi$ if
$\pi$ is stratified and if
no predicate~$p/i$ occurring in~$\pi$ belongs to the head of any rule in~$\Pi\setminus\pi$.
As a matter of fact,
a stratified normal subprogram~$\pi$ of~$\Pi$ has a unique answer set~$Y$ such that
$Y\subseteq X$ for any answer set~$X$ of~$\Pi$~\cite{liftur94a}.
A stratified subprogram~$\pi$ of~$\Pi$ is \emph{maximal} if every stratified
subprogram of~$\Pi$ is contained in~$\pi$.
Note that every logic program has a maximal stratified subprogram,
which is also easy to determine.
In fact, \gringo\ identifies the maximal stratified subprogram~$\pi$
of a logic program~$\Pi$,
and it fully evaluates the predicates~$p/i$ that occur in~$\pi$,
which we call \emph{domain predicates}.
In Section~\ref{subsec:gringo:condition},
we will see that domain predicates can serve particular purposes during grounding.

\begin{example}\label{ex:strat:max}
As the program in Line 1--4 of Example~\ref{ex:strat:prog} is stratified,
the maximal stratified subprogram consists of all rules.
After adding the rule in Line~5,
only the rules in Line~1--3 belong to the maximal stratified subprogram.
The corresponding domain predicates are
$\pred{fruit}/1$,
$\pred{foul}/1$,
$\pred{natural}/1$, and
$\pred{healthy}/1$, 
and the unique answer set for them is
$
Y
=
\{
\code{\pred{fruit}(\const{apple})},\linebreak[1]
\code{\pred{fruit}(\const{peach})},\linebreak[1]
\code{\pred{foul}(\const{peach})},\linebreak[1]
\code{\pred{natural}(\const{apple})},\linebreak[1]
\code{\pred{natural}(\const{peach})},\linebreak[1]
\code{\pred{healthy}(\const{apple})}
\}
$.
\eexample
\end{example}

One can check that the program consisting of Line~1--5
in Example~\ref{ex:strat:prog} is level-restricted, but not stratified.
Conversely, program
\begin{lstlisting}[numbers=none]
nat(0).  nat(s(X)) :- nat(X).
\end{lstlisting}
is stratified (take $\xi=\{\pred{nat}/1 \mapsto 0\}$),
but not level-restricted
($\lambda(\pred{nat}/1)>\lambda(\pred{nat}/1)$
 needed because of the second rule is impossible).
In fact, the program has a unique but infinite answer set, thus,
the program cannot be translated to a finite equivalent ground program.
For this reason, \gringo\ does not accept the given program.
Rather, in the first place, a program must be level-restricted in order
to ground it with \gringo.
If level-restrictedness is granted,
the maximal stratified subprogram is exploited in a second stage
to optimize grounding and to make use of domain predicates.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "guide"
%%% End:

### 2008-11-04 doc/documentation/options.tex
correct mini-bug
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@580 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 or do not record unfounded sets at all
-(with option \code{no}).
+(with argument \code{no}).
 \item[\code{--contraction \textit{n}}]~\\

### 2008-11-08 doc/documentation/restrictions.tex
polished restrictedness notions
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@590 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 must be bound to a finite set of ground terms via a predicate not
-subject to positive recursion through the rule.
+subject to positive recursion through that rule.
 The notion of level-restrictedness is complemented by stratification,
@@
 class of logic programs having unique answer sets.
-The formal definition of level-restricted programs
-that can be grounded by \gringo, \clingo, and \iclingo\
+The formal definition of level-restricted programs,
+which can be grounded by \gringo\ (stand-alone or embedded in \clingo\ and \iclingo),
 is provided in Section~\ref{subsec:lambda}.
@@
 only finitely many finite answer sets.
-Unfortunately, such properties are undecidable in general~\cite{daeigovo01a}.
+Unfortunately, such a property is undecidable in general~\cite{daeigovo01a}.
 Instead of semantic properties, grounders do thus impose rather simple syntactic
@@
 we find an atom~$A$ in the body of~$r$ with
-$V\in\vars{A}$, where $\vars{A}$ is that set of variables occurring in~$A$,
+$V\in\vars{A}$ (where $\vars{A}$ is that set of variables occurring in~$A$)
 such that the potentially derivable ground instances of~$A$ are limited (see below).
@@
 some answer set.
-Beyond outputting only ``relevant'' rules,
+Beyond outputting only ``relevant rules,''
 grounders can apply answer-set preserving simplifications,
@@
 of a logic program~$\Pi$,
-and it fully evaluates the predicates~$p/i$ that occur in~$\pi$,
+and it fully evaluates the predicates~$p/i$ not belonging to the head of any rule in~$\Pi\setminus\pi$,
 which we call \emph{domain predicates}.
@@
  needed because of the second rule is impossible).
-In fact, the program has a unique but infinite answer set, thus,
-the program cannot be translated to a finite equivalent ground program.
-For this reason, \gringo\ does not accept the given program.
+In fact, the program has a unique but infinite answer set.
+Thus,
+the program cannot be translated to a finite equivalent ground program,
+which is why \gringo\ does not accept it.
 Rather, in the first place, a program must be level-restricted in order

### 2008-11-09 doc/documentation/language.tex
polished input languages
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@591 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 The head~$A_0$ of a fact or a rule is an \emph{atom} of the form
-$p\code{(}t_1\code{,}\dots\code{,}t_n\code{)}$,
+$p\code{(}t_1\code{,}\dots\code{,}t_m\code{)}$,
 where $p$ is the name of some predicate, that is, a sequence of letters and digits
@@
 holds if the complement of the proposition is objectively derived~\cite{gellif91a}.
-Classical negation, indicated by symbol \code{-}, is permitted in front of atoms.
+Classical negation, indicated by symbol ``\code{-},'' is permitted in front of atoms.
 That is, if~$A$ is an atom, then~$\code{-}A$ is the complement of~$A$.
@@
 
-Disjunctive logic programs permit connective~\code{|} between atoms in rule heads.
+Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
 An additional case of translation~$\psi$ from Section~\ref{subsec:semantics}
@@
 instead of disjunction, unless the latter is required for complexity reasons
-(see~\cite{eitpol06a} for an implementation methodology).
+(see~\cite{eitpol06a} for an implementation methodology in disjunctive ASP).
 
@@
 Finally, note that~\code{==} and~\code{!=} also apply to
-general terms over functions of non-zero arity as well as to
+compound terms over functions with non-zero arity as well as to
 mixed integer and non-integer arguments.%
@@
 
-Importantly, a comparison predicate cannot bind variables,
+Importantly, a built-in comparison predicate cannot bind variables,
 that is, it does not belong to $\binders{\var{X}}{r}$ for any variable~\var{X}.
@@
 Built-in predicate~\code{=} can be used in the body of a rule
-(or on the right-hand side of a condition)
+(or on right-hand sides of conditions introduced in Section~\ref{subsec:gringo:condition})
 to assign a term on its right-hand side to a variable on its left-hand side.
-Note that ``\code{\var{X} = }$t$'' belongs to~$\binders{\var{X}}{r}$,
+Note that ``\code{\var{X} \!=\! }$t$'' belongs to~$\binders{\var{X}}{r}$,
 provided that~$t$ is variable-free or that its variables are bound
-by atoms other than assignments mentioning~\var{X} on their right-hand side.
-Thus, cyclic assignments over otherwise unbound variables are prohibited.
+by atoms other than assignments with~\var{X} on the right-hand side.
+Cyclic assignments over otherwise unbound variables are thus excluded.
 
@@
 %
-Line~3 contains four assignments, where the left-hand sides directly or indirectly
+Line~3 contains four assignments, where the right-hand sides directly or indirectly
 depend on~\var{X} and~\var{Y}.
@@
 
-As with built-in comparison functions,
+As with built-in arithmetic functions,
 an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
@@
 
-Several conditions may be connected via~``\code{:},''
-as in the following additional rules:
+Composite conditions can also be constructed via~``\code{:},''
+as in the additional rules:
 %
@@
 \begin{lstlisting}[firstnumber=8]
-on_week :- day(mon), day(tues), day(wed), day(thurs), day(fri).
+weekdays :- day(mon), day(tues), day(wed), day(thurs), day(fri).
 \end{lstlisting}
@@
 The atoms in the body of this rule follow from facts, so that
-the rule can be simplified to a fact \code{\pred{on\_week}.}
+the rule can be simplified to a fact \code{\pred{weekdays}.}
 (as done by \gringo).
@@
 of the options within the same body (or within the same condition),
-while they are expanded to multiple rules when occurring in the head.
+while they are expanded to multiple rules 
+(or multiple literals connected via ``\code{,}'')
+when occurring in the head (or in front of a condition).
 
@@
    \var{M}\mapsto\const{1},\var{N}\mapsto\const{2}\}$.
-First note that \pred{mix}/$2$ and \pred{-mix}/$2$ each admit four options,
+Note that \pred{mix}/$2$ and \pred{-mix}/$2$ each admit four options,
 corresponding to the cross product of $\{\const{a},\const{b}\}$ substituted
@@
 the instances for \pred{-mix}/$2$ jointly belong to the body.
-The (repeated) body also contains two instance each of \pred{sym}/$1$ and of \pred{num}/$1$.
+The (repeated) body also contains two instances each of \pred{sym}/$1$ and of \pred{num}/$1$.
 We thus get the rules:%
@@
 while an upper bound behaves monotonically for \const{min} (Item~8).
-Vice versa, the opposite comparison is antimonotone, that is,
+In turn, the opposite comparison is antimonotone, that is,
 by default it evaluates to \textit{true} unless too many or improper literals hold.
@@
 so that $L_i\code{=}\const{1}$ can simply be written as $L_i$.
-For instance, the following sets of (weighted) literals are the same
-when combined with any kinds of aggregate operation and bounds:
+For instance, the following (multi)sets of (weighted) literals are the same
+when combined with any kind of aggregate operation and bounds:
 \begin{lstlisting}[numbers=none,escapechar=@]
@@
 )$,
-where $\{L_1,\dots,L_m\}=\{L_1,\dots,L_n\}$ is obtained by dropping
+where $\{L_1,\dots,L_m\}=\{L_i \mid 1\leq i\leq n\}$ is obtained by dropping
 repeated literals.
 Of course, the use of~$l$ and~$u$ is optional also with \const{count}.
-As an example, note that the following aggregate atoms express the same:
+As an example, note that the next aggregate atoms express the same:
 \begin{lstlisting}[numbers=none,escapechar=@]
@@
 can be incorporated as usual within aggregates,
-where intervals and pooling are expanded like in the body of a rule.%
+where intervals and pooling are expanded locally.%
 \footnote{%
@@
   also built-in comparison predicates (cf.\ Section~\ref{subsec:gringo:comp})
-  are permitted on the right-hand side of a condition only.}
+  are permitted on the right-hand side of conditions only.}
+That is, an interval gives rise to multiple literals connected via~``\code{,}''
+within the same aggregate.
+The same applies to pooling in front of a condition,
+while it turns into a composite condition chained by~``\code{:}'' on the right-hand side.
 The notions of level-restrictedness and stratification
@@
 considering the predicates of all atoms (except for those on the right-hand side of a condition)
-in an aggregate that is the head of a rule~$r$,
-and by assuming $A\in\nbody{r}$ for all atoms~$A$ occurring in an aggregate in the body of~$r$.
+in an aggregate atom occurring as the head of a rule~$r$,%
+\footnote{%
+  The rules in a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
+  cannot have aggregate atoms in the head.}
+and by assuming $A\in\nbody{r}$ for all atoms~$A$ of an aggregate in the body of~$r$.
 Finally, note that aggregates without bounds are also permitted on the
@@
 Consider a situation where an informatics student
-wants to enroll for a number of courses at the beginning of the new term.
+wants to enroll for a number of courses at the beginning of a new term.
 In the university calendar, eight courses are found eligible,
@@
 %
-Note that courses~\const{4} and~\const{5} count three times because
+Note that courses~\const{4} and~\const{7} count three times because
 they are eligible for three subject areas, viz., there are three
@@
 \code{\const{course}(\var{4},\var{S},\const{3})} and
-\code{\const{course}(\var{5},\var{S},\const{4})}, respectively.
+\code{\const{course}(\var{7},\var{S},\const{4})}, respectively.
 Comparing the above ground instance,
@@
 (probably rather unnatural for a user to assign).
-Rather than restricting to a single optimization statements,
-a sequence of them is thus permitted,
+Thus, rather than restricting to a single optimization statement,
+a sequence of them is permitted,
 where a statement provided later on takes priority over previous ones.
@@
 To illustrate optimization, we consider a hotel booking situation
-where we want to chose one among five available hotels.
+where we want to choose one among five available hotels.
 The hotels are identified via numbers assigned in descending order of stars.
@@
 Line~6--8 contribute optimization statements in inverse order of significance,
-according to which we want to chose the best hotel to book.
+according to which we want to choose the best hotel to book.
 The most significant optimization statement in Line~8 states that
@@
 After considering the language of logic programs,
-we now introduce external features going beyond the contents of a program.
+we now introduce features going beyond the contents of a program.
 
@@
 In order to selectively include the atoms of a certain predicate in the output,
-one may use the \code{\#show} declarative (in which~\code{\#} is again optional),
+one may use the \code{\#show} declarative (in which~\code{\#} is again optional).
 Here are some examples:
@@
 This locality can be undermined by using
-\code{\#domain} declaratives (\code{\#}~is optional)
+\code{\#domain} declarations (\code{\#}~is optional)
 that globally associate variable names to atoms.
@@
 while~$n$ specifies a number of answer sets to compute.
-As \clasp\ and \clingo\ provide command line option
+As \clasp, \clingo, and \iclingo\ provide command line option
 \code{--number} (cf.\ Section~\ref{subsec:opt:clasp})
@@
 %
-Via ``\code{\const{\#base}.}'',
+Via ``\code{\const{\#base}.},''
 the subsequent part of a logic program is declared as static, that is,
@@
 \begin{lstlisting}[numbers=none]
-gringo [options] [files] | clasp [number] [options]
+gringo [ options | filenames ] | clasp [ number | options ]
 \end{lstlisting}
@@
 If a logic program in \lparse's output format has been stored in a \code{file},
-then it can be redirected into \clasp\ as follows:
+it can be redirected into \clasp\ as follows:%
+\footnote{%
+  The same is achieved by using option \code{--file} or its short form \code{-f}
+  (cf.\ Section~\ref{subsec:opt:clasp}).}
 %
 \begin{lstlisting}[numbers=none]
-clasp [number] [options] < file
+clasp [ number | options ] < file
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-clasp [number] [options] --dimacs < file
+clasp [ number | options ] --dimacs < file
 \end{lstlisting}

### 2008-11-09 doc/documentation/examples.tex
polished examples
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@593 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 in the head expand to multiple rules, which are facts here.
-Thus, the instance contains six nodes and 17 directed edges among them.
+Thus, the instance contains six nodes and 17 directed edges.
 
@@
 exactly one outgoing and exactly one incoming edge, respectively,
-that belong to the cycle.
+belonging to the cycle.
 By inserting the available edges, for node~\const{1},
@@
 requirement w.r.t.\ the reduct (cf.\ Section~\ref{subsec:semantics})
-is a particular feature of answer set semantics,%
+is a particular feature of answer set semantics,
+which is hard to mimic in either
+Boolean Satisfiability~\cite{bihemawa08a} or
+Constraint Programming~\cite{robewa06a}.
+In our present problem, this feature makes sure that all nodes are reached
+by a global cycle from node~\code{1}, thus, excluding isolated subcycles.
+In fact, the Test in Line~8 stipulates that every node in the given graph
+is reached, that is, the instances of \pred{cycle}/$2$ in an answer set
+must be the edges of a Hamiltonian cycle.%
 \marginlabel{%
@@
         \mbox{~}examples/graph.lp}}
-which is hard to mimic in either
-Boolean Satisfiability~\cite{bihemawa08a} or
-Constraint Programming~\cite{robewa06a}.
-In our present problem, this feature makes sure that all nodes are reached
-by a global cycle from node~\code{1}, thus, excluding isolated subcycles.
-In fact, the Test in Line~8 stipulates that every node in the given graph
-is reached, that is, the instances of \pred{cycle}/$2$ in an answer set
-must be the edges of a Hamiltonian cycle.
 Finally, the additional Display part in Line~10 and~11 states that answer
@@
 and answer sets for the above 
-part of the encoding correspond to Hamiltonian cycles.
+part of the encoding correspond to Hamiltonian cycles,
+that is, candidates for a minimum-cost round trip.
 
@@
 an instance is given by a set of facts, here,
-over predicates \pred{block}/$0$ (declaring blocks),
-\pred{init}/$1$ (defining the initial state),
+over predicates \pred{block}/$1$ (declaring blocks),
+\pred{init}/$1$ (defining the initial state), and
 \pred{goal}/$1$ (specifying the goal state).
@@
 
-Our Blocks-World Planning encoding for \iclingo\ makes use of statements
+Our Blocks-World Planning encoding for \iclingo\ makes use of declaratives
 \const{\#base}, \const{\#cumulative}, and \const{\#volatile},
@@
 must be moved to a location~\var{Y} (different from~\var{X}) at each step~\code{t}.
-The integrity constraint in Line~11-13 is used to test whether
+The integrity constraint in Line~11--13 is used to test whether
 moving block~\var{X} to location~\var{Y} is possible at step~\code{t}
@@
 at step~\code{t-1}.
-Finally, the Define part in Line~15 propagates
+Finally, the Define rule in Line~15 propagates
 a move to the state at step~\code{t},
@@
   To observe the ground program dealt with internally \iclingo\
-  at a step $t$, invoke:\\
+  at a step $n$, invoke:\\
   \code{\mbox{~}gringo -t \textbackslash\\
-        \mbox{~}--ifixed=$t$ \textbackslash\\
+        \mbox{~}--ifixed=$n$ \textbackslash\\
         \mbox{~}examples/blocks.lp \textbackslash\\
@@
 
-Finally, let us stress an important prerequisite for obtaining
+Finally, let us stress important prerequisites for obtaining
 a well-defined incremental computation result from \iclingo.
@@
 Furthermore, ground instances of head atoms in the volatile part
-must not occur in the cumulative part,
-and those of the cumulative part must in turn not be used in the static part.
+must not occur in the static and cumulative parts,
+and those of the cumulative part must not be used in the static part.
 Finally, ground instances of head atoms in either the cumulative or the volatile part
 must be different for each pair of distinct steps.
-This is granted for our encoding because both atoms over \pred{move}/$3$
+This is the case for our encoding because both atoms over \pred{move}/$3$
 and those over \pred{holds}/$2$ include~\const{t} as an argument in the
 heads of the rules in Line~9, 15, and~16--17.
-Notably, the smallest step number to replace~\const{t} with is~\const{1},
-so that there is no clash with the ground atoms over \pred{holds}/$2$
+As the smallest step number to replace~\const{t} with is~\const{1},
+there also is no clash with the ground atoms over \pred{holds}/$2$
 obtained from the head of the static rule in Line~5.
-Further details on these requirements and their formal background can
+Further details on the sketched requirements and their formal background can
 be found in~\cite{gekakaosscth08a}.
-Arguably, many problems including some mutable bound are naturally encoded
-such that the above conditions apply.
+Arguably, many problems including some mutable bound can be encoded
+such that the prerequisites apply.
 Some attention should of course be spent on
@@
 fulfilled in steps~\const{1} and~\const{2}, \iclingo\ 
-stops its incremental computation after finding some answer set in step~\const{3}.
+stops its incremental computation after finding an answer set in step~\const{3}.
 The scheme of iterating steps until finding at least one answer set

### 2010-07-02 doc/guide/background.tex
- made bingo the new gringo trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3149 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-03 doc/documentation/background.tex
@@
 we give a brief introduction to Answer Set Programming,
-which constitutes the basic framework of the tools we describe here.
+which constitutes the basic framework for the tools we describe here.
 This framework deals with logic programs,
@@
 However, we want to stress that the syntax we use for logic programs
-admits (uninterpreted) function symbols of non-zero arity,
+admits (uninterpreted) function symbols with non-zero arity,
 whose full support by our tools is rather exceptional and thus innovative.
@@
 Answer Set Programming (ASP)
-\cite{baral03a,gelleo02a,lifschitz02a,martru99a,niemela99a}
+\cite{ankolisc05a,baral03a,gelleo02a,lifschitz02a,martru99a,niemela99a}
 emerged in the late 1990s as a declarative paradigm
@@
   Instead of \code{neg\mus flies(X)}, one could write
-  \code{-flies(X)} in order to make explicit that
-  instances amount to the classical (or strong) negation of
+  \code{-flies(X)} to make explicit that
+  instances (obtained for~\var{X}) amount to the classical (or strong) negation of
   \code{flies(X)}.}
@@
 Variables are local to rules, that is,
-a variable name refers to a different variable in a distinct rule.
+a variable name refers to different variables in distinct rules.
 Finally, every variable is universally quantified and thus
@@
 %\addtocounter{footnote}{-1}%
-\footnotetext{Described in this document.}
+\footnotetext{Described in this guide.}
 
@@
 \lstinputlisting[nolol]{examples/bird.lp}
-\lstinputlisting[firstnumber=3]{examples/bfall.lp}
+\lstinputlisting[firstnumber=3,xrightmargin=-8pt]{examples/bfall.lp}
 %
@@
 \lstinputlisting[nolol]{examples/bird.lp}
-\lstinputlisting[firstnumber=3]{examples/bfeff.lp}
+\lstinputlisting[numbers=none]{examples/bfeff.lp}
 %
@@
 Applying the complementary principle that ``\textit{false} \& something'' evaluates to ``\textit{false},''
-the rule with \code{not \pred{neg\mus flies}(\const{tux})} in the body has been dropped completely.
+the rule with \code{not \pred{neg\mus flies}(\const{tux})} in the body (Line~3) has been dropped completely.
 After performing these simplifications, the ground program computed by \gringo\ is still
@@
   (rather than just one, as done by default).
-
+\\
   Alternatively, one may use \clingo\ to compute the answer sets.
@@
 grounder \gringo, solver \clasp, and their bondings in \clingo\ and \iclingo,
-this document aims at enabling the reader to make (better) use of ASP.
+this guide aims at enabling the reader to make (better) use of ASP.
 
@@
 Consult, e.g., \cite{lloyd87a} for a detailed introduction.
-The language of logic programs is composed from sets
+The language of a logic program is composed from sets
 %
@@
 
-A \emph{rule}~$r$ is an expression of the form $\mathit{Head}~\code{:-}~\mathit{Body}\code{.}$
+A \emph{rule}~$r$ is an expression of the form $\mathit{Head}~\code{:-}~\mathit{Body}\code{.}$,
 where $\mathit{Head}$ and $\mathit{Body}$ are composed from atoms in $\mathcal{A}$
@@
 For $0\leq k\leq n$, every~$A_k$ is an atom, that is, the head is an atom and
-the body a (possibly empty) conjunction of atoms that may be preceded by ``\code{not}.''
+the body is a (possibly empty) conjunction of atoms that may be preceded by ``\code{not}.''
 As mentioned in Section~\ref{subsec:asp},
@@
 The particular structure of normal programs admits a rather intuitive 
-reformulation of their semantics in the next section.
+characterization of their semantics in the next section.
 
@@
 %
-\begin{lstlisting}[numbers=none]
+\begin{lstlisting}[firstnumber=3]
 flies(tux)    :- bird(tux),    not neg_flies(tux).
@@
 need not be disjoint because atoms and terms are also distinguished
-by the context in which they occur,
+by the context they occur in,
 and a predicate or function symbol may be used with different arities.
@@
 relative to a set~$X$ of atoms is defined recursively as follows:
-\begin{itemize}
+\begin{itemize}\addtolength{\itemsep}{-3pt}
 \item $\reduct{F}{X}=\bot$ ~if $X\not\models F$,
@@
 As a matter fact, this implies that any maximal negative subformula 
-of form $(G\rightarrow \bot)$ is replaced by either $\bot$ (if $X\models G$)
+of the form $(G\rightarrow \bot)$ is replaced by either $\bot$ (if $X\models G$)
 or $(\bot\rightarrow\bot)=\top$ (if $X\not\models G$).
@@
 body atoms have been derived,
-while negative bodies are evaluated w.r.t.~$X$.
+while negative bodies are merely evaluated w.r.t.~$X$.
 Using this operator, we can equivalently define $X\subseteq\ground{\mathcal{A}}$
@@
 is the unique rule with head \code{\pred{flies}(\const{tweety})} that ``fires.''
-No further rule applies in step~$i=3$,
+No further rules apply in step~$i=3$,
 so that the set of atoms derived for~$i=2$ equals $\bigcup_{0\leq i}T^i_{\Pi,X}$.
@@
 We have above seen two definitions of answer sets that start
-from an answer set candidate $X\subseteq\ground{\mathcal{A}}$
-that is then verified in some way.
+from an answer set candidate $X\subseteq\ground{\mathcal{A}}$,
+which is then verified in some way.
 Such definitions must not be confused with the computation of answer sets,

### 2010-07-02 doc/guide/errors.tex
- made bingo the new gringo trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3149 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-03 doc/documentation/errors.tex
@@
 \section{Errors and Warnings}\label{sec:error:warn}
+
+\begin{newstuff}
+	\begin{itemize}
+		\item Needs rewrite and maybe bingo needs to be extended at some points
+	\end{itemize}
+\end{newstuff}
 
@@
 in the command line, and change them in a way such that a constant
-that is to be replaced does not (transitively) occur within a term
+that is to be replaced does not (transitively) occur in a term
 to be inserted.
@@
 statements over the same constant~\code{\textit{c}} are contained in the input,
-while only the first among them will be considered.
+but only the first of them will be considered.
 

### 2010-07-02 doc/guide/examples.tex
- made bingo the new gringo trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3149 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-09 doc/documentation/examples.tex
@@
 We exemplarily solve the following problems in ASP:
-$N$-Coloring (Section~\ref{subsec:ex:color}),
 Traveling Salesperson (Section~\ref{subsec:ex:tsp}), and
 Blocks-World Planning (Section~\ref{subsec:ex:block}).
-While the first problem could likewise be solved within neighboring paradigms,
-the second one requires checking reachability,
-something that is rather hard to encode in either
+The first problem demonstrates the advanced modeling capacities of ASP.
+It requires checking reachability,
+something that is rather hard to encode in neighboring paradigms like
 Boolean Satisfiability~\cite{bihemawa08a} or
@@
 illustrates incremental solving with \iclingo.
-
-\subsection{$N$-Coloring}\label{subsec:ex:color}
-
-As already mentioned in Section~\ref{subsec:asp},
-it is custom in ASP to provide a \emph{uniform}
-problem definition~\cite{martru99a,niemela99a,schlipf95a}.
-We follow this methodology and separate the encoding
-from an instance of the following problem:
-given a (directed) graph, decide whether each node can be assigned
-one of~$N$ colors such that any pair of adjacent nodes is colored differently.
-Note that this problem is NP-complete for~$N\geq 3$
-(see, e.g.,~\cite{papadimitriou94a}),
-and thus it seems unlikely that a worst-case polynomial algorithm
-can be found.
-In view of this,
-it is convenient to reduce the particular problem to
-a declarative problem solving paradigm like ASP,
-where efficient off-the-shelf tools like \gringo\ and \clasp\
-are ready to solve the problem reasonably well.
-Such a reduction is now exemplified.
-
-\subsubsection{Problem Instance}\label{subsec:color:instance}
-\input{figures/graph}
-
-We consider directed graphs specified via facts over predicates
-\pred{node}/$1$ and \pred{edge}/$2$.%
-\footnote{%
-  Directedness is not an issue in $N$-Coloring,
-  but we will reuse our directed example graph in Section~\ref{subsec:ex:tsp}.}
-The graph shown in Figure~\ref{fig:graph} is represented by the following set of facts:
-%
-\lstinputlisting{examples/graph.lp}
-%
-Recall from Section~\ref{subsec:lang:gringo} that~``\code{..}'' and~``\code{;}''
-in the head expand to multiple rules, which are facts here.
-Thus, the instance contains six nodes and 17 directed edges.
-
-\subsubsection{Problem Encoding}\label{subsec:color:encoding}
-
-We now proceed by encoding $N$-coloring via non-ground rules that are
-independent of particular instances.
-Typically, an encoding consists of a \emph{Generate}, a \emph{Define},
-and a \emph{Test} part~\cite{lifschitz02a}.
-As $N$-Coloring has a rather simple pattern, the following encoding does
-not contain any Define part:
-%
-\lstinputlisting{examples/color.lp}
-%
-In Line~2, we use the \code{\#const} declarative,
-described in Section~\ref{subsec:gringo:meta},
-to install~\const{3} as default value for constant~\const{n} that is to be replaced
-with the number~$N$ of colors.
-(The default value can be overridden by invoking \gringo\ with option
- \code{--const n=$N$}.)
-The Generate rule in Line~4 makes use of the \const{count} aggregate
-(cf.\ Section~\ref{subsec:gringo:aggregate}).
-For our example graph and~\const{1} substituted for~\var{X},
-we obtain the following ground rule:%
-\marginlabel{%
-  The full ground program is obtained by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\
-    \mbox{~}examples/color.lp \textbackslash\\
-    \mbox{~}examples/graph.lp}}
-%
-\begin{lstlisting}[numbers=none]
-1 { color(1,1), color(1,2), color(1,3) } 1.
-\end{lstlisting}
-%
-Note that~\code{\pred{node}(\const{1})} has been removed from the body,
-as it is derived via a corresponding fact,
-and similar ground instances are obtained for the other nodes~\const{2} to~\const{6}.
-Furthermore, for each instance of \pred{edge}/2,
-we obtain~\code{n} ground instances of the integrity constraint in Line~6,
-prohibiting that the same color~\var{C} is assigned to the adjacent nodes.
-Given~\code{\const{n}=\const{3}},
-we get the following ground instances due to \code{\pred{edge}(\const{1},\const{2})}:
-%
-\begin{lstlisting}[numbers=none]
-:- color(1,1), color(2,1).
-:- color(1,2), color(2,2).
-:- color(1,3), color(2,3).
-\end{lstlisting}
-%
-Again note that \code{\pred{edge}(\const{1},\const{2})},
-derived via a fact, has been removed from the body.
-
-\subsubsection{Problem Solution}\label{subsec:color:solution}
-\input{figures/color}
-
-Provided that a given graph is colorable with~\const{n} colors,
-a solution can be read off an answer set of the program consisting
-of the instance and the encoding.
-For the graph in Figure~\ref{fig:graph},
-the following answer set can be computed:%
-\marginlabel{%
-  To find an answer set, invoke:\\
-  \code{\mbox{~}gringo \textbackslash\\
-        \mbox{~}examples/color.lp \textbackslash\\
-        \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
-        \mbox{~}clasp}\\
-  or alternatively:\\
-  \code{\mbox{~}clingo \textbackslash\\
-        \mbox{~}examples/color.lp \textbackslash\\
-        \mbox{~}examples/graph.lp}}
-%
-\begin{lstlisting}[numbers=none]
-Answer: 1
-... color(1,2) color(2,1) color(3,1) color(4,3) color(5,2) color(6,3)
-\end{lstlisting}
-%
-Note that we have omitted the six instances of \pred{node}/$1$ and the
-17 instances of \pred{edge}/$2$ in order to emphasize the actual solution,
-which is depicted in Figure~\ref{fig:color}.
-Such output projection can also be specified within a logic program file by
-using the declaratives \code{\#hide} and \code{\#show},
-described in Section~\ref{subsec:gringo:meta}.
 
@@
 as specified in Line~23.
-Note that, for the meta-statements with Display functionality (Line~25-26),
+Note that, for the meta-statements with Display functionality (Line~25--26),
 it is actually unimportant whether they belong to a static, cumulative, or
@@
 putting rules into the right program parts.
+
+\begin{newstuff}
+	\begin{itemize}
+		\item allows for saving some predicates now
+	\end{itemize}
+\end{newstuff}
 
@@
 %%% End:
+
+\begin{newstuff}
+	\begin{itemize}
+		\item we need a better example to show the benefits of bingo simplified gbie?
+	\end{itemize}
+\end{newstuff}

### 2010-07-02 doc/guide/language.tex
- made bingo the new gringo trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3149 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-09 doc/documentation/language.tex
@@
 \clingo\ returns answer sets.
-Both \gringo\ and \clingo\ can handle level-restricted input programs
+Both \gringo\ and \clingo\ can handle safe input programs
 (cf.\ Section~\ref{subsec:lambda}),
@@
 $p\code{(}t_1\code{,}\dots\code{,}t_m\code{)}$,
-where $p$ is the name of some predicate, that is, a sequence of letters and digits
+where $p$ is the name of some predicate, that is, a sequence of letters, digits and underscores 
 starting with a lowercase letter,
@@
   For such an atom~$p$, parentheses after the name are skipped, e.g.,
-  \pred{p42X} could be an atom without arguments.
-  Also note that lowercase letters include ``\code{\mus},''
-  which can thus be used at any position within a predicate name.}
+  \pred{p42X} could be an atom without arguments.}
 A \emph{term} $t$ starting with an uppercase letter followed by
-a sequence of letters and digits (e.g., \var{X08x15}) is a variable name,
+a sequence of letters, digits and underscores (e.g., \var{X\_08x15}) is a variable name,
 and integers are written as sequences of digits possibly preceded by ``\code{-}.''
@@
 will be provided in Section~\ref{sec:examples}.
+
+\begin{newstuff}
+\subsubsection{Anonymous Variables}
+Variables that start with an underscore are so called \emph{anonymous variables}.
+Each such variable is treated like a distinct variable name.
+They are especially helpful to mark places where we do not care about the value of the variable.
+\begin{example}
+The following example demonstrates that such variables come in handy
+if we want to project out some variables.
+\begin{lstlisting}
+p(1,1,1).
+p(2,1,2).
+p(X) :- p(X,_A,_A).
+\end{lstlisting}
+Note that both occurrences of \code{\_A} are treated as distinct variables (we get both~\code{p(1)} and~\code{p(2)}).
+\end{example}
+\end{newstuff}
 
@@
 %
-The concept of level-restrictedness (cf.\ Section~\ref{subsec:lambda}) is extended
-to disjunctive programs by,
-for a disjunctive rule~$r$, every predicate~$p/i$
-of some atom in the head of~$r$, and each $V\in\vars{r}$,
-requiring that there is some $A\in\binders{V}{r}$ such that
-$\lambda(p/i)>\lambda(q/j)$ for predicate~$q/j$ of~$A$.
+The concept of safe programs (cf.\ Section~\ref{subsec:safe}) directly applies
+to disjunctive programs.
 Furthermore, a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
@@
 \code{*} (multiplication),
-\code{/} or \code{div} (integer division),
-\code{mod} (modulo function),
-\code{abs} (absolute value),
-\code{\textasciitilde} (bitwise complement),
-\code{\&}  (bitwise AND),
-\code{?}   (bitwise OR), and
-\code{\^} (bitwise exclusive OR).
+\code{/} or \code{\#div} (integer division),
+\code{\#mod} (modulo function),
+\code{\#abs} (absolute value), and
+\code{\#pow} (power function).
 
@@
 
+\begin{newstuff}
 It is important to note that variables
-in the scope of an arithmetic function are not necessarily bound 
-(in the sense of Section~\ref{subsec:lambda}) by a corresponding atom.
-For instance, atom \code{\pred{p}(\var{X}+1,\var{X})} belongs to
-$\binders{\var{X}}{r}$, while atom \code{\pred{p}(\var{X}+1,\var{Y})}
-does not belong to $\binders{\var{X}}{r}$ because it contains~\var{X}
+in the scope of an arithmetic function are not bound 
+(in the sense of Section~\ref{subsec:safe}) by a corresponding (positive) atom.
+Only edges for variables that are not in the scope of arithmetic functions
+are added to the variable graph~$G=(V,E)$.
+For instance, atom \code{\pred{p}(\var{X}+1,\var{X})} provides edge~$(\$, X)$,
+while atom \code{\pred{p}(\var{X}+1,\var{Y})} provides edge~$(\$, Y)$,
+but does not provide edge~$(\$,X)$ because it contains~\var{X}
 only in the scope of an arithmetic function.
+
+Note that previous version of \gringo\ also supported the arithmetic functions
+\code{\textasciitilde} (bitwise complement),
+\code{\&}  (bitwise AND),
+\code{?}   (bitwise OR), and
+\code{\^} (bitwise exclusive OR).
+These have been removed because there are scarcely application applications using them.
+Furthermore, with \gringo's new Lua support\comment{Reference to lua section}\ arbitrary arithmetic functions can be implemented.
+\end{newstuff}
 
@@
 
-Importantly, a built-in comparison predicate cannot bind variables,
-that is, it does not belong to $\binders{\var{X}}{r}$ for any variable~\var{X}.
-Also note that built-in predicates are not considered by level mappings~$\xi$
-in the context of stratification (cf.\ Section~\ref{subsec:strat}), or equivalently,
-one may assume that built-in predicates are mapped to level~$0$ and all other
-predicates to levels greater than~$0$.
+\begin{newstuff}
+Importantly, built-in comparison predicates do not provide edges to variable graphs (see Section~\ref{subsec:safe}).
+\end{newstuff}
 
@@
 
-Built-in predicate~\code{=} can be used in the body of a rule
+Built-in predicate~\code{:=}%
+\footnote{Note that there is also the built-in predicate~\code{=},
+which, in contrast to~\code{:=}, only permits variables on its left hand side.
+Apart from that both literals are semantically the same.
+Predicate~\code{=} is mainly kept for compatibility with \lparse{} and older \gringo{} versions.} 
+can be used in the body of a rule
 (or on right-hand sides of conditions introduced in Section~\ref{subsec:gringo:condition})
-to assign a term on its right-hand side to a variable on its left-hand side.
-Note that ``\code{\var{X} \!=\! }$t$'' belongs to~$\binders{\var{X}}{r}$,
-provided that~$t$ is variable-free or that its variables are bound
-by atoms other than assignments with~\var{X} on the right-hand side.
-Cyclic assignments over otherwise unbound variables are thus excluded.
-
+to unify a term on its right-hand side to a term on its left-hand side.
+
+%Note that the variable~$X$ in ``\code{\var{X} \!=\! }$t$'' is treated like a positive occurence if
+%the variables on the right-hand side are bound positively minus cyclic assignments.
 \begin{example}\label{ex:assign}
@@
 \end{example}
+
+\begin{newstuff}
+For an assignemt literal~$A$, the variable graph (see Section~\ref{subsec:safe}) contains edges from 
+the variables of the right hand sight to the variables of the left hand side of~$A$.
+If a variable on the right hand side occurs in the scope of an arithmetic function,
+then no edge is added.
+If the right hand side does not contain variables then edges start with~$\$$.
+For instance, literal \code{f(X,X+Y):=f(U,V)} provides the edges $(U,X)$ and $(V,X)$.
+Intuitively, this means the assignment can be used to bind $X$ if $U$ and $V$ are bound by some other literal.
+\end{newstuff}
 
@@
 
+\begin{newstuff}
 As with built-in arithmetic functions,
 an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
-cannot be used to bind the variable.
+does not contribute edges in the variable graph, i.e., it will never bind variables.
+\end{newstuff}
 
@@
 collections of terms within a single rule.
-This is particularly useful for encoding conjunctions or disjunctions over
+This is particularly useful for encoding conjunctions%
+\footnote{Note that conditions also permit comparison predicates in front of the first colon.}
+or disjunctions over
 arbitrarily many ground atoms as well as for the compact representation of aggregates
@@
 while they are expanded to multiple rules 
-(or multiple literals connected via ``\code{,}'')
+(or multiple literals connected via~``\code{,}'')
 when occurring in the head (or in front of a condition).
@@
 %
-Finally, we note that pooling is also possible with arguments of built-in predicates.
+Finally, we note that pooling is possible in every term.
+Although, it is most useful in predicates.
 \eexample
@@
 
+\begin{newstuff}
+There is a second pooling operator \code{;;} that is only permitted in predicates.
+It is used to seperate arguments of predicates,
+e.g., \code{p(a,b;;c,d)} expands to \code{p(a,b), p(c,d)}.
+Note that this operator corresponds to \lparse's~\code{;} operator.
+\end{newstuff}
+
 \subsubsection{Aggregates}\label{subsec:gringo:aggregate}
-
-An aggregate is an operation on a multiset of weighted literals that evaluates to an integer.
-In combination with arithmetic comparisons, we can extract a truth value from
-an aggregate's evaluation, thus, obtaining an aggregate atom.
-Customizing the notation in~\cite{ferraris05a},
-we consider ground \emph{aggregate atoms} of the form:
-%
-\begin{equation}\label{eq:aggregate:atom}
-  l~ 
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-\end{equation}
-%
-where~$l$ and~$u$ are integers each of which can possibly be omitted,
-$\mathrm{op}$ is a function from multisets of integers to an integer,
-each $L_i$ is a literal of the form~$A$ or $\code{not}~A$
-for some atom~$A\in\ground{\mathcal{A}}$,
-and each~$w_i$ is an integer.
-The intuitive reading of an aggregate atom is that the value returned by~$\mathrm{op}$
-applied to all weights~$w_i$ such that~$L_i$ holds should be in-between
-lower bound~$l$ and upper bound~$u$ (inclusively).
-If~$l$ or~$u$ is omitted,
-it means that the aggregate's value is not constrained from below or above,
-respectively.
-
-For an aggregate atom~$\mathit{Agg}$ as in~(\ref{eq:aggregate:atom}),
-we let $\atom{\mathit{Agg}}=\{L_i\in\nolinebreak\ground{\mathcal{A}}\mid 1\leq i\leq n\}$
-denote the set of atoms occurring positively in~$\mathit{Agg}$.
-Currently, \gringo\ and \clingo\ support aggregate operations
-\const{sum}, \const{max}, and \const{min}.%
-\footnote{%
-  The additional operation \const{count} is a shorthand for a \const{sum}
-  aggregate in which all weights are~\const{1}.}
-We can now extend translations~$\phi$ and~$\psi$ from Section~\ref{subsec:semantics}
-to aggregate atoms, 
-using the approach in~\cite{ferraris05a} but
-applying the ``choice semantics'' for aggregates
-in heads of rules~\cite{ferlif05a,siniso02a}:%
-\footnote{%
-  The translation~$\phi$ we provide for \const{sum} aggregates
-  assumes that there are no negative weights for literals,
-  while the more sophisticated translation in~\cite{ferraris05a} also
-  captures negative weights.
-  All current solvers work with positive weights only, and thus
-  \gringo\ and \lparse\ use a compilation technique~\cite{siniso02a}
-  to eliminate negative weights, which are not permitted in
-  \lparse's output format~\cite{lparseManual}.
-  As this technique may lead to counterintuitive results~\cite{ferraris05a},
-  we strongly recommend not to use negative weights within \const{sum} aggregates.}
-\begin{enumerate}
-\item % op %
-$\phi[
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
- \top$,
-\item % l op u %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \phi[
-  l~
- &
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
- {} \\ 
- &
-  \phi[
-  l~
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ]
-  \wedge
-  \phi[
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ]\text{,}
- \end{array}$
-\item % l sum %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \phi[
-  l~
- &
-  \const{sum} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
- {} \\ 
- &
-  \big(
-  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},l\leq w_{i_1}+\dots+w_{i_j}}
-  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
-  \big)\text{,}
- \end{array}$
-\item % sum u %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \phi
-  [
- &
-  \const{sum} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
- {} \\ 
- &
-  \neg
-  \big(
-  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},u< w_{i_1}+\dots+w_{i_j}}
-  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
-  \big)\text{,}
- \end{array}$
-\item % l max %
- $\phi[
-  l~
-  \const{max} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
- \hspace{6pt}
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},l\leq w_i}
-  \phi[L_i]
-  \big)$,
-\item % max u %
- $\phi[
-  \const{max} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
-  \neg
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},u< w_i}
-  \phi[L_i]
-  \big)$,
-\item % l min %
- $\phi[
-  l~
-  \const{min} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
-  \neg
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},w_i< l}
-  \phi[L_i]
-  \big)
- $,
-\item % min u %
- $\phi[
-  \const{min} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
-  \hspace{6pt}
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},w_i\leq u}
-  \phi[L_i]
-  \big)
- $, and
-\item % psi %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \psi[
-  \lhd
- &~
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  \rhd
-  ] = 
-% {} \\ 
-% &
-   \phi[
-  \lhd~
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  \rhd
-  ]
-  \wedge
- {} \\ 
- &
-  \big(
-  \bigwedge_{A\in\atom{
-    \lhd~
-    \mathrm{op} 
-    \code{[}
-    L_1\code{=}w_1\code{,}
-    \dots\code{,}
-    L_n\code{=}w_n
-    \code{]}
-    \rhd
-  }}
-  (A\vee\neg A)
-  \big)\text{.}
- \end{array}$
-\end{enumerate}
-%
-The first item reflects that an unconstrained aggregate atom always holds,
-while aggregate atoms with a lower and an upper bound are split up into two
-conjunctively connected parts (Item~2).
-Item~3--8 specify these parts individually for~\const{sum},~\const{max}, and~\const{min}
-aggregates, respectively.
-Provided that only non-negative weights are used within \const{sum}s,
-all three aggregates are subject to one comparison that is monotone,
-meaning that by default it evaluates to \textit{false} so that truth must be established.
-For \const{sum} and \const{max} aggregates, these monotone comparisons are related
-to lower bounds (Item~3 and~5),
-while an upper bound behaves monotonically for \const{min} (Item~8).
-In turn, the opposite comparison is antimonotone, that is,
-by default it evaluates to \textit{true} unless too many or improper literals hold.
-This is the case for comparisons to upper bounds with aggregates
-\const{sum} and \const{max} (Item~4 and~6), while it concerns 
-the lower bound for \const{min} (Item~7).
-Finally, translation~$\psi$ of an aggregate atom in the head of a rule consists of two parts:
-first, $\phi$ is used like in the body of a rule
-(in Item~9, $\lhd$ and~$\rhd$ are used as ``wildcards'' for provided or omitted 
- lower and upper bounds),
-second, a disjunction $(A\vee\neg A)$ is added for each atom~$A$ occurring positively.
-Even though such disjunctions are tautological,
-they are important under answer set semantics (cf.\ Section~\ref{subsec:semantics}),
-as they permit deriving an arbitrary subset of atoms from an aggregate in the head
-(which explains the name  ``choice semantics'').
-Beyond this special treatment in rule heads,
-the truth values of aggregate atoms result from their literals as
-may be intuitively expected,
-and Item~1--8 basically do nothing but formalizing the standard meanings
-of aggregate operations.
-Note that the above translations merely define the semantics of aggregates,
-which does not imply that solvers unfold them in this way.
-In fact, \clasp\ natively supports aggregates without translating them.
-
-As regards syntactic representation,
-weight~\const{1} is considered a default,
-so that $L_i\code{=}\const{1}$ can simply be written as $L_i$.
-For instance, the following (multi)sets of (weighted) literals are the same
-when combined with any kind of aggregate operation and bounds:
-\begin{lstlisting}[numbers=none,escapechar=@]
-[a=1, not b=1, c=2]  @\textnormal{and}@
-[a,   not b,   c=2]@\textnormal{.}@
-\end{lstlisting}
-%
-Furthermore,
-keyword \const{sum} may be omitted, which in a sense makes \const{sum} the default
-aggregate operation.
-In fact, the following aggregate atoms are synonyms:
-\begin{lstlisting}[numbers=none,escapechar=@]
-2 sum [a, not b, c=2] 3  @\textnormal{and}@
-2     [a, not b, c=2] 3@\textnormal{.}@
-\end{lstlisting}
-By omitting keyword \const{sum}, we obtain the same notation as the one of 
-so-called ``weight constraints''~\cite{siniso02a,lparseManual}, which are actually aggregate atoms
-whose operation is addition.
-
-It is important to note that the (weighted) literals within an aggregate
-belong to a multiset.
-In particular, if there are multiple occurrences
-$L\code{=}w_1,\dots,L\code{=}w_k$
-of a literal~$L$, in combination with \const{min} and \const{max},
-it is not the same like having $L\code{=}w_1+\dots+w_k$.
-To see this, note that the program consisting of the facts:
-\begin{lstlisting}[numbers=none]
-2 max [a=2].  2 min [a=2].
-\end{lstlisting}
-has $\{\code{a}\}$ as its unique answer set, while there is no answer set for:
-\begin{lstlisting}[numbers=none]
-2 max [a,a].  2 min [a,a].
-\end{lstlisting}
-
-If literals ought not to be repeated,
-we can use \const{count} instead of \const{sum}.
-Syntactically, \const{count} requires curly instead of square
-brackets, and there must not be any weights within a \const{count} aggregate.
-Regarding semantics,
-% \begin{equation*}%\label{eq:aggregate:atom}
-$(
-  l~ 
-  \const{count} 
-  \code{\char`\{}
-  L_1\code{,}
-  \dots\code{,}
-  L_n
-  \code{\char`\}}
-  u
-)$ 
-%\end{equation*}
-reduces to
-$(
-  l~ 
-  \const{sum} 
-  \code{[}
-  L_1\code{=}\const{1}\code{,}
-  \dots\code{,}
-  L_m\code{=}\const{1}
-  \code{]}
-  u
-)$,
-where $\{L_1,\dots,L_m\}=\{L_i \mid 1\leq i\leq n\}$ is obtained by dropping
-repeated literals.
-Of course, the use of~$l$ and~$u$ is optional also with \const{count}.
-As an example, note that the next aggregate atoms express the same:
-\begin{lstlisting}[numbers=none,escapechar=@]
-1 sum   [a=1, not b=1]     1  @\textnormal{and}@ 
-1 count {a,a, not b,not b} 1@\textnormal{.}@  
-\end{lstlisting}
-Keyword \const{count} can be omitted (like \const{sum}),
-so that the following are synonyms:
-\begin{lstlisting}[numbers=none,escapechar=@]
-1 count {a, not b} 1  @\textnormal{and}@ 
-1       {a, not b} 1@\textnormal{.}@  
-\end{lstlisting}
-The last notation is similar to the one of so-called
-``cardinality constraints''~\cite{siniso02a,lparseManual},
-which are aggregate atoms using counting as their operation.
-
-After considering the syntax and semantics of ground aggregate atoms,
-we now turn our attention to non-ground aggregates.
-Regarding contained variables, an atom occurring in an aggregate behaves
-similar to an atom on the left-hand side of a condition
-(cf.\ Section~\ref{subsec:gringo:condition}).
-That is, any variable occurring within an aggregate is a priori local,
-and it must be bound via a variable of the same name that is global
-or that occurs on the right-hand side of a condition
-(with the atom containing the variable in front).
-As with local variables of conditions, global variables take priority
-during grounding, so that the names of local variables must be chosen
-with care to avoid accidental clashes.
-Beyond conditions (which are more or less the natural construct to use
-for instantiating variables within an aggregate),
-classical negation (cf.\ Section~\ref{subsec:gringo:negation}),
-built-in arithmetic functions (cf.\ Section~\ref{subsec:gringo:arith}),
-intervals (cf.\ Section~\ref{subsec:gringo:interval}), and 
-pooling (cf.\ Section~\ref{subsec:gringo:pool})
-can be incorporated as usual within aggregates,
-where intervals and pooling are expanded locally.%
-\footnote{%
-  Assignments (cf.\ Section~\ref{subsec:gringo:assign}) and (currently)
-  also built-in comparison predicates (cf.\ Section~\ref{subsec:gringo:comp})
-  are permitted on the right-hand side of conditions only.}
-That is, an interval gives rise to multiple literals connected via~``\code{,}''
-within the same aggregate.
-The same applies to pooling in front of a condition,
-while it turns into a composite condition chained by~``\code{:}'' on the right-hand side.
-The notions of level-restrictedness and stratification
-(cf.\ Section~\ref{subsec:lambda} and~\ref{subsec:strat}) are extended
-to programs with aggregates by
-considering the predicates of all atoms (except for those on the right-hand side of a condition)
-in an aggregate atom occurring as the head of a rule~$r$,%
-\footnote{%
-  The rules in a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
-  cannot have aggregate atoms in the head.}
-and by assuming $A\in\nbody{r}$ for all atoms~$A$ of an aggregate in the body of~$r$.
-Finally, note that aggregates without bounds are also permitted on the
-right-hand sides of assignments, but using this feature is only recommended
-for aggregates whose atoms belong to domain predicates because space blow-up
-can become a bottleneck otherwise.
-The following example, making exhaustive use of aggregates,
-nonetheless demonstrates this and other features.
-
-\begin{example}\label{ex:aggr}
-Consider a situation where an informatics student
-wants to enroll for a number of courses at the beginning of a new term.
-In the university calendar, eight courses are found eligible,
-and they are represented by the following facts:
-%
-\lstinputlisting[xrightmargin=-20pt,lastline=8]{examples/aggr.lp}
-%
-In an instance of \const{course}/$3$,
-the first argument is a number identifying one of the eight courses,
-and the third argument provides the course's contact hours per week.
-The second argument stands for a subject area:
-\const{1} corresponding to ``theoretical informatics,''
-\const{2}               to ``practical informatics,''
-\const{3}               to ``technical informatics,''
-and~\const{4}               to ``applied informatics.''
-For instance, atom \code{\const{course}(\const{1},\const{2},\const{5})}
-expresses that course~\const{1} accounts for~\const{5} contact hours per week
-that may be credited to subject area~\const{2} (``practical informatics'').
-Observe that a single course is usually eligible for multiple
-subject areas.
-
-After specifying the above facts,
-the student starts to provide personal constraints on the courses to enroll.
-The first condition is that~\const{3} to~\const{6} courses should be enrolled:
-%
-\lstinputlisting[nolol,firstline=11,lastline=11,firstnumber=9]{examples/aggr.lp}
-%
-Instantiating the above \const{count} aggregate
-yields the following ground rule:%
-\marginlabel{%
-  The full ground program is obtained by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/aggr.lp}}
-%
-\begin{lstlisting}[firstnumber=9,stepnumber=9]
-3 { enroll(1), enroll(2), enroll(3), enroll(4), 
-    enroll(5), enroll(6), enroll(7), enroll(8) } 6.
-\end{lstlisting}
-%
-Observe that an instance of atom~\code{\const{enroll}(\var{C})} is included for each
-instantiation of~\var{C} such that \code{\const{course}(\var{C},\var{S},\var{H})}
-holds for some values of~\var{S} and~\var{H}.
-Duplicates resulting from distinct values for~\var{S} are removed, thus,
-obtaining the above set of ground atoms.
-
-The next constraints of the student regard the subject areas of enrolled courses:
-%
-\lstinputlisting[nolol,firstline=13,lastline=15,firstnumber=10]{examples/aggr.lp}
-%
-Each of the three integrity constraints above contains a \const{sum} aggregate,
-using default weight~\const{1} for literals.
-Recalling that \const{sum} aggregates operate on multisets,
-duplicates are not removed.
-Thus, the integrity constraint in Line~10 is instantiated as follows:
-%
-\begin{lstlisting}[firstnumber=10,stepnumber=10]
-:- [ enroll(1) = 1, enroll(1) = 1, 
-     enroll(2) = 1, enroll(2) = 1,
-     enroll(3) = 1, enroll(3) = 1,
-     enroll(4) = 1, enroll(4) = 1, enroll(4) = 1,
-     enroll(5) = 1, enroll(5) = 1,
-     enroll(6) = 1, enroll(6) = 1,
-     enroll(7) = 1, enroll(7) = 1, enroll(7) = 1,
-     enroll(8) = 1, enroll(8) = 1 ] 10.
-\end{lstlisting}
-%
-Note that courses~\const{4} and~\const{7} count three times because
-they are eligible for three subject areas, viz., there are three
-distinct instantiations for~\var{S} in
-\code{\const{course}(\var{4},\var{S},\const{3})} and
-\code{\const{course}(\var{7},\var{S},\const{4})}, respectively.
-Comparing the above ground instance,
-the meaning of the integrity constraint in Line~10 is that the 
-number of eligible subject areas over all enrolled courses
-must be more than~\const{10}.
-Similarly, the integrity constraint in Line~11 expresses the requirement
-that at most one course of subject area~\const{2} (``practical informatics'')
-is not enrolled,
-while Line~12 stipulates that the enrolled courses
-amount to less than six nominations of
-subject area~\const{3} (``technical informatics'') 
-or~\const{4} (``applied informatics'').
-Also note that, given the facts in Line~1--8,
-we could equivalently have used \const{count} rather than \const{sum} in Line~11,
-but not in Line~10 and~12.
-
-The remaining constraints of the student deal with contact hours.
-To express them, we first introduce an auxiliary rule and a fact:
-%
-\lstinputlisting[nolol,firstline=17,lastline=18,firstnumber=13]{examples/aggr.lp}
-%
-The rule in Line~13 projects instances of \pred{course}/$3$ to
-\pred{hours}/$2$, thereby, dropping courses' subject areas.
-This is used to not consider the same course multiple times within the following
-integrity constraints:
-%
-\lstinputlisting[nolol,firstline=20,lastline=22,firstnumber=15]{examples/aggr.lp}
-%
-As Line~15 shows,
-we may use default negation via ``\code{not}'' in front of aggregate atoms,
-and bounds may be specified in terms of variables.
-In fact, by instantiating~\var{M} to~\const{20},
-we obtain the following ground instance of the integrity constraint in Line~15:
-%
-\begin{lstlisting}[firstnumber=15,stepnumber=15]
-:- not 18 [ enroll(1) = 5, enroll(2) = 4, 
-            enroll(3) = 6, enroll(4) = 3,
-            enroll(5) = 4, enroll(6) = 2,
-            enroll(7) = 4, enroll(8) = 5 ] 20.
-\end{lstlisting}
-%
-The above integrity constraint states that the \const{sum} of contact hours per week
-must lie in-between~\const{18} and~\const{20}.
-Note that the \const{min} and \const{max} aggregates in Line~16 and~17, respectively,
-work on the same (multi)set of weighted literals as in Line~15.
-While the integrity constraint in Line~16 stipulates that any course to enroll
-must include more than~\const{2} contact hours,
-the one in Line~17 prohibits enrolling for courses of~\const{6} or more contact hours.
-Of course, the last two requirements could also be formulated as follows:
-%
-\begin{lstlisting}[firstnumber=16]
-:- enroll(C), hours(C,H), H <= 2.
-:- enroll(C), hours(C,H), H >= 6.
-\end{lstlisting}
-
-Finally, the following rules illustrate the use of aggregates within assignments:
-%
-\lstinputlisting[nolol,firstline=24,lastline=25,firstnumber=18]{examples/aggr.lp}
-%
-Note that the above aggregates have already been used in Line~9 and~15, respectively,
-where keywords \const{count} and \const{sum} have been omitted for convenience.
-These keywords can be dropped here too, and we merely include them to show the
-more verbose notations of \const{count} and \const{sum} aggregates.
-However, the usage of aggregates in the last two lines is different from before,
-as they now serve to assign an integer to a variable~\var{N}.
-In this context, bounds are not permitted, and so none are provided in Line~18 and~19.
-The effect of these two lines is that the student can read off the number of
-courses to enroll and the amount of contact hours per week from instances of
-\pred{courses}/$1$ and \pred{hours}/$1$ belonging to an answer set.%
-\marginlabel{%
-  To compute the unique answer set of the program, invoke:\\
-  \code{\mbox{~}gringo \textbackslash\\
-        \mbox{~}examples/aggr.lp | \textbackslash\\
-        \mbox{~}clasp -n 0}\\
-  or alternatively:\\
-  \code{\mbox{~}clingo -n 0 \textbackslash\\
-        \mbox{~}examples/aggr.lp}}
-%
-In fact, running \clasp\ shows the student that a unique
-collection of~\const{5} courses to enroll satisfies all requirements:
-the courses~\const{1}, \const{2}, \const{4}, \const{5}, and~\const{7},
-amounting to~\const{20} contact hours per week.
-
-Although the above program does not reflect this possibility,
-it should be noted that (as has been mentioned in Section~\ref{subsec:gringo:condition})
-multiple literals may be connected via ``\code{:}'' in order to construct
-composite conditions within an aggregate.
-As before, the predicates of atoms on the right-hand side of such conditions
-must be either domain predicates or built-in.
-Furthermore, the usage of non-domain predicates within an aggregate
-on the right-hand side of an assignment (like \pred{enroll}/$1$ in Line~18 and~19 above)
-is not recommended in general because the space blow-up may be significant.
-\eexample
-\end{example}
+\input{aggregates}
 
@@
 
+\begin{newstuff}
+	\begin{itemize}
+		\item \#'s are missing again
+		\item priorities can be attatched
+			\begin{itemize}
+				\item low/high integer corresponds to low/high priority
+				\item if no priority given, the n-th minimize/maximize gets priority n
+					  (for conformity with lparse)
+			\end{itemize}
+	\end{itemize}
+\end{newstuff}
+
 \subsubsection{Meta-Statements}\label{subsec:gringo:meta}
@@
 \code{p/$n$} to the output via ``\code{\#show p/$n$.}''
+
+\begin{newstuff}
+	\begin{itemize}
+		\item \#hide/show statements in predicate form have a new meaning now
+		\item \#hide p(X,X) will only hide p(1,1), p(2,2), p(a,a)
+		\item they can also have conditions now
+	\end{itemize}
+\end{newstuff}
+
+\begin{newstuff}
+\paragraph{External Statements.}
+\com{write me!!!}
+\end{newstuff}
 
@@
 
+\begin{newstuff}
+	\begin{itemize}
+		\item for performance reasons \#const statements are evaluated in order of appearance now
+	\end{itemize}
+\end{newstuff}
+
 \paragraph{Domain Declarations.}
@@
 Observe that the resulting program is level-restricted (and stratified).
+\begin{newstuff}
+	\begin{itemize}
+		\item they are still there don't use them they are evil
+	\end{itemize}
+\end{newstuff}
 
@@
 \end{lstlisting}
+\begin{newstuff}
+	\begin{itemize}
+		\item not yet there don't use them they are evil
+	\end{itemize}
+\end{newstuff}
+
+\begin{newstuff}
+\subsubsection{Lua Support}\label{subsec:lang:lua}
+\com{write me!}
+\end{newstuff}
 
@@
 
+\begin{newstuff}
+	\begin{itemize}
+		\item rules won't be instantiated using the same (global) substitution twice
+		\item :- \{ p(X) \}. in cumulative part won't be regrounded even though domain of p/1 changed
+	\end{itemize}
+\end{newstuff}
 \subsection{Input Language of \clasp}\label{subsec:lang:clasp}

### 2010-07-02 doc/guide/options.tex
- made bingo the new gringo trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3149 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-04 doc/documentation/options.tex
@@
 and it performs binder-splitting in rule instantiation.
+
+\begin{newstuff}
+	\begin{itemize}
+		\item stats is not yet implemented
+		\item verbose takes level now
+		\item debug corresponds to verbose=3
+		\item aspils is gone
+		\item bindersplit is gone (not worth implementing, can be done in preprocessor)
+	\end{itemize}
+\end{newstuff}
 
@@
 
+\begin{newstuff}
+	\begin{itemize}
+		\item the subsubsections below are probably outdated, benny?
+	\end{itemize}
+\end{newstuff}
+
 \subsubsection{General Options}\label{subsec:clasp:general}
@@
 \item[\code{--file,-f \textit{filename}}]~\\
-Read from file \code{\textit{filename}} (rather than from standard input).
+Read from file \code{\textit{filename}} (rather than from the standard input).
 \item[\code{--number,-n \textit{n}}]~\\
@@
 \item[\code{--seed \textit{n}}]~\\
-Use seed~\code{\textit{n}} for random number generator.
-(This may be used to obtain a reproducible behavior
- when running \clasp\ with randomized features.)
+Use seed~\code{\textit{n}} (rather than~\code{1}) for random number generator.
 \item[\code{--brave}]~\\
@@
 \begin{lstlisting}[numbers=none]
-clasp 1 --trans-ext=no --eq=5 --sat-prepro=no --rand-watches=yes
-        --lookback=yes --lookahead=no --heuristic=Berkmin
-        --rand-freq=0.0 --rand-prob=no
+clasp 1 --seed=1 --trans-ext=no --eq=5 --sat-prepro=no
+        --rand-watches=yes --lookback=yes --lookahead=no
+        --heuristic=Berkmin --rand-freq=0.0 --rand-prob=no
         --restarts=100,1.5 --shuffle=0,0 --deletion=3.0,1.1,3.0
@@
 numeric argument~\code{1} instructs \clasp\ to terminate immediately
-after finding an answer set.
-Together with \code{--lookback=yes}, \code{--restarts=100,1.5} lets \clasp\ apply a geometric
+after finding an answer set,
+and with \code{--lookback=yes}, \code{--restarts=100,1.5} lets \clasp\ apply a geometric
 restart policy.

### 2010-07-02 doc/guide/restrictions.tex
- made bingo the new gringo trunk
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3149 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-08 doc/documentation/restrictions.tex
@@
 
+\begin{oldstuff}
 In view of function symbols with non-zero arity,
@@
 and rather be looked up later on to find out details.
+\end{oldstuff}
 
+\begin{newstuff}
+In order to apply a grounding algorithm to a logic program,
+we have to impose certain restrictions on the input format.
+We will allow for a very general class of logic programs,
+so called safe programs,
+that give us a lot of modeling flexibility.
+In fact, \gringo-3.0's new input language is Turing-complete.
+This implies that in contrast to previous versions\comment{refer to appendix},
+the grounding is no longer guaranteed to be finite
+and thus the grounding algorithm will not necessarily halt.
+On first sight this might seem to be a disadvantage but is compensated by the fact 
+that much more interesting problems, often in a more concise way, can be represented now.
+For example, we can run some calculation till a certain fix-point is reached.
+This has not been possible before and frequently workarounds were needed, e.g, 
+large domains that provide the necessary values to perform the desired calculation.
+With safe programs such values can be calculated on demand,
+which often prevents space explosion, and also keeps the encoding more compact.
+The formal definition of safe programs programs,
+which can be grounded by \gringo\ (stand-alone or embedded in \clingo\ and \iclingo),
+is provided in Section~\ref{subsec:safe}.
+Section~\ref{subsec:strat} introduces stratified programs and the related
+concept of domain predicates, which can serve particular purposes during grounding.
+Both sections may be skimmed or even skipped upon the first reading,
+and rather be looked up later on to find out details.
+\end{newstuff}
+
+\begin{oldstuff}
 \subsection{Level-Restricted Logic Programs}\label{subsec:lambda}
@@
 will be discussed in Section~\ref{sec:language}.
+\end{oldstuff}
+
+\begin{newstuff}
+\subsection{Safe Logic Programs}\label{subsec:safe}
+
+The main task of a grounder is to substitute the variables in a logic program~$\Pi$
+by terms such that the result is a finite equivalent ground program~$\alt{\Pi}$.%
+\footnote{$\Pi$ and~$\alt{\Pi}$ are equivalent if they have the same answer sets.}
+Of course, a necessary condition for this is that~$\Pi$ possesses
+only finitely many finite answer sets.
+Unfortunately, such a property is undecidable in general~\cite{daeigovo01a}.
+Instead of semantic properties, grounders do thus impose rather simple syntactic
+conditions to guarantee the existence of finite equivalent ground programs
+(which are then also computed).
+For instance, the \dlv\ system~\cite{dlv03a} requires programs to be \emph{safe} for
+establishing finiteness in the absence of functions with non-zero arity and under
+limited arithmetic.
+Grounder \lparse~\cite{lparseManual} deals with \emph{$\omega$-restricted}
+programs~\cite{syrjanen01a}, which are more restrictive than safe programs
+but guarantee finiteness also in the presence of functions with non-zero arity.
+Previous \gringo~\cite{gescth07a} versions required programs to be \emph{$\lambda$-restricted},
+a property more general than $\omega$-restrictedness but likewise applicable
+to programs over the same languages.
+Finally, recent \gringo\ versions and, e.g., \dlvcomplex~\comment{add reference} use \emph{safe programs}, 
+which are a strict super set of \emph{$\lambda$-restricted} programs 
+and, in constrast to \dlv\, do not limit arithmetics and also allow for using function symbols.
+In this setting, it is no longer possibile to check
+whether there is a finite grounding for a given program and 
+thus a grounding algorithm need not necessarily hold on some inputs.
+
+A logic program is safe if every variable in each rule occurs in some positive predicate of the rule.
+In this case, we can apply \emph{semi-naive evaluation} \comment{add reference} to ground the logic program.
+Whenever we ground a rule and see a new head, we can use this head to bind variables
+in corresponding positive predicates to obtain new ground rules.
+This process is repeated until no more new ground rules will be generated.
+The sketched approach is justified by the $\subseteq$-minimality of answer sets
+in the sense of Section~\ref{subsec:semantics},
+as it allows us to restrict the attention to rules whose bodies are derivable w.r.t.\
+some answer set.
+
+\begin{example}\label{ex:safe}
+Consider the following logic program:
+%
+\lstinputlisting{examples/zigzagx.lp}
+%
+The set~$\mathcal{P}$ of predicates contains
+\pred{zig}/$1$, \pred{zag}/$1$,
+and \pred{zigzag}/$2$.
+The rules~$r$ with head predicates \pred{zig}/$1$ or \pred{zag}/$1$ in Line~1 and~2
+are ground ($\vars{r}=\nolinebreak\emptyset$).
+Hence, we immediately get the two ground rules
+\begin{lstlisting}[firstnumber=1]
+zigzag(0).
+zigzag(1).
+\end{lstlisting}
+There are two new heads $\pred{zig}(0)$ and $\pred{zag}(1)$ that can be used to instantiate the third rule
+\begin{lstlisting}[firstnumber=3]
+zigzag(0,1) :- zig(0), zag(1).
+\end{lstlisting}
+This allows use to use $\pred{zigzag}(0,1)$ to instantiate the last rule
+\begin{lstlisting}[firstnumber=4]
+zigzag(1,0) :- zigzag(0,1).
+\end{lstlisting}
+We get a new head $\pred{zigzag}(1,0)$ that will be used to instantiate the same rule again
+\begin{lstlisting}[firstnumber=5]
+zigzag(0,1) :- zigzag(1,0).
+\end{lstlisting}
+At this point, no new heads have been generated and the grounding process stops.
+\end{example}
+
+\begin{example}\label{ex:unsafe}
+Note that safety is a very intuitive concept.
+Consider for example the following unsafe program
+\begin{lstlisting}[firstnumber=1]
+a(X) :- not b(Y).
+\end{lstlisting}
+Using an infinite Herbrand base, the above program neither has a finite instantiation nor a finite answer set
+and we cannot apply the grounding algorithm sketched above.\comment{Refer to literature that covers such programs?}
+\end{example}
+
+The fundamental property of a safe
+program is that it admits using the grounding algorithm sketched above 
+and that it might yield a finite ground instantiation, not to be confused with
+the (full) ground instantiation because $\ground{\Pi}$ may still be infinite.
+Furthermore, safety provides a syntactic criterion that is not
+difficult to check.
+In fact, \gringo\ performs such a check and accepts an input program only if it is safe.
+To generalize the notion of safety to the rich input language of \gringo,
+we formalize safety using a directed \emph{variable graph} $G_r=(V \cup \{\$\},E)$
+w.r.t. a rule~$r$.
+The set~$V$ corresponds to the variables that occur in the rule and
+the set of edges~$E \subseteq (V \cup \{\$\}) \times V$ connect variables 
+or the special start node $\$$ with other variables.
+Only positive predicates contribute to edges of the graph.
+For every variable~$X$ that occurs in a positive literal of the rule there is an edge~$(\$,X)\in E$.
+A rule~$r$ is safe if there is a level mapping $\lambda:V\cup\{\$\}\rightarrow\mathbb{N}$ 
+and for every~$X \in V$ exits a~$Y \in V\cup\{\$\}$ 
+such that $\lambda(X) < \lambda(Y)$ and $(X,Y) \in E$ for $G_r=(V \cup \{\$\}, E)$.
+Finally, a logic program $\Pi$ is safe if all its rules are safe.
+In Section~\ref{sec:language}, extensions to this graph will be dicussed
+whenever new language constructs are introduced.
+\end{newstuff}
 
@@
 predicates that are subject to stratification can serve as domain predicates (see below).
-As with level-restrictedness, stratification can be characterized in terms
+Stratification can be characterized in terms
 of level mappings, here, witnessing the absence of recursion through ``\code{not}.''
@@
 \begin{example}\label{ex:strat:max}
-As the program in Line 1--4 of Example~\ref{ex:strat:prog} is stratified,
+As the program in Line~1--4 of Example~\ref{ex:strat:prog} is stratified,
 the maximal stratified subprogram consists of all rules.
@@
 
-One can check that the program consisting of Line~1--5
-in Example~\ref{ex:strat:prog} is level-restricted, but not stratified.
-Conversely, program
+Consider the safe, stratified (take $\xi=\{\pred{nat}/1 \mapsto 0\}$) program
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-is stratified (take $\xi=\{\pred{nat}/1 \mapsto 0\}$),
-but not level-restricted
-($\lambda(\pred{nat}/1)>\lambda(\pred{nat}/1)$
- needed because of the second rule is impossible).
 In fact, the program has a unique but infinite answer set.
 Thus,
-the program cannot be translated to a finite equivalent ground program,
-which is why \gringo\ does not accept it.
-Rather, in the first place, a program must be level-restricted in order
-to ground it with \gringo.
-If level-restrictedness is granted,
-the maximal stratified subprogram is exploited in a second stage
-to optimize grounding and to make use of domain predicates.
-
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "guide"
-%%% End:
+the program cannot be translated to a finite equivalent ground program.
+Nevertheless, \gringo\ would accept it, but it will not halt (assuming infinite memory).
+Stratified subprograms can be exploited during instantiation
+to optimize the grounding, e.g. facts in rule bodies need not be printed, and to make use of domain predicates, e.g., in aggregates~\comment{reference aggregates}.

### 2010-07-02 app/clasp/clasp_options.cpp

@@ -1,470 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#include "clasp_options.h"
-#include "program_opts/value.h"  
-#include <clasp/satelite.h>
-#include <sstream>
-#include <algorithm>
-
-using namespace ProgramOptions;
-using namespace std;
-
-namespace Clasp { namespace {
-
-bool mapSeed(const std::string& s, int& i) {
-	if (ProgramOptions::parseValue(s, i, 1)) {
-		if    (i > 0) { Clasp::srand(i); }
-		return i > 0 || i == -1;
-	}
-	return false;
-}
-} // end unnamed namespace
-
-/////////////////////////////////////////////////////////////////////////////////////////
-// Parseing & Mapping of options
-/////////////////////////////////////////////////////////////////////////////////////////
-bool parsePositional(const std::string& t, std::string& out) {
-	int num;
-	if (ProgramOptions::parseValue(t, num, 1)) { out = "number"; }
-	else                                       { out = "file";   }
-	return true;
-}
-
-bool parseValue(const std::string&, SolveOptionsWrapper&, int)      { return false; }
-bool parseValue(const std::string&, SolverStrategiesWrapper&, int)  { return false; }
-bool parseValue(const std::string&s , ApiOptions::ExtRuleMode& i, int){
-	std::string temp = toLower(s);
-	bool b = temp == "all";
[…]

### 2010-07-02 app/clingo/clingo_app.h

@@ -1,0 +1,502 @@
+// Copyright (c) 2010, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+// Copyright (c) 2009, Benjamin Kaufmann
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <program_opts/app_options.h>
+#include <gringo/output.h>
+#include <gringo/inclit.h>
+#include <gringo/converter.h>
+#include <gringo/streams.h>
+#include "clasp/clasp_options.h"
+#include "clasp/clasp_output.h"
+#include "clasp/smodels_constraints.h"
+#include "gringo/gringo_app.h"
+#include "clingo/clingo_options.h"
+#include "clingo/claspoutput.h"
+#include "clingo/timer.h"
+#include <iomanip>
+
+#define CLASP_VERSION "1.3.2"
+
+// (i)Clingo application, i.e. gringo+clasp
+template <bool ICLINGO>
+class ClingoApp : public GringoApp, public Clasp::ClaspFacade::Callback
+{
+public:
+	/** returns a singleton instance */
+	static ClingoApp& instance();
+
+protected:
+	// AppOptions interface
+	void initOptions(ProgramOptions::OptionGroup& root, ProgramOptions::OptionGroup& hidden)
+	{
+		config_.solver = &solver_;
+		cmdOpts_.setConfig(&config_);
+		cmdOpts_.initOptions(root, hidden);
+		clingo_.initOptions(root, hidden);
+		generic.verbose = 1;
+		GringoApp::initOptions(root, hidden);
+	}
+
+	void addDefaults(std::string& defaults)
+	{
+		cmdOpts_.addDefaults(defaults);
+		clingo_.addDefaults(defaults);
[…]

### 2010-07-02 app/gringo_app.cpp

@@ -1,565 +1,0 @@
-// 
-// Copyright (c) 2009, Benjamin Kaufmann
-// 
-// This file is part of gringo. See http://www.cs.uni-potsdam.de/gringo/ 
-// 
-// gringo is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// gringo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with gringo; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#include "gringo_app.h"
-#include <gringo/grounder.h>
-#include <gringo/gringoexception.h>
-#include <gringo/smodelsoutput.h>
-#include <gringo/lparseoutput.h>
-#include <gringo/pilsoutput.h>
-#include <gringo/lparseconverter.h>
-#include <gringo/lparseparser.h>
-#include <gringo/domain.h>
-#if defined(WITH_CLASP)
-#include <clasp/reader.h>
-#include <clasp/smodels_constraints.h>
-#include <gringo/claspoutput.h>
-#include <iomanip>
-#endif
-#include <iostream>
-#include <fstream>
-#include <sstream>
-#include <signal.h>
-using namespace std;
-namespace gringo {
-
-/////////////////////////////////////////////////////////////////////////////////////////
-// Application - common stuff
-/////////////////////////////////////////////////////////////////////////////////////////
-Application::Application() {}
-Application& Application::instance() {
-#if !defined(WITH_CLASP)
-	static GringoApp app;
-#else
-	static ClingoApp app;
-#endif
-	return app;
-}
-
-void Application::printWarnings() {
-	while (!messages.warning.empty()) {
-		cerr << "Warning: " << messages.warning.back() << endl;
-		messages.warning.pop_back();
-	}
-}
-
[…]

### 2010-07-02 app/main.cpp

@@ -1,35 +1,0 @@
-// 
-// Copyright (c) 2009, Benjamin Kaufmann
-// 
-// This file is part of gringo. See http://www.cs.uni-potsdam.de/gringo/ 
-// 
-// gringo is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// gringo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with gringo; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#include "gringo_app.h"
-#include <iostream>
-/////////////////////////////////////////////////////////////////////////////////////////
-// main - entry point
-/////////////////////////////////////////////////////////////////////////////////////////
-// #define CHECK_HEAP
-int main(int argc, char** argv) {
-	try {
-		return gringo::Application::instance().run(argc, argv);
-	} 
-	catch (...) {
-		std::cerr << "\n*** INTERNAL ERROR: please contact the authors of gringo" << std::endl;
-		return EXIT_FAILURE;
-	}
-	return EXIT_SUCCESS;
-}

### 2010-07-02 banane.sh

@@ -1,0 +1,44 @@
+#!/bin/bash
+
+svn update
+REV=$(svn status -v | grep -o "^ *[0-9]* *[0-9]*" | head -n1 | grep -o "[0-9]*$")
+make debug release static target=all || exit 1
+
+prefix=/home/wv/bin/linux/32
+public_html=${HOME}/public_html/bingo/bingo-x86-linux
+rm -rf $public_html
+mkdir -p $public_html
+
+
+for target in bingo clabingo iclabingo
+do
+	for variant in debug static
+	do
+		case $variant in
+			debug)
+				rm -f /home/wv/bin/linux/32/$target-dbg*
+				cp build/$variant/bin/$target $prefix/$target-dbg-r$REV
+				chmod g+w $prefix/$target-dbg-r$REV
+				(cd $prefix; rm -f $target-dbg-banane; ln -s $target-dbg-r$REV $target-dbg-banane)
+				;;
+			static)
+				name=$public_html/$target
+				cp build/$variant/bin/$target $name
+				strip $name
+				(
+					cd $public_html
+					cd ..
+					rm -f bingo-x86-linux.tar.bz2
+					tar -cjf bingo-x86-linux.tar.bz2 bingo-x86-linux
+				)
+				name=$target-r$REV
+				cp build/$variant/bin/$target $prefix/$name
+				chmod g+w $prefix/$name
+				strip $prefix/$name
+				(cd $prefix; rm -f $target-banane; ln -s $name $target-banane)
+				;;
+		esac
+	done
+done
+
+rm -rf $public_html

### 2010-07-02 libclasp/clasp/pod_vector.h

@@ -1,58 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-
-#ifndef CLASP_POD_VECTOR_H_INCLUDED
-#define CLASP_POD_VECTOR_H_INCLUDED
-#include <vector>
-#include <clasp/util/pod_vector.h>
-namespace Clasp {
-
-#ifdef _DEBUG
-	template <class Type>
-	struct PodVector {
-		typedef std::vector<Type> type;
-		static void destruct(type& t) {t.clear();}
-	};
-#else
-	template <class Type>
-	struct PodVector {
-		typedef bk_lib::pod_vector<Type> type;
-		static void destruct(type& t) {
-			for (typename type::size_type i = 0, end = t.size(); i != end; ++i) {
-				t[i].~Type();
-			}
-			t.clear();
-		}
-	};
-#endif
-
-template <class T>
-void releaseVec(T& t) {
-	T().swap(t);
-}
-
-template <class T>
-void shrinkVecTo(T& t, typename T::size_type j) {
-	t.erase(t.begin()+j, t.end());
-}
-
-}
-
-#endif

### 2010-07-02 libclasp/clasp/util/platform.h

@@ -1,81 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-
-#ifndef CLASP_PLATFORM_H_INCLUDED
-#define CLASP_PLATFORM_H_INCLUDED
-
-#ifdef _MSC_VER
-#pragma once
-#endif
-
-#if defined(_MSC_VER) && _MSC_VER >= 1200
-#include <basetsd.h>
-typedef UINT8     uint8;
-typedef UINT16    uint16;
-typedef INT32     int32;
-typedef UINT32    uint32;
-typedef UINT64    uint64;
-typedef INT64     int64;
-typedef UINT_PTR  uintp;
-#define PRIu64 "llu"
-#define PRId64 "lld"
-template <unsigned> struct Uint_t;
-template <> struct Uint_t<sizeof(uint8)>  { typedef uint8  type; };
-template <> struct Uint_t<sizeof(uint16)> { typedef uint16 type; };
-template <> struct Uint_t<sizeof(uint32)> { typedef uint32 type; };
-template <> struct Uint_t<sizeof(uint64)> { typedef uint64 type; };
-#define BIT_MASK(x,n) ( static_cast<Uint_t<sizeof((x))>::type>(1) << (n) )
-#elif defined(__GNUC__) && __GNUC__ >= 3
-#if !defined(__STDC_FORMAT_MACROS)
-#define __STDC_FORMAT_MACROS
-#endif
-#include <inttypes.h>
-typedef uint8_t	    uint8;
[…]

### 2010-07-02 libclasp/src/heuristics.cpp

@@ -1,532 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#include <clasp/heuristics.h>
-#include <algorithm>
-namespace Clasp { namespace {
-
-// our own rng for heuristics
-uint32 heuRand_g = 1;
-inline uint32 heuRand() {
-	return( ((heuRand_g = heuRand_g * 214013L + 2531011L) >> 16) & 0x7fff );
-}
-}
-/////////////////////////////////////////////////////////////////////////////////////////
-// Lookback selection strategies
-/////////////////////////////////////////////////////////////////////////////////////////
-uint32 momsScore(const Solver& s, Var v) {
-	uint32 sc;
-	if (s.numBinaryConstraints()) {
-		uint32 s1 = s.estimateBCP(posLit(v), 0) - 1;
-		uint32 s2 = s.estimateBCP(negLit(v), 0) - 1;
-		sc = ((s1 * s2)<<10) + (s1 + s2);
-	}
-	else {
-		// problem does not contain binary constraints - fall back to counting watches
-		uint32 s1 = s.numWatches(posLit(v));
-		uint32 s2 = s.numWatches(negLit(v));
-		sc = ((s1 * s2)<<10) + (s1 + s2);
-	}
-	return sc;
-}
-
[…]

### 2010-07-02 libclasp/src/lookahead.cpp

@@ -1,398 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#include <clasp/lookahead.h>
-#include <algorithm>
-namespace Clasp { 
-/////////////////////////////////////////////////////////////////////////////////////////
-// Lookahead scoring
-/////////////////////////////////////////////////////////////////////////////////////////
-uint32 ScoreLook::countNant(const Solver& s, const Literal* b, const Literal *e) const {
-	uint32 sc = 0;
-	for (; b != e; ++b) { sc += s.nant(b->var()); }
-	return sc;
-}
-void ScoreLook::scoreLits(const Solver& s, const Literal* b, const Literal *e) {
-	assert(b < e);
-	uint32 sc = !nant ? uint32(e-b) : countNant(s, b, e);
-	Var v     = b->var();
-	score[v].setScore(*b, sc);
-	if (addDeps) {
-		if ((score[v].testedBoth() || mode == score_max) && greater(v, best)) {
-			best = v;
-		}
-		for (; b != e; ++b) {
-			v = b->var();
-			if ( (s.type(v) & types) != 0) {
-				if (!score[v].seen()) { deps.push_back(v); }
-				score[v].setDepScore(*b, sc);
-				score[v].setSeen(*b);
-			}
-		}
-	}
-}
-void ScoreLook::clearDeps() {
[…]

### 2010-07-02 libclasp/src/preprocessor.cpp

@@ -1,823 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#include <clasp/preprocessor.h>
-#include <clasp/program_builder.h>
-namespace Clasp { namespace {
-
-struct LessBodySize {
-	LessBodySize(const BodyList& bl) : bodies_(&bl) {}
-	bool operator()(Var b1, Var b2 ) const {
-		return (*bodies_)[b1]->size() < (*bodies_)[b2]->size()
-			|| ((*bodies_)[b1]->size() == (*bodies_)[b2]->size() && (*bodies_)[b1]->rtype() == BASICRULE && (*bodies_)[b2]->rtype() != BASICRULE);
-	}
-private:
-	const BodyList* bodies_;
-};
-}
-
-Literal Preprocessor::newBodyLit(PrgBodyNode* n) {
-	if      (!backprop_||n->value()!=value_false) { 
-		                      return posLit(prg_->vars_.add(Var_t::body_var)); }
-	else if (false_ == 0) { return posLit(false_ = prg_->vars_.add(Var_t::body_var)); }
-	else                  { return posLit(false_); }
-}
-
-// If the program is defined incrementally, marks atoms from previous steps as supported
-// Pre: Atoms that are true/false have their value-member set
-void Preprocessor::updatePreviouslyDefinedAtoms(Var startAtom, bool strong) {
-	for (Var i = 0; i != startAtom; ++i) {
-		PrgAtomNode* a = prg_->atoms_[i];
[…]

### 2010-07-02 libclasp/src/satelite.cpp

@@ -1,695 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#include <clasp/satelite.h>
-#include <clasp/clause.h>
-#include <ctime>
-
-#ifdef _MSC_VER
-#pragma warning (disable : 4200) // nonstandard extension used : zero-sized array
-#endif
-namespace Clasp { namespace SatElite {
-
-inline uint64 abstractLit(Literal p) {
-	assert(p.var() > 0); 
-	return uint64(1) << ((p.var()-1) & 63); 
-}
-
-// A clause class optimized for the SatElite preprocessor.
-// Uses 1-literal watching to implement forward-subsumption
-class Clause {
-public:
-	static Clause* newClause(const LitVec& lits) {
-		void* mem = ::operator new( sizeof(Clause) + (lits.size()*sizeof(Literal)) );
-		return new (mem) Clause(lits);
-	}
-	bool simplify(Solver& s) {
-		uint32 i;
-		for (i = 0; i != size_ && s.value(lits_[i].var()) == value_free; ++i) {;}
-		if      (i == size_)          { return false; }
-		else if (s.isTrue(lits_[i]))  { return true;  }
-		uint32 j = i++;
-		for (; i != size_; ++i) {
-			if (s.isTrue(lits_[i]))   { return true; }
-			if (!s.isFalse(lits_[i])) { lits_[j++] = lits_[i]; }
-		}
-		size_ = j;
-		return false;
-	}
-	void destroy() {
[…]

### 2010-07-02 libclasp/src/solve_algorithms.cpp

@@ -1,292 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#include <clasp/solve_algorithms.h>
-#include <clasp/solver.h>
-#include <clasp/smodels_constraints.h>
-#include <cmath>
-using std::log;
-namespace Clasp { 
-
-/////////////////////////////////////////////////////////////////////////////////////////
-// Enumerator
-/////////////////////////////////////////////////////////////////////////////////////////
-Enumerator::Enumerator(Report* r) : numModels_(1), report_(r), progress_(0), mini_(0), limits_(0), restartOnModel_(false)  {}
-Enumerator::~Enumerator()     { if (mini_) mini_->destroy(); delete limits_; }
-Enumerator::ProgressReport::ProgressReport()  {}
-Enumerator::ProgressReport::~ProgressReport() {}
-Enumerator::Report::Report()  {}
-void Enumerator::setReport(Report* r)                   {  report_   = r; }
-void Enumerator::enableProgressReport(ProgressReport* r){  progress_ = r; }
-void Enumerator::setMinimize(MinimizeConstraint* min)   {  mini_ = min;   }
-void Enumerator::setRestartOnModel(bool r)              { restartOnModel_ = r; }
-void Enumerator::setSearchLimit(int64 maxC, int64 maxR) {
-	delete limits_; limits_ = 0;
-	if (maxC > 0 || maxR > 0) {
-		limits_ = new SearchLimits();
[…]

### 2010-07-02 libclasp/tests/enumerator_test.cpp

@@ -1,127 +1,0 @@
-#include <cppunit/TestFixture.h>
-#include <cppunit/TestAssert.h>
-#include <cppunit/extensions/HelperMacros.h>
-#include "common.h"
-#include <clasp/solver.h>
-#include <clasp/program_builder.h>
-#include <clasp/unfounded_check.h>
-#include <clasp/smodels_constraints.h>
-#include <clasp/model_enumerators.h>
-#include <sstream>
-using namespace std;
-namespace Clasp { namespace Test {
-	
-class EnumeratorTest : public CppUnit::TestFixture {
-  CPPUNIT_TEST_SUITE(EnumeratorTest);
-	CPPUNIT_TEST(testMiniProject);
-	CPPUNIT_TEST(testProjectBug);
-	CPPUNIT_TEST(testTerminateRemovesWatches);
-	CPPUNIT_TEST_SUITE_END();	
-public:
-	void testMiniProject() {
-		solver.strategies().symTab.reset(new AtomIndex);
-		builder.startProgram(*solver.strategies().symTab, 0, 0)
-			.setAtomName(1, "a").setAtomName(2, "b").setAtomName(3, "_x")
-			.startRule(CHOICERULE).addHead(1).addHead(2).addHead(3).endRule()
-			.startRule().addHead(3).addToBody(1, false).endRule()
-			.startRule().addHead(3).addToBody(2, false).endRule()
-			.startRule(OPTIMIZERULE).addToBody(3, true).endRule()
-		;
-		CPPUNIT_ASSERT_EQUAL(true, builder.endProgram(solver, true));
-		MinimizeConstraint* m = builder.createMinimize(solver, false);
-		BacktrackEnumerator enumerator(0);
-		enumerator.setMinimize(m);
-		enumerator.setEnableProjection(true);
-		enumerator.init(solver, 0);
-		AtomIndex& index = *solver.strategies().symTab;
-		solver.assume(index[1].lit);
-		solver.propagate();
-		solver.assume(index[2].lit);
-		solver.propagate();
-		solver.assume(index[3].lit);
-		solver.propagate();
-		CPPUNIT_ASSERT(solver.numVars() == solver.numAssignedVars());
-		CPPUNIT_ASSERT_EQUAL(false, enumerator.backtrackFromModel(solver));
-	}
-
-	void testProjectBug() {
-		solver.strategies().symTab.reset(new AtomIndex);
[…]

### 2010-07-02 libclasp/tests/rule_test.cpp

@@ -1,540 +1,0 @@
-#include <cppunit/TestFixture.h>
-#include <cppunit/TestAssert.h>
-#include <cppunit/extensions/HelperMacros.h>
-#include <clasp/program_builder.h>
-#include <clasp/solver.h>
-#include <utility>
-namespace Clasp { namespace Test {
-
-class RuleTest : public CppUnit::TestFixture {
-
-  CPPUNIT_TEST_SUITE(RuleTest);
-	CPPUNIT_TEST(testHashIgnoresOrder);
-	CPPUNIT_TEST(testRemoveDuplicateInNormal);
-	CPPUNIT_TEST(testMergeDuplicateInExtended);
-	CPPUNIT_TEST(testContraNormal);
-	CPPUNIT_TEST(testNoContraExtended);
-	CPPUNIT_TEST(testContraExtended);
-	CPPUNIT_TEST(testMultiSimplify);
-	CPPUNIT_TEST(testCardinalityIfAllWeightsEqual);
-	CPPUNIT_TEST(testNormalIfMinWeightNeeded);
-	CPPUNIT_TEST(testSelfblockNormal);
-	CPPUNIT_TEST(testTautNormal);
-	CPPUNIT_TEST_SUITE_END();
-
-public:
-	void setUp() {
-		rs.clear();
-	}
-
-	void testHashIgnoresOrder() {
-		PrgRule r1, r2, r3;
-		RuleState rs1, rs2, rs3;
-		PrgRule::RData rd1, rd2, rd3;
-		r1.setType(BASICRULE).addHead(1).addToBody(10, false).addToBody(20, true).addToBody(25, true);
-		r2.setType(BASICRULE).addHead(1).addToBody(20, true).addToBody(25, true).addToBody(10, false);
-		r3.setType(BASICRULE).addHead(1).addToBody(25, true).addToBody(10, false).addToBody(20, true);
-		rd1 = r1.simplify(rs1);
-		rd2 = r2.simplify(rs2);
-		rd3 = r3.simplify(rs3);
-		CPPUNIT_ASSERT(rd1.hash == rd2.hash && rd2.hash == rd3.hash);
-	}
-
-	void testRemoveDuplicateInNormal() {
-		// a :- b, b, not c -> a :- b, not c.
-		rule.setType(BASICRULE).addHead(1).addToBody(2, true).addToBody(2, true).addToBody(3, false);
-		res = rule.simplify(rs);
[…]

### 2010-07-02 libclasp/tests/smodels_constraints_test.cpp

@@ -1,550 +1,0 @@
-#include <cppunit/TestFixture.h>
-#include <cppunit/TestAssert.h>
-#include <cppunit/extensions/HelperMacros.h>
-#include <algorithm>
-#include <clasp/smodels_constraints.h>
-#include <clasp/solver.h>
-using namespace std;
-
-namespace Clasp { namespace Test {
-class CardTest : public CppUnit::TestFixture {
-	CPPUNIT_TEST_SUITE(CardTest);
-	CPPUNIT_TEST(testAssertTriviallySat);
-	CPPUNIT_TEST(testAssertTriviallyUnSat);
-	CPPUNIT_TEST(testAssertNotSoTriviallySat);
-	CPPUNIT_TEST(testAssertNotSoTriviallyUnSat);
-
-	CPPUNIT_TEST(testTrivialBackpropTrue);
-	CPPUNIT_TEST(testTrivialBackpropFalse);
-	CPPUNIT_TEST(testTrivialBackpropFalseWeight);
-	
-	CPPUNIT_TEST(testForwardTrue);
-	CPPUNIT_TEST(testForwardFalse);
-	CPPUNIT_TEST(testBackwardTrue);
-	CPPUNIT_TEST(testBackwardFalse);
-
-	CPPUNIT_TEST(testForwardTrueConflict);
-	CPPUNIT_TEST(testForwardFalseConflict);
-	CPPUNIT_TEST(testBackwardTrueConflict);
-	CPPUNIT_TEST(testBackwardFalseConflict);
-
-	CPPUNIT_TEST(testReasonBug);
-	CPPUNIT_TEST(testWeightReasonAfterBackprop);
-	CPPUNIT_TEST(testOrderBug);
-	CPPUNIT_TEST(testBackwardAfterForward);
-	
-	CPPUNIT_TEST(testSimplify);
-
-	CPPUNIT_TEST(testAssertWeightTriviallySat);
-	CPPUNIT_TEST(testAssertWeightTriviallyUnSat);
-	CPPUNIT_TEST(testAssertWeightNotSoTriviallySat);
-	CPPUNIT_TEST(testAssertWeightNotSoTriviallyUnSat);
-	CPPUNIT_TEST(testWeightForwardTrue);
-	CPPUNIT_TEST(testWeightForwardFalse);
-	CPPUNIT_TEST(testWeightBackwardTrue);
-	CPPUNIT_TEST(testWeightBackwardFalse);
[…]

### 2010-07-02 libgringo/gringo/constterm.h

@@ -1,0 +1,38 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/term.h>
+
+class ConstTerm : public Term
+{
+public:
+	ConstTerm(const Loc &loc, const Val &v);
+	Val val(Grounder *grounder) const;
+	void normalize(Lit *parent, const Ref &ref, Grounder *grounder, Expander *expander, bool unify) { (void)parent; (void)ref; (void)grounder; (void)expander; (void)unify; }
+	bool unify(Grounder *grounder, const Val &v, int binder) const;
+	void vars(VarSet &vars) const;
+	void visit(PrgVisitor *visitor, bool bind);
+	void print(Storage *sto, std::ostream &out) const;
+	bool constant() const { return true; }
+	Term *clone() const;
+private:
+	Val val_;
+};
+

### 2010-07-02 libgringo/gringo/functionterm.h

@@ -1,54 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#ifndef FUNCTIONTERM_H
-#define FUNCTIONTERM_H
-
-#include <gringo/gringo.h>
-#include <gringo/term.h>
-
-namespace gringo
-{
-	class FunctionTerm : public Term
-	{
-	public:
-		enum FunctionType { PLUS, MINUS, POWER, TIMES, DIVIDE, MOD, ABS, BITOR, BITXOR, BITAND, COMPLEMENT };
-	public:
-		FunctionTerm(FunctionType type, Term *a, Term *b = 0);
-		FunctionTerm(const FunctionTerm &f);
-		virtual void getVars(VarSet &vars) const;
-		virtual bool isComplex();
-		virtual void print(const GlobalStorage *g, std::ostream &out) const;
-		virtual Value getValue(Grounder *g);
-		virtual Value getConstValue(Grounder *g);
-		virtual Term* clone() const;
-		virtual void preprocess(Literal *l, Term *&p, Grounder *g, Expandable *e);
-		virtual bool unify(const GlobalStorage *g, const Value& t, const VarVector& vars, ValueVector& subst) const
-		{
-			FAIL(true);
-		}
-		virtual void addIncParam(Grounder *g, Term *&p, const Value &v);
-		virtual ~FunctionTerm();
-	protected:
-		FunctionType type_;
-		Term *a_;
-		Term *b_;
-	};
-}
-
-#endif
-

### 2010-07-02 libgringo/gringo/gringolexer.h

@@ -1,49 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#ifndef GRINGOLPARSELEXER_H
-#define GRINGOLPARSELEXER_H
-
-#include <gringo/gringo.h>
-
-namespace gringo
-{
-	class GrinGoLexer
-	{
-	public:
-		GrinGoLexer(std::istream *in = &std::cin);
-		std::string getToken();
-		void reset(std::istream *i = &std::cin);
-		int getLine();
-		void step();
-		int getColumn();
-		virtual ~GrinGoLexer();
-
-	protected:
-		void fill(int n);
-
-	protected:
-		std::istream *in;
-		int bufmin;
-		int bufsize;
-		char *buffer, *start, *offset, *cursor, *limit, *marker, *eof;
-		unsigned int line;
-	};
-}
-
-#endif
-

### 2010-07-02 libgringo/gringo/inclit.h

@@ -1,0 +1,66 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/printer.h>
+#include <gringo/lit.h>
+
+struct IncConfig
+{
+	int  incBegin;
+	int  incEnd;
+	bool incBase;
+	bool incVolatile;
+
+	IncConfig()
+		: incBegin(1)
+		, incEnd(1)
+		, incBase(false)
+		, incVolatile(true)
+	{ }
+};
+
+class IncLit : public Lit
+{
+public:
+	class Printer : public ::Printer
+	{
+	public:
+		virtual void print() { }
+		virtual ~Printer() { }
+	};
+
+public:
+	IncLit(const Loc &loc, IncConfig &config_, bool cumulative, uint32_t varId);
+	void normalize(Grounder *g, Expander *expander);
+	bool fact() const { return true; }
+	bool isFalse(Grounder *g);
+	bool forcePrint() { return !cumulative_; }
+	void accept(::Printer *v);
+	void index(Grounder *g, Groundable *gr, VarSet &bound);
+	void visit(PrgVisitor *visitor);
+	bool match(Grounder *grounder);
+	void print(Storage *sto, std::ostream &out) const;
+	Lit *clone() const;
+	~IncLit();
+private:
+	IncConfig         &config_;
+	bool               cumulative_;
+	clone_ptr<VarTerm> var_;
+};

### 2010-07-02 libgringo/gringo/index_set.h

@@ -1,0 +1,93 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <vector>
+#include <boost/functional/hash.hpp>
+#include <boost/unordered/unordered_set.hpp>
+
+template <typename T>
+class index_set
+{
+private:
+	struct cmp;
+	typedef std::vector<T> val_vec;
+	typedef boost::unordered_set<uint32_t, cmp, cmp> idx_set;
+	struct cmp
+	{
+		cmp(const index_set<T> &set);
+		const T &val(uint32_t i) const;
+		size_t operator()(uint32_t i) const;
+		bool operator()(uint32_t a, uint32_t b) const;
+		const index_set<T> &set_;
+	};
+public:
+	index_set();
+	uint32_t insert(const T &val);
+	const T &get(uint32_t idx) const;
+private:
+	const T *lastVal_;
+	val_vec  vals_;
+	idx_set  index_;
+};
+
+template <typename T>
+inline index_set<T>::index_set()
+	: index_(42, cmp(*this), cmp(*this))
+{
+}
+
+template <typename T>
+inline index_set<T>::cmp::cmp(const index_set<T> &set)
+	: set_(set)
+{
+}
+
+template <typename T>
+inline const T &index_set<T>::cmp::val(uint32_t i) const
+{
+	return i < set_.vals_.size() ? set_.vals_[i] : *set_.lastVal_;
+}
+
+template <typename T>
+inline size_t index_set<T>::cmp::operator()(uint32_t i) const
+{
+	return boost::hash_value(val(i));
+}
+
+template <typename T>
+inline bool index_set<T>::cmp::operator()(uint32_t a, uint32_t b) const
+{
[…]

### 2010-07-02 libgringo/gringo/litdep.h

@@ -1,0 +1,121 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/prgvisitor.h>
+
+namespace LitDep
+{
+	class LitNode;
+	struct LitNodeCmp
+	{
+		bool operator()(LitNode *a, LitNode *b);
+	};
+
+	typedef std::priority_queue<LitNode*, std::vector<LitNode*>, LitNodeCmp> LitQueue;
+
+	class VarNode
+	{
+	private:
+		typedef std::vector<LitNode*> LitNodeVec;
+	public:
+		VarNode(VarTerm *var);
+		VarTerm *var() const;
+		void depend(LitNode *litNode);
+		void provide(LitNode *litNode);
+		void reset();
+		bool done();
+		void propagate(LitQueue &queue);
+	private:
+		bool       done_;
+		VarTerm    *var_;
+		LitNodeVec provide_;
+	};
+
+	class LitNode
+	{
+	private:
+		typedef std::vector<VarNode*> VarNodeVec;
+	public:
+		LitNode(Lit *lit);
+		void depend(VarNode *varNode);
+		void provide(VarNode *varNode);
+		void reset();
+		void check(LitQueue &queue);
+		void propagate(LitQueue &queue);
+		bool done();
+		void score(double score) { score_ = score; }
+		double score() const { return score_; }
+		Lit *lit() const { return lit_; }
+	private:
+		Lit       *lit_;
+		uint32_t   done_;
+		uint32_t   depend_;
+		double     score_;
+		VarNodeVec provide_;
+	};
+
[…]

### 2010-07-02 libgringo/gringo/lparselexer.h

@@ -1,38 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#ifndef LPARSELEXER_H
-#define LPARSELEXER_H
-
-#include <gringo/gringo.h>
-#include <gringo/gringolexer.h>
-
-namespace gringo
-{
-	class LparseLexer : public GrinGoLexer
-	{
-	public:
-		LparseLexer();
-		int lex(std::string *&val);
-		virtual ~LparseLexer();
-	private:
-		void backwardWarning();
-	};
-}
-
-#endif
-

### 2010-07-02 libgringo/gringo/optimize.h

@@ -1,0 +1,75 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/printer.h>
+#include <gringo/statement.h>
+#include <gringo/predlit.h>
+
+typedef boost::shared_ptr<PredLitSet> PredLitSetPtr;
+
+class Optimize : public Statement
+{
+private:
+	class PrioLit;
+	typedef std::vector<std::pair<int32_t, int32_t> > PrioVec;
+	friend PrioLit* new_clone(const PrioLit& a);
+public:
+	class Printer : public ::Printer
+	{
+	public:
+		void print(PredLitRep *l) { (void)l; assert(false); }
+		virtual void begin(bool maximize, bool set) = 0;
+		virtual void print(PredLitRep *l, int32_t weight, int32_t prio) = 0;
+		virtual void end() = 0;
+		virtual ~Printer() { }
+	};
+public:
+	Optimize(const Loc &loc, PredLit *head, Term *weight, Term *prio, LitPtrVec &body, bool maximize);
+	Term *weight();
+	Term *prio();
+	PredLit *head() const { return head_.get(); }
+	LitPtrVec &body() { return body_; }
+	void append(Lit *lit);
+	void ground(Grounder *g);
+	bool grounded(Grounder *g);
+	void normalize(Grounder *g);
+	void init(Grounder *g, const VarSet &bound);
+	void visit(PrgVisitor *visitor);
+	void print(Storage *sto, std::ostream &out) const;
+	bool maximize() { return maximize_; }
+	bool set() { return set_; }
+	void uniques(PredLitSetPtr uniques) { uniques_ = uniques; set_ = true; }
[…]

### 2010-07-02 libgringo/gringo/output.h

@@ -1,235 +1,113 @@
-// Copyright (c) 2008, Roland Kaminski
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
 //
-// This file is part of GrinGo.
+// This file is part of gringo.
 //
-// GrinGo is free software: you can redistribute it and/or modify
+// gringo is free software: you can redistribute it and/or modify
 // it under the terms of the GNU General Public License as published by
 // the Free Software Foundation, either version 3 of the License, or
 // (at your option) any later version.
 //
-// GrinGo is distributed in the hope that it will be useful,
+// gringo is distributed in the hope that it will be useful,
 // but WITHOUT ANY WARRANTY; without even the implied warranty of
 // MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
 // GNU General Public License for more details.
 //
 // You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
 
-#ifndef OUTPUT_H
-#define OUTPUT_H
+#pragma once
 
 #include <gringo/gringo.h>
-#include <gringo/value.h>
+#include <gringo/printer.h>
 
-namespace gringo
+class Output
 {
-	namespace NS_OUTPUT
-	{
-		typedef std::vector<Object*> ObjectVector;
+private:
+	typedef std::map<Signature, bool> ShowMap;
+	typedef std::set<Signature> ExternalSet;
+	typedef boost::ptr_vector<boost::nullable<Printer> > PrinterVec;
+public:
+	Output(std::ostream *out);
+	virtual void initialize() { }
+	virtual void finalize() { }
+	virtual void addDomain(Domain *d) { (void)d; }
+	std::ostream &out() const { return *out_; }
+	Storage *storage() const { return s_; }
+	void storage(Storage *s) { s_ = s; }
+	void show(bool s);
+	void show(uint32_t nameId, uint32_t arity, bool s);
+	virtual void doShow(bool s) { (void) s; }
+	virtual void doShow(uint32_t nameId, uint32_t arity, bool s) { (void)nameId; (void)arity; (void) s; }
+	bool show(uint32_t nameId, uint32_t arity);
+	void external(uint32_t nameId, uint32_t arity);
[…]

### 2010-07-02 libgringo/gringo/parser.h

@@ -1,0 +1,116 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/lexer_impl.h>
+#include <gringo/locateable.h>
+#include <gringo/optimize.h>
+
+class Parser : public LexerImpl
+{
+private:
+	struct DomStm
+	{
+		DomStm(const Loc &loc, uint32_t idx, uint32_t id, const VarSigVec &vars);
+		Loc       loc;
+		uint32_t  idx;
+		uint32_t  id;
+		VarSigVec vars;
+	};
+	typedef std::pair<Loc,std::string> ErrorTok;
+	typedef std::vector<ErrorTok> ErrorVec;
+	typedef boost::ptr_map<uint32_t, Term> ConstMap;
+	typedef std::list<DomStm> DomStmList;
+
+public:
+	typedef std::multimap<uint32_t, DomStm*> DomStmMap;
+	typedef boost::iterator_range<DomStmMap::iterator> DomStmRng;
+
+public:
+	struct Token
+	{
+		Loc loc() const { return Loc(file, line, column); }
+		uint32_t file;
+		uint32_t line;
+		uint32_t column;
+		union
+		{
+			int number;
+			uint32_t index;
+		};
+	};
+	enum iPart { IPART_BASE, IPART_CUMULATIVE, IPART_VOLATILE };
+
+public:
+	Parser(Grounder *g, IncConfig &config, Streams &streams);
+	int lex();
+	std::string errorToken();
+	void syntaxError();
+	void include(uint32_t filename);
+	void parseError();
+	void parse();
+	int  level() const { return level_; }
+	void nextLevel() { level_ ++; }
[…]

### 2010-07-02 libgringo/gringo/poolterm.h

@@ -1,0 +1,43 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/term.h>
+
+class PoolTerm : public Term
+{
+public:
+	PoolTerm(const Loc &loc, Term *a, Term *b);
+	Val val(Grounder *grounder) const { (void)grounder; assert(false); return Val::create(); }
+	bool constant() const { assert(false); return false; }
+	bool unify(Grounder *grounder, const Val &v, int binder) const { (void)grounder; (void)v; (void)binder; assert(false); return false; }
+	void vars(VarSet &v) const { (void)v; assert(false); }
+	void visit(PrgVisitor *visitor, bool bind) { (void)visitor; (void)bind; assert(false); }
+	void normalize(Lit *parent, const Ref &ref, Grounder *g, Expander *expander, bool unify);
+	void print(Storage *sto, std::ostream &out) const;
+	Term *clone() const;
+	~PoolTerm();
+private:
+	PoolTerm(const Loc &loc, Term *a);
+private:
+	mutable clone_ptr<Term> a_;
+	clone_ptr<Term>         b_;
+	bool                    clone_;
+};
+

### 2010-07-02 libgringo/gringo/printable.h

@@ -1,61 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#ifndef PRINTABLE_H
-#define PRINTABLE_H
-
-#include <gringo/gringo.h>
-
-namespace gringo
-{
-	typedef std::pair<GlobalStorage *, std::ostream> OutputStream;
-	/**
-	 * Interface for Printable objects like terms, literals or statements.
-	 * This interface is used to print out an internal representation of the problem instance.
-	 */
-	class Printable
-	{
-	public:
-		/**
-		 * \brief Constructor - does nothing
-		 */
-		Printable();
-		/**
-		 * \brief Virtual destructor
-		 */
-		virtual ~Printable();
-		/**
-		 * \brief Prints the object on the given stream
-		 * \param stream The output stream
-		 */
-		virtual void print(const GlobalStorage *g, std::ostream &out) const = 0;
-	};
-	
-	inline const std::pair<const GlobalStorage *, const Printable *> pp(const GlobalStorage *g, const Printable *p)
-	{
-		return std::make_pair(g, p);
-	}
-	
-	inline std::ostream &operator<<(std::ostream &out, const std::pair<const GlobalStorage *, const Printable *> &p)
-	{
-		p.second->print(p.first, out);
-		return out;
-	}
-}
-
-#endif
-

### 2010-07-02 libgringo/gringo/rangelit.h

@@ -1,0 +1,41 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/lit.h>
+
+class RangeLit : public Lit
+{
+public:
+	RangeLit(const Loc &loc, VarTerm *var, Term *a, Term *b);
+	void normalize(Grounder *g, Expander *expander);
+	bool fact() const { return true; }
+	void accept(Printer *v);
+	void index(Grounder *g, Groundable *gr, VarSet &bound);
+	void visit(PrgVisitor *visitor);
+	bool match(Grounder *grounder);
+	void print(Storage *sto, std::ostream &out) const;
+	Lit *clone() const;
+	~RangeLit();
+private:
+	clone_ptr<VarTerm> var_;
+	clone_ptr<Term>    a_;
+	clone_ptr<Term>    b_;
+};
+

### 2010-07-02 libgringo/gringo/rangeliteral.h

@@ -1,57 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#ifndef RANGELITERAL_H
-#define RANGELITERAL_H
-
-#include <gringo/gringo.h>
-#include <gringo/literal.h>
-
-namespace gringo
-{
-	class RangeLiteral : public Literal
-	{
-	public:
-		RangeLiteral(Variable *var, Term *lower, Term *upper);
-		RangeLiteral(const RangeLiteral &r);
-		virtual SDGNode *createNode(SDG *dg, SDGNode *prev, DependencyAdd todo);
-		virtual void createNode(LDGBuilder *dg, bool head);
-		virtual void createNode(StatementChecker *dg, bool head, bool delayed);
-		virtual void getVars(VarSet &vars) const;
-		virtual bool checkO(LiteralVector &unsolved);
-		virtual void reset();
-		virtual void finish();
-		virtual bool solved();
-		virtual bool isFact(Grounder *g);
-		virtual void preprocess(Grounder *g, Expandable *e, bool head);
-		virtual IndexedDomain *createIndexedDomain(Grounder *g, VarSet &index);
-		virtual bool match(Grounder *g);
-		virtual NS_OUTPUT::Object *convert();
-		virtual Literal* clone() const;
-		virtual void print(const GlobalStorage *g, std::ostream &out) const;
-		virtual double heuristicValue();
-		virtual void addIncParam(Grounder *g, const Value &v);
-		virtual ~RangeLiteral();
-	protected:
-		Variable *var_;
-		Term *lower_;
-		Term *upper_;
-	};
-}
-
-#endif
-

### 2010-07-02 libgringo/gringo/stmdep.h

@@ -1,0 +1,132 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/prgvisitor.h>
+
+namespace StmDep
+{
+	/*
+	 * NOTE:
+	 * negative: a literal is negative if it is preceded by a not or inside a negative literal
+	 * aggregates: for the sake of simplicity all aggregates are considered to be negative
+	 * exception choice semantics: any aggregate in the head not preceeded by not is not negative
+	 */
+	class Node
+	{
+	public:
+		enum Type { DOM, POS, NEG };
+	public:
+		Node();
+		virtual Node *node(uint32_t i) = 0;
+		virtual Type type(uint32_t i) = 0;
+		virtual Node *next() = 0;
+		virtual bool root() = 0;
+		virtual void addToComponent(Grounder *g) { (void)g; }
+		virtual void print(Storage *sto, std::ostream &out) const = 0;
+		virtual const Loc &loc() const = 0;
+		void mark() { visited_ = 1; }
+		bool marked() const { return visited_ >= 1; }
+		void visited(uint32_t visited) { visited_ = visited + 1; }
+		uint32_t visited() const { return visited_ - 1; }
+		bool hasComponent() { return component_ > 0; }
+		void component(uint32_t component) { component_ = component + 1; }
+		uint32_t component() const { return component_ - 1; }
+	private:
+		uint32_t visited_;
+		uint32_t component_;
+	};
+
+	class StmNode;
+	class PredNode : public Node
+	{
+		friend class Builder;
+	private:
[…]

### 2010-07-02 libgringo/gringo/storage.h

@@ -1,0 +1,44 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+
+class Storage
+{
+public:
+	Storage(Output *output);
+	uint32_t index(const Func &f);
+	const Func &func(uint32_t i);
+	uint32_t index(const std::string &s);
+	const std::string &string(uint32_t i);
+	Domain *domain(uint32_t nameId, uint32_t arity);
+	const DomainMap &domains() const { return doms_; }
+	Output *output() const { return output_; }
+	virtual void print(std::ostream &out, uint32_t type, uint32_t index);
+	virtual int compare(uint32_t type, uint32_t a, uint32_t b);
+	~Storage();
+private:
+	index_set<std::string> strings_;
+	index_set<Func>        funcs_;
+	DomainMap              doms_;
+	Output                *output_;
+};
+
+inline void Storage::print(std::ostream &, uint32_t, uint32_t) { assert(false); }
+inline int Storage::compare(uint32_t, uint32_t, uint32_t) { assert(false); return 0; }

### 2010-07-02 libgringo/gringo/sumaggrlit.h

@@ -1,0 +1,71 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/aggrlit.h>
+
+//! Sum or Count Aggregate.
+/** Examples:
+  * <pre> 1 [ q(X) : p(X) ] 2
+  * X \#count{ q(1), q(2), not p }</pre>
+  */
+class SumAggrLit : public AggrLit
+{
+public:
+	class Printer : public AggrLit::Printer
+	{
+	public:
+		virtual void begin(bool head, bool sign, bool count) = 0;
+		virtual void lower(int32_t l) = 0;
+		virtual void upper(int32_t u) = 0;
+		virtual void end() = 0;
+		virtual ~Printer() { }
+	};
+public:
+	/** \param count whether it is a count aggregate (false for sum aggregate).
+	  */
+	SumAggrLit(const Loc &loc, CondLitVec &conds, bool count);
+
+	bool match(Grounder *grounder);
+	boost::tuple<int32_t, int32_t, int32_t> matchAssign(Grounder *grounder);
+	void setValue(int32_t value) { lowerBound_ = upperBound_ = value; }
+	void index(Grounder *g, Groundable *gr, VarSet &bound);
+	void accept(::Printer *v);
+	Lit *clone() const;
+	void print(Storage *sto, std::ostream &out) const;
+	tribool accumulate(Grounder *g, const Val &weight, Lit &lit) throw(const Val*);
+private:
+	//! whether all evaluated literals are facts
+	bool    factOnly_;
+	//! whether the literal is a #count aggregate
+	bool	count_;
+	//! sum of evaluated weight literals that are facts
+	int32_t fixed_;
[…]

### 2010-07-02 libgringo/gringo/term.h

@@ -1,101 +1,71 @@
-// Copyright (c) 2008, Roland Kaminski
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
 //
-// This file is part of GrinGo.
+// This file is part of gringo.
 //
-// GrinGo is free software: you can redistribute it and/or modify
+// gringo is free software: you can redistribute it and/or modify
 // it under the terms of the GNU General Public License as published by
 // the Free Software Foundation, either version 3 of the License, or
 // (at your option) any later version.
 //
-// GrinGo is distributed in the hope that it will be useful,
+// gringo is distributed in the hope that it will be useful,
 // but WITHOUT ANY WARRANTY; without even the implied warranty of
 // MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
 // GNU General Public License for more details.
 //
 // You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
 
-#ifndef TERM_H
-#define TERM_H
+#pragma once
 
 #include <gringo/gringo.h>
-#include <gringo/printable.h>
+#include <gringo/locateable.h>
 
-namespace gringo
+class Term : public Locateable
 {
-	/**
-	 * \brief Base class of all Terms
-	 *
-	 * Terms are variables, strings or functions. They are used by most literals. 
-	 * During grounding every term will be evaluated to a value.
-	 */
-	class Term : public Printable
+public:
+	typedef std::pair<Term*, TermPtrVec*> Split;
+	class Ref
 	{
 	public:
-		/**
-		 * \brief Constructor - does nothing
-		 */
-		Term();
-		/**
-		 * \brief Retrieve all Variables in term
-		 * \param vars stores the result
-		 */
-		virtual void getVars(VarSet &vars) const = 0;
-		/**
-		 * \brief Return true if the Term is complex
-		 *
-		 * Everything except variables, ids, strings has to be complex
-		 * \return Returns true if the term is complex
-		 */
-		virtual bool isComplex() = 0;
-		/**
-		 * \brief Returns the Value of the term under the current substitution
-		 * \return The value
-		 */
-		virtual Value getValue(Grounder *g) = 0;
-		/**
[…]

### 2010-07-02 libgringo/gringo/timesaggregate.h

@@ -1,39 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#ifndef TIMESAGGREGATE_H
-#define TIMESAGGREGATE_H
-
-#include <gringo/gringo.h>
-#include <gringo/aggregateliteral.h>
-
-namespace gringo
-{
-	class TimesAggregate : public AggregateLiteral
-	{
-	public:
-		TimesAggregate(ConditionalLiteralVector *literals);
-		TimesAggregate(const TimesAggregate &a);
-		virtual Literal *clone() const;
-		virtual void doMatch(Grounder *g);
-		virtual void print(const GlobalStorage *g, std::ostream &out) const;
-		virtual NS_OUTPUT::Object *convert();
-		virtual ~TimesAggregate();
-	};
-}
-
-#endif

### 2010-07-02 libgringo/gringo/variable.h

@@ -1,50 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#ifndef VARIABLE_H
-#define VARIABLE_H
-
-#include <gringo/gringo.h>
-#include <gringo/term.h>
-#include <gringo/value.h>
-
-namespace gringo
-{
-	class Variable : public Term
-	{
-	public:
-		Variable(Grounder *g, int id);
-		Variable(const Variable &c);
-		virtual void print(const GlobalStorage *g, std::ostream &out) const;
-		virtual void getVars(VarSet &vars) const;
-		virtual bool isComplex();
-		virtual void preprocess(Literal *l, Term *&p, Grounder *g, Expandable *e);
-		virtual Value getConstValue(Grounder *g);
-		virtual Value getValue(Grounder *g);
-		int getUID();
-		virtual Term* clone() const;
-		bool unify(const GlobalStorage *g, const Value& t, const VarVector& vars, ValueVector& vals) const;
-		virtual void addIncParam(Grounder *g, Term *&p, const Value &v);
- 		virtual ~Variable();
-	protected:
-		int id_;
-		int uid_;
-	};
-}
-
-#endif
-

### 2010-07-02 libgringo/gringo/varterm.h

@@ -1,0 +1,52 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#pragma once
+
+#include <gringo/gringo.h>
+#include <gringo/term.h>
+
+class VarTerm : public Term
+{
+public:
+	VarTerm(const Loc &loc);
+	VarTerm(const Loc &loc, uint32_t nameId);
+	bool anonymous() const;
+	uint32_t nameId() const { return nameId_; }
+	uint32_t index() const { return index_; }
+	uint32_t level() const { return level_; }
+	void normalize(Lit *parent, const Ref &ref, Grounder *grounder, Expander *expander, bool unify) { (void)parent; (void)ref; (void)grounder; (void)expander; (void)unify; }
+	void index(uint32_t index, uint32_t level, bool local) { level_ = level; index_ = index; local_ = local; }
+	Val val(Grounder *grounder) const;
+	bool constant() const { return false; }
+	bool unify(Grounder *grounder, const Val &v, int binder) const;
+	void vars(VarSet &vars) const;
+	void visit(PrgVisitor *visitor, bool bind);
+	void print(Storage *sto, std::ostream &out) const;
+	Term *clone() const;
+private:
+	uint32_t nameId_;
+	uint32_t index_;
+	uint32_t level_ : 31;
+	uint32_t local_ : 1;
+};
+
+inline VarTerm* new_clone(const VarTerm& a)
+{
+	return static_cast<VarTerm*>(a.clone());
+}
+

### 2010-07-02 libgringo/src/argterm.cpp

@@ -1,0 +1,53 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include <gringo/argterm.h>
+
+ArgTerm::ArgTerm(const Loc &loc, Term *term, TermPtrVec &terms)
+	: Term(loc)
+	, term_(term)
+	, terms_(terms.release())
+{
+}
+
+void ArgTerm::print(Storage *sto, std::ostream &out) const
+{
+	term_->print(sto, out);
+	out << ";;";
+	bool comma = false;
+	foreach(const Term &term, terms_)
+	{
+		if(comma) out << ",";
+		else comma = true;
+		term.print(sto, out);
+	}
+}
+
+Term::Split ArgTerm::split() 
+{ 
+	return Split(term_.release(), &terms_);
+}
+
+Term *ArgTerm::clone() const
+{
+	return new ArgTerm(*this);
+}
+
+ArgTerm::~ArgTerm()
+{
+}
+

### 2010-07-02 libgringo/src/assignmentliteral.cpp

@@ -1,179 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/assignmentliteral.h>
-#include <gringo/term.h>
-#include <gringo/variable.h>
-#include <gringo/value.h>
-#include <gringo/indexeddomain.h>
-#include <gringo/dlvgrounder.h>
-#include <gringo/grounder.h>
-#include <gringo/literaldependencygraph.h>
-#include <gringo/statementchecker.h>
-
-using namespace gringo;
-
-AssignmentLiteral::AssignmentLiteral(Variable *c, Term *t) : Literal(), c_(c), t_(t)
-{
-
-}
-
-SDGNode *AssignmentLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo)
-{
-	return 0;
-}
-
-void AssignmentLiteral::createNode(LDGBuilder *dg, bool head)
-{
-	assert(!head);
-	VarSet needed, provided;
-	t_->getVars(needed);
-	c_->getVars(provided);
-	dg->createNode(this, head, needed, provided);
-}
-
-void AssignmentLiteral::createNode(StatementChecker *dg, bool head, bool delayed)
-{
-	assert(!head && !delayed);
-	VarSet needed, provided;
-	c_->getVars(provided);
-	t_->getVars(needed);
-	dg->createNode(needed, provided);
-}
-
-void AssignmentLiteral::print(const GlobalStorage *g, std::ostream &out) const
-{
-	out << pp(g, c_) << " = " << pp(g, t_);
-}
-
[…]

### 2010-07-02 libgringo/src/constant.cpp

@@ -1,122 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/constant.h>
-#include <gringo/grounder.h>
-
-using namespace gringo;
-		
-Constant::Constant(const Value &v) : value_(v)
-{
-}
-
-void Constant::getVars(VarSet &vars) const
-{
-}
-
-bool Constant::isComplex()
-{
-	return false;
-}
-
-void Constant::print(const GlobalStorage *g, std::ostream &out) const
-{
-	value_.print(g, out);
-}
-
-Value Constant::getConstValue(Grounder *g)
-{
-	if(value_.type_ == Value::STRING)
-		return g->getConstValue(value_.uid_);
-	else
-		return value_;
-}
-
-Value Constant::getValue(Grounder *g)
-{
-	return value_;
-}
-
-Constant::~Constant()
-{
-}
-
-void Constant::preprocess(Literal *l, Term *&p, Grounder *g, Expandable *e)
-{
-	if(value_.type_ == Value::STRING)
-		value_ = g->getConstValue(value_.uid_);
-}
-
-Constant::Constant(const Constant &c) :  value_(c.value_)
-{
-}
-
-Term* Constant::clone() const
-{
-	return new Constant(*this);
-}
-
-namespace
-{
-	class IncTerm : public Term
-	{
-	public:
-		IncTerm() {}
-		IncTerm(const IncTerm &c) {}
-		Term* clone() const
-		{
-			return new IncTerm(*this);
-		}
-
-		bool unify(const GlobalStorage *g, const Value& t, const VarVector& vars, ValueVector& subst) const 
-		{
[…]

### 2010-07-02 libgringo/src/funcsymbol.cpp

@@ -1,86 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/funcsymbol.h>
-#include <gringo/value.h>
-#include <gringo/globalstorage.h>
-
-using namespace gringo;
-
-FuncSymbol::FuncSymbol(int s, const ValueVector& args) : name_(s), args_(args)
-{
-}
-
-FuncSymbol::~FuncSymbol()
-{
-}
-
-bool FuncSymbol::operator==(const FuncSymbol& a) const
-{
-	if (name_ != a.name_)
-		return false;
-	if (args_.size() != a.args_.size())
-		return false;
-
-	for (ValueVector::const_iterator i = args_.begin(), j = a.args_.begin(); i != args_.end(); ++i,++j)
-		if (!i->equal(*j))
-			return false;
-	return true;
-}
-
-size_t FuncSymbol::getHash() const
-{
-	size_t hash = 0;
-	size_t x = 0;
-
-	Value::VectorHash t;
-	hash = t(args_);
-
-	hash = (hash << 4) + size_t(name_);
-	if((x = hash & 0xF0000000L) != 0)
-	{
-		hash ^= (x >> 24);
-	}
-	hash &= ~x;
-	return hash;
-}
-
-void FuncSymbol::print(const GlobalStorage *g, std::ostream& out) const
-{
-	out << *g->getString(name_) << "(";
-	bool comma = false;
-	for (ValueVector::const_iterator i = args_.begin(); i != args_.end(); i++)
-	{
-		if(!comma)
-			comma = true;
-		else
-			out << ",";
-		i->print(g, out);
-	}
-	out << ")";
-}
-
-int FuncSymbol::getName() const
-{
-	return name_;
-}
-
-const ValueVector& FuncSymbol::getValues() const
-{
-	return args_;
-}
-

### 2010-07-02 libgringo/src/globalstorage.cpp

@@ -1,99 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/globalstorage.h>
-#include <gringo/domain.h>
-
-using namespace gringo;
-
-GlobalStorage::GlobalStorage()
-{
-}
-
-Domain *GlobalStorage::getDomain(int uid) const
-{
-	return domains_[uid];
-}
-
-DomainVector *GlobalStorage::getDomains() const
-{
-	return const_cast<DomainVector*>(&domains_);
-}
-
-GlobalStorage::~GlobalStorage()
-{
-	stringHash_.clear();
-	funcHash_.clear();
-	for(StringVector::iterator it = strings_.begin(); it != strings_.end(); it++)
-		delete *it;
-	for(FuncSymbolVector::iterator it = funcs_.begin(); it != funcs_.end(); it++)
-		delete *it;
-	for(DomainVector::iterator it = domains_.begin(); it != domains_.end(); it++)
-		delete *it;
-}
-
-const std::string *GlobalStorage::getString(int uid) const
-{
-	return strings_[uid];
-}
-
-const FuncSymbol  *GlobalStorage::getFuncSymbol(int uid) const
-{
-	return funcs_[uid];
-}
-
-int GlobalStorage::createString(const std::string &s2)
-{
-	return createString(new std::string(s2));
-}
-
-int GlobalStorage::createString(std::string *s)
-{
-	std::pair<StringHash::iterator, bool> res = stringHash_.insert(std::make_pair(s, strings_.size()));
-	if(!res.second)
-		delete s;
-	else
-		strings_.push_back(s);
-	return res.first->second;
-}
-
-int GlobalStorage::createFuncSymbol(FuncSymbol* fn)
-{
[…]

### 2010-07-02 libgringo/src/literaldependencygraph.cpp

@@ -1,265 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/literaldependencygraph.h>
-#include <gringo/literal.h>
-
-using namespace gringo;
-
-///////////////////////////////////// LiteralNode ///////////////////////////////////////////
-
-LDG::LiteralNode::LiteralNode(Literal *l, bool head) : l_(l), done_(0), in_(0), head_(head)
-{
-}
-
-///////////////////////////////////// VarNode ///////////////////////////////////////////
-
-LDG::VarNode::VarNode(int var) : done_(false), var_(var)
-{
-}
-
-///////////////////////////////////// Literaldata ///////////////////////////////////////////
-
-LDG::LiteralData::LiteralData(LiteralNode *n) : n_(n)
-{
-}
-
-///////////////////////////////////// LiteralDependencyGraph ///////////////////////////////////////////
-
-LDG::LDG() : sorted_(false)
-{
-}
-
-bool LDG::hasVarNodes() const
-{
-	return varNodes_.size() > 0;
-}
-
-const VarVector &LDG::getGlobalVars() const
-{
-	return globalVars_;
-}
-
-const VarVector &LDG::getParentVars() const
-{
-	return parentVars_;
-}
-
-void LDG::start(LiteralList &list)
-{
-	assert(globalVars_.size() == varNodes_.size());
[…]

### 2010-07-02 libgringo/src/literalstatement.cpp

@@ -1,124 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/literalstatement.h>
-#include <gringo/statementdependencygraph.h>
-#include <gringo/literaldependencygraph.h>
-#include <gringo/statementchecker.h>
-#include <gringo/conditionalliteral.h>
-#include <gringo/output.h>
-#include <gringo/grounder.h>
-#include <gringo/evaluator.h>
-#include <gringo/gringoexception.h>
-
-using namespace gringo;
-		
-LiteralStatement::LiteralStatement(Literal *lit, bool preserveOrder) : lit_(lit), preserveOrder_(preserveOrder)
-{
-}
-
-void LiteralStatement::buildDepGraph(SDG *dg)
-{
-	// the order of optimize statements is important
-	SDGNode *prev = dg->createStatementNode(this, preserveOrder_);
-	lit_->createNode(dg, prev, Literal::ADD_NOTHING);
-}
-
-void LiteralStatement::getVars(VarSet &vars) const
-{
-	// noone needs to know the vars in this statement
-	//for(ConditionalLiteralVector::iterator it = literals_.begin(); it != literals_.end(); it++)
-	//	(*it)->getVars(vars, VARS_ALL);
-}
-
-bool LiteralStatement::checkO(LiteralVector &unsolved)
-{
-	return lit_->checkO(unsolved);
-}
-
-bool LiteralStatement::check(Grounder *g, VarVector &free)
-{
-	StatementChecker s;
-	lit_->createNode(&s, false, false);
-
-	if(s.check())
-		return true;
-	else
-	{
-		s.getFreeVars(free);
-		return false;
-	}
-}
-
[…]

### 2010-07-02 libgringo/src/lparseconverter.cpp

@@ -1,93 +1,194 @@
-// Copyright (c) 2008, Roland Kaminski
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
 //
-// This file is part of GrinGo.
+// This file is part of gringo.
 //
-// GrinGo is free software: you can redistribute it and/or modify
+// gringo is free software: you can redistribute it and/or modify
 // it under the terms of the GNU General Public License as published by
 // the Free Software Foundation, either version 3 of the License, or
 // (at your option) any later version.
 //
-// GrinGo is distributed in the hope that it will be useful,
+// gringo is distributed in the hope that it will be useful,
 // but WITHOUT ANY WARRANTY; without even the implied warranty of
 // MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
 // GNU General Public License for more details.
 //
 // You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
 
-#include <gringo/lparseconverter.h>
-#include "lparseconverter_impl.h"
-#include <gringo/plainlparselexer.h>
-#include <gringo/grounder.h>
-#include <gringo/funcsymbol.h>
+#include <gringo/lparseoutput.h>
+#include <gringo/predlitrep.h>
+#include <gringo/domain.h>
+#include <gringo/storage.h>
+#include <gringo/rule.h>
+#include <gringo/sumaggrlit.h>
+#include <gringo/junctionaggrlit.h>
 
-using namespace gringo;
-
-void *lparseconverterAlloc(void *(*mallocProc)(size_t));
-void lparseconverterFree(void *p, void (*freeProc)(void*));
-void lparseconverter(void *yyp, int yymajor, std::string* yyminor, LparseConverter *pParser);
-
-LparseConverter::LparseConverter(std::istream* in) : GrinGoParser(), output_(0)
+namespace lparseconverter_impl
 {
-	lexer_  = new PlainLparseLexer();
-        pParser = lparseconverterAlloc (malloc);
-	streams_.push_back(in);
+
[…]

### 2010-07-02 libgringo/src/normalrule.cpp

@@ -1,562 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/normalrule.h>
-#include <gringo/literal.h>
-#include <gringo/predicateliteral.h>
-#include <gringo/statementdependencygraph.h>
-#include <gringo/literaldependencygraph.h>
-#include <gringo/statementchecker.h>
-#include <gringo/grounder.h>
-#include <gringo/output.h>
-#include <gringo/dlvgrounder.h>
-#include <gringo/evaluator.h>
-#include <gringo/output.h>
-#include <gringo/gringoexception.h>
-#include <gringo/indexeddomain.h>
-#include <gringo/variable.h>
-
-using namespace gringo;
-
-NormalRule::NormalRule(Literal *head, LiteralVector *body) : Statement(), head_(head), body_(body), grounder_(0), ground_(1), base_(0), last_(0), lambda_(0), delta_(0), isGround_(0)
-{
-}
-
-void NormalRule::buildDepGraph(SDG *dg)
-{
-	SDGNode *rn = dg->createStatementNode(this);
-	if(head_)
-		head_->createNode(dg, rn, Literal::ADD_HEAD_DEP);
-	else
-	{
-		// we have an integrity constraint " :- B", which can be modeled by "r :- B, not r",
-		// where r is a unique identifier
-		// this means the rule depends negativly on itself
-		rn->addDependency(rn, true);
-	}
-	if(body_)
-	{
[…]

### 2010-07-02 libgringo/src/optimize.cpp

@@ -1,0 +1,259 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include <gringo/optimize.h>
+#include <gringo/lit.h>
+#include <gringo/term.h>
+#include <gringo/output.h>
+#include <gringo/predlit.h>
+#include <gringo/grounder.h>
+#include <gringo/prgvisitor.h>
+#include <gringo/litdep.h>
+#include <gringo/exceptions.h>
+#include <gringo/instantiator.h>
+
+class Optimize::PrioLit : public Lit
+{
+public:
+	PrioLit(const Loc &loc, Term *weight, Term *prio) : Lit(loc), weight_(weight), prio_(prio) { }
+	bool fact() const { return true; }
+	bool match(Grounder *grounder) { (void)grounder; return true; }
+	void addDomain(bool fact) { (void)fact; assert(false); }
+	void index(Grounder *g, Groundable *gr, VarSet &bound) { (void)g; (void)gr; (void)bound; }
+	bool edbFact() const { return false; }
+	void normalize(Grounder *g, Expander *expander) { (void)g; (void)expander; assert(false); }
+	void visit(PrgVisitor *visitor)
+	{
+		visitor->visit(weight_.get(), false);
+		visitor->visit(prio_.get(), false);
+	}
+	void print(Storage *sto, std::ostream &out) const
+	{
+		weight_->print(sto, out);
+		out << "@";
+		prio_->print(sto, out);
+	}
[…]

### 2010-07-02 libgringo/src/output/reifiedoutput_impl.cpp

@@ -1,0 +1,386 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include "output/reifiedoutput_impl.h"
+#include <gringo/plainoutput.h>
+#include <gringo/domain.h>
+#include <gringo/predlit.h>
+#include <gringo/func.h>
+#include <gringo/storage.h>
+
+namespace reifiedoutput_impl
+{
+	void DisplayPrinter::print(PredLitRep *l)
+	{
+		out() << (show() ? "show" : "hide") << "(";
+		output_->val(output_->symbol(l)).print(output_->storage(), out());
+		out() << ").\n";
+	}
+
+	void ExternalPrinter::print(PredLitRep *l)
+	{
+		out() << "external(";
+		output_->val(output_->symbol(l)).print(output_->storage(), out());
+		out() << ").\n";
+	}
+
+	void RulePrinter::begin()
+	{
+		output_->startSet();
+		output_->pushDep();
+	}
+
+	void RulePrinter::endHead()
+	{
+		output_->startSet();
+		output_->pushDep(2);
+	}
+
+	void RulePrinter::print(PredLitRep *l)
+	{
+		size_t symbol = output_->symbol(l);
+		if(!l->sign()) { output_->addDep(symbol); }
+		output_->addToSet(symbol);
+	}
+
+	void RulePrinter::end()
+	{
+		uint32_t body = output_->addSet();
+		output_->popSet();
+		const ReifiedOutput::Set &head = output_->getSet();
[…]

### 2010-07-02 libgringo/src/rangeliteral.cpp

@@ -1,200 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/rangeliteral.h>
-#include <gringo/variable.h>
-#include <gringo/indexeddomain.h>
-#include <gringo/dlvgrounder.h>
-#include <gringo/grounder.h>
-#include <gringo/literaldependencygraph.h>
-#include <gringo/statementchecker.h>
-
-using namespace gringo;
-
-RangeLiteral::RangeLiteral(Variable *var, Term *lower, Term *upper) : Literal(), var_(var), lower_(lower), upper_(upper)
-{
-}
-
-SDGNode *RangeLiteral::createNode(SDG *dg, SDGNode *prev, DependencyAdd todo)
-{
-	return 0;
-}
-
-void RangeLiteral::createNode(LDGBuilder *dg, bool head)
-{
-	assert(!head);
-	VarSet needed, provided;
-	upper_->getVars(needed);
-	lower_->getVars(needed);
-	var_->getVars(provided);
-	dg->createNode(this, head, needed, provided);
-}
-
-void RangeLiteral::createNode(StatementChecker *dg, bool head, bool delayed)
-{
-	assert(!head && !delayed);
-	VarSet needed, provided;
-	var_->getVars(provided);
-	lower_->getVars(needed);
-	upper_->getVars(needed);
-	dg->createNode(needed, provided);
-}
-
-void RangeLiteral::print(const GlobalStorage *g, std::ostream &out) const
-{
[…]

### 2010-07-02 libgringo/src/rule.cpp

@@ -1,0 +1,207 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include <gringo/rule.h>
+#include <gringo/lit.h>
+#include <gringo/instantiator.h>
+#include <gringo/index.h>
+#include <gringo/printer.h>
+#include <gringo/litdep.h>
+#include <gringo/grounder.h>
+#include <gringo/output.h>
+
+Rule::Rule(const Loc &loc, Lit *head, LitPtrVec &body)
+	: Statement(loc)
+	, head_(head)
+	, body_(body.release())
+	, grounded_(false)
+{
+	if(head) head->head(true);
+}
+
+namespace
+{
+	class RuleHeadExpander : public Expander
+	{
+	public:
+		RuleHeadExpander(const Loc &loc, LitPtrVec &body, Grounder *g);
+		void expand(Lit *lit, Type type);
+	private:
+		const Loc &loc_;
+		LitPtrVec &body_;
+		Grounder  *g_;
+	};
+
+	class RuleBodyExpander : public Expander
+	{
+	public:
+		RuleBodyExpander(LitPtrVec &body);
+		void expand(Lit *lit, Type type);
+	private:
+		LitPtrVec &body_;
+	};
+
+	RuleHeadExpander::RuleHeadExpander(const Loc &loc, LitPtrVec &body, Grounder *g)
+		: loc_(loc)
+		, body_(body)
+		, g_(g)
+	{
+	}
+
+	void RuleHeadExpander::expand(Lit *lit, Expander::Type type)
+	{
+		switch(type)
+		{
+			case RANGE:
[…]

### 2010-07-02 libgringo/src/storage.cpp

@@ -1,0 +1,65 @@
+// Copyright (c) 2009, Roland Kaminski <kaminski@cs.uni-potsdam.de>
+//
+// This file is part of gringo.
+//
+// gringo is free software: you can redistribute it and/or modify
+// it under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+//
+// gringo is distributed in the hope that it will be useful,
+// but WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+// GNU General Public License for more details.
+//
+// You should have received a copy of the GNU General Public License
+// along with gringo.  If not, see <http://www.gnu.org/licenses/>.
+
+#include <gringo/storage.h>
+#include <gringo/domain.h>
+#include <gringo/output.h>
+#include <gringo/func.h>
+
+Storage::Storage(Output *output) 
+	: output_(output)
+{
+	output_->storage(this);
+}
+
+uint32_t Storage::index(const Func &f)
+{
+	return funcs_.insert(f);
+}
+
+const Func &Storage::func(uint32_t i)
+{
+	return funcs_.get(i);
+}
+
+uint32_t Storage::index(const std::string &s)
+{
+	return strings_.insert(s);
+}
+
+const std::string &Storage::string(uint32_t i)
+{
+	return strings_.get(i);
+}
+
+Domain *Storage::domain(uint32_t nameId, uint32_t arity)
+{
+	Signature sig(nameId, arity);
+	DomainMap::iterator i = doms_.find(sig);
+	if(i == doms_.end())
+	{
+		Domain *dom = new Domain(nameId, arity, doms_.size());
+		output_->addDomain(dom);
+		return doms_.insert(sig, dom).first->second;
+	}
+	else return i->second;
+}
+
+Storage::~Storage()
+{
+}
+

### 2010-07-02 libgringo/src/term.cpp

@@ -1,29 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/term.h>
-
-using namespace gringo;
-
-Term::Term()
-{
-}
-
-Term::~Term()
-{
-}
-

### 2010-07-02 libgringo/src/timesaggregate.cpp

@@ -1,173 +1,0 @@
-// Copyright (c) 2008, Roland Kaminski
-//
-// This file is part of GrinGo.
-//
-// GrinGo is free software: you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation, either version 3 of the License, or
-// (at your option) any later version.
-//
-// GrinGo is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-//
-// You should have received a copy of the GNU General Public License
-// along with GrinGo.  If not, see <http://www.gnu.org/licenses/>.
-
-#include <gringo/timesaggregate.h>
-#include <gringo/conditionalliteral.h>
-#include <gringo/term.h>
-#include <gringo/value.h>
-#include <gringo/output.h>
-
-
-using namespace gringo;
-
-namespace
-{
-	// very simple class to prevent integers from overflowing
-	// in theory templates could be used here
-	// NOTE: INT_MIN+1 is already treated as overflow
-	class SafeInt
-	{
-	public:
-		SafeInt(int v) : value(v) { }
-		operator int() const { return value; }
-		SafeInt operator*(const int b) const
-		{
-			if(value == 0 || b == 0)
-				return 0;
-			if(value < 0 && b < 0)
-				return (value < INT_MAX / b) ? INT_MAX : (value * b);
-			if(value < 0)
-				return (value < (-INT_MAX) / b) ? -INT_MAX : (value * b);
-			else if(b < 0)
-				return (b < (-INT_MAX) / value) ? -INT_MAX : (value * b);
-			else
-				return (value > INT_MAX / b) ? INT_MAX : (value * b);
-		}
-		bool overflow() const
-		{
-			if(value >= 0)
-				return value >= INT_MAX;
-			else
-				return value <= -INT_MAX;
-		}
-		int value;
-	};
-	SafeInt operator *(const SafeInt &a, const SafeInt &b)
-	{
-		return a * b.value;
-	}
[…]

### 2010-07-02 libprogram_opts/program_opts/app_options.h

@@ -1,81 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#ifndef APP_OPTIONS_H_INCLUDED
-#define APP_OPTIONS_H_INCLUDED
-
-#ifdef _MSC_VER
-#pragma warning (disable : 4200) // nonstandard extension used : zero-sized array
-#pragma once
-#endif
-
-#include <string>
-#include <utility>
-#include "program_options.h"
-
-typedef std::vector<std::string> StringSeq;
-struct Messages {
-	std::string error;
-	StringSeq   warning;
-	void clear() { error.clear(); warning.clear(); }
-};
-
-/////////////////////////////////////////////////////////////////////////////////////////
-// generic options
-/////////////////////////////////////////////////////////////////////////////////////////
-// Extends group "Basic Options"
-struct GenericOptions {
-	GenericOptions();
-	void initOptions(ProgramOptions::OptionGroup& root, ProgramOptions::OptionGroup& hidden);
-	void addDefaults(std::string& def);
-	StringSeq  input;    // positional options (list of files)
-	int        verbose;  // verbosity level, default:  0
-	bool       help;     // print help and exit
-	bool       version;  // print version and exit
-	bool       validateOptions(ProgramOptions::OptionValues&, Messages&) { return true; }
-};
-
-/////////////////////////////////////////////////////////////////////////////////////////
-// Interface for parsing application options
-/////////////////////////////////////////////////////////////////////////////////////////
-class AppOptions {
-public:
-	AppOptions() {}
[…]

### 2010-07-02 libprogram_opts/program_opts/errors.h

@@ -1,91 +1,0 @@
-//
-//  ProgramOptions
-//  (C) Copyright Benjamin Kaufmann, 2004 - 2005
-//  Permission to copy, use, modify, sell and distribute this software is 
-//  granted provided this copyright notice appears in all copies. 
-//  This software is provided "as is" without express or implied warranty, 
-//  and with no claim as to its suitability for any purpose.
-//
-//  ProgramOptions is a scaled-down version of boost::program_options
-//  see: http://boost-sandbox.sourceforge.net/program_options/html/
-// 
-#ifndef PROGRAM_OPTIONS_ERRORS_H_INCLUDED
-#define PROGRAM_OPTIONS_ERRORS_H_INCLUDED
-#include <stdexcept>
-namespace ProgramOptions
-{
-
-	//! thrown when a value has an unexpected type
-	class BadValue : public std::logic_error
-	{
-	public:
-		BadValue(const std::string& msg)
-			: std::logic_error(msg)
-		{}
-		~BadValue() throw () {}
-		std::string getOptionName() const
-		{
-			return optionName_;
-		}
-	private:
-		std::string optionName_;
-	};
-
-	//! thrown when an option has an illegal name
-	class BadOptionName : public std::logic_error
-	{
-	public:
-		BadOptionName(const std::string& name)
-			: std::logic_error(std::string("bad option name: ").append(name))
-		{}
-		~BadOptionName() throw () {}
-	};
-
-	//! thrown when more than one option in a group has the same name
-	class DuplicateOption : public std::logic_error
-	{
-	public:
-		DuplicateOption(const std::string& name, const std::string& grp)
-			: std::logic_error(grp + " - " + "Duplicate Option: '" + name + "'")
-			, name_(name)
-			, grpDesc_(grp)
-		{}
-		const std::string& getOptionName() const {return name_;}
-		const std::string& getGroupName() const {return grpDesc_;}
-		~DuplicateOption() throw() {}
-	private:
-		const std::string name_;
-		const std::string grpDesc_;
-	};
-
-	//! thrown when an unknown option is requested
-	class UnknownOption : public std::logic_error
-	{
-	public:
-		UnknownOption(const std::string& name)
-			: std::logic_error(std::string("unknown option: ").append(name))
-		{}
-		~UnknownOption() throw() {}
-	};
-
-	//! thrown when there's ambiguity amoung several possible options.
-	class AmbiguousOption : public std::logic_error
-	{
-	public:
[…]

### 2010-07-02 libprogram_opts/src/app_options.cpp

@@ -1,89 +1,0 @@
-// 
-// Copyright (c) 2006-2007, Benjamin Kaufmann
-// 
-// This file is part of Clasp. See http://www.cs.uni-potsdam.de/clasp/ 
-// 
-// Clasp is free software; you can redistribute it and/or modify
-// it under the terms of the GNU General Public License as published by
-// the Free Software Foundation; either version 2 of the License, or
-// (at your option) any later version.
-// 
-// Clasp is distributed in the hope that it will be useful,
-// but WITHOUT ANY WARRANTY; without even the implied warranty of
-// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-// GNU General Public License for more details.
-// 
-// You should have received a copy of the GNU General Public License
-// along with Clasp; if not, write to the Free Software
-// Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-//
-#include <program_opts/app_options.h>
-#include <program_opts/value.h>
-#include <sstream>
-#include <algorithm>
-#include <cctype>
-
-using namespace ProgramOptions;
-using namespace std;
-
-/////////////////////////////////////////////////////////////////////////////////////////
-// Generic Options - independent of concrete system
-/////////////////////////////////////////////////////////////////////////////////////////
-GenericOptions::GenericOptions() 
-	: verbose(0)
-	, help(false)
-	, version(false) {}
-
-namespace {
-bool mapVLevel(const std::string& s, int& level) {
-	if (s.empty()) return !!(level = 2);
-	else return ProgramOptions::parseValue(s, level, 1);
-}
-}
-
-void GenericOptions::initOptions(OptionGroup& root, OptionGroup& hidden) {
-	OptionGroup basic("Basic Options");
-	basic.addOptions()
-		("help,h"   , bool_switch(&help),    "Print help information and exit")
-		("version,v", bool_switch(&version), "Print version information and exit")    
-		("verbose,V", value<int>(&verbose)->setImplicit()->parser(mapVLevel),   "Verbosity level", "<n>")
-	;
-	root.addOptions(basic, true);
-	hidden.addOptions()
-		("file,f", value<StringSeq>(&input)->setComposing(), "Input files\n")
-	;
-}
-
-void GenericOptions::addDefaults(std::string&) {}
-
-/////////////////////////////////////////////////////////////////////////////////////////
-// Parsing & Validation of command line
[…]

### 2010-07-02 libprogram_opts/src/program_options.cpp

@@ -1,759 +1,0 @@
-//
-//  ProgramOptions
-//  (C) Copyright Benjamin Kaufmann, 2004 - 2005
-//  Permission to copy, use, modify, sell and distribute this software is 
-//  granted provided this copyright notice appears in all copies. 
-//  This software is provided "as is" without express or implied warranty, 
-//  and with no claim as to its suitability for any purpose.
-//
-//  ProgramOptions is a scaled-down version of boost::program_options
-//  see: http://boost-sandbox.sourceforge.net/program_options/html/
-// 
-
-#ifdef _MSC_VER
-#pragma warning (disable : 4786)
-#pragma warning (disable : 4503)
-#endif
-#include <program_opts/program_options.h>
-#include <program_opts/errors.h>
-#include <cassert>
-#include <cstring>
-#include <iostream>
-#include <climits>
-#include <sstream>
-using namespace std;
-
-namespace ProgramOptions {
-
-	namespace {
-
-		void defaultFormat(std::ostream& os, const ProgramOptions::Option& o)
-		{
-			stringstream temp;
-			temp << "  ";
-			temp << "--" << o.longName();
-			if (o.getValue()->isImplicit() && !o.argDescription().empty()) {
-				temp << "[=" << o.argDescription() << "]";
-			}
-			if (!o.shortName().empty()) {
-				temp << ",-" << o.shortName();
-			}
-			if (!o.getValue()->isImplicit())
-				temp << (o.shortName().empty()?'=':' ') << o.argDescription();
-
-			os << temp.str();
-			if (!o.description().empty())
-			{
-				if (temp.str().size() < size_t(22)) {
-					for(size_t pad = 22 - temp.str().size(); pad > 0; --pad) {
-						os.put(' ');
-					}
-				}
-				if (temp.str().size() <= size_t(22)) os << " ";
-
-				os << ": " << o.description();
-			}
-		}
-
-	}
-	///////////////////////////////////////////////////////////////////////////////
-	// class ValueBase
-	///////////////////////////////////////////////////////////////////////////////
-	ValueBase::ValueBase() {}
-	ValueBase::~ValueBase() {}
-
-	///////////////////////////////////////////////////////////////////////////////
-	// class Option
-	///////////////////////////////////////////////////////////////////////////////
[…]

### 2010-07-09 doc/guide/aggregates.tex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3171 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-09 doc/documentation/language.tex
@@
-\section{Input Languages}\label{sec:language}
+\com{Rewrote much here}
 
-This section provides an overview of the input languages of
-grounder \gringo, combined grounder and solver \clingo,
-incremental grounder and solver \iclingo, and of solver \clasp.
-The joint input language of \gringo\ and \clingo\ is detailed in
-Section~\ref{subsec:lang:gringo}.
-It is extended by \iclingo\ with a few directives described in Section~\ref{subsec:lang:iclingo}.
-Finally, Section~\ref{subsec:lang:clasp} is dedicated to the inputs
-handled by \clasp.
-
-\subsection{Input Language of \gringo\ and \clingo}\label{subsec:lang:gringo}
-
-The tool \gringo~\cite{gescth07a} is a grounder capable of translating
-logic programs provided by users into equivalent ground programs.
-The output of \gringo\ can be piped into solver \clasp~\cite{gekanesc07b},
-which then computes answer sets.
-System \clingo\ internally couples \gringo\ and \clasp, thus,
-it takes care of both grounding and solving.
-% The input of \gringo\ and \clingo\ is the same,
-% but 
-In contrast to \gringo\ outputting ground programs, 
-\clingo\ returns answer sets.
-Both \gringo\ and \clingo\ can handle level-restricted input programs
-(cf.\ Section~\ref{subsec:lambda}),
-usually specified in one or more text files whose names are passed via the command line
-in an invocation of either \gringo\ or \clingo.
-We below provide a description of constructs belonging to the input language of \gringo\ and \clingo.
-
-\subsubsection{Normal Programs and Integrity Constraints}\label{subsec:gringo:normal}
-
-In the previous sections, we have already seen a number of normal programs.
-Now also considering integrity constraints, we get the following basic rule types:
-
-\begin{tabular}{rl@{}l}
-\textbf{Fact:} & $A_0$&\code{.}
-\\
-\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
-\\
-\textbf{Integrity Constraint:} & & $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
-\end{tabular}
-\\
-
-\noindent
-The head~$A_0$ of a fact or a rule is an \emph{atom} of the form
-$p\code{(}t_1\code{,}\dots\code{,}t_m\code{)}$,
-where $p$ is the name of some predicate, that is, a sequence of letters and digits
-starting with a lowercase letter,
-and any $t_i$ is a term.%
-\footnote{An atom~$p$ without arguments is simply a sequence of letters and digits
-  (starting with a lowercase letter).
-  For such an atom~$p$, parentheses after the name are skipped, e.g.,
-  \pred{p42X} could be an atom without arguments.
-  Also note that lowercase letters include ``\code{\mus},''
-  which can thus be used at any position within a predicate name.}
-A \emph{term} $t$ starting with an uppercase letter followed by
-a sequence of letters and digits (e.g., \var{X08x15}) is a variable name,
-and integers are written as sequences of digits possibly preceded by ``\code{-}.''
-In addition, a term can have the same syntactic structure as an atom
-(e.g., \code{\pred{p}(\const{a},\const{1},\const{f}(\var{X}))}
- can be either an atom or a term, depending on where it occurs in a rule),
-and functions can be nested within a term.
-There are also built-in constructs
-(cf.\ Section~\ref{subsec:gringo:arith} and~\ref{subsec:gringo:comp})
-having particular representations.
-Finally, any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
-for an atom~$A$.
-
-In Section~\ref{subsec:semantics},
-we have already provided a translation~$\psi$
-from the head of a rule to a propositional formula.
-By viewing an integrity constraint
-$\code{:-}~\mathit{Body}\code{.}$ as a shorthand for $\bot~\code{:-}~\mathit{Body}\code{.}$,
-we can now explain the semantics of integrity constraints by adding the following case:
-\begin{itemize}
-\item $\psi[\bot]=\bot$.
-\end{itemize}
-Note the role of integrity constraints is to eliminate answer set candidates.
-In fact, given an integrity constraint~$\code{:-}~\mathit{Body}\code{.}$,
-any answer set~$X$ is such that $X\not\models\phi[\mathit{Body}]$, or
-in other words, some literal~$L$ in $\mathit{Body}$ must be unsatisfied w.r.t.~$X$.
-Elaborate examples on the usage of facts, rules, and integrity constraints
-will be provided in Section~\ref{sec:examples}.
-
-\subsubsection{Classical Negation}\label{subsec:gringo:negation}
-
-In logic programs, connective \code{not} expresses default negation,
-that is, a literal $\code{not}~A$ is assumed to hold unless~$A$ is derived.
-In contrast, the classical (or strong) negation of some proposition
-holds if the complement of the proposition is objectively derived~\cite{gellif91a}.
-Classical negation, indicated by symbol ``\code{-},'' is permitted in front of atoms.
-That is, if~$A$ is an atom, then~$\code{-}A$ is the complement of~$A$.
-Semantically, $\code{-}A$ is simply a new atom,
-with the additional condition that~$A$ and~$\code{-}A$ must not jointly hold.
-A logic program~$\Pi$ containing complementary atoms is thus translated
-to propositional theory
-$\Phi[\Pi \cup \{\code{:-}~A\code{,}\code{-}A \mid A\in\ground{\mathcal{A}}\}]$.%
-\footnote{Recall that $\ground{\mathcal{A}}$ consists of all variable-free atoms
-built from predicates and functions in~$\Pi$.}
-Observe that classical negation is merely a syntactic feature that can be
-implemented via integrity constraints whose effect is to eliminate
-any answer set candidate containing complementary atoms.
-
-\begin{example}\label{ex:flies:neg}
-The following logic program reformulates the one in Example~\ref{ex:flies}:
-%
-\lstinputlisting[nolol]{examples/bird.lp}
-\lstinputlisting[firstnumber=3]{examples/flycn.lp}
-%
-Logically, classical negation is reflected by
-(implicit) integrity constraints as follows:%
-\marginlabel{By invoking\\
-  \code{\mbox{~}gringo -t \textbackslash\\
-        \mbox{~}examples/bird.lp \textbackslash\\
-        \mbox{~}examples/flycn.lp}
-the reader can observe that \gringo\ indeed produces the integrity constraint
-in Line~7.}
-%
-\begin{lstlisting}[firstnumber=6]
-:- flies(tux),    -flies(tux).
-:- flies(tweety), -flies(tweety).
-\end{lstlisting}
-The additional integrity constraints do not yet change the semantics of the
-program, which still has the answer sets already provided in Example~\ref{ex:flies:ground}
-(of course, identifying
-\code{\pred{neg\mus flies}(\const{tux})} with
-\code{\pred{-flies}(\const{tux})} and
-\code{\pred{neg\mus flies}(\const{tweety})} with
-\code{\pred{-flies}(\const{tweety})}).
-This situation changes if we add the following fact:
-\begin{lstlisting}[firstnumber=8]
-flies(tux).
-\end{lstlisting}
-While the program from Example~\ref{ex:flies} still admits two answer sets,
-both containing 
-\code{\pred{flies}(\const{tux})} and
-\code{\pred{neg\mus flies}(\const{tux})},
-there no longer is any answer set for our new program using classical negation.
-In fact, answer set candidates that contain both
-\code{\pred{flies}(\const{tux})} and
-\code{\pred{-flies}(\const{tux})} violate the integrity constraint in Line~6.
-\eexample
-\end{example}
-
-\subsubsection{Disjunction}\label{subsec:gringo:disjunction}
-
-Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
-An additional case of translation~$\psi$ from Section~\ref{subsec:semantics}
-reflects the semantics of disjunction:
-\begin{itemize}
-\item $\psi[G\code{|}H]=\psi[G]\vee\psi[H]$.
-\end{itemize}
-%
-The concept of level-restrictedness (cf.\ Section~\ref{subsec:lambda}) is extended
-to disjunctive programs by,
-for a disjunctive rule~$r$, every predicate~$p/i$
-of some atom in the head of~$r$, and each $V\in\vars{r}$,
-requiring that there is some $A\in\binders{V}{r}$ such that
-$\lambda(p/i)>\lambda(q/j)$ for predicate~$q/j$ of~$A$.
-Furthermore, a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
-must not contain any rule with a (non-trivial) disjunctive head.
-
-The following rule combines the ones in Line~3 and~4 from Example~\ref{ex:flies:neg}:
-\begin{lstlisting}[numbers=none]
-flies(X) | -flies(X) :- bird(X).
-\end{lstlisting}
-In general, the use of disjunction however increases
-computational complexity~\cite{eitgot95a}.
-This is why \clingo%
-\footnote{Run as a monolithic system performing both grounding and solving.}
-and solvers like 
-\assat~\cite{linzha04a},
-\clasp~\cite{gekanesc07b},
-\nomorepp~\cite{angelinesc05c},
-\smodels~\cite{siniso02a}, and
-\smodelscc~\cite{warsch04a}
-do not work on disjunctive programs.
-Rather,
-\claspD~\cite{drgegrkakoossc08a},
-\cmodels~\cite{gilima06a,lierler05a}, or
-\gnt~\cite{janisesiyo06a}
-needs to be used for solving a disjunctive program.%
-\footnote{System \dlv~\cite{dlv03a} also deals with disjunctive programs,
-  but it uses a different syntax than presented here.}
-We thus suggest to use ``choice constructs'' (cf.\ Section~\ref{subsec:gringo:aggregate})
-instead of disjunction, unless the latter is required for complexity reasons
-(see~\cite{eitpol06a} for an implementation methodology in disjunctive ASP).
-
-\subsubsection{Built-In Arithmetic Functions}\label{subsec:gringo:arith}
-
-\gringo\ and \clingo\ support a number of arithmetic functions that
-are evaluated during grounding.
-The following symbols are used for these functions:
-\code{+} (addition),
-\code{-} (subtraction),
-\code{*} (multiplication),
-\code{/} or \code{div} (integer division),
-\code{mod} (modulo function),
-\code{abs} (absolute value),
-\code{\textasciitilde} (bitwise complement),
-\code{\&}  (bitwise AND),
-\code{?}   (bitwise OR), and
-\code{\^} (bitwise exclusive OR).
-
-\begin{example}\label{ex:arith:fun}
-The usage of arithmetic functions is illustrated by the logic program:%
-\marginlabel{%
-  The unique answer set of the program,
-  obtained after evaluating all arithmetic functions,
-  can be inspected by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithf.lp}}
-%
-\lstinputlisting{examples/arithf.lp}
-%
-Note that variables~\var{L} and~\var{R} are instantiated to~\const{7} and~\const{2},
-respectively, before arithmetic evaluations.
-Consecutive and non-separative (e.g., before ``\code{(}'')
-spaces can also be dropped,
-while spaces around tokens \code{div} and \code{mod} are mandatory.
-Furthermore, the argument of function \code{abs} must be enclosed in parentheses.
-In Line~9, observe that there is a unary version of \code{-},
-``\code{- \var{R}}'' standing for ``\code{0 - \var{R}}.''
-The four bitwise functions apply to signed integers,
-using the complement on two of a negative integer.
-\eexample
-\end{example}
-
-It is important to note that variables
-in the scope of an arithmetic function are not necessarily bound 
-(in the sense of Section~\ref{subsec:lambda}) by a corresponding atom.
-For instance, atom \code{\pred{p}(\var{X}+1,\var{X})} belongs to
-$\binders{\var{X}}{r}$, while atom \code{\pred{p}(\var{X}+1,\var{Y})}
-does not belong to $\binders{\var{X}}{r}$ because it contains~\var{X}
-only in the scope of an arithmetic function.
-
-\subsubsection{Built-In Comparison Predicates}\label{subsec:gringo:comp}
-
-The following built-in predicates permit term comparisons
-within the bodies of rules and on the right-hand side of conditions
-(cf.\ Section~\ref{subsec:gringo:condition}):
-\code{==} (equal),
-\code{!=} (not equal),
-\code{<} (less than),
-\code{<=} (less than or equal),
-\code{>} (greater than),
-\code{>=} (greater than or equal).
-
-\begin{example}\label{ex:arith:pred}
-The usage of comparison predicates is illustrated by the logic program:%
-\marginlabel{%
-  The unique answer set of the program is obtained via call:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithc.lp}}
-%
-\lstinputlisting{examples/arithc.lp}
-%
-The last two lines hint at the fact that arithmetic functions are evaluated
-before comparison predicates, so that the latter actually compare integers.
-
-All comparison predicates can also be used with constants,
-as in the next program:%
-\marginlabel{%
-  As above, invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/symbc.lp}
-  yields the unique answer set of the program in terms of facts.}
-%
-\lstinputlisting{examples/symbc.lp}
-%
-Finally, note that~\code{==} and~\code{!=} also apply to
-compound terms over functions with non-zero arity as well as to
-mixed integer and non-integer arguments.%
-\footnote{%
-  Such functionalities are currently not supported for the other four
-  comparison predicates,
-  where (after instantiation and arithmetic evaluation)
-  both arguments must be either integers or (symbolic) constants.
-}
-\eexample
-\end{example}
-
-Importantly, a built-in comparison predicate cannot bind variables,
-that is, it does not belong to $\binders{\var{X}}{r}$ for any variable~\var{X}.
-Also note that built-in predicates are not considered by level mappings~$\xi$
-in the context of stratification (cf.\ Section~\ref{subsec:strat}), or equivalently,
-one may assume that built-in predicates are mapped to level~$0$ and all other
-predicates to levels greater than~$0$.
-
-\subsubsection{Assignments}\label{subsec:gringo:assign}
-
-Built-in predicate~\code{=} can be used in the body of a rule
-(or on right-hand sides of conditions introduced in Section~\ref{subsec:gringo:condition})
-to assign a term on its right-hand side to a variable on its left-hand side.
-Note that ``\code{\var{X} \!=\! }$t$'' belongs to~$\binders{\var{X}}{r}$,
-provided that~$t$ is variable-free or that its variables are bound
-by atoms other than assignments with~\var{X} on the right-hand side.
-Cyclic assignments over otherwise unbound variables are thus excluded.
-
-\begin{example}\label{ex:assign}
-The next program demonstrates how terms can be assigned to variables:%
-\marginlabel{%
-  The unique answer set of the program is obtained via call:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/assign.lp}}
-%
-\lstinputlisting{examples/assign.lp}
-%
-Line~3 contains four assignments, where the right-hand sides directly or indirectly
-depend on~\var{X} and~\var{Y}.
-These two variables are bound in Line~4 via atoms of predicate \pred{num}/$1$.
-Also observe the different usage and role of built-in comparison predicate~\code{==}.
-\eexample
-\end{example}
-
-\subsubsection{Intervals}\label{subsec:gringo:interval}
-
-In Line~1 of Example~\ref{ex:assign},
-there are five facts~\code{\pred{num}($k$)}
-over consecutive integers~$k$.
-For a more compact representation,
-\gringo\ and \clingo\ support integer intervals of the form $i$\code{..}$j$,
-where~$i$ and~$j$ are integers.
-Such an interval represents each integer~$k$ such that $i\leq k\leq j$,
-and intervals are expanded during grounding.
-
-\begin{example}\label{ex:int}
-The next program makes use of integer intervals:
-%
-\lstinputlisting{examples/int.lp}
-%
-The facts in Line~1 and~2 are expanded as follows:
-%
-\begin{lstlisting}[numbers=none]
-num(1).    num(2).   num(3).   num(4).   num(5).
-top5(5).  top5(6).  top5(7).  top5(8).  top5(9).
-\end{lstlisting}
-%
-By instantiating~\var{X} to~\const{9},
-the rule in Line~4 becomes:
-%
-\begin{lstlisting}[numbers=none]
-top5num(1..5,5..9) :- num(5..9), top5(1..5), top(9).
-\end{lstlisting}
-%
-It is expanded to the cross product
-$(\code{\const{1}..\const{5}})\times
- (\code{\const{5}..\const{9}})\times
- (\code{\const{5}..\const{9}})\times
- (\code{\const{1}..\const{5}})$
-of intervals:
-%
-\begin{lstlisting}[numbers=none,escapechar=@]
-top5num(1,5) :- num(5), top5(1), top(9).
-top5num(2,5) :- num(5), top5(1), top(9).
-        @\rlap{\vdots}@
-top5num(5,5) :- num(5), top5(1), top(9).
-top5num(1,6) :- num(5), top5(1), top(9).
-top5num(2,6) :- num(5), top5(1), top(9).
-        @\rlap{\vdots}@  @\rlap{\vdots}@
-top5num(5,9) :- num(5), top5(1), top(9).
-top5num(1,5) :- num(6), top5(1), top(9).
-top5num(2,5) :- num(6), top5(1), top(9).
-        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
-top5num(5,9) :- num(9), top5(1), top(9).
-top5num(1,5) :- num(5), top5(2), top(9).
-top5num(2,5) :- num(5), top5(2), top(9).
-        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@         @\rlap{\vdots}@
-top5num(5,9) :- num(9), top5(4), top(9).
-top5num(1,5) :- num(5), top5(5), top(9).
-top5num(2,5) :- num(5), top5(5), top(9).
-        @\rlap{\vdots}@  @\rlap{\vdots}@
-top5num(5,9) :- num(5), top5(5), top(9).
-top5num(1,5) :- num(6), top5(5), top(9).
-top5num(2,5) :- num(6), top5(5), top(9).
-        @\rlap{\vdots}@  @\rlap{\vdots}@          @\rlap{\vdots}@
-top5num(5,9) :- num(9), top5(5), top(9).
-\end{lstlisting}
-%
-Note that only the rules with
-\code{\pred{num}(\const{5})} and
-\code{\pred{top5}(\const{5})}
-\marginlabel{%
-  Again the unique answer set is obtained via call:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/int.lp}}
-%
-in the body actually contribute to the unique answer set
-of the above program by deriving all atoms \code{\pred{top5num}($m$,$n$)}
-for $\const{1}\leq m\leq\const{5}$ and $\const{5}\leq n\leq\const{9}$.
-\eexample
-\end{example}
-
-As with built-in arithmetic functions,
-an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
-cannot be used to bind the variable.
-
-\subsubsection{Conditions}\label{subsec:gringo:condition}
-
-Conditions allow for instantiating variables to
-collections of terms within a single rule.
-This is particularly useful for encoding conjunctions or disjunctions over
-arbitrarily many ground atoms as well as for the compact representation of aggregates
-(cf.\ Section~\ref{subsec:gringo:aggregate}).
-The symbol~``\code{:}'' is used to formulate conditions.
-
-\begin{example}\label{ex:cond}
-The following program uses conditions in a rule body and in a rule head:
-%
-\lstinputlisting{examples/cond.lp}
-%
-We are particularly interested in the rules in Line~5 and~6,
-instantiated as follows:%
-%
-\begin{lstlisting}[firstnumber=5]
-meet :- available(jane), available(john).
-on(mon) | on(tues) | on(wed) | on(thurs) | on(fri) :- meet.
-\end{lstlisting}
-%
-The conjunction in Line~5 is obtained by replacing~\var{X} in
-\marginlabel{%
-  The reader can reproduce these ground rules by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/cond.lp}}
-\code{\pred{available}(\var{X})} with all ground terms~$t$ such that
-\code{\pred{person}($t$)} holds, namely, $t=\const{jane}$ and $t=\const{john}$.
-Furthermore, the condition in the head of the rule in Line~6 turns into
-a disjunction over all ground instances of
-\code{\pred{on}(\var{X})} where~\var{X} is substituted by some term~$t$
-such that
-\code{\pred{day}($t$)} holds.
-That is, conditions in the body and in the head of a rule
-are expanded to different basic language constructs.
-
-Composite conditions can also be constructed via~``\code{:},''
-as in the additional rules:
-%
-\lstinputlisting[firstnumber=7]{examples/twocond.lp}
-%
-Observe that we may use the same atom, viz., \code{\pred{day}(\var{X})},
-both on the left-hand and on the right-hand side of~``\code{:}.''
-Furthermore, negative literals like \code{not \pred{weekend}(\var{X})}
-can occur on both sides of a condition.
-Note that literals on the right-hand side of a condition are connected conjunctively, that is,
-all of them must hold for ground instances of an atom in front
-of the condition.
-Thus, the instantiated rule in Line~8 looks as follows:
-%
-\begin{lstlisting}[firstnumber=8]
-weekdays :- day(mon), day(tues), day(wed), day(thurs), day(fri).
-\end{lstlisting}
-%
-The atoms in the body of this rule follow from facts, so that
-the rule can be simplified to a fact \code{\pred{weekdays}.}
-(as done by \gringo).
-\eexample
-\end{example}
-
-There are three important issues about the correct usage of conditions.
-First, all predicates of atoms on the right-hand side of a condition
-must be either domain predicates (cf.\ Section~\ref{subsec:strat}) or built-in,
-which is due to the fact that conditions are evaluated during grounding.%
-\footnote{%
-  The bodies of rules in a stratified subprogram (cf.\ Section~\ref{subsec:strat})
-  may contain conditions.
-  For a rule~$r$ and
-  $L_0\code{:}L_1\code{:}\dots\code{:}L_n$
-  in the body of~$r$,
-  let $A_0\in\nbody{r}$ if $L_0=\code{not}~A_0$
-  and $L_0\in\pbody{r}$ otherwise,
-  and for $1\leq i\leq n$, 
-  assume $A_i\in\nbody{r}$ for atom~$A_i$
-  such that $L_i=A_i$ or $L_i=\code{not}~A_i$, respectively.
-  }
-Second, any variable occurring within a condition is considered as \emph{local},
-that is, a condition cannot be used to bind variables outside itself.
-In turn, variables outside conditions are \emph{global}, and each variable
-within an atom in front of a condition must occur on the right-hand side or
-be global.
-Third, global variables take priority over local ones, that is,
-they are instantiated first.
-As a consequence, a local variable that also occurs globally is substituted by a term
-before the ground instances of a condition are determined.
-Hence, the names of local variables must be chosen with care,
-making sure that they do not accidentally match the names of global variables.
-
-\subsubsection{Pooling}\label{subsec:gringo:pool}
-
-Symbol~``\code{;}'' allows for pooling alternative terms to be used as
-argument within an atom, thus, specifying rules more compactly.
-An atom written in the form \code{\pred{p}($\dots$,X;Y,$\dots$)}
-abbreviates two options:
-\code{\pred{p}($\dots$,X,$\dots$)} and \code{\pred{p}($\dots$,Y,$\dots$)}.
-Pooled arguments in an atom of a rule body 
-(or on the right-hand side of a condition) are expanded to a conjunction
-of the options within the same body (or within the same condition),
-while they are expanded to multiple rules 
-(or multiple literals connected via ``\code{,}'')
-when occurring in the head (or in front of a condition).
-
-\begin{example}\label{ex:pool}
-The following logic program makes use of pooling:
-%
-\lstinputlisting{examples/pool.lp}
-%
-Let us consider instantiations of the rule in Line~3 obtained with substitution 
-$\{\var{A}\mapsto\const{a},\linebreak[1]\var{B}\mapsto\const{b},
-   \var{M}\mapsto\const{1},\var{N}\mapsto\const{2}\}$.
-Note that \pred{mix}/$2$ and \pred{-mix}/$2$ each admit four options,
-corresponding to the cross product of $\{\const{a},\const{b}\}$ substituted
-for~\var{A} and~\var{B}, respectively, together with $\{\const{1},\const{2}\}$
-substituted for~\var{M} and~\var{N}.
-While the instances obtained for \pred{mix}/$2$ give rise to four rules,
-the instances for \pred{-mix}/$2$ jointly belong to the body.
-The (repeated) body also contains two instances each of \pred{sym}/$1$ and of \pred{num}/$1$.
-We thus get the rules:%
-\marginlabel{%
-  Simplified versions of these rules are produced via call:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/pool.lp}}
-%
-\begin{lstlisting}[numbers=none]
-mix(a,1) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
-  not -mix(1,b), not -mix(2,a), not -mix(2,b).
-mix(a,2) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
-  not -mix(1,b), not -mix(2,a), not -mix(2,b).
-mix(b,1) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
-  not -mix(1,b), not -mix(2,a), not -mix(2,b).
-mix(b,2) :- sym(a),sym(b), num(1),num(2), not -mix(1,a),
-  not -mix(1,b), not -mix(2,a), not -mix(2,b).
-\end{lstlisting}
-%
-Finally, we note that pooling is also possible with arguments of built-in predicates.
-\eexample
-\end{example}
-
-\subsubsection{Aggregates}\label{subsec:gringo:aggregate}
-
-An aggregate is an operation on a multiset of weighted literals that evaluates to an integer.
-In combination with arithmetic comparisons, we can extract a truth value from
+An aggregate is an operation on a multiset of weighted literals that evaluates to a value,
+depending on the type of aggregate this might be integers but, 
+e.g., \const{\#min} and \const{\#max} aggregates also permit function symbols.
+In combination with comparisons, we can extract a truth value from
 an aggregate's evaluation, thus, obtaining an aggregate atom.
@@
 %
-where~$l$ and~$u$ are integers each of which can possibly be omitted,
-$\mathrm{op}$ is a function from multisets of integers to an integer,
+where~$l$ and~$u$ are values each of which can possibly be omitted,
+$\mathrm{op}$ is a function from multisets of values to a value,
 each $L_i$ is a literal of the form~$A$ or $\code{not}~A$
 for some atom~$A\in\ground{\mathcal{A}}$,
-and each~$w_i$ is an integer.
+and each~$w_i$ is a value.
 The intuitive reading of an aggregate atom is that the value returned by~$\mathrm{op}$
@@
 
-For an aggregate atom~$\mathit{Agg}$ as in~(\ref{eq:aggregate:atom}),
+For an aggregate atom~$\mathit{Agg}$ as in Equation(\ref{eq:aggregate:atom}),
 we let $\atom{\mathit{Agg}}=\{L_i\in\nolinebreak\ground{\mathcal{A}}\mid 1\leq i\leq n\}$
@@
 Currently, \gringo\ and \clingo\ support aggregate operations
-\const{sum}, \const{max}, and \const{min}.%
+\const{\#sum}, \const{\#min}, and \const{\#max}, \const{\#avg}, \const{\#even}, and \const{\#odd}.%
 \footnote{%
@@
 in heads of rules~\cite{ferlif05a,siniso02a}:%
+
+\newcommand\shiftleft{\kern\leftshift&\kern-\leftshift}
+
+\newlength{\leftshift} 
+\setlength{\leftshift} {2em}
+%\settowidth{\leftshift}{$\phi^{+}[ l~ \const{\#sum} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ]$}
+
+\begin{align}
+  % l op u (head) %
+  \shiftleft
+  \psi[ \lhd~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}\rhd ] 
+  \label{eq:aggr_head} \\
+   &= \phi[ \triangleleft~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}\triangleright ] \notag \\
+   &\wedge \bigl(\bigwedge\nolimits_{A\in atom(\triangleleft~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}\triangleright)}(A\vee\neg A)\bigr) \notag \\
+  % l op u %
+  \shiftleft
+  \phi[ l~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} u ]
+  \label{eq:aggr_lu} \\
+  &= \phi[ l~ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] \wedge{} 
+    \phi[ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} u ] \notag \\
+  % op %
+  \shiftleft
+  \phi[ \mathrm{op} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] = \top
+  \label{eq:aggr}
+\end{align}
+Translation~$\psi$ of an aggregate atom in the head of a rule consists of two parts:
+first, $\phi$ is used as defined in the following
+(in Equation (\ref{eq:aggr_head}), $\lhd$ and~$\rhd$ are used as ``wildcards'' for provided or omitted 
+ lower and upper bounds),
+second, a disjunction $(A\vee\neg A)$ is added for each atom~$A$ occurring positively.
+Even though such disjunctions are tautological,
+they are important under answer set semantics (cf.\ Section~\ref{subsec:semantics}),
+as they permit deriving an arbitrary subset of atoms from an aggregate in the head
+(which explains the name  ``choice semantics'').
+Equation \ref{eq:aggr_lu} splits up an aggregate with lower and an upper into the conjunction of two
+conjunctively connected parts.
+Note that this split agrees with the semtaics in \cite{ferraris05a} independend of the aggregate $\mathrm{op}$ used.
+Finally, Equation (\ref{eq:aggr}) reflects that an unconstrained aggregate atom always holds.
+
+\paragraph{Sum Aggregates}
+First, we show how monotone aggregates can be translated into a nested expression.
+Using this translation we show how to transfom antimonotone sum aggregates and 
+also give a translation that treats negative weights.
 \footnote{%
@@
   we strongly recommend not to use negative weights within \const{sum} aggregates.}
-\begin{enumerate}
-\item % op %
-$\phi[
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
- \top$,
-\item % l op u %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \phi[
-  l~
- &
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
- {} \\ 
- &
-  \phi[
-  l~
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ]
-  \wedge
-  \phi[
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ]\text{,}
- \end{array}$
-\item % l sum %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \phi[
-  l~
- &
-  \const{sum} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
- {} \\ 
- &
-  \big(
-  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},l\leq w_{i_1}+\dots+w_{i_j}}
-  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
-  \big)\text{,}
- \end{array}$
-\item % sum u %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \phi
-  [
- &
-  \const{sum} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
- {} \\ 
- &
-  \neg
-  \big(
-  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},u< w_{i_1}+\dots+w_{i_j}}
-  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
-  \big)\text{,}
- \end{array}$
-\item % l max %
- $\phi[
-  l~
-  \const{max} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
- \hspace{6pt}
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},l\leq w_i}
-  \phi[L_i]
-  \big)$,
-\item % max u %
- $\phi[
-  \const{max} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
-  \neg
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},u< w_i}
-  \phi[L_i]
-  \big)$,
-\item % l min %
- $\phi[
-  l~
-  \const{min} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
-  \neg
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},w_i< l}
-  \phi[L_i]
-  \big)
- $,
-\item % min u %
- $\phi[
-  \const{min} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
-  \hspace{6pt}
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},w_i\leq u}
-  \phi[L_i]
-  \big)
- $, and
-\item % psi %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \psi[
-  \lhd
- &~
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  \rhd
-  ] = 
-% {} \\ 
-% &
-   \phi[
-  \lhd~
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  \rhd
-  ]
-  \wedge
- {} \\ 
- &
-  \big(
-  \bigwedge_{A\in\atom{
-    \lhd~
-    \mathrm{op} 
-    \code{[}
-    L_1\code{=}w_1\code{,}
-    \dots\code{,}
-    L_n\code{=}w_n
-    \code{]}
-    \rhd
-  }}
-  (A\vee\neg A)
-  \big)\text{.}
- \end{array}$
-\end{enumerate}
-%
-The first item reflects that an unconstrained aggregate atom always holds,
-while aggregate atoms with a lower and an upper bound are split up into two
-conjunctively connected parts (Item~2).
+\begin{align}
+  % l sum+ %
+  \shiftleft
+  \phi[ l~ \const{\#sum}^{+} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] 
+  \label{eq:sum_plus}\\
+  &= \bigvee\nolimits_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},l\leq w_{i_1}+\dots+w_{i_j}}
+   (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}]) \notag 
+\end{align}
+Equation (\ref{eq:sum_plus}) translates a monotone aggregate~$\const{\#sum}^{+}$
+with a lower bound and only positive weights into a nested expressions.
+
+\begin{align}
+  % l sum %
+  \shiftleft
+  \phi[l~\const{\#sum}~\code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ]
+  \label{eq:sum_neg} \\
+  &= \phi [ l + \phi^{+}(w_1) + \dots + \phi^{+}(w_n) ~\const{\#sum}^{+}
+    \code{[} \phi^{+}(L_1\code{=}w_1)\code{,} \dots\code{,} \phi^{+}(L_n\code{=}w_n) \code{]} ] \notag \\
+  % weight lit %
+  \shiftleft
+  \phi^{+}(L=w) = 
+    \left\{\begin{aligned}
+      L=w & \mbox{ if } w \geq 0 \\
+      \overline{L}=-w &  \mbox{ otherwise}\\
+    \end{aligned}\right. \notag \\
+  % weight %
+  \shiftleft
+  \phi^{+}(w) =-\min\{0,w\} \notag
+\end{align}
+Equation (\ref{eq:sum_neg}) removes all negative weights from a sum aggregate with a lower bound 
+and then applies the transformation described above.
+Whenever there is a negative weight, the sign of both literal and weight is inverted and the weight itself is subtracted from the bound.
+Note that this transformation produces counter-intuitive results as soon as the aggregate is used recursively.
+Nevertheless, there are scenarios where negative weights might be helpfull.
+As long as the aggregate is stratified, the transformation corresponds with the intuitive meaning, e.g.,
+you can use negative weights to compare two sets $A=\{a,b,c\}$ and $B=\{d,e\}$:
+\begin{lstlisting}[numbers=none]
+subseteq(A,B) :- #sum [ a=1,b=1,c=1,d=-1,e=-1 ] ] 0.
+\end{lstlisting}
+
+\begin{align}
+  % sum u %
+  \phi[ \const{\#sum} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}~u ] 
+  &= \neg \phi[  u+1~\const{\#sum} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}]
+  \label{eq:sum_u}
+\end{align}
+Finally, Equation (\ref{eq:sum_u}) translates sum aggregates with an upper bound using the
+translation for sum aggregates with lower bounds described above.
+Note that again negative literals might cause problems.
+Even in the case that all weights are negative (and as such the sum aggregate itself is monotone),
+the above translation is used, which might lead to counter-intuitive results.
+
+\paragraph{Max and Min Aggregates} 
+We define the sematics of max aggregates using the transformation for sum aggregates described above.
+First, we translate max aggregates with upper and lower bound, respectively, into corresponding sum aggregates.
+Then we give the semantics of min aggregates using the translation for max aggregates.
+Note that in constrast to sum aggregates, max aggregates allow for using non-integer weights as long as there
+is some strict total ordering defined on them.
+\begin{align}
+  % l max
+  \shiftleft
+  \phi[l~\const{\#max} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ]
+  \label{eq:max_l}\\
+  &= \phi[1~\const{\#sum} \code{[} L_1\code{=}\mathrm{ge}(w_1, l)\code{,} \dots\code{,} L_n\code{=}\mathrm{ge}(w_1, l) \code{]} ]  \notag \\
+  % max(w,l)
+  \shiftleft
+  \mathrm{ge}(a, b) = \left\{\!\begin{aligned}
+       1 &\mbox{ if } a \geq b \\
+       0 &\mbox{ otherwise}
+     \end{aligned}\right. \notag
+\end{align}
+A (monotone) max aggregate with a lower bound is translated into a sum by setting all weights 
+that are greater equal to the lower bound to one and requiring that the sum of weights is at least one.
+This transformation preserves the monotonicity of the aggregate and holds independendly of the type of weights used.
+
+\begin{align}
+  % max u %
+  \shiftleft
+  \phi[\const{\#max} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} u]
+  \label{eq:max_u} \\
+  &= \phi[\const{\#sum} \code{[} L_1\code{=}\mathrm{ge}(w_1, l)\code{,} \dots\code{,} L_n\code{=}\mathrm{ge}(w_1, l) \code{]} 0]  \notag \\
+  \shiftleft
+  \mathrm{gt}(a, b) = \left\{\!\begin{aligned}
+       1 &\mbox{ if } a > b \\
+       0 &\mbox{ otherwise}
+     \end{aligned}\right. \notag
+\end{align}
+Similarly to Equation~(\ref{eq:max_l}), an (antimonotone) max aggregate with an upper bound can be translated into a sum with an upper bound.
+Again the monotonicity of the aggregate is preserved.
+
+\begin{align}
+  % min u %
+  \shiftleft
+  \phi[ \const{\#min} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}~u ] 
+  \label{eq:min_u}  \\
+  &= \phi[1~\const{\#sum} \code{[} L_1\code{=}\mathrm{ge}(l,w_1)\code{,} \dots\code{,} L_n\code{=}\mathrm{ge}(l,w_1) \code{]}]
+  \notag \\
+  % l min %
+  \shiftleft
+  \phi[ l~ \const{\#min} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ]
+  \label{eq:min_l} \\
+  &= \phi[\const{\#sum} \code{[} L_1\code{=}\mathrm{gt}(l,w_1)\code{,} \dots\code{,} L_n\code{=}\mathrm{gt}(l,w_1) \code{]} 0]
+  \notag
+\end{align}
+Analogous to the definition of max aggregates, 
+Equations (\ref{eq:min_l}) and (\ref{eq:min_u}) define the semantics of min aggregates.
+
+\paragraph{Average Aggregates}
+Avergate can be directly translated into sum aggregates by modifying the weights of its literals.
+To see how the translation works consider:
+\begin{align*}
+\mathrm{avg}\code{[}a_1,\dots,a_n\code{]} &\prec b \\
+a_1+\dots+a_n               &\prec n \cdot b \\
+a_1 - b + \dots + a_n - b   &\prec 0
+\end{align*}
+where $\prec$ is some relation.
+To avoid problems with empty average aggregates we define that $\mathrm{avg}\code{[} \code{]} \prec b$ iff $0 \prec 0$.
+Thus the translation can turn an average aggregate into a sum by subtracting the bound from all its weights.
+A strong point of this translation is that it preserves the monotonicity of the average aggregate.
+This translation is applied in Equations (\ref{eq:avg_l}) and (\ref{eq:avg_u}).
+
+\begin{align}
+  % l avg %
+  \shiftleft
+  \phi[ l~\const{\#avg} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] 
+  \label{eq:avg_l} \\
+  &= \phi[ \const{\#sum} \code{[} L_1\code{=}w_1-l \code{,} \dots\code{,} L_n\code{=}w_n-l \code{]}~0 ] 
+  \notag \\
+  % avg u %
+  \shiftleft
+  \phi[ \const{\#avg} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]}~u ] = 
+  \label{eq:avg_u} \\
+  &= \phi[ \const{\#sum} \code{[} L_1\code{=}w_1-u \code{,} \dots\code{,} L_n\code{=}w_n-u \code{]}~0 ]
+  \notag
+\end{align}
+
+\paragraph{Parity Aggregates}
+We also support the two parity aggregates $\code{\#even}$ and $\code{\#odd}$.
+Intuitively, an even or odd aggregate is true if the sum of its weights modulo two is one or zero, respectively.
+Note that we do not write bounds
+Even and odd aggregates are transformed into nested expressions:
+\begin{align}
+  % odd %
+  \shiftleft
+  \phi[ \const{\#odd} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] \\
+  &=\left\{\begin{aligned}
+    \bot & \mbox{ if } n = 0 \\
+    \phi[ \const{\#odd} \code{[} L_2\code{=}w_2-u \code{,} \dots\code{,} L_n\code{=}w_n-u \code{]}~0 ] & \mbox{ if } 2 | w_1  \\
+    \neg L_1 \wedge \phi[ \const{\#odd} \code{[} L_2\code{=}w_2 \code{,} \dots\code{,} L_n\code{=}w_n \code{]}~0 ] & \\
+    \vee L_1 \wedge \neg \phi[ \const{\#odd} \code{[} L_2\code{=}w_2 \code{,} \dots\code{,} L_n\code{=}w_n \code{]}~0 ] & \mbox{ otherwise}
+  \end{aligned}\right. \notag
+\end{align}
+
+\begin{align}
+  % even %
+  \shiftleft
+  \phi[ \const{\#even} \code{[} L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n \code{]} ] \\
+  &=\left\{\begin{aligned}
+    \top & \mbox{ if } n = 0 \\
+    \phi[ \const{\#even} \code{[} L_2\code{=}w_2-u \code{,} \dots\code{,} L_n\code{=}w_n-u \code{]}~0 ] & \mbox{ if } 2 | w_1  \\
+    \neg L_1 \wedge \phi[ \const{\#even} \code{[} L_2\code{=}w_2 \code{,} \dots\code{,} L_n\code{=}w_n \code{]}~0 ] & \\
+    \vee L_1 \wedge \neg \phi[ \const{\#even} \code{[} L_2\code{=}w_2 \code{,} \dots\code{,} L_n\code{=}w_n \code{]}~0 ] & \mbox{ otherwise}
+  \end{aligned}\right. \notag
+\end{align}
+
 Item~3--8 specify these parts individually for~\const{sum},~\const{max}, and~\const{min}
@@
 the lower bound for \const{min} (Item~7).
-Finally, translation~$\psi$ of an aggregate atom in the head of a rule consists of two parts:
-first, $\phi$ is used like in the body of a rule
-(in Item~9, $\lhd$ and~$\rhd$ are used as ``wildcards'' for provided or omitted 
- lower and upper bounds),
-second, a disjunction $(A\vee\neg A)$ is added for each atom~$A$ occurring positively.
-Even though such disjunctions are tautological,
-they are important under answer set semantics (cf.\ Section~\ref{subsec:semantics}),
-as they permit deriving an arbitrary subset of atoms from an aggregate in the head
-(which explains the name  ``choice semantics'').
+% <cut here>
 Beyond this special treatment in rule heads,
@@
   also built-in comparison predicates (cf.\ Section~\ref{subsec:gringo:comp})
-  are permitted on the right-hand side of conditions only.}
+  are permitted on the right-hand sides of conditions only.}
 That is, an interval gives rise to multiple literals connected via~``\code{,}''
@@
 
-\subsubsection{Optimization}\label{subsec:gringo:optimize}
-
-Optimization statements extend the basic question of
-whether a set of atoms is an answer set to
-whether it is an optimal answer set.
-To support this reasoning mode, \gringo\ and \clingo\ adopt the
-optimization statements of \lparse~\cite{lparseManual},
-indicated via keywords \const{maximize} and \const{minimize}.
-Syntactically, a \const{maximize} or \const{minimize} statement
-is similar to a fact whose head is a \const{count} or a \const{sum}
-aggregate (without bounds),
-viz.,
-``$\textrm{opt}\code{\char`\{}\dots\code{\char`\}}\code{.}$'' or
-``$\textrm{opt}\code{[}\dots\code{]}\code{.}$'' where
-$\textrm{opt}\in\{\const{maximize},\linebreak[1]\const{minimize}\}$.
-As an optimization statement does not admit a body, any (local) variable
-in it must also occur in an atom (over a domain or built-in predicate)
-on the right-hand side of a condition (cf.\ Section~\ref{subsec:gringo:condition})
-within the optimization statement.
-In multiset notation (square brackets),
-weights may be provided as with \const{sum} aggregates. %,
-% where~\const{1} is used as default for omitted weights.
-In set notation (curly brackets),
-duplicates of literals are removed
-as with \const{count} aggregates.
-
-The semantics of an optimization statement is intuitive:
-an answer set is \emph{optimal} if the sum of weights
-(using~\const{1} for unsupplied weights)
-of literals that hold is maximal or minimal, as required by the statement,
-among all answer sets of the given program.
-This definition is sufficient if a single optimization statement is specified
-along with a logic program.
-Unfortunately, the syntax used by \lparse\ and adopted by \gringo\ and \clingo\
-does not provide any means to explicitly declare priorities among multiple
-optimization statements.
-A technique how to compile multiple optimization statements into a single one
-is discussed in~\cite{siniso02a},
-but it can lead to large integers
-(probably rather unnatural for a user to assign).
-Thus, rather than restricting to a single optimization statement,
-a sequence of them is permitted,
-where a statement provided later on takes priority over previous ones.
-That is, the last optimization statement in program is the most significant one,
-and previous ones are only considered if answer sets agree on the sum of 
-weights of its literals.
-In this way, multiple optimization statements are arranged into a total order,
-which is the inverse order of occurrence.
-In contrast to the declarative answer set semantics, the semantics of
-optimization statements is thus order-dependent,
-requiring special care when providing more than one of them.
-
-\begin{example}\label{ex:opt}
-To illustrate optimization, we consider a hotel booking situation
-where we want to choose one among five available hotels.
-The hotels are identified via numbers assigned in descending order of stars.
-Of course, the more stars a hotel has the more it costs per night.
-As an ancillary information, we know that hotel~\const{4} is located
-on a main street, which is why we expect its rooms to be noisy.
-This knowledge is specified in Line~1--5 of the following program:
-%
-\lstinputlisting[xrightmargin=-26pt]{examples/opt.lp}
-%
-Line~6--8 contribute optimization statements in inverse order of significance,
-according to which we want to choose the best hotel to book.
-The most significant optimization statement in Line~8 states that
-avoiding noise is our main priority.
-The secondary optimization criterion in Line~7 consists of
-minimizing the cost per star.
-Finally, the third optimization statement in Line~6 specifies that we want
-to maximize the number of stars among hotels that are otherwise indistinguishable.
-The optimization statements in Line~6--8 are instantiated as follows:%
-\marginlabel{%
-  The full ground program is obtained by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/opt.lp}}
-%
-\begin{lstlisting}[firstnumber=6,breaklines,breakindent=48pt]
-maximize [ hotel(1) = 5, hotel(2) = 4, hotel(3) = 3,    hotel(4) = 3, hotel(5) = 2 ].
-minimize [ hotel(1) = 34, hotel(2) = 35, hotel(3) = 30,   hotel(4) = 25, hotel(5) = 30 ].
-minimize [ noisy = 1 ].
-\end{lstlisting}
-If we now use \clasp\ to compute an optimal answer set,%
-\marginlabel{%
-  To compute the unique optimal answer set, invoke:\\
-  \code{\mbox{~}gringo \textbackslash\\
-        \mbox{~}examples/opt.lp | \textbackslash\\
-        \mbox{~}clasp -n 0}\\
-  or alternatively:\\
-  \code{\mbox{~}clingo -n 0 \textbackslash\\
-        \mbox{~}examples/opt.lp}}
-we find that hotel~\const{4} is not eligible because it implies \pred{noisy}.
-Thus, hotel~\const{3} and~\const{5} remain as optimal w.r.t.\ the second most
-significant optimization statement in Line~7.
-This tie is broken via the least significant optimization statement in Line~6
-because hotel~\const{3} has one star more than hotel~\const{5}.
-We thus decide to book hotel~\const{3} offering~\const{3} stars
-to cost~\const{90} per night.
-\eexample
-\end{example}
-
-\subsubsection{Meta-Statements}\label{subsec:gringo:meta}
-
-After considering the language of logic programs,
-we now introduce features going beyond the contents of a program.
-
-\paragraph{Comments.}
-To keep records of the contents of a logic program,
-a logic program file may include comments.
-A comment until the end of a line is initiated by symbol ``\code{\%},''
-and a comment within one or over multiple lines is enclosed
-by ``\code{\%*}'' and ``\code{*\%}.''
-As an abstract example, consider:
-%
-\begin{lstlisting}[numbers=none,escapechar=@]
-@\textit{logic program}@  %* enclosed comment *%  @\textit{logic program}@
-@\textit{logic program}@  % comment till end of line
-@\textit{logic program}@
-%*
-comment over multiple lines
-*%
-@\textit{logic program}@
-\end{lstlisting}
-
-\paragraph{Hiding Predicates.}
-Sometimes, one may be interested only in a subset of the atoms belonging
-to an answer set.
-In order to suppress the atoms of ``irrelevant'' predicates from the output,
-the \code{\#hide} declarative (in which~\code{\#} is optional) can be used.
-The meanings of the following statements are indicated via accompanying comments:
-%
-\begin{lstlisting}[numbers=none]
-#hide.          % Suppress all atoms in output
- hide.          % Same as "#hide."
-#hide p/3.      % Suppress all atoms of predicate p/3 in output
- hide p/3.      % Same as "#hide p/3."
-#hide p(X,Y,Z). % Same as "#hide p/3."
- hide p(X,Y,Z). % Same as "#hide p/3."
-\end{lstlisting}
-%
-In order to selectively include the atoms of a certain predicate in the output,
-one may use the \code{\#show} declarative (in which~\code{\#} is again optional).
-Here are some examples:
-%
-\begin{lstlisting}[numbers=none]
-#show p/3.      % Include all atoms of predicate p/3 in output
- show p/3.      % Same as "#show p/3."
-#show p(X,Y,Z). % Same as "#show p/3."
- show p(X,Y,Z). % Same as "#show p/3."
-\end{lstlisting}
-%
-A typical usage of \code{\#hide} and \code{\#show} is to hide all predicates
-via ``\code{\#hide.}'' and to selectively re-add atoms of certain predicates
-\code{p/$n$} to the output via ``\code{\#show p/$n$.}''
-
-\paragraph{Constant Replacement.}
-Constants appearing in a logic program may actually be placeholders for
-concrete values to be provided by a user.
-An example of this will be given in Section~\ref{subsec:ex:color}.
-Via the \code{\#const} declarative (\code{\#}~is optional),
-one may define a default value to be inserted for a constant.
-Such a default value can still be overridden via command line option
-\code{--const} (cf.\ Section~\ref{subsec:opt:gringo}).
-Syntactically, \code{\#const} must be followed by an assignment having
-a (symbolic) constant on the left-hand side and a term on the right-hand side.
-Some exemplary \code{\#const} declarations are:
-%
-\begin{lstlisting}[numbers=none]
-#const x = 42.
- const y = f(g,h).
-\end{lstlisting}
-
-\paragraph{Domain Declarations.}
-Usually, variable names are local to a rule,
-where they must be bound via appropriate atoms (cf.\ Section~\ref{subsec:lambda}).
-This locality can be undermined by using
-\code{\#domain} declarations (\code{\#}~is optional)
-that globally associate variable names to atoms.
-An associated atom is then simply added to the body of a rule in which
-such a predefined variable name occurs in.
-The following is a made-up example:
-%
-\begin{lstlisting}[numbers=none]
-p(1,1). p(1,2).
-#domain p(X,Y).
- domain p(Y,Z).
-q(Z,X) :- not p(Z,X).
-\end{lstlisting}
-%
-The above program is a priori not level-restricted because
-variables~\var{X} and~\var{Z} are unbound in the last rule.
-However, as they belong to \code{\#domain} declarations,
-\gringo\ and \clingo\ expand the last rule to:
-%
-\begin{lstlisting}[numbers=none]
-q(Z,X) :- p(X,Y), p(Y,Z), not p(Z,X).
-\end{lstlisting}
-%
-Observe that the resulting program is level-restricted (and stratified).
-
-\paragraph{Compute Statements.}
-These statements are artefacts supported for backward compatibility.
-Although we strongly recommend to avoid compute statements,
-we now describe their syntax.
-A compute statement is of the form
-``$\const{\#compute}~n \code{\char`\{}\dots\code{\char`\}}\code{.}$''
-(\code{\#}~and non-negative integer $n$ are optional),
-where the ``$\code{\char`\{}\dots\code{\char`\}}$'' part
-is similar to a \const{count} aggregate.
-The meaning is that all literals contained in
-``$\code{\char`\{}\dots\code{\char`\}}$'' must hold w.r.t.\ answer sets
-that are to be computed,
-while~$n$ specifies a number of answer sets to compute.
-As \clasp, \clingo, and \iclingo\ provide command line option
-\code{--number} (cf.\ Section~\ref{subsec:opt:clasp})
-to specify how many answer sets are to be computed,
-they simply ignore~$n$.
-Furthermore,
-the ``$\code{\char`\{}\dots\code{\char`\}}$'' part can equivalently
-be expressed in terms of integrity constraints,
-as indicated in the comments provided along with the following example:
-%
-\begin{lstlisting}[xrightmargin=-20pt,numbers=none]
-q(1;2).
-{ p(1..5) }.
-#compute 0 { p(X) : q(X) }.       % :- 1 { not p(X) : q(X) }.
- compute   { not p(X) : X=4..5 }. % :- 1 { p(X) : X=4..5 }.
-\end{lstlisting}
-
-\subsection{Input Language of \iclingo}\label{subsec:lang:iclingo}
-
-System \iclingo~\cite{gekakaosscth08a} extends \clingo\ by an
-\emph{incremental} computation mode that incorporates both grounding and solving.
-Hence, its input language includes all constructs described
-in Section~\ref{subsec:lang:gringo}.%
-\footnote{%
-  In its current version, \iclingo\ has no built-in support for
-  classical negation (cf.\ Section~\ref{subsec:gringo:negation}).
-  That is, complementary atoms may jointly belong to answer sets,
-  unless explicitly prohibited by integrity constraints.}
-In addition, \iclingo\ deals with statements of the following form:
-%
-\begin{lstlisting}[numbers=none,escapechar=@]
-#base.
-#cumulative @\textit{constant}@.
-#volatile   @\textit{constant}@.
-\end{lstlisting}
-%
-Via ``\code{\const{\#base}.},''
-the subsequent part of a logic program is declared as static, that is,
-it is processed only once at the beginning of an incremental computation.
-In contrast, ``\code{\const{\#cumulative} \textit{constant}.}''
-and ``\code{\const{\#volatile} \textit{constant}.}'' are used to declare
-a (symbolic) \code{\textit{constant}} as a placeholder for incremental step numbers.
-In the parts of a logic program below a \const{\#cumulative} statement,
-\code{\textit{constant}} is in each step replaced with the current step number,
-and the resulting rules, facts, and integrity constraints are accumulated
-over a whole incremental computation.
-While the replacement of \code{\textit{constant}} is similar,
-a logic program part below a \const{\#volatile} statement is local to steps,
-that is, all rules, facts, and integrity constraints computed in one step
-are dismissed before the next incremental step.
-Note that the type of a logic program part (static, cumulative, or volatile)
-is determined by the last
-\const{\#base}, \const{\#cumulative}, or \const{\#volatile} statement
-preceding it.
-
-During an incremental computation, all static program parts are grounded first,
-while cumulative and volatile parts are grounded step-wise,
-replacing \code{\textit{constant}}s with successive step numbers starting from~\const{1}.
-After a grounding step, \clasp\ is usually invoked via an internal interface
-(like with \clingo), and the incremental computation stops after a step
-in which at least one answer set has been found by \clasp.
-This default behavior can be readapted via command line 
-options (cf.\ Section~\ref{subsec:opt:iclingo}).
-For obtaining a well-defined incremental computation result,
-it is important that (ground) head atoms within static, cumulative, and volatile program parts
-are distinct from each other, and they must also be different from step to step
-(see~\cite{gekakaosscth08a} for details).
-In Section~\ref{subsec:ex:block},
-we will provide a typical example in which these conditions naturally hold.
-
-\subsection{Input Language of \clasp}\label{subsec:lang:clasp}
-
-Solver \clasp~\cite{gekanesc07b} works on logic programs in \lparse's
-output format~\cite{lparseManual}.
-This numerical format, which is not supposed to be human-readable,
-is output by \gringo\ and can be piped into \clasp.
-Such an invocation of \clasp\ looks as follows:
-%
-\begin{lstlisting}[numbers=none]
-gringo [ options | filenames ] | clasp [ number | options ]
-\end{lstlisting}
-%
-Note that \code{number} may be provided to specify a maximum number of answer sets
-to be computed, where~\code{0} makes \clasp\ compute all answer sets.
-This maximum number can also be set via
-option \code{--number} or its abbreviation \code{-n}
-(cf.\ Section~\ref{subsec:opt:clasp}).
-By default, \clasp\ computes one answer set (if it exists).
-If a logic program in \lparse's output format has been stored in a \code{file},
-it can be redirected into \clasp\ as follows:%
-\footnote{%
-  The same is achieved by using option \code{--file} or its short form \code{-f}
-  (cf.\ Section~\ref{subsec:opt:clasp}).}
-%
-\begin{lstlisting}[numbers=none]
-clasp [ number | options ] < file
-\end{lstlisting}
-%
-Via option \code{--dimacs},
-\clasp\ can also be instructed to compute models of a propositional formula
-in DIMACS/CNF format~\cite{dimacs}.
-If such a formula is contained in \code{file},
-then \clasp\ can be invoked in the following way:
-%
-\begin{lstlisting}[numbers=none]
-clasp [ number | options ] --dimacs < file
-\end{lstlisting}
-%
-Finally, \clasp\ may be used as a library, as done within \clingo\ and \iclingo.
-
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "guide"
-%%% End:
+\begin{newstuff}
+	\begin{itemize}
+		\item there is also $\#$avg, $\#$times, $\#$even, $\#$odd
+		\item bingo will support everything but times
+		\item heads of stratified aggregates can bind their variables
+		\item a \# has to be put in front of every aggregate $\Rightarrow$ adjust examples
+	\end{itemize}
+\end{newstuff}

### 2010-07-09 doc/guide/quickstart.tex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3171 3e02f596-3131-0410-8352-e04eeaec37cc

\section{Quickstart}

In ASP it is custom to provide a \emph{uniform}
problem definition~\cite{martru99a,niemela99a,schlipf95a}.
Following this methodology we separate the encoding
from an instance of the following problem:
given a (directed) graph, decide whether each node can be assigned
one of~$N$ colors such that any pair of adjacent nodes is colored differently.
Note that this problem is NP-complete for~$N\geq 3$
(see, e.g.,~\cite{papadimitriou94a}),
and thus it seems unlikely that a worst-case polynomial algorithm
can be found.
In view of this,
it is convenient to reduce the particular problem to
a declarative problem solving paradigm like ASP,
where efficient off-the-shelf tools like \gringo\ and \clasp\
are ready to solve the problem reasonably well.
Such a reduction is now exemplified.

\subsection{Problem Instance}\label{subsec:color:instance}
\input{figures/graph_intro}

We consider directed graphs specified via facts over predicates
\pred{node}/$1$ and \pred{edge}/$2$.%
The predicate \pred{color}/$1$ denotes the colors that shall be used to color the graph.
Figure~\ref{fig:graph_intro} depicts the graph specified by the following set of facts:
%
\lstinputlisting{examples/graph_intro.lp}

\subsection{Problem Encoding}\label{subsec:color:encoding}

We now proceed by encoding $N$-coloring via non-ground rules that are
independent of particular instances.
Typically, an encoding consists of a \emph{Generate}, a \emph{Define},
and a \emph{Test} part~\cite{lifschitz02a}.
As $N$-Coloring has a rather simple pattern, the following encoding does
not contain any Define part:
%
\lstinputlisting{examples/color.lp}
%
The Generate rule in Line~2 makes use of the \const{count} aggregate
(cf.\ Section~\ref{subsec:gringo:aggregate}).
For our example graph and~\const{1} substituted for~\var{X},
we obtain the following ground rule:%
\marginlabel{%
  The full ground program is obtained by invoking:\\
  \code{\mbox{~}gringo -t \textbackslash\\
    \mbox{~}examples/color.lp \textbackslash\\
    \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
1 { color(1,red), color(1,green), color(1,blue) } 1.
\end{lstlisting}
%
The \code{1} on the left hand side (right hand side) specifies that at least (at most) on atom within the curly brackets has to be true
that is~\code{\pred{node}(\const{1})} must have exactly one color.
Note that~\code{\pred{node}(\const{1})} has been removed from the body,
as it is derived via a corresponding fact and \code{color(1,C)} has been expanded using 
the colors \code{red}, \code{green}, and \code{blue} given by \pred{color}/$1$ specified in the instance.
Similar ground instances are obtained for the other nodes~\const{2} to~\const{6}.
With these six rules alone we already have a coloring the graph,
which does not yet ensure that two adjacent nodes have different colors.
This property is asserted by the rule in line four in the test part.
For each instance of \pred{edge}/2,
we obtain~\code{n} ground instances of the integrity constraint in Line~6,
prohibiting that the same color~\var{C} is assigned to the adjacent nodes.
We get the following ground instances due to \code{\pred{edge}(\const{1},\const{2})}:
%
\begin{lstlisting}[numbers=none]
:- color(1,red), color(2,red).
:- color(1,green), color(2,green).
:- color(1,blue), color(2,blue).
\end{lstlisting}
%
Again note that \code{\pred{edge}(\const{1},\const{2})},
derived via a fact, has been removed from the body.

\subsection{Problem Solution}\label{subsec:color:solution}
\input{figures/color}

Provided that a given graph is colorable with the colors given in the instance,
a solution can be read off an answer set of the program consisting
of the instance and the encoding.
For the graph in Figure~\ref{fig:graph_intro},
the following answer set can be computed:%
\marginlabel{%
  To find an answer set, invoke:\\
  \code{\mbox{~}gringo \textbackslash\\
        \mbox{~}examples/color.lp \textbackslash\\
        \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
        \mbox{~}clasp}\\
  or alternatively:\\
  \code{\mbox{~}clingo \textbackslash\\
        \mbox{~}examples/color.lp \textbackslash\\
        \mbox{~}examples/graph.lp}}
%
\begin{lstlisting}[numbers=none]
Answer: 1
... color(1,red)  color(2,green) color(3,green)
    color(4,blue) color(5,red)   color(6,blue)
\end{lstlisting}
%
Note that we have omitted the instances of \pred{color}/$1$, \pred{node}/$1$ 
and \pred{edge}/$2$ in order to emphasize the actual solution,
which is depicted in Figure~\ref{fig:color}.
Such output projection can also be specified within a logic program file by
using the declaratives \code{\#hide} and \code{\#show},
described in Section~\ref{subsec:gringo:meta}.

### 2010-08-10 doc/guide/errors.tex
- updated the guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3297 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-03 doc/documentation/errors.tex
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-syntax error on line @\textit{L}@ column @\textit{C}@ unexpected token: @\textit{T}@
+ERROR: parsing failed:
+        @\textit{File}@:@\textit{Line}@:@\textit{Column}@: unexpected token: @\textit{Token}@
 \end{lstlisting}
 %
-To correct this error, please investigate the line~\code{\textit{L}}
+To correct this error, please investigate the line~\code{\textit{Line}}
 and check whether something looks strange there
@@
 
-The next error occurs if an input program is not level-restricted
-(cf.\ Section~\ref{subsec:lambda}):
+The next error occurs if an input program is not safe:
 %
 \begin{lstlisting}[numbers=none,escapechar=@]
-the following rule cannot be grounded, \
-weakly restricted variables: { @\textit{V}@, ... }
-        @\textit{rule}@
+ERROR: unsafe variables in:
+@\textit{File}@:@\textit{Line}@:@\textit{Column}@: @\textit{Rule}@
+        @\textit{File}@:@\textit{Line}@:@\textit{Column}@: @\textit{Var}@
+        ...
 \end{lstlisting}
 %
-Along with the error message, the \code{\textit{rule}} and the
-name~\code{\textit{V}} of at least one variable causing the problem
+Along with the error message, the \code{\textit{Rule}} and the
+name~\code{\textit{Var}} of at least one variable causing the problem
 are reported.
 The first action to take usually consists of checking whether
-variable~\code{\textit{V}} is actually in the scope of any atom
+variable~\code{\textit{Var}} is actually in the scope of any atom
 (in the positive body of \code{\textit{rule}}) that can bind it.%
@@
   any variable.}
-If~\code{\textit{V}} is a local variable belonging to an atom~$A$ 
+If~\code{\textit{Var}} is a local variable belonging to an atom~$A$ 
 on the left-hand side of a condition (cf.\ Section~\ref{subsec:gringo:condition})
@@
 an atom over some domain predicate might be included in a condition
-to bind~\code{\textit{V}}.
+to bind~\code{\textit{Var}}.
 In particular, if~$A$ itself is over a domain predicate,
 the problem is often easily fixed by writing ``\code{$A$:$A$}.''
-If the problem is not as simple,
-then the predicates of all atoms that in principle may be used
-to bind~\code{\textit{V}} are involved in positive recursion
-through the predicate of some atom in the head of \code{\textit{rule}}.
-In this case, the positive bodies of some of the rules subject to recursion
-must be augmented with additional atoms over non-recursive predicates that
-can bind variables, in order to eventually establish level-restrictedness.
-As breaking recursion is a rather global matter, the reported
-\code{\textit{rule}} is not necessarily the best place to tackle it.
 
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-the following rule cannot be grounded, \
-non domain predicates : { @\textit{p}@(...), ... }
-        @\textit{rule}@
+ERROR: unstratified predicate in:
+@\textit{File}@:@\textit{Line}@:@\textit{Column}@: @\textit{Rule}@
+        @\textit{File}@:@\textit{Line}@:@\textit{Column}@: @\textit{Predicate}@/@\textit{Arity}@
 \end{lstlisting}
 %
-The problem is that an atom \code{\textit{p}(...)} such
-that its predicate \code{\textit{p}/\!\textit{i}} is not a domain predicate
-(cf.\ Section~\ref{subsec:strat}) is used on the right-hand side of a condition
-within \code{\textit{rule}}.
+The problem is that an atom \code{\textit{Predicate}(...)} such
+that its predicate \code{\textit{Predicate}/\!\textit{Arity}} is not a domain predicate
+(cf.\ Section~\ref{subsec:gringo:condition}) is used on the right-hand side of a condition
+within \code{\textit{Rule}}.
 The error is corrected by either removing the atom or
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-trying to convert string to int
-trying to convert functionsymbol to int
+ERROR: cannot convert @\textit{Term}@ to integer in:
+        @\textit{File}@:@\textit{Line}@:@\textit{Column}@: @\textit{Literal}@
 \end{lstlisting}
@@
 has occurred in the scope of some built-in arithmetic function.
-Unfortunately, no context information is provided yet,
-which is left as an issue to future work
-(cf.\ Section~\ref{sec:future}).
-
-The below errors are related to built-in comparison predicates
-(cf.\ Section~\ref{subsec:gringo:comp}):
-%
-\begin{lstlisting}[numbers=none,escapechar=@]
-comparing different types
-comparing function symbols
-\end{lstlisting}
-%
-The first error indicates that one of the built-in comparison predicates
-\code{<}, \code{<=}, \code{>}, and \code{>=}
-has been applied to an integer or a (symbolic) constant
-and a term of a different sort,
-while the second error means that compound terms have occurred in such a comparison.
-The mere reason for these errors is that the required functionalities
-have not been implemented yet,
-so that the errors are likely to disappear in the future
-(cf.\ Section~\ref{sec:future}).
-
-An error occurs if a constant~\code{\textit{c}}
-ought to be replaced with a term~\code{\textit{t}}
-containing~\code{\textit{c}}:
-%
-\begin{lstlisting}[numbers=none,escapechar=@]
-cyclic constant definition
-\end{lstlisting}
-%
-To resolve such an error, check the occurrences of \const{\#const}
-in the input as well as \code{--const} or \code{-c} options
-in the command line, and change them in a way such that a constant
-that is to be replaced does not (transitively) occur within a term
-to be inserted.
-
-The next error may occur if a statement from the input language
-of \iclingo\ (cf.\ Section~\ref{subsec:lang:iclingo})
-is used in the input of \gringo, \clingo, or
-\iclingo\ run non-incrementally
-(via one of options \code{--text}, \code{--lparse}, \code{--aspils},
- or \code{--clingo}):
-%
-\begin{lstlisting}[numbers=none,escapechar=@]
-A fixed number of incremental steps is needed to ground the program.
-\end{lstlisting}
-%
-This problem is solved by providing a step number
-via option \code{--ifixed}.
-
-The following error may be reported by \iclingo:
-%
-\begin{lstlisting}[numbers=none,escapechar=@]
-The following statement cant be used with the incremental interface:
-        @\textit{rule}@
-\end{lstlisting}
-%
-The \code{\textit{rule}} contains a \const{\#maximize},
-\const{\#minimize}, or \const{\#compute} statement,
-which are currently not supported within incremental computations.
-Their support by \iclingo\ is an issue to future work
-(cf.\ Section~\ref{sec:future}).
-
-The next error may occur if \emph{ASPils}~\cite{gejaosscth08a}
-output is requested via option \code{--aspils}:
-%
-\begin{lstlisting}[numbers=none,escapechar=@]
-@\textit{e}@ not allowed in this normal form, \ 
-please choose normal form @\textit{n}@.
-\end{lstlisting}
-%
-The chosen normal form (in-between~\code{1} and~\code{7})
-does not cover the input language expression~\code{\textit{e}},
-and one among the normal forms~\code{\textit{n}} suggested
-in the error message ought to be provided as an argument instead.
 
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-Input Error at line @\textit{L}@: Symbol table expected!
+ERROR: Read Error: Line 2, Compute Statement expected!
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-Input Error at line @\textit{L}@: Atom out of bounds
+ERROR: Read Error: Line @\textit{Line}@, Atom out of bounds
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-Input Error at line @\textit{L}@: Unsupported rule type!
+ERROR: Read Error: Line @\textit{Line}@, Unsupported rule type!
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-sorry clasp cannot handle disjunctive rules!
+ERROR: Error: clasp cannot handle disjunctive rules use option --shift!
 \end{lstlisting}
@@
 (cf.\ Section~\ref{sec:future}).
+Furthermore, if your program is head-cycle free, 
+you might want to try \gringo's \code{--shift} option (see \ref{subsec:opt:gringo}).
 
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-unknown option: @\textit{o}@
-ambiguous option: '@\!\textit{o}@' could be:
-  @\textit{o1}@
-  @\textit{o2}@
+ERROR: unknown option: @\textit{Option}@
+ERROR: ambiguous option: '@\!\textit{Option}@' could be:
+  @\textit{Option1}@
+  @\textit{Option2}@
   ...
-'@\!\textit{a}@': invalid value for Option '@\!\textit{o}@'
+ERROR: '@\!\textit{Arg}@': invalid value for Option '@\!\textit{Option}@'
 \end{lstlisting}
 %
-The first error means that a provided option~\code{\textit{o}}
+The first error means that a provided option~\code{\textit{Option}}
 could not be expanded to one that is recognized,
-while the second error expresses that the result of expanding~\code{\textit{o}}
+while the second error expresses that the result of expanding~\code{\textit{Option}}
 is ambiguous.
-Finally, the third error occurs if a provided argument~\code{\textit{a}}
-is invalid for option~\code{\textit{o}}.
+Finally, the third error occurs if a provided argument~\code{\textit{Arg}}
+is invalid for option~\code{\textit{Option}}.
 In either case, option \code{--help} can be used to see 
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-Warning: @\textit{p}@/@\!\textit{i}@ is never defined.
-Warning: sum with negative bounds in the body of a rule
-Warning: multiple definitions of #const @\textit{c}@
+% warning: @\textit{p}@/@\!\textit{i}@ is never defined
 \end{lstlisting}
 %
-The first one, stating that a predicate \code{\textit{p}/\!\textit{i}}
+This warning, states that a predicate \code{\textit{p}/\!\textit{i}}
 has occurred in some rule body, but not in the head of any rule,
 might point at a mistyped predicate.
-The second warning means that negative weights within a \const{sum} aggregate
-(cf.\ Section~\ref{subsec:gringo:aggregate})
-have been compiled away~\cite{siniso02a},
-which is semantically delicate~\cite{ferraris05a}.
-Thus, we strongly recommend to avoid negative weights
-within a \const{sum} aggregate in the positive body of a rule.
-Finally, the third warning recognizes that multiple \const{\#const}
-statements over the same constant~\code{\textit{c}} are contained in the input,
-while only the first among them will be considered.
-
-The next four warnings may be issued by \iclingo:
-\begin{lstlisting}[numbers=none,escapechar=@]
-Warning: There are no #base, #cumulative or #volatile sections.
-Warning: Option ifixed will be ignored!
-Warning: There are statements not within a #base, \
-         #cumulative or #volatile section.
-         These Statements are put into the #base section.
-Warning: Classical negation is not handled correctly \
-         in combination with the incremental output.
-         You have to add rules like: :- a, -a. on your own! \
-         (at least for now)
-\end{lstlisting}
-%
-For omitting the first warning, one may use option \code{--clingo},
-as the input program contains no statements particular to \iclingo\
-(cf.\ Section~\ref{subsec:lang:iclingo}).
-Similarly, option \code{--clingo} may be used to suppress the second warning,
-stating that a provided option \code{--ifixed} is ignored in incremental mode
-(if \iclingo\ is not run as a grounder 
- via one of options \code{--text}, \code{--lparse}, or \code{--aspils}).
-The third warning states that rules have been inserted into the static
-program part without a 	preceding \const{\#base} statement.
-This warning can be avoided by ending an encoding with \const{\#base.}
-and by supplying an instance file afterwards, as illustrated in Section~\ref{subsec:ex:block}.
-The fourth warning refers to the fact that there currently is no
-built-in support for classical negation (cf.\ Section~\ref{subsec:gringo:negation})
-by \iclingo.
-Hence, integrity constraints of the form ``$\code{:-}~A\code{,}\code{-}A\code{.}$''
-have to be added explicitly by a user.
 

### 2010-08-10 doc/guide/examples.tex
- updated the guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3297 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-09 doc/documentation/examples.tex
@@
 
-\subsection{$N$-Coloring}\label{subsec:ex:color}
+\subsection{\texorpdfstring{$N$}{N}-Coloring}\label{subsec:ex:color}
 
-As already mentioned in Section~\ref{subsec:asp},
+As already mentioned in Section~\ref{sec:quickstart},
 it is custom in ASP to provide a \emph{uniform}
@@
 (see, e.g.,~\cite{papadimitriou94a}),
-and thus it seems unlikely that a worst-case polynomial algorithm
+and thus it seems unlikely that a worst-case polynomial time algorithm
 can be found.
@@
   \code{\mbox{~}gringo -t \textbackslash\\
-    \mbox{~}examples/color.lp \textbackslash\\
-    \mbox{~}examples/graph.lp}}
+    \mbox{~}\attach{examples/color.lp}{color.lp} \attach{examples/graph.lp}{graph.lp}}}
 %
@@
   To find an answer set, invoke:\\
-  \code{\mbox{~}gringo \textbackslash\\
-        \mbox{~}examples/color.lp \textbackslash\\
-        \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
-        \mbox{~}clasp}\\
+  \code{\mbox{~}gringo \attach{examples/color.lp}{color.lp} \textbackslash \\
+  \mbox{~}\attach{examples/graph.lp}{graph.lp} | clasp}\\
   or alternatively:\\
   \code{\mbox{~}clingo \textbackslash\\
-        \mbox{~}examples/color.lp \textbackslash\\
-        \mbox{~}examples/graph.lp}}
+        \mbox{~}\attach{examples/color.lp}{color.lp} \attach{examples/graph.lp}{graph.lp}}}
 %
@@
   \code{\mbox{~}gringo -t \textbackslash\\
-    \mbox{~}examples/ham.lp \textbackslash\\
-    \mbox{~}examples/min.lp \textbackslash\\
-    \mbox{~}examples/costs.lp \textbackslash\\
-    \mbox{~}examples/graph.lp}}
+    \mbox{~}\attach{examples/ham.lp}{ham.lp} \attach{examples/min.lp}{min.lp} \textbackslash\\
+    \mbox{~}\attach{examples/costs.lp}{costs.lp} \attach{examples/graph.lp}{graph.lp}}}
 %
@@
 in which case a ground program is called \emph{non-tight}~\cite{erdlif03a,fages94a}.
-The ``correct'' treatment of (positive) recursion due to the justification
-requirement w.r.t.\ the reduct (cf.\ Section~\ref{subsec:semantics})
+The ``correct'' treatment of (positive) recursion 
 is a particular feature of answer set semantics,
@@
   \code{\mbox{~}gringo \textbackslash\\
-        \mbox{~}examples/ham.lp \textbackslash\\
-        \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
+        \mbox{~}\attach{examples/ham.lp}{ham.lp} \attach{examples/graph.lp}{graph.lp} |\rlap{\:\textbackslash}\\
         \mbox{~}clasp -n 0}\\
@@
   \code{\mbox{~}clingo -n 0 \textbackslash\\
-        \mbox{~}examples/ham.lp \textbackslash\\
-        \mbox{~}examples/graph.lp}}
+        \mbox{~}\attach{examples/ham.lp}{ham.lp} \attach{examples/graph.lp}{graph.lp}}}%
 Finally, the additional Display part in Line~10 and~11 states that answer
@@
   \code{\mbox{~}gringo \textbackslash\\
-    \mbox{~}examples/ham.lp \textbackslash\\
-    \mbox{~}examples/min.lp \textbackslash\\
-    \mbox{~}examples/costs.lp \textbackslash\\
-    \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
+    \mbox{~}\attach{examples/ham.lp}{ham.lp} \attach{examples/min.lp}{min.lp} \textbackslash\\
+    \mbox{~}\attach{examples/costs.lp}{cost.lp} \attach{examples/graph.lp}{graph.lp} |\rlap{\:\textbackslash}\\
     \mbox{~}clasp -n 0}\\
@@
   \code{\mbox{~}clingo -n 0 \textbackslash\\
-    \mbox{~}examples/ham.lp \textbackslash\\
-    \mbox{~}examples/min.lp \textbackslash\\
-    \mbox{~}examples/costs.lp \textbackslash\\
-    \mbox{~}examples/graph.lp}}
+    \mbox{~}\attach{examples/ham.lp}{ham.lp} \attach{examples/min.lp}{min.lp} \textbackslash\\
+    \mbox{~}\attach{examples/costs.lp}{cost.lp} \attach{examples/graph.lp}{graph.lp}}}
 %
@@
   \code{\mbox{~}gringo \textbackslash\\
-    \mbox{~}examples/ham.lp \textbackslash\\
-    \mbox{~}examples/min.lp \textbackslash\\
-    \mbox{~}examples/costs.lp \textbackslash\\
-    \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
+    \mbox{~}\attach{examples/ham.lp}{ham.lp} \attach{examples/min.lp}{min.lp} \textbackslash\\
+    \mbox{~}\attach{examples/costs.lp}{costs.lp} \attach{examples/graph.lp}{graph} |\rlap{\:\textbackslash}\\
     \mbox{~}clasp -n 0 \textbackslash\\
@@
     \mbox{~}--opt-all \textbackslash\\
-    \mbox{~}examples/ham.lp \textbackslash\\
-    \mbox{~}examples/min.lp \textbackslash\\
-    \mbox{~}examples/costs.lp \textbackslash\\
-    \mbox{~}examples/graph.lp}}
+    \mbox{~}\attach{examples/ham.lp}{ham.lp} \attach{examples/min.lp}{min.lp} \textbackslash\\
+    \mbox{~}\attach{examples/costs.lp}{costs.lp} \attach{examples/graph.lp}{graph}}}%
 After obtaining only the second answer given above,
@@
 \footnote{%
-  Blocks-World instances ``\code{examples/world$i$.lp}'' for
-  $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$
+  Blocks-World instances \code{world$i$.lp} for $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$
   are adaptations of the instances provided at~\cite{erdemBW}.}
@@
 if it is not moved to any other location~\var{Y}.
-Note that we require atoms \code{\pred{block}(\var{X})} and \code{\pred{location}(\var{Z})}
-to bind variables~\var{X} and~\var{Z} in the body of the latter rule,
-as recursive predicate \pred{holds}/$2$ cannot be used for this purpose
-(cf.\ the definition of level-restrictedness in Section~\ref{subsec:lambda}).
 
@@
 This is also the default of \iclingo\
-(as well as of \gringo\ and \clingo\ that can be used for non-incremental computations),
-but applying this default behavior triggers a warning (cf.\ Section~\ref{subsec:warn}).%
+(as well as of \gringo\ and \clingo\ that can be used for non-incremental computations).%
 \marginlabel{%
@@
         \mbox{~}--ifixed=$n$ \textbackslash\\
-        \mbox{~}examples/blocks.lp \textbackslash\\
-        \mbox{~}examples/world$i$.lp}
-  where $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$.}
+        \mbox{~}\attach{examples/blocks.lp}{blocks.lp} \attach{examples/world0.lp}{world0.lp}}\\
+  Furthermore you can try: \\
+  \code{\attach{examples/world1.lp}{world1.lp}},  
+  \code{\attach{examples/world2.lp}{world2.lp}}, \\
+  \code{\attach{examples/world3.lp}{world3.lp}}, 
+  \code{\attach{examples/world4.lp}{world4.lp}}.}
 
@@
   \code{\mbox{~}iclingo -n 0 \textbackslash\\
-        \mbox{~}examples/blocks.lp \textbackslash\\
-        \mbox{~}examples/world0.lp}}
+        \mbox{~}\attach{examples/blocks.lp}{blocks.lp} \attach{examples/world0.lp}{world0.lp}}
+  Furthermore you can try: \\
+  \code{\attach{examples/world1.lp}{world1.lp}},  
+  \code{\attach{examples/world2.lp}{world2.lp}}, \\
+  \code{\attach{examples/world3.lp}{world3.lp}}, 
+  \code{\attach{examples/world4.lp}{world4.lp}}.}
 %
@@
   \code{\mbox{~}gringo --ifixed=$n$ \textbackslash\\
-        \mbox{~}examples/blocks.lp \textbackslash\\
-        \mbox{~}examples/world$i$.lp |\rlap{\textbackslash}\\
+        \mbox{~}\attach{examples/blocks.lp}{blocks.lp} \textbackslash\\
+        \mbox{~}\attach{examples/world0.lp}{world0.lp} |\rlap{\textbackslash}\\
         \mbox{~}clasp -n 0}\\
@@
         \mbox{~}--ifixed=$n$ \textbackslash\\
-        \mbox{~}examples/blocks.lp \textbackslash\\
-        \mbox{~}examples/world$i$.lp}
-  where $i\in\{\code{0},\code{1},\code{2},\code{3},\code{4}\}$.}
+        \mbox{~}\attach{examples/blocks.lp}{blocks.lp} \attach{examples/world0.lp}{world0.lp}}}
 enabling \gringo\ or \clingo\ to generate the ground program present
@@
 
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "guide"
-%%% End:
+% %%% Local Variables: 
+% %%% mode: latex
+% %%% TeX-master: "guide"
+% %%% End:

### 2010-08-10 doc/guide/introduction.tex
- updated the guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3297 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-03 doc/documentation/introduction.tex
@@
 combinations thereof within integrated systems \clingo\ and \iclingo.
-All these tools are published under GNU General Public License(s)~\cite{GNUgpl}.
+All these tools are written in \texttt{C++} and published under GNU General Public License(s)~\cite{GNUgpl}.
 Source packages as well as precompiled binaries for Linux and Windows are available at~\cite{potassco}.
-Note that there currently are two source packages:
-one containing \clasp, and another one grouping \gringo, \clingo, and \iclingo.
 For building one of the tools from sources,
@@
 If you nonetheless encounter problems in the building process,
-please do not hesitate to contact the authors of this guide.
+please use the potassco mailing list 
+\href{mailto:potassco-users@lists.sourceforge.net}{potassco-users@lists.sourceforge.net}.
 
@@
 \clingo\ and \iclingo\ are all available,
-one usually provides the filenames of input text files to either
+one usually provides the file names of input text files to either
 \gringo, \clingo, or \iclingo, while the output of \gringo\ is
@@
 \begin{lstlisting}[numbers=none]
-gringo  [ options | filenames ] | clasp [ options | number ]
-clingo  [ options | filenames | number ]
-iclingo [ options | filenames | number ]
+gringo  [ options | files ] | clasp [ options | number ]
+clingo  [ options | files | number ]
+iclingo [ options | files | number ]
 \end{lstlisting}
@@
 The outline is as follows.
-In Section~\ref{sec:background},
-we describe the basics of Answer Set Programming,
-and we formally introduce the syntax and semantics of logic programs.
-Section~\ref{sec:restricted} details restrictedness notions,
-important when dealing with logic programs containing first-order variables.
+In Section~\ref{sec:quickstart},
+an introductory example is given
+that serves both as guideline on how to model problems using logic programs
+and also as an example on how compact on concise the modeling language of \gringo is.
 The probably most important part for a user, Section~\ref{sec:language},
@@
 as well as input-related errors and warnings that may be reported.
-Finally, we conclude with some remarks on future work in Section~\ref{sec:future}.
+During the guide we forgo most of the theoretical background in favor of
+of small intuitive examples and informal descriptions.
+%For a more theoretical background, the interested reader is referred
+%to Appendix~\ref{sec:background} where technical details are covered.
 

### 2010-08-10 doc/guide/language.tex
- rearanged math terms
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3214 3e02f596-3131-0410-8352-e04eeaec37cc
- in clingo/iclingo all atoms get a default name (0) by now to make them accessible via the atomindex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3215 3e02f596-3131-0410-8352-e04eeaec37cc
- updated the guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3297 3e02f596-3131-0410-8352-e04eeaec37cc

--- 2008-11-09 doc/documentation/language.tex
@@
 it takes care of both grounding and solving.
-% The input of \gringo\ and \clingo\ is the same,
-% but 
 In contrast to \gringo\ outputting ground programs, 
 \clingo\ returns answer sets.
-Both \gringo\ and \clingo\ can handle level-restricted input programs
-(cf.\ Section~\ref{subsec:lambda}),
-usually specified in one or more text files whose names are passed via the command line
+Usually logic programs are specified in one or more text files whose names are passed via the command line
 in an invocation of either \gringo\ or \clingo.
@@
 
-In the previous sections, we have already seen a number of normal programs.
-Now also considering integrity constraints, we get the following basic rule types:
-
-\begin{tabular}{rl@{}l}
+\begin{figure}
+\begin{rail}
+	constant    : ('\_'*) '[a-z]' ('[A-Za-z0-9]'*);
+	variable    : ('\_'*) '[A-Z]' ('[A-Za-z0-9]'*);
+	simpleterm  : ('integer' | constant | variable | '\_' | '\#supremum' | '\#infimum');
+	function    : constant '(' simpleterm (',' simpleterm*) ')';
+	term        : simpleterm | function;
+\end{rail}
+\caption{Terms\label{fig:terms}}
+\end{figure}
+
+Every logic program is constructed from terms.
+An overview of \gringo{} terms is depicted in Figure \ref{fig:terms}.
+The most basic terms are integers, constants, and variables.
+Furthermore, there are some special variables and constants.
+An anonymous variable denoted by `\_' is simialar to a normal variable 
+but each occurrence is treated like a different variable 
+(intuitively a new unique variable name is substituted).
+Additionally, there are the two special constants `\#supremum' and `\#infimum'
+representing the largest and the smallest possible values, respectively,
+which behave essentially like constants.
+Finally, there are function symbols which are composed of other terms.
+A term that does not contain any (anonymous) variables is called a ground term.
+More complex terms involving arithmetics and other constructs
+are introduced later on.
+
+\begin{tabular}{rl@{}l}\label{eq:normal:rule}
+\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
+\\
 \textbf{Fact:} & $A_0$&\code{.}
-\\
-\textbf{Rule:} & $A_0$& $\code{:-}~L_1\code{,}\dots\code{,}L_n$\code{.}
 \\
@@
 \noindent
-The head~$A_0$ of a fact or a rule is an \emph{atom} of the form
-$p\code{(}t_1\code{,}\dots\code{,}t_m\code{)}$,
-where $p$ is the name of some predicate, that is, a sequence of letters and digits
-starting with a lowercase letter,
-and any $t_i$ is a term.%
-\footnote{An atom~$p$ without arguments is simply a sequence of letters and digits
-  (starting with a lowercase letter).
-  For such an atom~$p$, parentheses after the name are skipped, e.g.,
-  \pred{p42X} could be an atom without arguments.
-  Also note that lowercase letters include ``\code{\mus},''
-  which can thus be used at any position within a predicate name.}
-A \emph{term} $t$ starting with an uppercase letter followed by
-a sequence of letters and digits (e.g., \var{X08x15}) is a variable name,
-and integers are written as sequences of digits possibly preceded by ``\code{-}.''
-In addition, a term can have the same syntactic structure as an atom
-(e.g., \code{\pred{p}(\const{a},\const{1},\const{f}(\var{X}))}
- can be either an atom or a term, depending on where it occurs in a rule),
-and functions can be nested within a term.
-There are also built-in constructs
-(cf.\ Section~\ref{subsec:gringo:arith} and~\ref{subsec:gringo:comp})
-having particular representations.
-Finally, any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
-for an atom~$A$.
-
-In Section~\ref{subsec:semantics},
-we have already provided a translation~$\psi$
-from the head of a rule to a propositional formula.
-By viewing an integrity constraint
-$\code{:-}~\mathit{Body}\code{.}$ as a shorthand for $\bot~\code{:-}~\mathit{Body}\code{.}$,
-we can now explain the semantics of integrity constraints by adding the following case:
-\begin{itemize}
-\item $\psi[\bot]=\bot$.
-\end{itemize}
-Note the role of integrity constraints is to eliminate answer set candidates.
-In fact, given an integrity constraint~$\code{:-}~\mathit{Body}\code{.}$,
-any answer set~$X$ is such that $X\not\models\phi[\mathit{Body}]$, or
-in other words, some literal~$L$ in $\mathit{Body}$ must be unsatisfied w.r.t.~$X$.
+The head~$A_0$ of a fact or a rule is an \emph{atom} of the same form
+as a function symbol or constant.
+Any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
+for an atom~$A$ where the connective \code{not} corresponds to default negation.
+The set of literals $\{L_1,\dots,L_n\}$ is called the body of the rule.
+Facts have an empty body.
+Throughout this section we will further extend the predicates that can be used 
+in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
+Furthermore, \gringo{} expects rules to be save, i.e., 
+all variables that appear in a rule have to appear in some some positive literal (a literal not preceded by \code{not}) in the body.
+If a variable appears positively in some predicate, then we say that this predicate binds the variable.
+
+Intuitively, the head of a rule has to be true whenever all its body literals are true.
+In ASP every predicate needs some derivation, i.e., 
+a predicate cannot be true if there is no body deriving it.
+This implies that only predicates appearing in some head can appear in answer sets.
+Furthermore, derivations
+\footnote{There are extensions like disjunctions 
+that go beyond simple derivability and also require minimality w.r.t. a reduct. 
+We will not cover the semantics of such constraints in this guide.}%
+have to be acyclic,
+a feature that is important to model reachability.
+As a simple example consider the program \code{a :- b. b :- a.}
+The only answer set to this program is the empty set.
+Adding either \code{a.} or \code{b.} to the program results in the answer set $\{a,b\}$.
+Finally, note that default negation is ignored when checking for acyclic derivations 
+(we do not need a reason for a predicate being false).
+Default negation can be used to express choices, e.g., the program \code{a :- not b. b :- not a.} 
+has the two answer sets $\{a\}$ and $\{b\}$.
+But in practice it is never needed to express choices this way.
+For example in the introductory example in Section~\ref{sec:quickstart} we used a cardinality 
+constraint, which provides a much more readable way to introduce choices.
+
+A fact has an empty body and thus its associated head predicate is always true 
+and will appear in all answer sets.
+On the other hand, integrity constraints eliminate answer set candidates.
+They are merely tests that discard unwanted answer sets.
+That is, there will be no answer sets that satisfy all literals an integrity constraint.
 Elaborate examples on the usage of facts, rules, and integrity constraints
@@
 with the additional condition that~$A$ and~$\code{-}A$ must not jointly hold.
-A logic program~$\Pi$ containing complementary atoms is thus translated
-to propositional theory
-$\Phi[\Pi \cup \{\code{:-}~A\code{,}\code{-}A \mid A\in\ground{\mathcal{A}}\}]$.%
-\footnote{Recall that $\ground{\mathcal{A}}$ consists of all variable-free atoms
-built from predicates and functions in~$\Pi$.}
 Observe that classical negation is merely a syntactic feature that can be
@@
 \begin{example}\label{ex:flies:neg}
-The following logic program reformulates the one in Example~\ref{ex:flies}:
+Consider a logic program comprising the following facts:
 %
@@
   \code{\mbox{~}gringo -t \textbackslash\\
-        \mbox{~}examples/bird.lp \textbackslash\\
-        \mbox{~}examples/flycn.lp}
+        \mbox{~}\attach{examples/bird.lp}{bird.lp}~\attach{examples/flycn.lp}{flycn.lp}}
 the reader can observe that \gringo\ indeed produces the integrity constraint
@@
 \end{lstlisting}
-The additional integrity constraints do not yet change the semantics of the
-program, which still has the answer sets already provided in Example~\ref{ex:flies:ground}
-(of course, identifying
-\code{\pred{neg\mus flies}(\const{tux})} with
-\code{\pred{-flies}(\const{tux})} and
-\code{\pred{neg\mus flies}(\const{tweety})} with
-\code{\pred{-flies}(\const{tweety})}).
-This situation changes if we add the following fact:
+The program has two answer sets.
+One contains \code{flies(tweety)} and the other contains \code{-flies(tweety)}.
+Let us now add a new fact to the program:
 \begin{lstlisting}[firstnumber=8]
@@
 \end{lstlisting}
-While the program from Example~\ref{ex:flies} still admits two answer sets,
-both containing 
-\code{\pred{flies}(\const{tux})} and
-\code{\pred{neg\mus flies}(\const{tux})},
-there no longer is any answer set for our new program using classical negation.
+There no longer is any answer set for our new program using classical negation.
 In fact, answer set candidates that contain both
@@
 Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
-An additional case of translation~$\psi$ from Section~\ref{subsec:semantics}
-reflects the semantics of disjunction:
-\begin{itemize}
-\item $\psi[G\code{|}H]=\psi[G]\vee\psi[H]$.
-\end{itemize}
-%
-The concept of level-restrictedness (cf.\ Section~\ref{subsec:lambda}) is extended
-to disjunctive programs by,
-for a disjunctive rule~$r$, every predicate~$p/i$
-of some atom in the head of~$r$, and each $V\in\vars{r}$,
-requiring that there is some $A\in\binders{V}{r}$ such that
-$\lambda(p/i)>\lambda(q/j)$ for predicate~$q/j$ of~$A$.
-Furthermore, a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
-must not contain any rule with a (non-trivial) disjunctive head.
-
-The following rule combines the ones in Line~3 and~4 from Example~\ref{ex:flies:neg}:
-\begin{lstlisting}[numbers=none]
-flies(X) | -flies(X) :- bird(X).
-\end{lstlisting}
+A disjunction is true if at least one of its atoms is true.
+Additionally, logic programs have to satisfy a minimality criterion, 
+which we will not detail in this guide.
+The simple program \code{a | b.} has the two answer sets $\{a\}$ and $\{b\}$ but
+does not admit the answer set ${a,b}$ because it is no minimal model.
+
 In general, the use of disjunction however increases
@@
 \code{+} (addition),
-\code{-} (subtraction),
+\code{-} (subtraction, unary minus),
 \code{*} (multiplication),
-\code{/} or \code{div} (integer division),
-\code{mod} (modulo function),
-\code{abs} (absolute value),
-\code{\textasciitilde} (bitwise complement),
+\code{/} or \code{\#div} (integer division),
+\code{\textbackslash} or \code{\#mod} (modulo function),
+\code{**} or \code{\#pow} (exponentation),
+\code{|$\cdot$|} or \code{\#abs} (absolute value),
 \code{\&} (bitwise AND),
-\code{?}   (bitwise OR), and
-\code{\^} (bitwise exclusive OR).
+\code{?} (bitwise OR),
+\code{\^} (bitwise exclusive OR), and
+\code{\textasciitilde} (bitwise complement).
 
@@
   can be inspected by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithf.lp}}
+  \code{\mbox{~}gringo -t \attach{examples/arithf.lp}{arithf.lp}}}
 %
@@
 spaces can also be dropped,
-while spaces around tokens \code{div} and \code{mod} are mandatory.
-Furthermore, the argument of function \code{abs} must be enclosed in parentheses.
-In Line~9, observe that there is a unary version of \code{-},
-``\code{- \var{R}}'' standing for ``\code{0 - \var{R}}.''
+while spaces after tokens \code{\#div} and \code{\#mod} are mandatory.
+Furthermore, the argument of function \code{\#abs}, \code{\#div}, and \code{\#mod} must be enclosed in parentheses.
 The four bitwise functions apply to signed integers,
@@
 It is important to note that variables
-in the scope of an arithmetic function are not necessarily bound 
-(in the sense of Section~\ref{subsec:lambda}) by a corresponding atom.
-For instance, atom \code{\pred{p}(\var{X}+1,\var{X})} belongs to
-$\binders{\var{X}}{r}$, while atom \code{\pred{p}(\var{X}+1,\var{Y})}
-does not belong to $\binders{\var{X}}{r}$ because it contains~\var{X}
-only in the scope of an arithmetic function.
+in the scope of an arithmetic function are not bound by a corresponding atom.
+For instance, the rule \code{p(X) :- p(X+1).} is not safe but \code{p(X-1) :- p(X).} is.
+Although, the latter might produce an infinite grounding 
+and \gringo{} will not necessarily halt when given such an input.
 
@@
 The following built-in predicates permit term comparisons
-within the bodies of rules and on the right-hand side of conditions
-(cf.\ Section~\ref{subsec:gringo:condition}):
+within the bodies of rules:
 \code{==} (equal),
@@
   The unique answer set of the program is obtained via call:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/arithc.lp}}
-%
-\lstinputlisting{examples/arithc.lp}
+  \code{\mbox{~}gringo -t \attach{examples/arithf.lp}{arithf.lp}}}
+%
+\lstinputlisting{examples/arithf.lp}
 %
@@
 
-All comparison predicates can also be used with constants,
+All comparison predicates can also be used with arbitrary ground terms,
 as in the next program:%
@@
   As above, invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/symbc.lp}
+  \code{\mbox{~}gringo -t \attach{examples/symbc.lp}{symbc.lp}}
   yields the unique answer set of the program in terms of facts.}
@@
 %
-Finally, note that~\code{==} and~\code{!=} also apply to
-compound terms over functions with non-zero arity as well as to
-mixed integer and non-integer arguments.%
-\footnote{%
-  Such functionalities are currently not supported for the other four
-  comparison predicates,
-  where (after instantiation and arithmetic evaluation)
-  both arguments must be either integers or (symbolic) constants.
-}
-\eexample
-\end{example}
-
-Importantly, a built-in comparison predicate cannot bind variables,
-that is, it does not belong to $\binders{\var{X}}{r}$ for any variable~\var{X}.
-Also note that built-in predicates are not considered by level mappings~$\xi$
-in the context of stratification (cf.\ Section~\ref{subsec:strat}), or equivalently,
-one may assume that built-in predicates are mapped to level~$0$ and all other
-predicates to levels greater than~$0$.
+Integers are compared in the usual way and constants are ordered lexicographically.
+Function symbols are compared first using their arity.
+If the arity differs, then the name of the function symbol is compared lexicographically.
+If again the name differs, then arguments are compared component wise.
+Finally, integers are always smaller than constants and constants are always smaller than function symbols.
+\eexample
+\end{example}
+
+Importantly, a built-in comparison predicate cannot bind variables, i.e.,
+when checking whether a rule is safe, 
+comparison predicates are not considered to be positive.
 
@@
 
-Built-in predicate~\code{=} can be used in the body of a rule
-(or on right-hand sides of conditions introduced in Section~\ref{subsec:gringo:condition})
-to assign a term on its right-hand side to a variable on its left-hand side.
-Note that ``\code{\var{X} \!=\! }$t$'' belongs to~$\binders{\var{X}}{r}$,
-provided that~$t$ is variable-free or that its variables are bound
-by atoms other than assignments with~\var{X} on the right-hand side.
-Cyclic assignments over otherwise unbound variables are thus excluded.
+The built-in predicates~\code{:=} and ~\code{=} can be used in the body of a rule
+to unify a term on their right-hand side 
+to a (non-ground) term or variable on its left-hand side, respectively.
 
@@
   The unique answer set of the program is obtained via call:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/assign.lp}}
+  \code{\mbox{~}gringo -t \attach{examples/assign.lp}{assign.lp}}}%
 %
@@
 \end{example}
+
+\begin{example}\label{ex:unify}
+The second program demonstrates the usage of \code{:=},
+which allows for terms on the left hand side:
+\marginlabel{%
+  The unique answer set of the program is obtained via call:\\
+  \code{\mbox{~}gringo -t \attach{examples/unify.lp}{unify.lp}}}%
+%
+\lstinputlisting{examples/unify.lp}
+%
+Here the term \code{f(a,X,X+1)} is unified with every function symbol provided by \code{sym}/$1$.
+Note the usage of \code{X+1} in the term.
+\gringo{} will not try to unify any term containing arithmetic 
+but in this example \code{X} occurs also directly as second argument of the argument
+and can thus be unified with.
+The term $X+1$ is merely a test that deferred and checked later.
+For example, the fourth line is equivalent to:
+\begin{lstlisting}[firstnumber=6]
+unifyf(X) :- f(a,X,Y) := F, sym(F), Y = X + 1.
+\end{lstlisting}%
+\eexample
+\end{example}
+
+Finally, note that assignments to some extend can bind variables.
+Of course cyclic assignments cannot bind variables.
+For example the rule \code{p(X) :- X = Y, Y = X.} is rejected by \gringo{}.
+Either \code{X} or \code{Y} has to be provided by some positive predicate in this case.
+Additionally, the unification is restricted to ground terms on the right hand side of the assignment, 
+that is all variables on the right hand side have to be by some other predicate.
 
@@
   Again the unique answer set is obtained via call:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/int.lp}}
+  \code{\mbox{~}gringo -t \attach{examples/int.lp}{int.lp}}}
 %
@@
   The reader can reproduce these ground rules by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/cond.lp}}
+  \code{\mbox{~}gringo -t \attach{examples/cond.lp}{cond.lp}}}
 \code{\pred{available}(\var{X})} with all ground terms~$t$ such that
@@
 First, all predicates of atoms on the right-hand side of a condition
-must be either domain predicates (cf.\ Section~\ref{subsec:strat}) or built-in,
-which is due to the fact that conditions are evaluated during grounding.%
-\footnote{%
-  The bodies of rules in a stratified subprogram (cf.\ Section~\ref{subsec:strat})
-  may contain conditions.
-  For a rule~$r$ and
-  $L_0\code{:}L_1\code{:}\dots\code{:}L_n$
-  in the body of~$r$,
-  let $A_0\in\nbody{r}$ if $L_0=\code{not}~A_0$
-  and $L_0\in\pbody{r}$ otherwise,
-  and for $1\leq i\leq n$, 
-  assume $A_i\in\nbody{r}$ for atom~$A_i$
-  such that $L_i=A_i$ or $L_i=\code{not}~A_i$, respectively.
-  }
+must be either domain predicates\comment{refer to appendix?},
+i.e., predicates that can be completely evaluated during grounding,
+or built-in (can be used on the left-hand side but only in a rule body),
+which is due to the fact that conditions are evaluated during grounding.
+
 Second, any variable occurring within a condition is considered as \emph{local},
@@
 \code{\pred{p}($\dots$,X,$\dots$)} and \code{\pred{p}($\dots$,Y,$\dots$)}.
-Pooled arguments in an atom of a rule body 
+Pooled arguments in any term of a rule body 
 (or on the right-hand side of a condition) are expanded to a conjunction
@@
   Simplified versions of these rules are produced via call:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/pool.lp}}
+  \code{\mbox{~}gringo -t \attach{examples/pool.lp}{pool.lp}}}
 %
@@
 \end{lstlisting}
-%
-Finally, we note that pooling is also possible with arguments of built-in predicates.
+\eexample
+\end{example}
+%
+Additionally, there is the \code{;;} operator for pooling, 
+which can only be used to separate arguments of predicates.
+This operator does not work on single terms but
+simply lists arguments of predicates.
+The rules for expanding the predicates are the same as for the \code{;} operator.
+
+\begin{example}\label{ex:sep}
+The following example show the difference between the \code{;} and \code{;;} operator:
+\marginlabel{%
+  Simplified versions of these rules are produced via call:\\
+  \code{\mbox{~}gringo -t \attach{examples/sep.lp}{sep.lp}}}
+%
+\lstinputlisting{examples/sep.lp}
+%
+The second line is expanded into the following:
+\begin{lstlisting}[firstnumber=2]
+p(X,Z) :- p(X,Y), p(Y,Z). 
+\end{lstlisting}%
+and the third line into:
+\begin{lstlisting}[firstnumber=3]
+p(X,Z) :- p(X,Y,Z), p(X,Y,Z).
+\end{lstlisting}%
+Clearly, the first variant is the desired expansion in this case 
+to calculate the transitive closure.
+Both operators have their usages in different scenarios 
+to keep the encoding more compact and readable.
 \eexample
@@
 
-An aggregate is an operation on a multiset of weighted literals that evaluates to an integer.
-In combination with arithmetic comparisons, we can extract a truth value from
-an aggregate's evaluation, thus, obtaining an aggregate atom.
-Customizing the notation in~\cite{ferraris05a},
-we consider ground \emph{aggregate atoms} of the form:
-%
-\begin{equation}\label{eq:aggregate:atom}
-  l~ 
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-\end{equation}
-%
-where~$l$ and~$u$ are integers each of which can possibly be omitted,
-$\mathrm{op}$ is a function from multisets of integers to an integer,
-each $L_i$ is a literal of the form~$A$ or $\code{not}~A$
-for some atom~$A\in\ground{\mathcal{A}}$,
-and each~$w_i$ is an integer.
-The intuitive reading of an aggregate atom is that the value returned by~$\mathrm{op}$
-applied to all weights~$w_i$ such that~$L_i$ holds should be in-between
-lower bound~$l$ and upper bound~$u$ (inclusively).
-If~$l$ or~$u$ is omitted,
-it means that the aggregate's value is not constrained from below or above,
-respectively.
-
-For an aggregate atom~$\mathit{Agg}$ as in~(\ref{eq:aggregate:atom}),
-we let $\atom{\mathit{Agg}}=\{L_i\in\nolinebreak\ground{\mathcal{A}}\mid 1\leq i\leq n\}$
-denote the set of atoms occurring positively in~$\mathit{Agg}$.
-Currently, \gringo\ and \clingo\ support aggregate operations
-\const{sum}, \const{max}, and \const{min}.%
-\footnote{%
-  The additional operation \const{count} is a shorthand for a \const{sum}
-  aggregate in which all weights are~\const{1}.}
-We can now extend translations~$\phi$ and~$\psi$ from Section~\ref{subsec:semantics}
-to aggregate atoms, 
-using the approach in~\cite{ferraris05a} but
-applying the ``choice semantics'' for aggregates
-in heads of rules~\cite{ferlif05a,siniso02a}:%
-\footnote{%
-  The translation~$\phi$ we provide for \const{sum} aggregates
-  assumes that there are no negative weights for literals,
-  while the more sophisticated translation in~\cite{ferraris05a} also
-  captures negative weights.
-  All current solvers work with positive weights only, and thus
-  \gringo\ and \lparse\ use a compilation technique~\cite{siniso02a}
-  to eliminate negative weights, which are not permitted in
-  \lparse's output format~\cite{lparseManual}.
-  As this technique may lead to counterintuitive results~\cite{ferraris05a},
-  we strongly recommend not to use negative weights within \const{sum} aggregates.}
-\begin{enumerate}
-\item % op %
-$\phi[
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
- \top$,
-\item % l op u %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \phi[
-  l~
- &
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
- {} \\ 
- &
-  \phi[
-  l~
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ]
-  \wedge
-  \phi[
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ]\text{,}
- \end{array}$
-\item % l sum %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \phi[
-  l~
- &
-  \const{sum} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
- {} \\ 
- &
-  \big(
-  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},l\leq w_{i_1}+\dots+w_{i_j}}
-  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
-  \big)\text{,}
- \end{array}$
-\item % sum u %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \phi
-  [
- &
-  \const{sum} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
- {} \\ 
- &
-  \neg
-  \big(
-  \bigvee_{\{i_1,\dots,i_j\}\subseteq\{1,\dots,n\},u< w_{i_1}+\dots+w_{i_j}}
-  (\phi[L_{i_1}]\wedge\dots\wedge \phi[L_{i_j}])
-  \big)\text{,}
- \end{array}$
-\item % l max %
- $\phi[
-  l~
-  \const{max} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
- \hspace{6pt}
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},l\leq w_i}
-  \phi[L_i]
-  \big)$,
-\item % max u %
- $\phi[
-  \const{max} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
-  \neg
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},u< w_i}
-  \phi[L_i]
-  \big)$,
-\item % l min %
- $\phi[
-  l~
-  \const{min} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  ] = 
-  \neg
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},w_i< l}
-  \phi[L_i]
-  \big)
- $,
-\item % min u %
- $\phi[
-  \const{min} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  u
-  ] = 
-  \hspace{6pt}
-  \big(
-  \bigvee_{i\in\{1,\dots,n\},w_i\leq u}
-  \phi[L_i]
-  \big)
- $, and
-\item % psi %
-$\begin{array}[t]{@{}r@{}l@{}}
-  \psi[
-  \lhd
- &~
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  \rhd
-  ] = 
-% {} \\ 
-% &
-   \phi[
-  \lhd~
-  \mathrm{op} 
-  \code{[}
-  L_1\code{=}w_1\code{,}
-  \dots\code{,}
-  L_n\code{=}w_n
-  \code{]}
-  \rhd
-  ]
-  \wedge
- {} \\ 
- &
-  \big(
-  \bigwedge_{A\in\atom{
-    \lhd~
-    \mathrm{op} 
-    \code{[}
-    L_1\code{=}w_1\code{,}
-    \dots\code{,}
-    L_n\code{=}w_n
-    \code{]}
-    \rhd
-  }}
-  (A\vee\neg A)
-  \big)\text{.}
- \end{array}$
-\end{enumerate}
-%
-The first item reflects that an unconstrained aggregate atom always holds,
-while aggregate atoms with a lower and an upper bound are split up into two
-conjunctively connected parts (Item~2).
-Item~3--8 specify these parts individually for~\const{sum},~\const{max}, and~\const{min}
-aggregates, respectively.
-Provided that only non-negative weights are used within \const{sum}s,
-all three aggregates are subject to one comparison that is monotone,
-meaning that by default it evaluates to \textit{false} so that truth must be established.
-For \const{sum} and \const{max} aggregates, these monotone comparisons are related
-to lower bounds (Item~3 and~5),
-while an upper bound behaves monotonically for \const{min} (Item~8).
-In turn, the opposite comparison is antimonotone, that is,
-by default it evaluates to \textit{true} unless too many or improper literals hold.
-This is the case for comparisons to upper bounds with aggregates
-\const{sum} and \const{max} (Item~4 and~6), while it concerns 
-the lower bound for \const{min} (Item~7).
-Finally, translation~$\psi$ of an aggregate atom in the head of a rule consists of two parts:
-first, $\phi$ is used like in the body of a rule
-(in Item~9, $\lhd$ and~$\rhd$ are used as ``wildcards'' for provided or omitted 
- lower and upper bounds),
-second, a disjunction $(A\vee\neg A)$ is added for each atom~$A$ occurring positively.
-Even though such disjunctions are tautological,
-they are important under answer set semantics (cf.\ Section~\ref{subsec:semantics}),
-as they permit deriving an arbitrary subset of atoms from an aggregate in the head
-(which explains the name  ``choice semantics'').
-Beyond this special treatment in rule heads,
-the truth values of aggregate atoms result from their literals as
-may be intuitively expected,
-and Item~1--8 basically do nothing but formalizing the standard meanings
-of aggregate operations.
-Note that the above translations merely define the semantics of aggregates,
-which does not imply that solvers unfold them in this way.
-In fact, \clasp\ natively supports aggregates without translating them.
+An aggregate is an operation on a multiset of weighted literals that evaluates to some value. 
+In combination with comparisons, we can extract a truth value from an aggregate's evaluation,
+thus, obtaining an aggregate atom.
+We consider aggregate atoms of the form:
+\[l\,\mathsf{op}\,\code{[}\, L_1\code{=}w_1\code{,} \dots\code{,} L_n\code{=}w_n {\,]\,} u\]
+
+An aggregate has a lower bound $l$, an upper bound $u$, an operation $\mathsf{op}$, 
+and a multiset of literal $L_i$ each assigned to a weight $w_i$.
+An aggregate is true if operation $\mathsf{op}$ applied to the multiset of weights 
+of true literals is between the bounds (inclusive).
+Currently, \gringo\ supports the aggregates 
+\code{\#sum} (the sum of weights), 
+\code{\#min} (the minimum weight), 
+\code{\#max} (the maximum weight), and
+\code{\#avg} (the average of all weights
+\footnote{The average aggregate over an empty set of weights is defined to be always true irrespective of any bounds.}%
+).
+Furthermore, there are three aggregates that are syntactically different.
+The first is the \code{\#count} aggregate:
+\[l\,\code{\#count}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}}\, u\]%
+which basically are \code{\#sum} aggregates with all weights set to one 
+and duplicate true literals counted only once.
+Finally, there are the two parity aggregates:
+\begin{align*}
+&\code{\#even}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}} & &
+\code{\#odd}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}}
+\end{align*}%
+These aggregates are true if the number of different true literals is even or odd, respectively.
 
@@
 Furthermore,
-keyword \const{sum} may be omitted, which in a sense makes \const{sum} the default
+keyword \const{\#sum} may be omitted, which in a sense makes \const{\#sum} the default
 aggregate operation.
@@
 \begin{lstlisting}[numbers=none,escapechar=@]
-2 sum [a, not b, c=2] 3  @\textnormal{and}@
+2 #sum [a, not b, c=2] 3  @\textnormal{and}@
 2      [a, not b, c=2] 3@\textnormal{.}@
 \end{lstlisting}
-By omitting keyword \const{sum}, we obtain the same notation as the one of 
+By omitting keyword \const{\#sum}, we obtain the same notation as the one of 
 so-called ``weight constraints''~\cite{siniso02a,lparseManual}, which are actually aggregate atoms
@@
 $L\code{=}w_1,\dots,L\code{=}w_k$
-of a literal~$L$, in combination with \const{min} and \const{max},
+of a literal~$L$, in combination with \const{\#min} and \const{\#max},
 it is not the same like having $L\code{=}w_1+\dots+w_k$.
@@
 \begin{lstlisting}[numbers=none]
-2 max [a=2].  2 min [a=2].
+2 #max [a=2].  2 #min [a=2].
 \end{lstlisting}
@@
 \begin{lstlisting}[numbers=none]
-2 max [a,a].  2 min [a,a].
+2 #max [a,a].  2 #min [a,a].
 \end{lstlisting}
@@
 If literals ought not to be repeated,
-we can use \const{count} instead of \const{sum}.
-Syntactically, \const{count} requires curly instead of square
-brackets, and there must not be any weights within a \const{count} aggregate.
+we can use \const{\#count} instead of \const{\#sum}.
+Syntactically, \const{\#count} requires curly instead of square
+brackets, and there must not be any weights within a \const{\#count} aggregate.
 Regarding semantics,
@@
 $(
-  l~ 
-  \const{count} 
+  l\,
+  \const{\#count} 
+  \,
   \code{\char`\{}
+  \,
   L_1\code{,}
@@
   L_n
+  \,
   \code{\char`\}}
+  \,
   u
@@
 $(
-  l~ 
+  l\,
   \const{sum}
+  \,
   \code{[}
+  \,
   L_1\code{=}\const{1}\code{,}
@@
   L_m\code{=}\const{1}
+  \,
   \code{]}
+  \,
   u
@@
 repeated literals.
-Of course, the use of~$l$ and~$u$ is optional also with \const{count}.
+Of course, the use of~$l$ and~$u$ is optional also with \const{\#count}.
 As an example, note that the next aggregate atoms express the same:
 \begin{lstlisting}[numbers=none,escapechar=@]
-1 sum   [a=1, not b=1]     1  @\textnormal{and}@ 
-1 count {a,a, not b,not b} 1@\textnormal{.}@  
-\end{lstlisting}
-Keyword \const{count} can be omitted (like \const{sum}),
+1 #sum   [a=1, not b=1]     1  @\textnormal{and}@ 
+1 #count {a,a, not b,not b} 1@\textnormal{.}@  
+\end{lstlisting}
+Keyword \const{\#count} can be omitted (like \const{\#sum}),
 so that the following are synonyms:
 \begin{lstlisting}[numbers=none,escapechar=@]
-1 count {a, not b} 1  @\textnormal{and}@ 
+1 #count {a, not b} 1  @\textnormal{and}@ 
 1        {a, not b} 1@\textnormal{.}@  
@@
 \footnote{%
-  Assignments (cf.\ Section~\ref{subsec:gringo:assign}) and (currently)
-  also built-in comparison predicates (cf.\ Section~\ref{subsec:gringo:comp})
-  are permitted on the right-hand side of conditions only.}
+  Assignments (cf.\ Section~\ref{subsec:gringo:assign}) 
+  are permitted on the right-hand sides of conditions only.}
 That is, an interval gives rise to multiple literals connected via~``\code{,}''
@@
 while it turns into a composite condition chained by~``\code{:}'' on the right-hand side.
-The notions of level-restrictedness and stratification
-(cf.\ Section~\ref{subsec:lambda} and~\ref{subsec:strat}) are extended
-to programs with aggregates by
-considering the predicates of all atoms (except for those on the right-hand side of a condition)
-in an aggregate atom occurring as the head of a rule~$r$,%
-\footnote{%
-  The rules in a (maximal) stratified subprogram (cf.\ Section~\ref{subsec:strat})
-  cannot have aggregate atoms in the head.}
-and by assuming $A\in\nbody{r}$ for all atoms~$A$ of an aggregate in the body of~$r$.
-Finally, note that aggregates without bounds are also permitted on the
+Finally, note that aggregates 
+\code{\#sum}, \code{\#count}, \code{\#min}, and \code{\#max}
+without bounds are also permitted on the
 right-hand sides of assignments, but using this feature is only recommended
@@
 %
-Instantiating the above \const{count} aggregate
+Instantiating the above \const{\#count} aggregate
 yields the following ground rule:%
@@
   The full ground program is obtained by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/aggr.lp}}
+  \code{\mbox{~}gringo -t \attach{examples/aggr.lp}{aggr.lp}}}
 %
@@
 using default weight~\const{1} for literals.
-Recalling that \const{sum} aggregates operate on multisets,
+Recalling that \const\#{sum} aggregates operate on multisets,
 duplicates are not removed.
@@
 %
-The above integrity constraint states that the \const{sum} of contact hours per week
+The above integrity constraint states that the \const{\#sum} of contact hours per week
 must lie in-between~\const{18} and~\const{20}.
-Note that the \const{min} and \const{max} aggregates in Line~16 and~17, respectively,
+Note that the \const{\#min} and \const{\#max} aggregates in Line~16 and~17, respectively,
 work on the same (multi)set of weighted literals as in Line~15.
@@
 Note that the above aggregates have already been used in Line~9 and~15, respectively,
-where keywords \const{count} and \const{sum} have been omitted for convenience.
+where keywords \const{\#count} and \const{\#sum} have been omitted for convenience.
 These keywords can be dropped here too, and we merely include them to show the
-more verbose notations of \const{count} and \const{sum} aggregates.
+more verbose notations of \const{\#count} and \const{\#sum} aggregates.
 However, the usage of aggregates in the last two lines is different from before,
@@
   To compute the unique answer set of the program, invoke:\\
-  \code{\mbox{~}gringo \textbackslash\\
-        \mbox{~}examples/aggr.lp | \textbackslash\\
+  \code{\mbox{~}gringo \attach{examples/aggr.lp}{aggr.lp} | \textbackslash\\
         \mbox{~}clasp -n 0}\\
   or alternatively:\\
-  \code{\mbox{~}clingo -n 0 \textbackslash\\
-        \mbox{~}examples/aggr.lp}}
+  \code{\mbox{~}clingo -n 0 \attach{examples/aggr.lp}{aggr.lp}}}
 %
@@
 
+\com{Should we mention stratifed aggregates that do not necessarily need conditions?}
+
 \subsubsection{Optimization}\label{subsec:gringo:optimize}
@@
 optimization statements of \lparse~\cite{lparseManual},
-indicated via keywords \const{maximize} and \const{minimize}.
-Syntactically, a \const{maximize} or \const{minimize} statement
-is similar to a fact whose head is a \const{count} or a \const{sum}
-aggregate (without bounds),
-viz.,
-``$\textrm{opt}\code{\char`\{}\dots\code{\char`\}}\code{.}$'' or
-``$\textrm{opt}\code{[}\dots\code{]}\code{.}$'' where
-$\textrm{opt}\in\{\const{maximize},\linebreak[1]\const{minimize}\}$.
+indicated via keywords \const{\#maximize} and \const{\#minimize}.
 As an optimization statement does not admit a body, any (local) variable
@@
 In multiset notation (square brackets),
-weights may be provided as with \const{sum} aggregates. %,
-% where~\const{1} is used as default for omitted weights.
+weights may be provided as with \const{\#sum} aggregates. %,
 In set notation (curly brackets),
@@
 as with \const{count} aggregates.
+Additionally, priorities can be associated with each literal.
+A (ground) optimize statement has the form:
+\[\mathsf{opt}\,\code{[}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\]
+\[\mathsf{opt}\,\code{\char`\{}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\]
+where $\mathsf{opt}$ is either \code{\#maximize} or \code{\#minimize} , $L_i$ are literals 
+with associates (integer) weights $w_i$ and (integer) priorities $p_i$.
 
@@
 along with a logic program.
-Unfortunately, the syntax used by \lparse\ and adopted by \gringo\ and \clingo\
-does not provide any means to explicitly declare priorities among multiple
-optimization statements.
-A technique how to compile multiple optimization statements into a single one
-is discussed in~\cite{siniso02a},
-but it can lead to large integers
-(probably rather unnatural for a user to assign).
-Thus, rather than restricting to a single optimization statement,
-a sequence of them is permitted,
-where a statement provided later on takes priority over previous ones.
-That is, the last optimization statement in program is the most significant one,
-and previous ones are only considered if answer sets agree on the sum of 
-weights of its literals.
-In this way, multiple optimization statements are arranged into a total order,
-which is the inverse order of occurrence.
-In contrast to the declarative answer set semantics, the semantics of
-optimization statements is thus order-dependent,
-requiring special care when providing more than one of them.
+If different priorities occur in the program, 
+then, depending on the type of optimize statement,
+answer sets whose sum of weights assigned to higher priorities
+is maximized or minimized, respectively.
+
+Note that for compatibility with \lparse, if multiple optimize statements are used, 
+default priorities are assigned.
+The $n$-th statement gets priority $n$, thus, the later statements have higher priorities.
+We suggest that if you want to use more than one optimization statement,
+to always specify priorities to make the program more readable and order independent.
 
@@
   The full ground program is obtained by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\\mbox{~}examples/opt.lp}}
-%
-\begin{lstlisting}[firstnumber=6,breaklines,breakindent=48pt]
-maximize [ hotel(1) = 5, hotel(2) = 4, hotel(3) = 3,    hotel(4) = 3, hotel(5) = 2 ].
-minimize [ hotel(1) = 34, hotel(2) = 35, hotel(3) = 30,   hotel(4) = 25, hotel(5) = 30 ].
-minimize [ noisy = 1 ].
+  \code{\mbox{~}gringo -t \attach{examples/opt.lp}{opt.lp}}}
+%
+\begin{lstlisting}[firstnumber=6,breakindent=0pt,escapechar=&]
+#maximize [ hotel(1)=5@1, hotel(2)=4@1, &\\&            hotel(3)=3@1, hotel(4)=3@1, hotel(5)=2@1 ].
+#minimize [ hotel(1)=34@2, hotel(2)=35@2, &\\&            hotel(3)=30@2, hotel(4)=25@2, hotel(5)=30@2 ].
+#minimize [ noisy=1@3 ].
 \end{lstlisting}
@@
   To compute the unique optimal answer set, invoke:\\
-  \code{\mbox{~}gringo \textbackslash\\
-        \mbox{~}examples/opt.lp | \textbackslash\\
+  \code{\mbox{~}gringo \attach{examples/opt.lp}{opt.lp} | \textbackslash\\
         \mbox{~}clasp -n 0}\\
   or alternatively:\\
-  \code{\mbox{~}clingo -n 0 \textbackslash\\
-        \mbox{~}examples/opt.lp}}
+  \code{\mbox{~}clingo -n 0 \attach{examples/opt.lp}{opt.lp}}}
 we find that hotel~\const{4} is not eligible because it implies \pred{noisy}.
@@
 In order to suppress the atoms of ``irrelevant'' predicates from the output,
-the \code{\#hide} declarative (in which~\code{\#} is optional) can be used.
+the \code{\#hide} declarative can be used.
 The meanings of the following statements are indicated via accompanying comments:
@@
 #hide.     % Suppress all atoms in output
- hide.          % Same as "#hide."
 #hide p/3. % Suppress all atoms of predicate p/3 in output
- hide p/3.      % Same as "#hide p/3."
-#hide p(X,Y,Z). % Same as "#hide p/3."
- hide p(X,Y,Z). % Same as "#hide p/3."
-\end{lstlisting}
-%
+#hide p(X,Y) : q(X). % Supress p/3 if the condition holds
+\end{lstlisting}
+Note that for the conditionals on the right-hand side, 
+the same conditions as described in Section \ref{subsec:gringo:condition} apply.
+
 In order to selectively include the atoms of a certain predicate in the output,
-one may use the \code{\#show} declarative (in which~\code{\#} is again optional).
+one may use the \code{\#show} declarative.
 Here are some examples:
@@
 #show p/3. % Include all atoms of predicate p/3 in output
- show p/3.      % Same as "#show p/3."
-#show p(X,Y,Z). % Same as "#show p/3."
- show p(X,Y,Z). % Same as "#show p/3."
+#showp(X,Y) : q(X). % Supress p/3 if the condition holds
 \end{lstlisting}
@@
 An example of this will be given in Section~\ref{subsec:ex:color}.
-Via the \code{\#const} declarative (\code{\#}~is optional),
+Via the \code{\#const} declarative,
 one may define a default value to be inserted for a constant.
@@
 #const x = 42.
- const y = f(g,h).
-\end{lstlisting}
+#const y = f(x,h).
+\end{lstlisting}%
+Note that (for efficiency reasons) constant declarations are order dependent.
+In the example above \code{x} would be replaced by \code{42}
+but when reversing the directives this would no longer be the case.
 
@@
 Usually, variable names are local to a rule,
-where they must be bound via appropriate atoms (cf.\ Section~\ref{subsec:lambda}).
+where they must be bound via appropriate atoms.
 This locality can be undermined by using
-\code{\#domain} declarations (\code{\#}~is optional)
+\code{\#domain} declarations 
 that globally associate variable names to atoms.
@@
 #domain p(X,Y).
- domain p(Y,Z).
+#domain p(Y,Z).
 q(Z,X) :- not p(Z,X).
@@
 %
-The above program is a priori not level-restricted because
+The above program is a priori not safe because
 variables~\var{X} and~\var{Z} are unbound in the last rule.
@@
 %
-Observe that the resulting program is level-restricted (and stratified).
+Observe that the resulting program is safe.
+Nevertheless, we suggest not to use domain statements because 
+in ASP it is common to use very short variable names and
+using domain statements will likely result in name clashes
+and undesired behavior.
 
 \paragraph{Compute Statements.}
-These statements are artefacts supported for backward compatibility.
+These statements are artifacts supported for backward compatibility.
 Although we strongly recommend to avoid compute statements,
@@
 ``$\const{\#compute}~n \code{\char`\{}\dots\code{\char`\}}\code{.}$''
-(\code{\#}~and non-negative integer $n$ are optional),
+(the non-negative integer $n$ is optional),
 where the ``$\code{\char`\{}\dots\code{\char`\}}$'' part
-is similar to a \const{count} aggregate.
+is similar to a \const{\#count} aggregate.
 The meaning is that all literals contained in
@@
 #compute 0 { p(X) : q(X) }.       % :- 1 { not p(X) : q(X) }.
- compute   { not p(X) : X=4..5 }. % :- 1 { p(X) : X=4..5 }.
-\end{lstlisting}
+#compute   { not p(X) : X=4..5 }. % :- 1 { p(X) : X=4..5 }.
+\end{lstlisting}
+
+\paragraph{External Statements.}
+External statements are used to mark certain atoms as external. This means 
+that those atoms are not subject of simplification and consequently will not be removed 
+from the logic program. There are two kinds of external directives, 
+global and local external statements.
+
+Global external statements have the form \code{\#external predicate/arity.} where 
+\code{predicate} refers to the name of a predicate and \code{arity} to the arity of the predicate.
+They mark complete predicates irrespective of any arguments as external.
+This means that nothing is known about the predicate and hence it cannot be used 
+for instantiation.
+\begin{example}
+Consider the following example:
+\begin{lstlisting}[numbers=none]
+#external q/1.
+p(1). p(2).
+r(X) :- q(X), p(X).
+\end{lstlisting}
+Here, the external predicate \code{q}/$1$ will not be used for simplification
+of the problem and hence two ground rules (excluding facts) will be printed.
+\eexample
+\end{example}
+
+Local external statements have the form \code{\#external predicate (: conditional)*.} where
+\code{predicate} is some non-ground predicate and \code{conditional} some conditional.
+In contrast to global external directives, local external statements precisely specify 
+which atoms are external and hence can be used for instantiation.
+\begin{example}
+Again, consider a similar example:
+\begin{lstlisting}[numbers=none]
+#external q(X) : p(X).
+p(1). p(2).
+r(X) :- q(X).
+\end{lstlisting}
+Here, the external predicate \code{q}/$1$ is used to bind variable \code{X}, 
+yielding the same rules as in the example above.
+\eexample
+\end{example}
+
+Furthermore, the \lparse\ output\cite{lparseManual} has been modified to include an additional table 
+that stores a list  of all external atoms. For compatibility, this table is only inserted if the 
+program actually contains external directives. It contains the respective atom indices terminated
+by a zero and is inserted directly after \lparse' compute statement.
+\begin{lstlisting}[numbers=none]
+  ...
+  0
+  ...
+  0
+  B+
+  ...
+  0
+  B-
+  ...
+  0
+  E
+  2
+  3
+  4
+  ...
+  0
+  1
+\end{lstlisting}
+
+\subsubsection{Integrated Scripting Language}\label{subsec:lang:lua}
+Utilizing the scripting language Lua\footnote{\url{http://www.lua.org}}, 
+\gringo's input language can be enriched by arbitrary arithmetical functions and implicit domains,
+answer sets can be intercepted and for example inserted into a database, or
+interactions between grounding and solving are possible when incrementally solving with \iclingo.
+We do not give an introduction to Lua here (there are numerous tutorial on the web), 
+but give some examples showing the capabilities of this integration.
+
+\begin{example}
+The first example shows basic Lua usage:
+\lstinputlisting{examples/luaf.lp}
+\marginlabel{%
+To compute the unique answer set, invoke:\\
+\code{%
+\mbox{~}gringo -t \attach{examples/luaf.lp}{luaf.lp}}
+}%
+In Line 3 we add a function that calculates the greatest common divisor of two numbers.
+This function is called in Line 20 and the result stored in predicate \code{gcd}/$3$.
+Note that Lua function calls look like function symbols but are preceded by ``\code{@}''.
+Regarding binding of variables, the same restrictions as with arithmetic \ref{subsec:gringo:arith} apply.
+
+In Line 9 we add function that emulates a range term.
+It returns a table containing all numbers in the interval $[\code{a},\code{b}]$.
+The values in this table are then successively replaced for the call \code{\@rng{X,Y}}.
+In fact, this function exactly behaves like a range term.
+\eexample
+\end{example}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{||l|l||}
+ \hline 
+ \code{Val.NUM}  & Type identifier for \gringo\ numbers.\\
+ \hline 
+ \code{Val.ID}   & Type identifier for \gringo\ strings.\\
+ \hline 
+ \code{Val.FUNC} & Type identifier for function symbols.\\
+ \hline 
+ \code{Val.SUP}  & Type identifier for \gringo's \code{\#supremum}.\\
+ \hline 
+ \code{Val.INF}  & Type identifier for \gringo's \code{\#infimum}.\\
+ \hline 
+ \code{Val.new(type[,value][,args])}  & Creates new ground terms (see Example \ref{ex:lua:values}).\\
+ \hline 
+ \code{Val.cmp(a,b)}  & Compares two ground \gringo\ terms. \\
+ \hline 
+ \code{Val.type(a)} & Returns the type of a term. \\
+ \hline 
+ \code{Val.name(f)} & Returns the name of function symbol \code{f}. \\
+ \hline 
+ \code{Val.args(f)} & Returns the arguments of function symbol \code{f} \\
+ \hline
+\end{tabular}
+\caption{The Val meta-table.\label{tab:val}}
+\end{table}
+
+\begin{example}\label{ex:lua:values}
+The second example shows how to create ground terms from within Lua:
+\lstinputlisting{examples/luav.lp}
+\marginlabel{%
+To compute the unique answer set, invoke:\\
+\code{%
+\mbox{~}gringo -t \attach{examples/luav.lp}{luav.lp}}
+}%
+Function \code{f} in Line 2 returns a tuple 
+its first member a string indicating the type of the argument 
+and the second reconstructs the value passed to function \code{f}.
+Note that in Line 4,6,9,11 and 13 the function \code{Val.new} is called.
+Its first argument indicates that a function symbol is to be created and 
+the second argument passes the arguments of the function symbol.
+We do not give a name for the function symbol,
+thus a tuple is created (this is equivalent to passing the empty string as name).
+You can infer from context, how \gringo's other in-built ground terms are created.
+Finally, note that \gringo integers and strings are directly mapped 
+to the respective Lua types.
+\eexample
+\end{example}
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{||l|p{6cm}||}
+ \hline 
+ \code{Assignment.begin(n,a)} & Starts iteration over an atom with name~\code{n} with arity~\code{a}. \\
+ \hline 
+ \code{Assignment.next()}     & Advances to the next atom in the assignment and returns false if there is none. \\
+ \hline 
+ \code{Assignment.args()}     & The arguments of the current atom. \\
+ \hline 
+ \code{Assignment.isTrue()}   & The current atom is true. \\
+ \hline 
+ \code{Assignment.isFalse()}  & The current atom is false. \\
+ \hline 
+ \code{Assignment.isUndef()}  & The current atom is undefined. \\
+ \hline 
+ \code{Assignment.level()}    & The decision level on which the current atom has been assigned. \\
+ \hline
+\end{tabular}
+\caption{The Assignment meta-table.\label{tab:assignment}}
+\end{table}
+
+\begin{example}
+The next example show some advanced usage also accessing \clasp's truth assignment:
+\lstinputlisting{examples/sql.lp}
+This example creates a database connection using SQLite\footnote{\url{http://www.sqlite.org/}}.
+The database connection is created Lines 3-4 using LuaSQL\footnote{\url{http://www.keplerproject.org/luasql/}}.
+Initially, a new table \code{test} is created if it does not already exists (Lines 5-7).
+There are two functions to access this table.
+The first function \code{query} in Line 9 selects everything from the table and 
+returns it in form of a Lua table, which is then used in the logic program
+to provide new instantiations of~\code{p}/$2$ in Line 42.
+The second function \code{insert} is a helper function that expects a predicate name
+and then inspects \clingo's (or \iclingo's) possibly partial assignment and
+inserts all true atoms into the database.
+It makes use of the \code{Assignment} meta-table.
+Next, note the three functions
+\code{onBeginStep},
+\code{onModel}, and
+\code{onEndStep} 
+in Line 35-37.
+The first function is called directly before solving but after preprocessing.
+At this point grounded facts are accessible via the \code{Assignment} meta-table (Table \ref{tab:assignment}),
+we call \code{insert} to insert all ground instantiation of \code{p}/$2$ into the \code{test} table.
+The same is done in the \code{onModel} function whenever a model is found.
+At this point \clingo's assignment is total and 
+we insert all instantiation of \code{q}/2 into the database that are contained in the answer set.
+Finally, in function \code{onEndStep} we print the number of tuples inserted during grounding and after solving.
+
+\marginlabel{%
+Try to invoke this program multiple times using:\\
+\code{%
+\mbox{~}clingo \attach{examples/sql.lp}{sql.lp}}
+}
+Note that the values from the answer set are inserted into the database.
+In fact the output of the program changes when it is consecutively called.
+Even more non-determinism can be added by using the option \code{--rand-freq=1.0}
+to induce 100 percent random decisions.
+\eexample
+\end{example}
 
@@
 Hence, its input language includes all constructs described
-in Section~\ref{subsec:lang:gringo}.%
-\footnote{%
-  In its current version, \iclingo\ has no built-in support for
-  classical negation (cf.\ Section~\ref{subsec:gringo:negation}).
-  That is, complementary atoms may jointly belong to answer sets,
-  unless explicitly prohibited by integrity constraints.}
+in Section~\ref{subsec:lang:gringo}.
 In addition, \iclingo\ deals with statements of the following form:
@@
 replacing \code{\textit{constant}}s with successive step numbers starting from~\const{1}.
+Note that due to \gringo's grounding algorithm,
+rules will not be grounded twice using the same substitution of global variables
+(the incremental constant is treated like a global constant here).
 After a grounding step, \clasp\ is usually invoked via an internal interface
@@
 
+\begin{example}
+For now, consider this simple Schur number example:
+\lstinputlisting{examples/inc.lp}
+\marginlabel{%
+  To calculate the Schur number, invoke:\\
+  \code{\mbox{~}iclingo \attach{examples/inc.lp}{inc.lp} \textbackslash\\
+  --istop=UNSAT
+  }\\
+  The solving stops when the largest number has been found.
+}
+%
+The Schur number~$n$ w.r.t. to a given number~$c$ is the largest integer 
+such that the interval~$[1..n]$ can be partitioned into $c$ sum-free sets.
+A set $S=\{c_1 ,\dots, c_k\}$ is sum-free if for each~$x_i$ and~$x_j$, 
+it holds that~$x_i + x_j \not \in S$.
+
+In the base part in Line 2 we specify the number of partitions.
+Then, in the cumulative part in Line 4 each fresh integer $k$
+is assigned to exactly one partition.
+In Line 5 we check whether the guessed partition is sum-free.
+Note that we put this check in the cumulative part and incrementally extend it.
+The idea here is to keep as much constraints in the cumulative part
+because \clasp\ applies nogood learning and only information learnt from 
+the cumulative part can be kept among further solving steps;
+all information learnt from the volatile part has to be forgotten.
+Additionally, we use the comparison  \code{X <= Y}, 
+which helps to avoid grounding some redundant rules.
+Furthermore, note that there appears no incremental constant in this rule,
+\gringo's grounding algorithm makes sure that no ground instantiation of this rule 
+is grounded twice just the new slice for the next step is instantiated.
+Finally, we break symmetries in Line 6, i.e.,
+number one is always assigned to partition one, number two to partition two 
+if it is not in partition one, and so on.
+
+We use option \code{--istop=UNSAT} to solve as long as 
+\iclingo\ is able to find a solution, i.e., there is a 
+valid partitioning. Once this is no longer possible, the grounding/solving 
+process stops.\eexample
+\end{example}
+
 \subsection{Input Language of \clasp}\label{subsec:lang:clasp}
@@
 
-%%% Local Variables: 
-%%% mode: latex
-%%% TeX-master: "guide"
-%%% End:
+Solver \clasp~\cite{gekanesc07b} works on logic programs in \lparse's
+output format~\cite{lparseManual}.
+This numerical format, which is not supposed to be human-readable,
+is output by \gringo\ and can be piped into \clasp.
+Such an invocation of \clasp\ looks as follows:
+%
+\begin{lstlisting}[numbers=none]
+gringo [ options | filenames ] | clasp [ number | options ]
+\end{lstlisting}
+%
+Note that \code{number} may be provided to specify a maximum number of answer sets
+to be computed, where~\code{0} makes \clasp\ compute all answer sets.
+This maximum number can also be set via
+option \code{--number} or its abbreviation \code{-n}
+(cf.\ Section~\ref{subsec:opt:clasp}).
+By default, \clasp\ computes one answer set (if it exists).
+If a logic program in \lparse's output format has been stored in a \code{file},
+it can be redirected into \clasp\ as follows:%
+\footnote{%
+  The same is achieved by using option \code{--file} or its short form \code{-f}
+  (cf.\ Section~\ref{subsec:opt:clasp}).}
+%
+\begin{lstlisting}[numbers=none]
+clasp [ number | options ] < file
+\end{lstlisting}
+%
+Via option \code{--dimacs},
+\clasp\ can also be instructed to compute models of a propositional formula
+in DIMACS/CNF format~\cite{dimacs}.
+If such a formula is contained in \code{file},
+then \clasp\ can be invoked in the following way:
+%
+\begin{lstlisting}[numbers=none]
+clasp [ number | options ] --dimacs < file
+\end{lstlisting}
+%
+Finally, \clasp\ may be used as a library, as done within \clingo\ and \iclingo.
+%

### 2010-08-10 doc/guide/options.tex
- updated the guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3297 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 Print version information and exit.
-\item[\code{--syntax}]~\\
-Print syntax information (about the input language) and exit.
-\item[\code{--stats}]~\\
-Print (extended) statistic information before termination.
-\item[\code{--verbose,-V}]~\\
+%\item[\code{--syntax}]~\\
+%Print syntax information (about the input language) and exit.
+%\item[\code{--stats}]~\\
+%Print (extended) statistic information before termination.
+\item[\code{--verbose[=n],-V=[n]}]~\\
 Print additional (progress) information during computation.
-\item[\code{--debug}]~\\
-Print internal representations of rules during grounding.
+Verbosity level one and two is reserved for clasp only.
+Level three prints internal representations of rules during grounding.
 (This may be used to identify either semantic errors in an input program or
@@
 Output ground program in (human-readable) text format.
+\item[\code{--reify}]~\\
+Output ground program in form of facts.
+(These facts can then be used to, i.e., write a meta-interpreter.)
 \item[\code{--lparse,-l}]~\\
 Output ground program in \lparse's numerical format~\cite{lparseManual}.
-\item[\code{--aspils,-a 1|2|3|4|5|6|7}]~\\
-Output ground program in \emph{ASPils} format~\cite{gejaosscth08a},
-where the argument specifies one of the seven normal forms
-defined in~\cite{gejaosscth08a}.
 \item[\code{--ground,-g}]~\\
@@
  but it leads to a syntax error (cf.\ Section~\ref{subsec:error}) otherwise.)
-\item[\code{--bindersplit yes|no}]~\\
-Enable or disable binder-splitting~\cite{gescth07a} in the instantiation of rules.
-(This option is included mainly for comparison purposes, and generally,
- it is recommended to keep the default argument value \code{yes}.)
 \item[\code{--ifixed \textit{n}}]~\\
@@
  including some mutable bound.)
+\item[\code{--dep-graph=filename}]~\\
+This option can be used to get the dependency graph in from
+of a dot\footnote{\url{http://www.graphviz.org/}} file of the program.
+\item[\code{--shift}]~\\
+Removes disjunction by shifting (see \cite{gellif91a}).
 \end{description}
@@
 \begin{lstlisting}[numbers=none]
-gringo --lparse --bindersplit=yes
+gringo --lparse
 \end{lstlisting}
 That is, \gringo\ usually outputs a ground program in \lparse's numerical format,
-dealt with by various solvers (cf.\ Section~\ref{subsec:asp}),
-and it performs binder-splitting in rule instantiation.
-
-\begin{newstuff}
-	\begin{itemize}
-		\item stats is not yet implemented
-		\item verbose takes level now
-		\item debug corresponds to verbose=3
-		\item aspils is gone
-		\item bindersplit is gone (not worth implementing, can be done in preprocessor)
-	\end{itemize}
-\end{newstuff}
+dealt with by various ASP solvers \cite{gekanesc07b,siniso02a,linzha04a}.
 
@@
 and in Section~\ref{subsec:opt:clasp}.
-In particular, (long) options \code{--help}, \code{--version}, and
-\code{--syntax} make \clingo\ print the desired information and exit,
-while \code{--text}, \code{--lparse}, and \code{--aspils}
+In particular, (long) options \code{--help} and \code{--version}
+make \clingo\ print the desired information and exit,
+while \code{--text}, \code{--lparse}, and \code{--reify}
 instruct \clingo\ to output a ground program (rather than solving it) like \gringo.
@@
 Finally, the default command line when invoking \clingo\
-consists of all \clasp\ defaults (cf.\ Section~\ref{subsec:opt:clasp})
-and option \code{--bindersplit=yes} of \gringo.
+consists of all \clasp\ defaults (cf.\ Section~\ref{subsec:opt:clasp}).
 
@@
 option \code{--ifixed} is ignored by \iclingo\ if not run as a grounder
-(via one of (long) options \code{--text}, \code{--lparse}, or \code{--aspils}).
+(via one of (long) options \code{--text}, \code{--lparse}, or \code{--reify}).
 However, option \code{--clingo} (see below) may be used to let \iclingo\
@@
 consists of all \clasp\ defaults, explained in the next section,
-option \code{--bindersplit=yes} of \gringo\ (cf.\ Section~\ref{subsec:opt:gringo}),
 along with \code{--istop=SAT}, \code{--iquery=1}, \code{--ilearnt=keep}, and
@@
 We thus categorize them according to their functionalities in the following description.
-
-\begin{newstuff}
-	\begin{itemize}
-		\item the subsubsections below are probably outdated, benny?
-	\end{itemize}
-\end{newstuff}
 

### 2010-08-10 doc/guide/quickstart.tex
- updated the guide
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3297 3e02f596-3131-0410-8352-e04eeaec37cc

@@
-\section{Quickstart}
-
+\include{figures/hanoi}%
+\section{Quickstart}\label{sec:quickstart}
+In this section we will demonstrate the expressive power 
+and the simple yet powerful modeling language of \gringo{}
+by looking at the simple Towers of Hanoi puzzle.
+It consists of three pegs and a set of discs of different sizes, 
+which can be put onto the pegs. 
+The goal is to move all pegs from the leftmost peg to the rightmost peg,
+where at each time only the topmost disc can be moved on top of another peg.
+Additionally, a disc may not be put on top of a smaller disc.
+We will ignore that there is an efficient algorithm to solve this problem 
+and just specify how a solution, in terms of a sequence of moves, has to look.
 In ASP it is custom to provide a \emph{uniform}
 problem definition~\cite{martru99a,niemela99a,schlipf95a}.
-Following this methodology we separate the encoding
+Following this methodology, we separate the encoding
 from an instance of the following problem:
-given a (directed) graph, decide whether each node can be assigned
-one of~$N$ colors such that any pair of adjacent nodes is colored differently.
-Note that this problem is NP-complete for~$N\geq 3$
-(see, e.g.,~\cite{papadimitriou94a}),
-and thus it seems unlikely that a worst-case polynomial algorithm
-can be found.
-In view of this,
-it is convenient to reduce the particular problem to
-a declarative problem solving paradigm like ASP,
+given an initial placement of the discs, a goal situation, and a number of moves $n$,
+decide whether there is a sequence of moves of length $n$ 
+that satisfies the conditions given above.
+We will see that this decision problem can be elegantly 
+specified by reducing it to a declarative problem solving paradigm like ASP,
 where efficient off-the-shelf tools like \gringo\ and \clasp\
@@
 
-\subsection{Problem Instance}\label{subsec:color:instance}
-\input{figures/graph_intro}
+\subsection{Problem Instance}
 
-We consider directed graphs specified via facts over predicates
-\pred{node}/$1$ and \pred{edge}/$2$.%
-The predicate \pred{color}/$1$ denotes the colors that shall be used to color the graph.
-Figure~\ref{fig:graph_intro} depicts the graph specified by the following set of facts:
-%
-\lstinputlisting{examples/graph_intro.lp}
+\begin{figure}[tb]
+\centering
+\hanoiInstance
+\caption{Towers of Hanoi Initial Situation\label{fig:toh_inst}}
+\end{figure}
 
-\subsection{Problem Encoding}\label{subsec:color:encoding}
+We consider a Towers of Hanoi instance specified via facts over predicates
+\pred{peg}/$1$ and \pred{disk}/$1$ that correspond to the pegs and disks in the puzzle.
+Discs are enumerated by consecutive integers beginning with one
+where a disc with a lower number is considered to be bigger than a list with a higher number.
+The pegs can have arbitrary names. 
+Furthermore, the predicates \pred{init\_on}/$2$ and \pred{goal\_on}/$2$ describe the initial and goal situation, respectively.
+Their first argument is the number of a disc and the second argument the peg 
+this disc is located in the initial or goal situation.
+Finally, the predicate \pred{moves}/$1$ specifies the number of moves within which the goal situation has to be reached.
+Note that the original puzzle had exactly three pegs and a fixed initial and goal situation.
+With ASP we can easily change this requirement and 
+the encoding represented in the following will work with an arbitrary number of pegs and any initial or goal situation.
+Figure~\ref{fig:toh_inst} depicts a possible instance (the dashed discs mark the goal situation) 
+corresponding to the ASP program given below.
 
-We now proceed by encoding $N$-coloring via non-ground rules that are
-independent of particular instances.
+\begin{lstlisting}
+peg(a;b;c).
+disk(1..4).
+init_on(1..4,a).
+goal_on(1..4,c).
+moves(15).
+\end{lstlisting}
+
+The `\code{;}' in the first line is some syntactic sugar (Section \ref{subsec:gringo:pool})
+that expands the statement into three facts
+\code{peg(a)}, \code{peg(b)}, and \code{peg(c)} representing the three pegs.
+Again, in the second line some syntactic sugar is used to 
+create the facts \code{disc(1)}, \code{disc(2)}, \code{disc(3)}, and \code{disc(4)}.
+Here the term \code{1..4}, an intervall (Section \ref{subsec:gringo:interval}), is successively replaced by \code{1}, \code{2}, \code{3}, and \code{4}.
+The initial and goal situation is specified in line three and four again using intervall.
+Finally, in the last line the number of moves to solve the problem is given.
+
+\subsection{Problem Encoding}
+
+We now proceed by encoding the Towers of Hanoi puzzle via non-ground rules (Section \ref{subsec:gringo:normal}), 
+i.e, rules with variables that are independent of particular instances.
 Typically, an encoding consists of a \emph{Generate}, a \emph{Define},
 and a \emph{Test} part~\cite{lifschitz02a}.
-As $N$-Coloring has a rather simple pattern, the following encoding does
-not contain any Define part:
-%
-\lstinputlisting{examples/color.lp}
-%
-The Generate rule in Line~2 makes use of the \const{count} aggregate
-(cf.\ Section~\ref{subsec:gringo:aggregate}).
-For our example graph and~\const{1} substituted for~\var{X},
-we obtain the following ground rule:%
-\marginlabel{%
-  The full ground program is obtained by invoking:\\
-  \code{\mbox{~}gringo -t \textbackslash\\
-    \mbox{~}examples/color.lp \textbackslash\\
-    \mbox{~}examples/graph.lp}}
-%
-\begin{lstlisting}[numbers=none]
-1 { color(1,red), color(1,green), color(1,blue) } 1.
+We will follow this paradigm and
+mark respective parts via comment lines beginning with \code{\%} in the encoding below.
+The variables \code{D}, \code{P}, \code{T}, and \code{M} are used
+to refer to disks, pegs, the \code{T}-th move in the sequence of moves, and the length of the sequence, respectively.
+
+\begin{lstlisting}
+% Generate
+1 { move(D,P,T) : disk(D) : peg(P) } 1 :- moves(M), T = 1..M.
+% Define
+move(D,T)   :- move(D,_,T).
+on(D,P,0)   :- init_on(D,P).
+on(D,P,T)   :- move(D,P,T).
+on(D,P,T+1) :- on(D,P,T), not move(D,T+1), not moves(T).
+blocked(D-1,P,T+1) :- on(D,P,T), disk(D), not moves(T).
+blocked(D-1,P,T)   :- blocked(D,P,T), disk(D).
+% Test
+:- move(D,P,T), blocked(D-1,P,T).
+:- move(D,T), on(D,P,T-1), blocked(D,P,T).
+:- goal_on(D,P), not on(D,P,M), moves(M). 
+:- not 1 { on(D,P,T) : peg(P) } 1, disk(D), moves(M), T = 1..M.
+#hide.
+#show move/3. 
+\end{lstlisting}%
+The Generate part consists of just one rule in Line 2.
+At each time point \code{T} at which a move is executed, we
+``guess'' exactly one move that puts an arbitrary disk to some arbitrary peg.
+The head of this rule is a so called cardinality constraint (Section \ref{subsec:gringo:aggregate}) that
+consists of a set that is expanded using the predicates behind the colons (Section \ref{subsec:gringo:condition}) and a lower and an upper bounds.
+The constraint is true if and only if the number of true literals within the set is between the upper and lower bound.
+Furthermore, the constraint is used in the head of a rule, 
+that is, it is not only a test but can derive (``guess'') new atoms, which in this case correspond to possible moves of discs.
+Note that at this point we have not constrained the moves.
+Up to now, any disc could be moved to any peg at each time point with out considering any problem constraints.
+
+Next follows the Define part, here we give rules that define new auxiliary predicates, 
+which as such do not touch the satisfiability of the problem but will be used in the Test part later on.
+The rule in Line 4 projects out the target peg of a move, i.e., the predicate \code{move}/$2$ can be used 
+if we only need the disc affected by a move but not its target location.
+We will use the predicate \code{on}/$3$ to capture the state of the Hanoi puzzle at each time point.
+Its first two argument give the location of a disc at the time point given by the third argument.
+The next rule in Line 5 infers the location of each disc in the initial state (time point 0).
+Then we model the state transition using the rules in Line 6 and 7.
+The first rule is quite straightforward and states that the moved disc changes its location
+Note the usage of \code{not moves(T)} here.
+This literal prevents deriving an infinite number of rules, which would be all useless 
+because the state no longer changes.
+The second rule makes sure that all discs that are not moved stay where they are.
+Finally, we define the auxiliary predicate \code{blocked}/$3$, 
+which marks positions w.r.t. pegs that cannot be moved from.
+First in Line 8, the position below a disc on some peg is blocked.
+Second in Line 9, the position directly below a blocked position is blocked.
+Note that we mark position zero to be blocked, too.
+This will be convenient later on to assert some redundant moves.
+
+Finally, there is the Test part building upon both Generate and Define part 
+asserting wrong moves that do not agree with the problem description.
+It consists solely of integrity constraints, which fail whenever all their literals are true.
+Its first integrity constraint in Line 11 asserts 
+that a disc that is blocked, i.e, with some disc on top, cannot be moved.
+Note the usage of \code{D-1} here, this way a disc cannot be put back to the same location again.
+The integrity constraint in Line 12 asserts that a disc can only be placed on top of a bigger disc.
+Line 13 asserts the goal situation.
+To make the the encoding more efficient, we add a redundant constraint in Line 14,
+which asserts that each disc at all time points is located on exactly one peg.
+Although, this constraint is implied by the constraints above,
+adding this additional domain knowledge greatly improves the speed with which the problem can be solved.
+Finally, the last two statements control which predicates are printed, when a satisfying model
+for the instance is found.
+Here we first hide all predicates (Line 15) and then explicitly show only the \code{move}/$3$ predicate (Line 16).
+
+\subsection{Problem Solution}
+
+Now we are ready to solve the encoded puzzle. 
+To find an answer set, invoke one of the following commands 
+(\clingo{}, or \gringo{} and \clasp{} have to be installed somewhere under the systems path for the commands below to work):
+\begin{lstlisting}[numbers=none,escapechar=\%]
+clingo %\attach{examples/inst_toh.lp}{inst\char`\_toh.lp}% %\attach{examples/enc_toh.lp}{enc\char`\_toh.lp}%
+gringo %\attach{examples/inst_toh.lp}{inst\char`\_toh.lp}% %\attach{examples/enc_toh.lp}{enc\char`\_toh.lp}% | clasp
 \end{lstlisting}
-%
-The \code{1} on the left hand side (right hand side) specifies that at least (at most) on atom within the curly brackets has to be true
-that is~\code{\pred{node}(\const{1})} must have exactly one color.
-Note that~\code{\pred{node}(\const{1})} has been removed from the body,
-as it is derived via a corresponding fact and \code{color(1,C)} has been expanded using 
-the colors \code{red}, \code{green}, and \code{blue} given by \pred{color}/$1$ specified in the instance.
-Similar ground instances are obtained for the other nodes~\const{2} to~\const{6}.
-With these six rules alone we already have a coloring the graph,
-which does not yet ensure that two adjacent nodes have different colors.
-This property is asserted by the rule in line four in the test part.
-For each instance of \pred{edge}/2,
-we obtain~\code{n} ground instances of the integrity constraint in Line~6,
-prohibiting that the same color~\var{C} is assigned to the adjacent nodes.
-We get the following ground instances due to \code{\pred{edge}(\const{1},\const{2})}:
-%
-\begin{lstlisting}[numbers=none]
-:- color(1,red), color(2,red).
-:- color(1,green), color(2,green).
-:- color(1,blue), color(2,blue).
-\end{lstlisting}
-%
-Again note that \code{\pred{edge}(\const{1},\const{2})},
-derived via a fact, has been removed from the body.
+Note that (depending on your viewer) you can right or double-click on file names marked with a red font 
+to safe the associated file to disc.
+In the latter of the document, this is possible with all examples.
 
-\subsection{Problem Solution}\label{subsec:color:solution}
-\input{figures/color}
-
-Provided that a given graph is colorable with the colors given in the instance,
-a solution can be read off an answer set of the program consisting
-of the instance and the encoding.
-For the graph in Figure~\ref{fig:graph_intro},
-the following answer set can be computed:%
-\marginlabel{%
-  To find an answer set, invoke:\\
-  \code{\mbox{~}gringo \textbackslash\\
-        \mbox{~}examples/color.lp \textbackslash\\
-        \mbox{~}examples/graph.lp |\rlap{\:\textbackslash}\\
-        \mbox{~}clasp}\\
-  or alternatively:\\
-  \code{\mbox{~}clingo \textbackslash\\
-        \mbox{~}examples/color.lp \textbackslash\\
-        \mbox{~}examples/graph.lp}}
-%
+The output of the solver (\clingo{} in this case) will look something like that:
 \begin{lstlisting}[numbers=none]
 Answer: 1
-... color(1,red)  color(2,green) color(3,green)
-    color(4,blue) color(5,red)   color(6,blue)
+move(4,b,1)  move(3,c,2)  move(4,c,3)  move(2,b,4)  \
+move(4,a,5)  move(3,b,6)  move(4,b,7)  move(1,c,8)  \
+move(4,c,9)  move(3,a,10) move(4,a,11) move(2,c,12) \
+move(4,b,13) move(3,c,14) move(4,c,15) 
+SATISFIABLE
+
+Models      : 1+    
+Time        : 0.010
+  Prepare   : 0.000
+  Prepro.   : 0.010
+  Solving   : 0.000
 \end{lstlisting}
-%
-Note that we have omitted the instances of \pred{color}/$1$, \pred{node}/$1$ 
-and \pred{edge}/$2$ in order to emphasize the actual solution,
-which is depicted in Figure~\ref{fig:color}.
-Such output projection can also be specified within a logic program file by
-using the declaratives \code{\#hide} and \code{\#show},
-described in Section~\ref{subsec:gringo:meta}.
+The first line indicates that an answer set follows in the line below (the \code{\textbackslash} marks a line wrap).
+Then the status follows, this might be either \code{SATISFIABLE}, \code{UNSATISFIABLE}, or \code{UNKNOW} 
+if the computation is interrupted.
+The \code{1+} right of \code{Models:} indicates that one answer set has been found and the \code{+} that 
+the whole search space has not yet been explored, so there might be further answer sets.
+Following that, there are some time measurements: 
+Beginning with total computation time, 
+which is split into 
+preparation time (grounding), 
+preprocessing time (\clasp{} has an internal preprocessor that tries to simplify the program), 
+and solving time (the time needed to find the answer excluding preparation and preprocessing time).
+More information about options and output can be found in Section~\ref{sec:options}.

### 2010-09-21 doc/guide/introduction.tex
- updated introduction
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3390 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 please use the potassco mailing list 
-\href{mailto:potassco-users@lists.sourceforge.net}{potassco-users@lists.sourceforge.net}.
+\href{mailto:potassco-users@lists.sourceforge.net}{potassco-users@lists.sourceforge.net}
+or consult the supporting pages at \href{http://potassco.sourceforge.net}{potassco.sourceforge.net}.
 
@@
 that serves both as guideline on how to model problems using logic programs
-and also as an example on how compact on concise the modeling language of \gringo is.
+and also as an example on how compact and concise the modeling language of \gringo\ is.
 The probably most important part for a user, Section~\ref{sec:language},
@@
 as well as input-related errors and warnings that may be reported.
-During the guide we forgo most of the theoretical background in favor of
+During the guide we forgo most of the theoretical background in favor
 of small intuitive examples and informal descriptions.
@@
 (a grounder that constitutes the traditional front end of solver \smodels~\cite{siniso02a}),
-Appendix~\ref{sec:lparse} lists the most prominent differences to our tools.
+Appendix~\ref{sec:lparse}\comment{Appendix not updated yet}\ lists the most prominent differences to our tools.
 Otherwise, \gringo, \clingo, and \iclingo\ should accept most inputs recognized by \lparse,
 while the input of solver \clasp\ can also be generated by \lparse\ instead of \gringo.
-Throughout this guide, we will provide quite a number of examples.
-Many of them can actually be run, and instructions how to accomplish this
+Throughout this guide, we provide quite a number of examples.
+Many of them can actually be run, and instructions on how to accomplish this
 (or sometimes meta-remarks)
@@
 through Potassco~\cite{potassco}.
-We sincerely hope that you will find it enjoyable and helpful!
+We hope that you will find it enjoyable and helpful!
 

### 2010-09-21 doc/guide/quickstart.tex
- updated quickstart section
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3391 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \section{Quickstart}\label{sec:quickstart}
-In this section we will demonstrate the expressive power 
+In this section we demonstrate the expressive power 
 and the simple yet powerful modeling language of \gringo{}
@@
 Additionally, a disc may not be put on top of a smaller disc.
-We will ignore that there is an efficient algorithm to solve this problem 
+We ignore that there is an efficient algorithm to solve this problem 
 and just specify how a solution, in terms of a sequence of moves, has to look.
@@
 from an instance of the following problem:
-given an initial placement of the discs, a goal situation, and a number of moves $n$,
+given an initial placement of the discs, a goal situation, and a number $n$,
 decide whether there is a sequence of moves of length $n$ 
@@
 Furthermore, the predicates \pred{init\_on}/$2$ and \pred{goal\_on}/$2$ describe the initial and goal situation, respectively.
-Their first argument is the number of a disc and the second argument the peg 
-this disc is located in the initial or goal situation.
+Their first argument is the number of a disc and the second argument is the peg 
+on which the disc is located in the initial or goal situation.
 Finally, the predicate \pred{moves}/$1$ specifies the number of moves within which the goal situation has to be reached.
@@
 With ASP we can easily change this requirement and 
-the encoding represented in the following will work with an arbitrary number of pegs and any initial or goal situation.
+the encoding represented in the following works with an arbitrary number of pegs and any initial or goal situation.
 Figure~\ref{fig:toh_inst} depicts a possible instance (the dashed discs mark the goal situation) 
@@
 and a \emph{Test} part~\cite{lifschitz02a}.
-We will follow this paradigm and
+We follow this paradigm and
 mark respective parts via comment lines beginning with \code{\%} in the encoding below.
@@
 Next follows the Define part, here we give rules that define new auxiliary predicates, 
-which as such do not touch the satisfiability of the problem but will be used in the Test part later on.
+which as such do not touch the satisfiability of the problem but are used in the Test part later on.
 The rule in Line 4 projects out the target peg of a move, i.e., the predicate \code{move}/$2$ can be used 
 if we only need the disc affected by a move but not its target location.
-We will use the predicate \code{on}/$3$ to capture the state of the Hanoi puzzle at each time point.
+We use the predicate \code{on}/$3$ to capture the state of the Hanoi puzzle at each time point.
 Its first two argument give the location of a disc at the time point given by the third argument.
@@
 Then we model the state transition using the rules in Line 6 and 7.
-The first rule is quite straightforward and states that the moved disc changes its location
+The first rule is quite straightforward and states that the moved disc changes its location.
 Note the usage of \code{not moves(T)} here.
 This literal prevents deriving an infinite number of rules, which would be all useless 
-because the state no longer changes.
+because the state no longer changes after the last move.
 The second rule makes sure that all discs that are not moved stay where they are.
@@
 Finally, there is the Test part building upon both Generate and Define part 
-asserting wrong moves that do not agree with the problem description.
+to rule out wrong moves that do not agree with the problem description.
 It consists solely of integrity constraints, which fail whenever all their literals are true.
-Its first integrity constraint in Line 11 asserts 
+The first integrity constraint in Line 11 asserts 
 that a disc that is blocked, i.e, with some disc on top, cannot be moved.
@@
 Line 13 asserts the goal situation.
-To make the the encoding more efficient, we add a redundant constraint in Line 14,
+To make the encoding more efficient, we add a redundant constraint in Line 14,
 which asserts that each disc at all time points is located on exactly one peg.
@@
 to safe the associated file to disc.
-In the latter of the document, this is possible with all examples.
+This is possible with all examples given in this document.
 
-The output of the solver (\clingo{} in this case) will look something like that:
+The output of the solver (\clingo{} in this case) looks something like that:
 \begin{lstlisting}[numbers=none]

### 2010-09-21 doc/guide/language.tex
- updated language section
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3394 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \clingo\ returns answer sets.
+
 Usually logic programs are specified in one or more text files whose names are passed via the command line
@@
 Facts have an empty body.
-Throughout this section we will further extend the predicates that can be used 
+Throughout this section we further extend the predicates that can be used 
 in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
-Furthermore, \gringo{} expects rules to be save, i.e., 
-all variables that appear in a rule have to appear in some some positive literal (a literal not preceded by \code{not}) in the body.
+Furthermore, \gringo{} expects rules to be safe, i.e., 
+all variables that appear in a rule have to appear in some positive literal (a literal not preceded by \code{not}) in the body.
 If a variable appears positively in some predicate, then we say that this predicate binds the variable.
@@
 Intuitively, the head of a rule has to be true whenever all its body literals are true.
-In ASP every predicate needs some derivation, i.e., 
-a predicate cannot be true if there is no body deriving it.
-This implies that only predicates appearing in some head can appear in answer sets.
-Furthermore, derivations
+In ASP every atom needs some derivation, i.e., 
+an atom cannot be true if there is no rule deriving it.
+This implies that only atoms appearing in some head can appear in answer sets.
+Furthermore, derivations%
 \footnote{There are extensions like disjunctions 
 that go beyond simple derivability and also require minimality w.r.t. a reduct. 
-We will not cover the semantics of such constraints in this guide.}%
+We do not cover the semantics of such constraints in this guide.}%
 have to be acyclic,
 a feature that is important to model reachability.
-As a simple example consider the program \code{a :- b. b :- a.}
+As a simple example, consider the program \code{a :- b. b :- a.}
 The only answer set to this program is the empty set.
@@
 Finally, note that default negation is ignored when checking for acyclic derivations 
-(we do not need a reason for a predicate being false).
+(we do not need a reason for an atom being false).
 Default negation can be used to express choices, e.g., the program \code{a :- not b. b :- not a.} 
@@
 constraint, which provides a much more readable way to introduce choices.
+\com{This paragraph contains a lot of information. Could someone help me making it more readable?}
 
 A fact has an empty body and thus its associated head predicate is always true 
-and will appear in all answer sets.
+and appears in all answer sets.
 On the other hand, integrity constraints eliminate answer set candidates.
 They are merely tests that discard unwanted answer sets.
-That is, there will be no answer sets that satisfy all literals an integrity constraint.
+That is, there are no answer sets that satisfy all literals an integrity constraint.
 Elaborate examples on the usage of facts, rules, and integrity constraints
-will be provided in Section~\ref{sec:examples}.
+are provided in Section~\ref{sec:examples}.
 
@@
 Additionally, logic programs have to satisfy a minimality criterion, 
-which we will not detail in this guide.
+which we do not detail in this guide.
 The simple program \code{a | b.} has the two answer sets $\{a\}$ and $\{b\}$ but
@@
 \gnt~\cite{janisesiyo06a}
-needs to be used for solving a disjunctive program.%
+need to be used for solving a disjunctive program.%
 \footnote{System \dlv~\cite{dlv03a} also deals with disjunctive programs,
@@
 The four bitwise functions apply to signed integers,
-using the complement on two of a negative integer.
+using the two's complement of a negative integer.
 \eexample
@@
 Although, the latter might produce an infinite grounding 
-and \gringo{} will not necessarily halt when given such an input.
+and \gringo{} not necessarily halts when given such an input.
 
@@
   The unique answer set of the program is obtained via call:\\
-  \code{\mbox{~}gringo -t \attach{examples/arithf.lp}{arithf.lp}}}
-%
-\lstinputlisting{examples/arithf.lp}
+  \code{\mbox{~}gringo -t \attach{examples/arithc.lp}{arithc.lp}}}
+%
+\lstinputlisting{examples/arithc.lp}
 %
@@
 \eexample
+\comment{add \#sup and \#inf?}
 \end{example}
@@
 depend on~\var{X} and~\var{Y}.
-These two variables are bound in Line~4 via atoms of predicate \pred{num}/$1$.
+These two variables are bound in Line~5 via atoms of predicate \pred{num}/$1$.
 Also observe the different usage and role of built-in comparison predicate~\code{==}.
@@
 Note the usage of \code{X+1} in the term.
-\gringo{} will not try to unify any term containing arithmetic 
+\gringo{} does not try to unify any term containing arithmetic 
 but in this example \code{X} occurs also directly as second argument of the argument
 and can thus be unified with.
-The term $X+1$ is merely a test that deferred and checked later.
+The term $X+1$ is merely a test that is deferred and checked later.
 For example, the fourth line is equivalent to:
 \begin{lstlisting}[firstnumber=6]
-unifyf(X) :- f(a,X,Y) := F, sym(F), Y = X + 1.
+unifyf(X) :- f(a,X,Y) := F, sym(F), Y == X + 1.
 \end{lstlisting}%
@@
 
-Finally, note that assignments to some extend can bind variables.
+Finally, note that assignments to some extent can bind variables.
 Of course cyclic assignments cannot bind variables.
@@
 Either \code{X} or \code{Y} has to be provided by some positive predicate in this case.
-Additionally, the unification is restricted to ground terms on the right hand side of the assignment, 
-that is all variables on the right hand side have to be by some other predicate.
+Additionally, unification is restricted to ground terms on the right hand side of the assignment, 
+that is, all variables on the right hand side have to be bound by some other predicate.
 
@@
 meet :- available(jane), available(john).
-on(mon) | on(tues) | on(wed) | on(thurs) | on(fri) :- meet.
+on(mon) | on(tue) | on(wed) | on(thu) | on(fri) :- meet.
 \end{lstlisting}
@@
 \begin{lstlisting}[firstnumber=8]
-weekdays :- day(mon), day(tues), day(wed), day(thurs), day(fri).
+weekdays :- day(mon), day(tue), day(wed), day(thu), day(fri).
 \end{lstlisting}
@@
 
-There are three important issues about the correct usage of conditions.
-First, all predicates of atoms on the right-hand side of a condition
+Note that there are three important issues about the correct usage of conditions:
+\begin{enumerate}
+\item All predicates of atoms on the right-hand side of a condition
 must be either domain predicates\comment{refer to appendix?},
 i.e., predicates that can be completely evaluated during grounding,
-or built-in (can be used on the left-hand side but only in a rule body),
-which is due to the fact that conditions are evaluated during grounding.
-
-Second, any variable occurring within a condition is considered as \emph{local},
-that is, a condition cannot be used to bind variables outside itself.
+or built-in, which is due to the fact that conditions are evaluated during grounding.
+\item Any variable occurring within a condition is considered as \emph{local},
+that is, a condition cannot be used to bind variables outside the condition.
 In turn, variables outside conditions are \emph{global}, and each variable
@@
 be global.
-Third, global variables take priority over local ones, that is,
+\item Global variables take priority over local ones, that is,
 they are instantiated first.
@@
 making sure that they do not accidentally match the names of global variables.
+\end{enumerate}
 
@@
 \[\mathsf{opt}\,\code{[}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\] 
-\[\mathsf{opt}\,\code{\char`\{}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\]
+\[\mathsf{opt}\,\code{\char`\{}\,L_1@p_1,\dots,L_n@p_n\,\code{\char`\}}\]
 where $\mathsf{opt}$ is either \code{\#maximize} or \code{\#minimize} , $L_i$ are literals 
@@
 The hotels are identified via numbers assigned in descending order of stars.
-Of course, the more stars a hotel has the more it costs per night.
+Of course, the more stars a hotel has, the more it costs per night.
 As an ancillary information, we know that hotel~\const{4} is located
@@
 and a comment within one or over multiple lines is enclosed
-by ``\code{\%*}'' and ``\code{*\%}.''
+in ``\code{\%*}'' and ``\code{*\%}.''
 As an abstract example, consider:
@@
 #show p/3. % Include all atoms of predicate p/3 in output
-#showp(X,Y) : q(X). % Supress p/3 if the condition holds
+#show(X,Y) : q(X). % Include p/3 if the condition holds
 \end{lstlisting}
@@
 concrete values to be provided by a user.
-An example of this will be given in Section~\ref{subsec:ex:color}.
+An example of this is given in Section~\ref{subsec:ex:color}.
 Via the \code{\#const} declarative,
@@
 a (symbolic) constant on the left-hand side and a term on the right-hand side.
-Some exemplary \code{\#const} declarations are:
+Some exemplary constant declarations are:
 %
@@
 in ASP it is common to use very short variable names and
-using domain statements will likely result in name clashes
+using domain statements likely results in name clashes
 and undesired behavior.
@@
 External statements are used to mark certain atoms as external. This means 
-that those atoms are not subject of simplification and consequently will not be removed 
+that those atoms are not subject of simplification and consequently are not removed 
 from the logic program. There are two kinds of external directives, 
@@
 \end{lstlisting}
-Here, the external predicate \code{q}/$1$ will not be used for simplification
-of the problem and hence two ground rules (excluding facts) will be printed.
+Here, the external predicate \code{q}/$1$ is not used for simplification
+of the problem and hence two ground rules (excluding facts) are printed.
 \eexample
@@
 by a zero and is inserted directly after \lparse' compute statement.
+\begin{example}
+The following listing gives schematic example output of \gringo\ when external are used:
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
+\eexample
+\end{example}
 
@@
 interactions between grounding and solving are possible when incrementally solving with \iclingo.
-We do not give an introduction to Lua here (there are numerous tutorial on the web), 
+We do not give an introduction to Lua here (there are numerous tutorials on the web), 
 but give some examples showing the capabilities of this integration.
@@
 Note that Lua function calls look like function symbols but are preceded by ``\code{@}''.
-Regarding binding of variables, the same restrictions as with arithmetic \ref{subsec:gringo:arith} apply.
-
-In Line 9 we add function that emulates a range term.
+Regarding binding of variables, the same restrictions as with arithmetic in Section \ref{subsec:gringo:arith} apply.
+
+In Line 9 we add a function that emulates a range term.
 It returns a table containing all numbers in the interval $[\code{a},\code{b}]$.
-The values in this table are then successively replaced for the call \code{\@rng{X,Y}}.
+The values in this table are then successively replaced for the call \code{\@rng(X,Y)}.
 In fact, this function exactly behaves like a range term.
@@
 Function \code{f} in Line 2 returns a tuple 
-its first member a string indicating the type of the argument 
+whose first member is a string indicating the type of the argument 
 and the second reconstructs the value passed to function \code{f}.
-Note that in Line 4,6,9,11 and 13 the function \code{Val.new} is called.
+Note that in Line 4, 6, 9, 11 and 13 the function \code{Val.new} is called.
 Its first argument indicates that a function symbol is to be created and 
@@
 thus a tuple is created (this is equivalent to passing the empty string as name).
-You can infer from context, how \gringo's other in-built ground terms are created.
-Finally, note that \gringo integers and strings are directly mapped 
+Similarly, \gringo's other in-built ground terms are created.
+Finally, note that \gringo\ integers and strings are directly mapped 
 to the respective Lua types.
@@
 to induce 100 percent random decisions.
+\com{Admittedly, this example is somewhat spacy. Better ideas?}
 \eexample
@@
 Note that due to \gringo's grounding algorithm,
-rules will not be grounded twice using the same substitution of global variables
+rules are not grounded twice using the same substitution of global variables
 (the incremental constant is treated like a global constant here).
@@
 In Section~\ref{subsec:ex:block},
-we will provide a typical example in which these conditions naturally hold.
+we provide a typical example in which these conditions naturally hold.
 
@@
   The solving stops when the largest number has been found.
-}
-%
+}%
 The Schur number~$n$ w.r.t. to a given number~$c$ is the largest integer 

### 2010-09-21 doc/guide/options.tex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3396 3e02f596-3131-0410-8352-e04eeaec37cc

@@
   Note that our description of command line options
-  is based on Version~2.0.2 of \gringo, \clingo, and \iclingo\
-  as well as Version~1.1.2 of \clasp.
+  is based on Version~3.0.x of \gringo, \clingo, and \iclingo\
+  as well as Version~1.3.x of \clasp.
   While it is rather unlikely that command line options will
@@
 (These facts can then be used to, i.e., write a meta-interpreter.)
+\comment{add a section about a meta-interpreter}
 \item[\code{--lparse,-l}]~\\
@@
 Compute supported models~\cite{apblwa88a} (rather than answer sets).
-\item[\code{--dimacs}]~\\
-Compute models of a propositional formula in DIMACS/CNF format~\cite{dimacs}.
 \item[\code{--trans-ext all|choice|weight|no}]~\\
@@
 Having introduced the general options of \clasp,
-let us note that the options below \code{--dimacs} in the above list
+let us note that the options below \code{--supp-models} in the above list
 are quite low-level and more or less an issue of fine-tuning.
@@
 but scarcely yields conflicts during search.
+For a brief overview about fine-tuning see \cite{gekasc09b}.
 

### 2010-09-21 doc/guide/language.tex
- added an index covering the language
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3398 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \subsubsection{Normal Programs and Integrity Constraints}\label{subsec:gringo:normal}
+\index{Terms}
+\index{Terms!Constants}
+\index{Terms!Variables}
+\index{Terms!Variables!Anonymous}
+\index{Terms!Functions}
+\index{Terms!\code{\#supremum}}
+\index{Terms!\code{\#infimum}}
 
@@
 
+\index{Statements!Rules}
+\index{Statements!Facts}
+\index{Statements!Integrity Constraints}
 \begin{tabular}{rl@{}l}\label{eq:normal:rule}
@@
 Any $L_j$ is a \emph{literal} of the form $A$ or $\code{not}~A$
-for an atom~$A$ where the connective \code{not} corresponds to default negation.
+for an atom~$A$ where the connective \code{not} corresponds to default negation\index{Literals!Default Negation}.
 The set of literals $\{L_1,\dots,L_n\}$ is called the body of the rule.
@@
 \subsubsection{Classical Negation}\label{subsec:gringo:negation}
+\index{Literals!Classical Negation}
 
@@
 \subsubsection{Disjunction}\label{subsec:gringo:disjunction}
-
+\index{Aggregates!Disjunction}
 Disjunctive logic programs permit connective~``\code{|}'' between atoms in rule heads.
@@
 \subsubsection{Built-In Arithmetic Functions}\label{subsec:gringo:arith}
+\index{Literals!Arithmetic Functions}
+\index{Literals!Arithmetic Functions!Addition, \code{+}}
+\index{Literals!Arithmetic Functions!Subtraction, \code{-}}
+\index{Literals!Arithmetic Functions!Multiplication, \code{*}}
+\index{Literals!Arithmetic Functions!Division, \code{/}}
+\index{Literals!Arithmetic Functions!Modulo, \code{\%}}
+\index{Literals!Arithmetic Functions!Exponentation, \code{**}}
+\index{Literals!Arithmetic Functions!Absolute Value, \code{\#abs}}
+\index{Literals!Arithmetic Functions!Bitwise AND, \code{\&}}
+\index{Literals!Arithmetic Functions!Bitwise OR, \code{?}}
+\index{Literals!Arithmetic Functions!Bitwise XOR, \code{\^}}
+\index{Literals!Arithmetic Functions!Bitwise Complement, \code{\textasciitilde}}
 
@@
 \subsubsection{Built-In Comparison Predicates}\label{subsec:gringo:comp}
+\index{Literals!Comparison Predicates}
+\index{Literals!Comparison Predicates!Equality, \code{==}}
+\index{Literals!Comparison Predicates!Inequality, \code{!=}}
+\index{Literals!Comparison Predicates!Less, \code{<}}
+\index{Literals!Comparison Predicates!Less or Equal, \code{<=}}
+\index{Literals!Comparison Predicates!Greater,\code{>}}
+\index{Literals!Comparison Predicates!Greater or Equal\code{>=}}
 
@@
 \subsubsection{Assignments}\label{subsec:gringo:assign}
+\index{Literals!Assignments}
+\index{Literals!Assignments!Term Unification, \code{:=}}
+\index{Literals!Assignments!Variable Assignment, \code{=}}
 
@@
 \subsubsection{Intervals}\label{subsec:gringo:interval}
+\index{Terms!Intervals}
 
@@
 \subsubsection{Conditions}\label{subsec:gringo:condition}
+\index{Aggregates!Conditions}
 
@@
 \subsubsection{Pooling}\label{subsec:gringo:pool}
+\index{Terms!Pooling}
 
@@
 \subsubsection{Aggregates}\label{subsec:gringo:aggregate}
+\index{Aggregates}
 
@@
 Currently, \gringo\ supports the aggregates 
-\code{\#sum} (the sum of weights), 
-\code{\#min} (the minimum weight), 
-\code{\#max} (the maximum weight), and
-\code{\#avg} (the average of all weights
+\code{\#sum} (the sum of weights), \index{Aggregates!Sum, \code{\#sum}}
+\code{\#min} (the minimum weight), \index{Aggregates!Minimum, \code{\#min}}
+\code{\#max} (the maximum weight), and \index{Aggregates!Maximum, \code{\#max}}
+\code{\#avg} (the average of all weights \index{Aggregates!Average, \code{\#avg}}
 \footnote{The average aggregate over an empty set of weights is defined to be always true irrespective of any bounds.}%
@@
 Furthermore, there are three aggregates that are syntactically different.
-The first is the \code{\#count} aggregate:
+The first is the \code{\#count}\index{Aggregates!Count, \code{\#count}} aggregate:
 \[l\,\code{\#count}\,\code{\char`\{}\, L_1\code{,} \dots\code{,} L_n \,\code{\char`\}}\, u\]%
@@
 Finally, there are the two parity aggregates:
+\index{Aggregates!Even Parity, \code{\#even}}%
+\index{Aggregates!Odd Parity, \code{\#odd}}%
 \begin{align*}
@@
 \subsubsection{Optimization}\label{subsec:gringo:optimize}
+\index{Statements!Optimize Statements}
 
@@
 A (ground) optimize statement has the form: 
+\index{Statements!Optimize Statements!\#minimize}%
+\index{Statements!Optimize Statements!\#maximize}%
 \[\mathsf{opt}\,\code{[}\,L_1=w_1@p_1,\dots,L_n=w_n@p_n\,\code{\char`\}}\] 
@@
 \subsubsection{Meta-Statements}\label{subsec:gringo:meta}
+\index{Meta-Statements}
 
@@
 \paragraph{Comments.}
+\index{Meta-Statements!Comments}
 To keep records of the contents of a logic program,
@@
 \paragraph{Hiding Predicates.}
+\index{Meta-Statements!Hiding Predicates, \code{\#show}, \code{\#hide}}%
 Sometimes, one may be interested only in a subset of the atoms belonging
@@
 \paragraph{Constant Replacement.}
+\index{Meta-Statements!Constant Replacement, \code{\#const}}%
 Constants appearing in a logic program may actually be placeholders for
@@
 \paragraph{Domain Declarations.}
+\index{Meta-Statements!Domain Declarations, \code{\#domain}}%
 Usually, variable names are local to a rule,
@@
 \paragraph{Compute Statements.}
+\index{Meta-Statements!Compute Statements, \code{\#compute}}%
 These statements are artifacts supported for backward compatibility.
@@
 \paragraph{External Statements.}
+\index{Meta-Statements!External Statements, \code{\#external}}%
 External statements are used to mark certain atoms as external. This means 
@@
 \subsubsection{Integrated Scripting Language}\label{subsec:lang:lua}
+\index{Lua}%
 Utilizing the scripting language Lua\footnote{\url{http://www.lua.org}}, 
@@
 Note that Lua function calls look like function symbols but are preceded by ``\code{@}''. 
+\index{Terms!Lua Function Call}%
+\index{Lua!Function Call}%
 Regarding binding of variables, the same restrictions as with arithmetic in Section \ref{subsec:gringo:arith} apply.
@@
 It returns a table containing all numbers in the interval $[\code{a},\code{b}]$.
-The values in this table are then successively replaced for the call \code{\@rng(X,Y)}.
+\index{Lua!Term Insertion}%
+The values in this table are then successively inserted for the call \code{\@rng(X,Y)}.
 In fact, this function exactly behaves like a range term.
@@
 \centering
+\index{Lua!Val Metatable}%
 \begin{tabular}{||l|l||}
@@
 \begin{table}[ht]
+\index{Lua!Assignment Metatable}%
 \centering
@@
 \subsection{Input Language of \iclingo}\label{subsec:lang:iclingo}
+\index{Incremental Grounding}%
 
@@
 %
+\index{Incremental Grounding!Base Part, \code{\#base}}%
+\index{Meta-Statements!Base Part, \code{\#base}}%
 Via ``\code{\const{\#base}.},''
@@
 is determined by the last
+\index{Incremental Grounding!Cumulative Part, \code{\#cumulative}}%
+\index{Meta-Statements!Cumulative Part, \code{\#cumulative}}%
+\index{Incremental Grounding!Volatile Part, \code{\#volatile}}%
+\index{Meta-Statements!Volatile Part, \code{\#volatile}}%
 \const{\#base}, \const{\#cumulative}, or \const{\#volatile} statement

### 2010-09-29 doc/guide/language.tex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3418 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \index{Statements!Integrity Constraints}
+Rules are defined as follows:\\
+
 \begin{tabular}{rl@{}l}\label{eq:normal:rule}
@@
 in a rule including comparison predicates (Section \ref{subsec:gringo:comp}) and aggregates (Section \ref{subsec:gringo:aggregate}).
-Furthermore, \gringo{} expects rules to be safe, i.e., 
+Furthermore, \gringo{} expects rules to be safe, i.e., \index{Safe Program}
 all variables that appear in a rule have to appear in some positive literal (a literal not preceded by \code{not}) in the body.
@@
 constraint, which provides a much more readable way to introduce choices.
-\com{This paragraph contains a lot of information. Could someone help me making it more readable?}
+%\com{This paragraph contains a lot of information. Could someone help me making it more readable?}
 
@@
 In contrast, the classical (or strong) negation of some proposition
-holds if the complement of the proposition is objectively derived~\cite{gellif91a}.
+holds if the complement of the proposition is derived~\cite{gellif91a}.
 Classical negation, indicated by symbol ``\code{-},'' is permitted in front of atoms.
@@
 
-It is important to note that variables
+\begin{Note}
+that it is important that variables
 in the scope of an arithmetic function are not bound by a corresponding atom.
@@
 and \gringo{} not necessarily halts when given such an input.
+\end{Note}
 
@@
 \eexample
-\comment{add \#sup and \#inf?}
-\end{example}
-
-Importantly, a built-in comparison predicate cannot bind variables, i.e.,
+%\comment{add \#sup and \#inf?}
+\end{example}
+
+\begin{Note}
+that a built-in comparison predicate cannot bind variables, i.e.,
 when checking whether a rule is safe, 
 comparison predicates are not considered to be positive.
+\end{Note}
 
@@
 
-Finally, note that assignments to some extent can bind variables.
+\begin{Note}
+that assignments to some extent can bind variables.
 Of course cyclic assignments cannot bind variables.
@@
 that is, all variables on the right hand side have to be bound by some other predicate.
+\end{Note}
 
@@
 
-As with built-in arithmetic functions,
+\begin{Note}
+that as with built-in arithmetic functions,
 an integer interval mentioning some variable (like~\var{X} in Line~4 of Example~\ref{ex:int})
 cannot be used to bind the variable.
+\end{Note}
 
@@
 
-Note that there are three important issues about the correct usage of conditions:
+\begin{Note}
+that there are three important issues about the correct usage of conditions:
 \begin{enumerate}
 \item All predicates of atoms on the right-hand side of a condition
-must be either domain predicates\comment{refer to appendix?},
+must be either domain predicates,%\comment{refer to appendix?}
 i.e., predicates that can be completely evaluated during grounding,
@@
 \end{enumerate}
+\end{Note}
 
@@
 
-\com{Should we mention stratifed aggregates that do not necessarily need conditions?}
+%\com{Should we mention stratifed aggregates that do not necessarily need conditions?}
 
@@
 #const y = f(x,h).
-\end{lstlisting}%
-Note that (for efficiency reasons) constant declarations are order dependent.
+\end{lstlisting}
+
+\begin{Note}
+that (for efficiency reasons) constant declarations are order dependent.
 In the example above \code{x} would be replaced by \code{42}
 but when reversing the directives this would no longer be the case.
+\end{Note}
 
@@
 Observe that the resulting program is safe.
-Nevertheless, we suggest not to use domain statements because 
+
+\begin{Note}
+that we suggest not to use domain statements because 
 in ASP it is common to use very short variable names and
@@
 and undesired behavior.
+\end{Note}
 
@@
 \end{lstlisting}
+
+\begin{Note}
+that compute statement are not needed in general.
+The same behavior can be achieved by using integrity constraints.
+In fact, compute statements exist mainly for compatibility reasons with \lparse.
+We suggest to not use them.
+\end{Note}
 
@@
 to induce 100 percent random decisions.
-\com{Admittedly, this example is somewhat spacy. Better ideas?}
+%\com{Admittedly, this example is somewhat spacy. Better ideas?}
 \eexample

### 2010-10-01 doc/guide/quickstart.tex
- updated language section
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3394 3e02f596-3131-0410-8352-e04eeaec37cc
fixed typo
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3428 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 Discs are enumerated by consecutive integers beginning with one
-where a disc with a lower number is considered to be bigger than a list with a higher number.
+where a disc with a lower number is considered to be bigger than a disc with a higher number.
 The pegs can have arbitrary names. 
@@
 
-The `\code{;}' in the first line is some syntactic sugar (Section \ref{subsec:gringo:pool})
+The ``\code{;}'' in the first line is some syntactic sugar (Section \ref{subsec:gringo:pool})
 that expands the statement into three facts

### 2010-10-01 doc/guide/options.tex
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/trunk/gringo@3418 3e02f596-3131-0410-8352-e04eeaec37cc

@@
 \clasp\ (Section~\ref{subsec:opt:clasp}).
-Each of these tools displays its available options
-when invoked with flag \code{--help} or \code{-h},
-respectively.%
+Each of these tools display their available options
+when invoked with flag \code{--help} or \code{-h}.%
 \footnote{%
@@
 %Print (extended) statistic information before termination.
-\item[\code{--verbose[=n],-V=[n]}]~\\
+\item[\code{--verbose[=n],-V}]~\\
 Print additional (progress) information during computation.
-Verbosity level one and two is reserved for clasp only.
+Verbosity level one and two are currently not used by \gringo.
 Level three prints internal representations of rules during grounding.
@@
 Output ground program in form of facts.
-(These facts can then be used to, i.e., write a meta-interpreter.)
-\comment{add a section about a meta-interpreter}
+(These facts can then be used to, i.e., write a \attach{examples/meta.lp}{meta-interpreter}.)
+%\comment{add a section about a meta-interpreter}
 \item[\code{--lparse,-l}]~\\
@@
  but it leads to a syntax error (cf.\ Section~\ref{subsec:error}) otherwise.)
-\item[\code{--ifixed \textit{n}}]~\\
+\item[\code{--ifixed=\textit{n}}]~\\
 Use \code{\textit{n}} as fix step number if the input program
@@
 Incremental ASP system \iclingo\ extends \clingo\ by interleaving
-grounding and solving for problems including a mutable bound,
-and an abstract invocation of \iclingo\ is as with \clingo:
+grounding and solving for problems including a mutable bound.
+An abstract invocation of \iclingo\ is as with \clingo:
 %
@@
 Print statistic information for each incremental solving step.
-\item[\code{--imin \textit{n}}]~\\
+\item[\code{--imin=\textit{n}}]~\\
 Perform at least \code{\textit{n}} incremental solving steps before termination.
@@
  set via \code{--istop}.)
-\item[\code{--imax \textit{n}}]~\\
+\item[\code{--imax=\textit{n}}]~\\
 Perform at most \code{\textit{n}} incremental solving steps before termination.
@@
  set via \code{--istop}.)
-\item[\code{--istop SAT|UNSAT}]~\\
+\item[\code{--istop=SAT|UNSAT}]~\\
 Terminate after an incremental solving step 
@@
 answer set has been found.
-\item[\code{--iquery \textit{n}}]~\\
+\item[\code{--iquery=\textit{n}}]~\\
 Start with incremental solving at step number~\code{\textit{n}}.
@@
  still accumulating static and cumulative rules for these steps.)
-\item[\code{--ilearnt keep|forget}]~\\
+\item[\code{--ilearnt=keep|forget}]~\\
 Maintain (\code{keep}) or delete (\code{forget}) learnt constraints
@@
 (This option configures the behavior of embedded \clasp.)
-\item[\code{--iheuristic keep|forget}]~\\
+\item[\code{--iheuristic=keep|forget}]~\\
 Maintain (\code{keep}) or delete (\code{forget}) heuristic information
@@
 Stand-alone \clasp~\cite{gekanesc07b} is a solver for ground programs in
-\lparse's numerical format~\cite{lparseManual}.
+\lparse's numerical format~\cite{lparseManual}. Beyond that, it can also
+be used as a SAT (on a simplified version of DIMACS/CNF format%
+\footnote{%
+\url{http://www.satcompetition.org/2009/format-benchmarks2009.html}
+}), or PB solver (on OPB format%
+\footnote{%
+\url{http://www.cril.univ-artois.fr/PB09/solver_req.html}
+}).
 An abstract invocation of \clasp\ looks as follows:
@@
 \begin{lstlisting}[numbers=none]
-clasp [ number | options ]
+clasp [ number | options | filename ]
 \end{lstlisting}
@@
  \code{--number} or its short form~\code{-n}.)
-If neither a filename (via long option \code{--file} or its short form \code{-f})
-nor an option that makes \clasp\ exit (see below) is provided,
-\clasp\ reads from the standard input.
+If neither a filename nor an option that makes \clasp\ exit (see below) is provided,
+\clasp\ reads from the standard input.%
+\footnote{%
+ In earlier versions of \clasp\ filenames had to be given via long option \code{--file} 
+ or its short form \code{-f}.}
 In fact, it is typical to use \clasp\ in a pipe with \gringo\ in the following way:
@@
 Print version information and exit.
+\item[\code{--pre}]~\\
+Run ASP preprocessor then print preprocessed input program and exit.
+\item[\code{--verbose[=n],-V}]~\\
+Configure printing of (progress) information during computation. 
+Argument $\code{\textit{n}}=\code{0}$ disables progress information, while
+$\code{\textit{n}}=\code{1}$ prints basic, and $\code{\textit{n}}=\code{2}$
+extended information.
 \item[\code{--stats}]~\\
 Print (extended) statistic information before termination.
-\item[\code{--file,-f \textit{filename}}]~\\
-Read from file \code{\textit{filename}} (rather than from the standard input).
 \item[\code{--number,-n \textit{n}}]~\\
@@
 (This is useful for benchmarking.)
-\item[\code{--seed \textit{n}}]~\\
+\item[\code{--time-limit=\textit{t}}]~\\
+Force termination after \code{\textit{t}} seconds.
+\item[\code{--search-limit=\textit{n},\textit{m}}]~\\
+Force termination after either \code{\textit{n}} 
+conflicts or \code{\textit{m}} restarts.
+\item[\code{--seed=\textit{n}}]~\\
 Use seed~\code{\textit{n}} (rather than~\code{1}) for random number generator.
+\item[\code{--solution-recording}]~\\
+Switch from backtrack-based enumeration~\cite{gekanesc07c} to
+enumeration based on solution recording. 
+Note that this  mode is prone to blow up 
+in space in view of an exponential number of solutions in the worst case.
+\item[\code{--restart-on-model}]~\\
+Restart the search from scratch (rather than doing enumeration~\cite{gekanesc07c})
+after finding an answer set. Note: This mode implies solution recording and hence
+has the same caveat. It is mainly useful when searching for an optimal 
+solution because, first, restarting may greatly speed up finding the optimum, and second,
+only the current optimum needs to be recorded.
+\item[\code{--project}]~\\
+Project answer sets to named atoms and only enumerate unique projected solutions~\cite{gekasc09a}.
 \item[\code{--brave}]~\\
@@
  than the best one found so far.)
-\item[\code{--opt-restart}]~\\
-Restart the search from scratch (rather than doing enumeration~\cite{gekanesc07c})
-after finding a provisionally optimal answer set.
-(This may speed up finding the optimum.)
-\item[\code{--opt-value \textit{n1}[,\textit{n2},\textit{n3}...]}]~\\
+\item[\code{--opt-value=\textit{n1}[,\textit{n2},\textit{n3}...]}]~\\
 Initialize objective function(s) to minimize with \code{\textit{n1}[,\textit{n2},\textit{n3}...]}.
-Depending on the reasoning mode,
-only equitable or strictly better answer sets are computed.
+\item[\code{--opt-ignore}]~\\
+Ignore any optimize statements of a logic program during computation.
+\item[\code{--opt-heu}]~\\
+Consider optimize statements in heuristics.
 \item[\code{--supp-models}]~\\
 Compute supported models~\cite{apblwa88a} (rather than answer sets).
-\item[\code{--trans-ext all|choice|weight|no}]~\\
+\item[\code{--trans-ext=all|choice|weight|dynamic|no}]~\\
 Compile extended rules~\cite{siniso02a} into normal rules of form (\ref{eq:normal:rule}).
@@
 while \code{all} means that both and \code{no} that none of them
-are subject to compilation.
-\item[\code{--eq \textit{n}}]~\\
+are subject to compilation. If argument \code{dynamic} is given, 
+\clasp\ heuristically decides whether or not to compile individual
+``weight rules''.
+\item[\code{--eq=\textit{n}}]~\\
 Run equivalence reasoning~\cite{gekanesc08a} for \code{\textit{n}} iterations,
@@
 run to fixpoint or do not run equivalence reasoning, respectively.
-\item[\code{--sat-prepro yes|no|\textit{n1}[,\textit{n2},\textit{n3}]}]~\\
+\item[\code{--backprop}]~\\
+Enable backpropagation in ASP-preprocessing.
+\item[\code{--sat-prepro=yes|no|\textit{n1}[,\textit{n2},\textit{n3}]}]~\\
 Run \emph{SatElite}-like preprocessing~\cite{eenbie05a}
@@
 or that \emph{SatElite}-like preprocessing is not to be run at all, respectively.
-\item[\code{--rand-watches yes|no}]~\\
-Initially choose watched literals randomly (with argument \code{yes})
-or systematically (with argument \code{no}).
 \end{description}
@@
 \begin{description}
-\item[\code{--lookback yes|no}]~\\
-Enable or disable lookback techniques (cf.\ Section~\ref{subsec:clasp:back}).
-(This option is included mainly for comparison purposes, and generally,
- it is recommended to keep the default argument value \code{yes}.)
-\item[\code{--lookahead atom|body|hybrid|auto|no}]~\\
+\item[\code{--lookahead=atom|body|hybrid|no}]~\\
 Apply failed-literal detection~\cite{freeman95}
 to atoms (with argument \code{atom}),
-to rule bodies (with argument \code{body}),
-to atoms and rule bodies like in \nomorepp~\cite{angelinesc05c} (with argument \code{hybrid}),
-or let \clasp\ pick one of the previous three possibilities
-based on program properties (with argument \code{auto}).
+to rule bodies (with argument \code{body}), or
+to atoms and rule bodies like in \nomorepp~\cite{angelinesc05c} (with argument \code{hybrid}).
 Failed-literal detection is switched off via argument~\code{no}.
-\item[\code{--initial-lookahead}]~\\
-Apply failed-literal detection (to atoms) in preprocessing,
-but not (necessarily) during search.
-\item[\code{--heuristic Berkmin|Vmtf|Vsids|Unit|None}]~\\
+\item[\code{--initial-lookahead=\textit{n}}]~\\
+Apply failed-literal detection in preprocessing,
+and for the first \code{\textit{n}} decisions during search.
+\item[\code{--heuristic=Berkmin|Vmtf|Vsids|Unit|None}]~\\
 Use \emph{BerkMin}-like decision heuristic~\cite{golnov02a}
@@
 (with argument~\code{None}).
-\item[\code{--rand-freq \textit{p}}]~\\
+\item[\code{--rand-freq=\textit{p}}]~\\
 Perform random (rather than heuristic) decisions
 with probability $0\leq\code{\textit{p}}\leq 1$.
-\item[\code{--rand-prob yes|no|\textit{n1},\textit{n2}}]~\\
+\item[\code{--rand-prob=yes|no|\textit{n1},\textit{n2}}]~\\
 Run \emph{Satzoo}-like random probing~\cite{een03a},
@@
 or that random probing is not to be run at all, respectively.
+\item[\code{--rand-watches=yes|no}]~\\
+Initially choose watched literals randomly (with argument \code{yes})
+or systematically (with argument \code{no}).
 \end{description}
-%
-Let us note that switching the above options can have dramatic effects
-(both positively and negatively) on the search performance of \clasp.
-Sticking to the default argument values \code{yes} for \code{--lookback}
-and \code{no} for \code{--lookahead} is generally recommended.
-If nonetheless performance bottlenecks are observed,
-it is worthwhile to give \code{Vmtf} and \code{Vsids} for \code{--heuristic}
-a try, in particular,
-when the program under consideration is huge 
-but scarcely yields conflicts during search.
-For a brief overview about fine-tuning see \cite{gekasc09b}.
 
@@
 
-The following options have an effect only if \code{--lookback}
-is turned on, that is, its argument value must be \code{yes}.
+The following options have an effect only if lookback techniques
+are turned on, that is, option \code{--no-lookback} is not used.
 %
 \begin{description}
+\item[\code{--no-lookback}]~\\
+Disable all lookback techniques.
+This option is included mainly for comparison purposes, and its use
+is not generally recommended.
 \item[\code{--restarts,-r \textit{n1}[,\textit{n2},\textit{n3}]|no}]~\\
@@
 Perform bounded restarts in answer set enumeration~\cite{gekanesc07c}.
-\item[\code{--save-progress}]~\\
+\item[\code{--reset-restarts}]~\\
+Reset restart strategy whenever an answer set is found.
+\item[\code{--save-progress[=\textit{n}]}]~\\
 Use cached (rather than heuristic) decisions~\cite{pipdar07a} if available.
+Cache decisions on backjumps $>$ \code{\textit{n}}.
 \item[\code{--shuffle,-s \textit{n1},\textit{n2}}]~\\
@@
 $\min\{(c/\code{\textit{n1}})*\code{\textit{n2}}^i,c*\code{\textit{n3}}\}$,
-where~$c$ is the initial number of constraints and~$i$ is the number of restarts
-performed so far.
+where~$c$ is the initial number of constraints and~$i$ is the number of restarts performed so far.
 \item[\code{--reduce-on-restart}]~\\
 Delete a portion of learnt constraints after every restart.
-\item[\code{--strengthen bin|tern|all|no}]~\\
+\item[\code{--estimate}]~\\
+Base the initial limit of learnt constraints on an estimate
+of the problem's complexity.
+\item[\code{--strengthen=bin|tern|all|no}]~\\
 Check binary (with argument \code{bin}),
@@
 Recursively apply strengthening, as proposed in~\cite{soreen05a}.
-\item[\code{--loops common|distinct|shared|no}]~\\
+\item[\code{--loops=common|distinct|shared|no}]~\\
 Learn loop nogood~\cite{gekanesc07a} per atom in an unfounded set~\cite{gerosc91a}
@@
 (with argument \code{no}).
-\item[\code{--contraction \textit{n}}]~\\
+\item[\code{--contraction=\textit{n}}]~\\
 Temporarily truncate learnt constraints over more than \code{\textit{n}} variables~\cite{ryan04a}.
@@
 %
-As with search options described in the previous section,
-switching lookback options may have a significant impact on the
-performance of \clasp.
-In particular, we suggest trying the universal restart sequence~\cite{lusizu93a}
+Let us note that switching the above search and lookback options can have dramatic effects
+(both positively and negatively) on the search performance of \clasp.
+If performance bottlenecks are observed, it is worthwhile to give 
+\code{Vmtf} and \code{Vsids} for \code{--heuristic} a try, in particular,
+when the program under consideration is huge 
+but scarcely yields conflicts during search.
+Furthermore, we suggest trying the universal restart sequence~\cite{lusizu93a}
 with different base units~\code{\textit{n1}} or even disabling restarts
@@
 in case that performance needs to be improved.
-
+For a brief overview on fine-tuning see \cite{gekasc09b}.
 Finally, let us consider the default command line of \clasp:
@@
 \begin{lstlisting}[numbers=none]
-clasp 1 --seed=1 --trans-ext=no --eq=5 --sat-prepro=no
-        --rand-watches=yes --lookback=yes --lookahead=no
-        --heuristic=Berkmin --rand-freq=0.0 --rand-prob=no
-        --restarts=100,1.5 --shuffle=0,0 --deletion=3.0,1.1,3.0
-        --strengthen=all --loops=common --contraction=250
+clasp 1 --verbose=1 --seed=1 --trans-ext=no --eq=5
+        --sat-prepro=no --lookahead=no
+        --heuristic=Berkmin --deletion=3.0,1.1,3.0
+        --rand-freq=0.0 --rand-watches=yes --rand-prob=no 
+        --restarts=100,1.5 --shuffle=0,0 --strengthen=all
+        --loops=common --contraction=250
 \end{lstlisting}
@@
 after finding an answer set,
-and with \code{--lookback=yes}, \code{--restarts=100,1.5} lets \clasp\ apply a geometric
+while \code{--restarts=100,1.5} lets \clasp\ apply a geometric
 restart policy.
